### 2023-02-09 Huston_Paper.tex
Adding Stuff

\documentclass[journal,onecolumn]{IEEEtran}

\hyphenation{op-tical net-works semi-conduc-tor}
\usepackage[square,sort&compress,numbers]{natbib}
\usepackage[margin=1in]{geometry}
\usepackage{textcomp} 
\usepackage{amssymb}  
\usepackage{bm}       
\usepackage{booktabs}
\usepackage{dcolumn}  
\usepackage{multirow} 
\usepackage{ragged2e}
\usepackage{graphicx} 
\usepackage{float}
\usepackage{rotating}
\usepackage{makecell}
\usepackage{multirow}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{listings}
\usepackage{array}
\usepackage[printonlyused]{acronym}
\usepackage[framed, numbered]{matlab-prettifier}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{times}
\usepackage{indentfirst}

\begin{document}
	%
	% paper title
	% Titles are generally capitalized except for words such as a, an, and, as,
	% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
	% not capitalized unless they are the first or last word of the title.
	% Linebreaks \\ can be used within to get better formatting as desired.
	% Do not put math or special symbols in the title.
	\title{Enhance Road Detection Data Processing of LiDAR Point Clouds to Specifically Identify Unmarked Gravel Rural Roads}
	%
	%
	% author names and IEEE memberships
	% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
	% a structure at a ~ so this keeps an author's name from being broken across
	% two lines.
	% use \thanks{} to gain access to the first footnote area
	% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
	% was not built to handle multiple paragraphs
	%

	\author{Rhett Huston, Jay Wilhelm}
%	{
%		and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
%		\thanks{M. Shell was with the Department
%			of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
%			GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
%		\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
%		\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}}

	% note the % following the last \IEEEmembership and also \thanks - 
	% these prevent an unwanted space from occurring between the last author name
	% and the end of the author line. i.e., if you had this:
	% 
	% \author{....lastname \thanks{...} \thanks{...} }
	%                     ^------------^------------^----Do not want these spaces!
	%
	% a space would be appended to the last name and could cause every name on that
	% line to be shifted left slightly. This is one of those "LaTeX things". For
	% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
	% "AB" then you have to do: "\textbf{A}\textbf{B}"
	% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
\markboth{Some Sort of Journal}%
{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for IEEE Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}

	Gravel roads that lack standardized features such as curbs or painted lines present detection and navigation challenges to autonomous vehicles. Global Positioning Service (GPS) and high resolution maps cannot be relied upon for navigation of rural roads, as some may only be width of the vehicle and GPS may not be accurate enough. Differentiation between gravel road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation. Normal Distribution Transform (NDT), based on scanning LiDAR data may be insufficient for navigating on gravel roads. This work will examine methods of classifying scanning LiDAR data for explicit detection of gravel road surface spatial and remission features. Exploration of terrain classification of gravel road surfaces with high resolution scanning LiDAR data will allow for predicting road boundary location enabling confident autonomous operations on gravel roads. The principal outcome of this work is a method for gravel road detection using LiDAR data for the purpose of predicting gravel road boundary locations. Detection of gravel road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.

\end{abstract}

%	
%	% Note that keywords are not normally used for peerreview papers.
%	\begin{IEEEkeywords}
%		IEEE, IEEEtran, journal, \LaTeX, paper, template.
%	\end{IEEEkeywords}
%	
%	
%	
%

\section{Introduction}

	% Autonomous vehicles have difficulty detecting rural unpaved gravel roads due to lack of standardized features such as curbs and painted lane lines.

	Autonomous vehicles have relied upon standardized road features such as curbs and painted lane lines or high definition maps paired with precision GPS for navigational solutions. Challenges arise when detecting unpaved gravel and chipseal roads in rural environments due to the lack of standardized road features and potentGPS signal unreliability. Gravel roads comprise 34.8\% \cite{road_stats_2} of all road surfaces in the United States Some nations have predominately unpaved road networks, such as India, where 70.7\% of all roads by mile are categorized as unpaved \cite{malik_lal_2019}. Detection of gravel roads may allow for expansion of current civilian autonomous vehicle capabilities as well as introduce opportunities for industries that operate in rural areas. Global Positioning Service (GPS) and high resolution maps cannot be relied upon for navigation of rural roads, as some gravel roads may only be width of the vehicle and GPS is not accurate enough or may not have a reliable signal for precision driving \cite{noauthor_gpsgov_nodate}. Current methods of road surface analysis rely on cameras and Light Detection and Ranging (LiDAR) sensors to detect standardized road features, however on gravel roads such attributes are non-existent \cite{skorseth_gravel_nodate}. Systems that rely on cameras may fail when analyzing gravel roads, as these lack visual cues such as painted lane markings \cite{crisman_scarf_1993} and may in fact closely resemble the surrounding terrain. LiDAR point cloud processing systems depend upon distinct geometric features, most commonly curbs \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_nodate,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja_road_nodate}, however as curbs are not installed on rural gravel roads \cite{skorseth_gravel_nodate} this method breaks down. Normal Distribution Transform (NDT) scan matching analytically compares two point cloud data sets to track current position, however this model relies on distinct geometric features, and suffers when distinguishing between terrain types making it insufficient for determining road boundaries \cite{biber_normal_2003}. Using 2D projection of planes unto a 3D point cloud of a road surface is an alternative approach, but likewise requires curbs or painted lines for road boundary definition \cite{fernandes_road_2014, borkar_robust_2009-1, guo_lane_2015}. Instead, using cameras and LiDAR may overcome difficulties in achieving real-time detection of gravel road surfaces using terrain classification to discover road boundaries, allowing for autonomous operations that do not rely upon match enabling processes and traditional visual cues.

	{Detection of gravel road surfaces may rely on the surface properties of the road without relying on distinct markings or topographical features. Surface roughness is a measurable property that may be used to characterize a gravel road, as they are typically consistent with gravel type used \cite{skorseth_gravel_nodate} which may be exploited in a search and compare function, and is distinct from grass and other ground types \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain}. Surface roughness properties may be derived from processing LiDAR data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}. Real time description of surface noise based on incoming LiDAR data, then subsequent comparison to a standard, is a possible alternative that may allow for computationally efficient road surface detection. Overlaying camera imagery data with LiDAR allows for assigning incoming LiDAR data with color information, potentially increasing efficiency by narrowing the search area. Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident. Detection of gravel road surfaces would increase the capabilities of self driving vehicles considerably, allowing autonomous operations on 1.5 million miles \cite{road_stats_2} of previously undetected rural roads. Problem Statement: Determine a processing method for increasing road detection precision of unmarked gravel rural roads.

\subsection{Literature Review}

	Current LiDAR based road surface detection models may be categorized into two methodologies. Firstly is the analysis of road surface properties, which considers aspects such as topology and surface roughness. Detection based on road surface smoothness was studied by segmenting the point cloud into candidate road surface regions and searching for elevation jumps \cite{liu_new_2013}. Secondly is the detection of roadside curbs, which relies upon the height difference between the road surface and the curb for road edge detection. LiDAR point density at a curb's upper and lower edges can be used to indicate road edges \cite{ibrahim_curb-based_2012}. Fernandes et al. propose a road surface detection method using LiDAR \cite{fernandes_road_2014}, by projecting a 2D reference plane unto the 3D LiDAR data. In this study, road surfaces are assumed to be flat regions between two elevated regions such as curbs. Cameras have been used to supplement LiDAR point cloud data, which can implement 3D colored elevation maps to match pre-stored geometry data in a modified Iterative Closest Point (ICP) algorithm \cite{manz_detection_2011}. Supervised Classification Applied to Road Following (SCARF) is an algorithm that detects road surfaces based on color differences between the road and the surrounding terrain \cite{crisman_scarf_1993}. It was found that difficulties arise when detecting road surfaces in less colored environments \cite{crisman_scarf_1993,manz_detection_2011}.

	Real time processing of rural gravel roads cannot rely on normalized, consistent features found on urban roads such as curbs, as rural roads lack these features \cite{skorseth_gravel_nodate}. As such, many of the proposed models are rendered unserviceable, as they rely on curbs or painted lines to offer distinctions in data sets \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_nodate,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja_road_nodate}.

	Further limitations are imposed when considering data set size and processing speed. Real time analysis of road surfaces dictates that trajectory updates to the vehicle in motion must have a rapid update rate, which prohibits the relatively lengthy collection of large data sets or any form of post-processing. Proposed methods \cite{yadav_extraction_2017,yadav_road_2018,yadav_rural_2018,yadav_pole-shaped_2015,miyazaki_line-based_2014,yang_semi-automated_2013,liu_new_2013,qiu_fast_2016} do not indicate the minimum number of points required for road surface detection, however large data sets numbering many millions of points are used in those studies. Computation efficiency of road surface detection is mentioned only by a few studies, such as the one completed by Yadav et al \cite{yadav_road_2018}. Principal component analysis on the height of each LiDAR data point was proposed to detect a straight, curb-less rural road with low computation time \cite{yadav_road_2018}. However a computational time of 24 minutes was required to process a 156 meter stretch of road, rendering real time detection impossible.

	Classification of the surrounding environment is an integral part to the proposed work, as differentiation of gravel road surfaces and surrounding terrain is necessary to determine road boundaries. Terrain classification is the analysis of the ground surface in order to specify the ground surface type \cite{laible_3d_2012,laible_terrain_2013,laible_map_building,rasmussen_combining_2002,reymann_improving_2015,walas_terrain_2014,wietrzykowski_boosting_2014,wang_road_nodate}. Current literature proposes multiple methods of terrain classification using LiDAR and RGB cameras, and may be roughly broken into two areas of application. Traversability or trafficability is the analysis of generally unstructured environments that includes obstacles or rough terrain types for autonomous vehicle path finding solutions  \cite{schilling_geometric_2017,ojeda_terrain_2006,coombs_driving_2000,stavens_self-supervised_nodate,belter_rough_2010,bartoszyk_terrain-aware_2017,noauthor_fusion_nodate,li_rugged_2019,wilson_terrain_2014,siva_robot_2019}. While this research may be useful for detection of obstacles on the road for the proposed work, it is assumed that the roads that will be studied will be free of obstacles.

	LiDAR and camera based real-time terrain classification was studied by projecting a 2D plane unto the point cloud data using a variant of Random Sample Consensus (RANSAC) called M-estimator Sample Consensus (MSAC) \cite{mijakovska_generating_2014} that is built for a more robust result \cite{laible_3d_2012,laible_map_building,laible_terrain_2013}. After segmenting the plane into cells, roughness and intensity histograms were analyzed on a per-cell basis. Markov Random Field (MRF) \cite{chellappa_classification_1985} and Conditional Random Field (CRF) \cite{wallach_conditional_nodate} were compared in their ability to compare individual cells with neighboring cells. Comparison with neighboring cells increases classification type accuracy, as adjacent cell are likely to have the same terrain type \cite{haselich_terrain_2011,zhao_fusion_2014}. It was found that CRF was better suited for classification.	Random Forest (RF) classification showed good results for identification of analysis of LiDAR and camera data \cite{breiman_random_2001}, and was used to predict cell terrain into one of five types, including gravel and grass. It was found that characterizing the LiDAR intensity as Gaussian distributed noise yielded poor results with only a 49.5\% true positive rate, however a low-resolution LiDAR was used. Higher resolution LiDAR that produces a high density point cloud, such as the Velodyne VLP-32 that will be used in the proposed work \cite{vlp_32c}, may allow for better terrain classification with Gaussian characterization of return intensity and spatial data. While capable of real-time classification of terrain type, this method was not employed in road identification and used a low resolution LiDAR system.

	Neural networks and similar machine learning tools may be used in terrain and road identification. Road classification with LiDAR spacial and remission features was completed with Neural Network classifier, Naive Bayes classifier, and Support Vector Machines (SVM) \cite{wang_road_nodate,wang_two-stage_2018}. Gravel roads were successfully identified with a 98.5\% true positive identification rate, however data collected required post processing. MATLAB's Neural Network tools were used in autonomous road identification using LiDAR and Camera data \cite{rasmussen_combining_2002}, however post processing was required, with unknown computational time.  Terrain classification was completed using Support Vector Machine (SVM), a classification algorithm where a line is drawn between two different categories to differentiate between them \cite{breiman_random_2001}. Histograms and averages for image hue, saturation, and color value, along with LiDAR intensity were used as points of classification for different terrain types. Although a 96\% true positive rate was reported for a rocky surface, the overall process was computationally expensive with 1.8 seconds per image.

	MLS and RANSAC were compared as methods to project a two dimensional plane unto a point cloud, facilitating examination of points by providing a frame of reference \cite{miller_method_nodate, gojcic_perfect_2019}. Method of least-squares analyses uses regression to approximate the best fit to a data set (\ref{eqn:MLS}). RANSAC is an iterative algorithm that is designed to handle large numbers of outliers in the sample data set \cite{derpanis_overview_nodate,yaniv_random_2010,fischler_random_1987,cantzler_random_nodate}.

	{Testing LiDAR data for normality was completed to determine its usefulness as a feature. Shapiro-Wilk analysis \cite{shapiro_analysis_1965, royston_extension_1982} tests the null hypothesis that a sample vector comes from a normally distributed population, using the test statistic in equation \ref{eq:shapiro_wilks}.}

	RDFs are a supervised machine learning method that operates on a set of decision trees \cite{ho_random_1995}. REFs have been shown to be useful with handling LiDAR data sets \cite{breiman_random_2001}, and have been employed in terrain classification \cite{laible_3d_2012,laible_map_building,laible_terrain_2013,khan_high_2011,reymann_improving_2015,schilling_geometric_2017, wietrzykowski_context-aware_2019}. Multiple decision trees are created with bootstrapped data sets - randomly selected data sub-sets from a training set. Classification may be determined by the majority results as opposed to the mean \cite{breiman_random_2001,ho_random_1995}. RDFs may be used to predict road surface area by detecting gravel terrain.

\section{Methodology}

	Required physical data was gathered by the Hexegon Chrysler Pacifica New Eagle DBW van (Figure \ref{fig:van}), built by AutonomouStuff, owned and operated by Ohio University. LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C (Figure \ref{fig:vlp32mount}). Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a high output data stream \cite{vlp_32c}. GPS data was recorded from the van's PwrPak7D-E2 GNSS \& INS enclosure manufactured by Novatel equipped with two GNSS-502 antennas manufactured by NavtechGPS. Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using Robotic Operating System (ROS). Data gathering locations included lots and unmarked roads near Ohio University in Athens, Ohio (Figure \ref{fig:Combined_Roads}).

	\begin{figure}[H] 
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/Van}
			\caption[Sensor Van]{}
			\label{fig:van}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/vlp_32_mount_2}
			\caption[VLP 32 on Van]{}
			\label{fig:vlp32mount}
		\end{subfigure}
		\caption[Experimental Apparatus]{Experimental Apparatus (a). Velodyne VLP-32 Scanning LiDAR is mounted on top of the vehicle (b).}
		\label{fig:Experimental_Apperatus}
	\end{figure}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/Bean_Hollow_Road_Camera}
			\caption[Bean Hollow Road Camera View]{}
			\label{fig:Bean_Cam_View}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/armig_2}
			\caption[Armig Road Camera View]{}
			\label{fig:Armig_Road_Camera_View}
		\end{subfigure}
		\caption[Armig Road \& Bean Hollow Road]{Road view of Bean Hollow Road, an unmarked chipseal road (a). Road view of Armig Road, and unmarked gravel road (b).}
		\label{fig:Combined_Roads}
	\end{figure}

	Scanning LiDAR training and verification data was extracted manually selecting appropriate consecutive points representing gravel, chipseal, foliage, and grass surfaces. Feature extraction was completed by performing mathematical functions on the gathered training data with reference to the projected plane.

	MATLAB's $TreeBagger$ function was used to create RDFs. Included features, tree depth, and number of trees were principal guiding factors in RDF creation. Features to be included in the training data set was determined by examination of feature usage in each RDF and validation accuracy, eliminating lesser-used features if it had negligible effect on the validation accuracy. Tree depth was determined by examination of the validation accuracy and out-of-bag error to determine a point of diminishing returns. RDF classification accuracy was verified (Table \ref{tab:Verification_Results}). Quadrant size inversely affects rate of classification, as smaller quadrants allow for greater resolution at the expense of more computational time.

	\begin{table}
		\centering
		\begin{tabular}{lc|c|c|c|c|}
			\multicolumn{1}{c}{}       & \multicolumn{1}{c}{} & \multicolumn{4}{c}{Actual}                                                                                           \\
			\multicolumn{1}{c}{}       & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Gravel} & \multicolumn{1}{c}{Chipseal} & \multicolumn{1}{c}{Foliage} & \multicolumn{1}{c}{Grass}  \\ 
			\cline{3-6}
			\multirow{4}{*}{\rotatebox[origin=c]{90}{Predicted}} 	& Gravel 	& 93\% 		& 27\% 		& 0\% 		& 35\% \\ 
			\cline{3-6}
			& Chipseal 	& 3\% 		& 67\% 		& 0\% 		& 9\% \\ 
			\cline{3-6}
			& Foliage 	& 0\% 		& 0\% 		& 100\% 	& 7\% \\ 
			\cline{3-6}
			& Grass 	& 4\% 		& 16\% 		& 0\% 		& 54\% \\
			\cline{3-6}
		\end{tabular}
		\caption{Verification accuracy table for the principal RDF used for this work}
		\label{tab:Verification_Results}
	\end{table}

	Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data. Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU. LiDAR origin was found using GPS longitude, latitude, and altitude data. Physical distances between the GPS, IMU, and LiDAR was rectitude by obtaining reference frames provided by AutonomouStuff. GPS coordinates were offset by the current orientation and converts the ground truth to the LiDAR frame. GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors. Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}).

	\begin{figure}
		\centering
		\includegraphics[width=0.7\linewidth]{figures/Compiled_PCD}
		\caption[Compiled Point Cloud Data]{Compiled Point Cloud Data}
		\label{fig:Compiled_PCD}
	\end{figure}

	Manually defined areas were projected unto consecutive LiDAR scans to score classification results. Classification results that were sympatric to manually defined areas were examined for accuracy by scoring the exact terrain classification and true/false-positive road surface detection. Drop off in positive road surface classification rates in side-of-road areas of 50\% or greater in side-of-road areas would be indicative that the random decision forest classification algorithm was adequate at detecting the road edge.

\section{Results}

	Scanning LiDAR, GPS, and IMU data was collected on closed lots and public roads near Athens Ohio using the Experimental Apparatus. Point cloud data was manually defined by area as one of four basic terrain types: gravel, chipseal, foliage, or grass. Gravel and chipseal data was gathered from closed lots and unmarked public roads (Figures \ref{fig:Road_Examples}), while grass (well-trimmed lawn grass) and foliage (leafy trees and bushes) was gathered from areas surrounding public roads.

	Testing consecutive scans was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road. Physical distances between the GPS, IMU, and LiDAR were rectified by measuring the distance between the modules. Rotational reference frames between the GPS, IMU, and LiDAR were rectified by applying appropriate rotational matrices. Transformation matrices were created by considering the GPS position and the IMU's roll, pitch, and yaw data. Point cloud data was then classified using the RDF derived in Objective I. Translation and rotation was applied to the terrain classification results. Consecutive scanning LiDAR data was compiled allowing for manually assigning areas representing road or side of road surfaces (Figure \ref{fig:Classified_PointCloud_1_ALL}). Side of road area were manually defined as an area roughly one meter outwards from the road edge for consideration of road edge detection.

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.45\linewidth]{figures/both_classification_things_sturbois_curve_1}
		\caption{Classified point cloud with projected manually defined areas}
		\label{fig:Classified_PointCloud_1_ALL}
	\end{figure}

	\begin{table}
		\centering
		\begin{tabular}{lll}
			& Road 		& Side-of-Road 	\\
			Gravel   	& 79.3\% 	& 0\%			\\
			Chipseal 	& 2.0\% 	& 0\%  			\\
			Foliage  	& 11.2\% 	& 0\% 			\\
			Grass    	& 7.5\% 	& 0\%
		\end{tabular}
		\caption[Sturbois Road Curve Example 1]{Accuracy for Sturbois Road and Side-of-Road analysis}
		\label{tab:Sturbois_Curve_1_Road_Results}
	\end{table}

	Quadrant size affects terrain classification accuracy at the expense of granularity of results and computational time. Smaller quadrants allow for greater resolution of terrain classification, however more computational time is required. Larger quadrants allow for swifter classification, however as consecutive scanning LiDAR points may include road and non-road surfaces, classification accuracy may suffer. Per-Quadrant classification rate data was gathered and plotted, the average rate being 2.97 quadrants per second. This was tested with a machine with 64 GB of DDR4 RAM and an Intel Core i7-770K @ 4.20GHz*8.

\section{Conclusion}

	The proposed work addresses the problem of road surface detection on unmarked gravel and chipseal roads. Current LiDAR and Camera based detection models are insufficient due to lack of distinct features typical of urban roads, such as painted line markings or curbs, or have excessive computational or storage requirements. The completed work balanced accuracy and efficiency by using less intensive analysis techniques of smaller point cloud data sets. The final deliverable of the described work was a method of detecting physical unmarked gravel and chipseal roads by using a terrain classification approach to predicting road surface area. The impact of this work is that autonomous vehicles using LiDAR may be able to detect unmarked gravel and chipseal road surfaces, allowing autonomous operations on 1.5 million miles of previously undetected rural roads.

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

% use appendices with more than one appendix
% then use \section to start each appendix
% you must declare a \section before using any
% \subsection or using \label (\appendices by itself
% starts a section numbered zero.)
%

\appendices
\section{Equations}

\begin{equation}\label{eqn:MLS}
	\left[ {\begin{array}{cc}
			\sum_{i=1}^{m} x_i z_i \\
			\sum_{i=1}^{m} y_i z_i \\
			\sum_{i=1}^{m} z_i \\

	\end{array} } \right]
	=
	\left[ {\begin{array}{ccc}
			\sum_{i=1}^{m} x_i^2 		& \sum_{i=1}^{m} x_i y_i 		& \sum_{i=1}^{m} x_i \\
			\sum_{i=1}^{m} x_i y_i 		& \sum_{i=1}^{m} y_i^2 			& \sum_{i=1}^{m} y_i \\
			\sum_{i=1}^{m} x_i 			& \sum_{i=1}^{m} y_i 			& \sum_{i=1}^{m} 1   \\
	\end{array} } \right]
	\left[ {\begin{array}{cc}
			A\\
			B\\
			C\\
	\end{array} } \right]
\end{equation}

\begin{equation}
	W = (\sum_{i=1}^{n} a_i x_{(i)})^2 / \sum_{i=1}^{n}(x_i - x_m)^2
	\label{eq:shapiro_wilks}
\end{equation}

{where $x_i$ is the $i_{th}$ order statistic and $x_m$ is the sample mean \cite{shapiro_analysis_1965}. Coefficients $a_{i}$ are given by equation \ref{eq:shapiro_wilks_coeff}.}

\begin{equation}
	(a_1, a_2, ...,a_n) = m^{T} V^{-1} / C
	\label{eq:shapiro_wilks_coeff}
\end{equation}

{where C is a vector norm in equation \ref{eq:shapiro_wilks_vector_norm} \cite{davis_covariance_1977}.}

\begin{equation}
	C = |mV^{-1}| = (m^{T} V^{-1} V^{-1} m)^{1/2}
	\label{eq:shapiro_wilks_vector_norm}
\end{equation}

{where $m$ is the vector of expected values in the normal order statistics: $m = (m_1, m_2,...,m_n)^{T}$ and $V$ is the covariance matrix of the normal order statistics. Shapiro-Wilks was used to test the variance of LiDAR spatial and remission data for normality.}

% you can choose not to have a title for an appendix
%% if you want by leaving the argument blank
%\section{}
%Appendix two text goes here.

% use section* for acknowledgment
\section*{Acknowledgment}

The authors would like to thank the United States Dairy Farmers for their continuous cheese production. This work would not happen without them.

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
\newpage
\fi

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)

	\bibliographystyle{ieeetr}  
	\bibliography{resources.bib}

% that's all folks
\end{document}

### 2023-02-20 Huston_Paper.tex
updates

@@
 \usepackage{indentfirst}
+\usepackage{multicol}
+\usepackage{enumitem}
 
@@
 
-	Gravel roads that lack standardized features such as curbs or painted lines present detection and navigation challenges to autonomous vehicles.
+	{Gravel and chipseal roads that lack standardized features such as curbs or painted lines present detection and navigation challenges to autonomous vehicles.
+Differentiation between road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation.
 Global Positioning Service (GPS) and high resolution maps cannot be relied upon for navigation of rural roads, as some may only be width of the vehicle and GPS may not be accurate enough.
-Differentiation between gravel road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation.
-Normal Distribution Transform (NDT), based on scanning LiDAR data may be insufficient for navigating on gravel roads.
-This work will examine methods of classifying scanning LiDAR data for explicit detection of gravel road surface spatial and remission features.
-Exploration of terrain classification of gravel road surfaces with high resolution scanning LiDAR data will allow for predicting road boundary location enabling confident autonomous operations on gravel roads.
-The principal outcome of this work is a method for gravel road detection using LiDAR data for the purpose of predicting gravel road boundary locations.
-Detection of gravel road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.
+Normal Distribution Transform (NDT) based on scanning LiDAR data may be insufficient for navigating on gravel roads.
+This work will examine a method of classifying scanning LiDAR data spatial and remission features for explicit detection of unmarked gravel and chipseal road surfaces.
+Exploration of terrain classification using high resolution scanning LiDAR data of these road surfaces will allow for predicting road boundary location potentially enabling confident autonomous operations on gravel roads.
+The principal outcome of this work is a method for gravel road detection using LiDAR data for the purpose of predicting road boundary locations.
+Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.}
 
@@
 	% Autonomous vehicles have difficulty detecting rural unpaved gravel roads due to lack of standardized features such as curbs and painted lane lines. 
+	%	Autonomous vehicles have relied upon standardized road features such as curbs and painted lane lines or high definition maps paired with precision GPS for navigational solutions.
 
-	Autonomous vehicles have relied upon standardized road features such as curbs and painted lane lines or high definition maps paired with precision GPS for navigational solutions.
-Challenges arise when detecting unpaved gravel and chipseal roads in rural environments due to the lack of standardized road features and potentGPS signal unreliability.
-Gravel roads comprise 34.8\% \cite{road_stats_2} of all road surfaces in the United States Some nations have predominately unpaved road networks, such as India, where 70.7\% of all roads by mile are categorized as unpaved \cite{malik_lal_2019}.
-Detection of gravel roads may allow for expansion of current civilian autonomous vehicle capabilities as well as introduce opportunities for industries that operate in rural areas.
+	% Current methods of road surface analysis rely on cameras and Light Detection and Ranging (LiDAR) sensors to detect standardized road features, however on gravel roads such attributes are non-existent \cite{skorseth_gravel_nodate}.
+
+	{LiDAR has been used by autonomous vehicles for road edge detection purposes, however challenges arise in environments with unpaved gravel and chipseal roads due to the lack of standardized road features such as curbs and painted lane lines.
+Gravel roads comprise 34.8\% \cite{road_stats_2} of all road surfaces in the United States.
+Some nations have predominately unpaved road networks, such as India, where 70.7\% of all roads by mile are categorized as unpaved \cite{malik_lal_2019}.
+Detection of unmarked roads may allow for expansion of current civilian autonomous vehicle capabilities as well as introduce opportunities for industries that operate in rural areas.
 Global Positioning Service (GPS) and high resolution maps cannot be relied upon for navigation of rural roads, as some gravel roads may only be width of the vehicle and GPS is not accurate enough or may not have a reliable signal for precision driving \cite{noauthor_gpsgov_nodate}.
-Current methods of road surface analysis rely on cameras and Light Detection and Ranging (LiDAR) sensors to detect standardized road features, however on gravel roads such attributes are non-existent \cite{skorseth_gravel_nodate}.
-Systems that rely on cameras may fail when analyzing gravel roads, as these lack visual cues such as painted lane markings \cite{crisman_scarf_1993} and may in fact closely resemble the surrounding terrain.
+Systems that rely on cameras may fail when analyzing gravel roads, as these lack visual cues such as painted lane markings \cite{crisman_scarf_1993} and may in some circumstances closely resemble surrounding terrain.
 LiDAR point cloud processing systems depend upon distinct geometric features, most commonly curbs \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_nodate,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja_road_nodate}, however as curbs are not installed on rural gravel roads \cite{skorseth_gravel_nodate} this method breaks down.
@@
 Using 2D projection of planes unto a 3D point cloud of a road surface is an alternative approach, but likewise requires curbs or painted lines for road boundary definition \cite{fernandes_road_2014, borkar_robust_2009-1, guo_lane_2015}.
-Instead, using cameras and LiDAR may overcome difficulties in achieving real-time detection of gravel road surfaces using terrain classification to discover road boundaries, allowing for autonomous operations that do not rely upon match enabling processes and traditional visual cues.
+Instead, using a LiDAR forward approach may overcome difficulties in achieving real-time detection of unmarked roads by using terrain classification to discover road type, boundaries, and intersections.}
 
-	{Detection of gravel road surfaces may rely on the surface properties of the road without relying on distinct markings or topographical features.
+	%Real time description of surface noise based on incoming LiDAR data, then subsequent comparison to a standard, is a possible alternative that may allow for computationally efficient road surface detection.
+Overlaying camera imagery data with LiDAR allows for assigning incoming LiDAR data with color information, potentially increasing efficiency by narrowing the search area.
+
+	{Detection of unmarked road surfaces may rely on the roughness properties without relying on distinct markings or topographical features.
 Surface roughness is a measurable property that may be used to characterize a gravel road, as they are typically consistent with gravel type used \cite{skorseth_gravel_nodate} which may be exploited in a search and compare function, and is distinct from grass and other ground types \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain}.
-Surface roughness properties may be derived from processing LiDAR data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}.
-Real time description of surface noise based on incoming LiDAR data, then subsequent comparison to a standard, is a possible alternative that may allow for computationally efficient road surface detection.
-Overlaying camera imagery data with LiDAR allows for assigning incoming LiDAR data with color information, potentially increasing efficiency by narrowing the search area.
-Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.
-Detection of gravel road surfaces would increase the capabilities of self driving vehicles considerably, allowing autonomous operations on 1.5 million miles \cite{road_stats_2} of previously undetected rural roads.
-Problem Statement: Determine a processing method for increasing road detection precision of unmarked gravel rural roads.
+Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}.
+Implementing a terrain classification model using Random Decision Forests (RDF) using scanning LiDAR data to exploit surface roughness was studied for road surface detection in this study.}
 
-\subsection{Literature Review}
+	{Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.
+Problem Statement: Determine a processing method for increasing road detection precision of unmarked gravel rural roads.}
 
-	Current LiDAR based road surface detection models may be categorized into two methodologies.
+\subsection{Related Work}
+
+	{Current LiDAR based road surface detection models may be categorized into two methodologies.
 Firstly is the analysis of road surface properties, which considers aspects such as topology and surface roughness.
@@
 Supervised Classification Applied to Road Following (SCARF) is an algorithm that detects road surfaces based on color differences between the road and the surrounding terrain \cite{crisman_scarf_1993}.
-It was found that difficulties arise when detecting road surfaces in less colored environments \cite{crisman_scarf_1993,manz_detection_2011}.
+It was found that difficulties arise when detecting road surfaces in less colored environments \cite{crisman_scarf_1993,manz_detection_2011}.}
 
-	Real time processing of rural gravel roads cannot rely on normalized, consistent features found on urban roads such as curbs, as rural roads lack these features \cite{skorseth_gravel_nodate}.
-As such, many of the proposed models are rendered unserviceable, as they rely on curbs or painted lines to offer distinctions in data sets \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_nodate,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja_road_nodate}.
+	{Real time processing of rural gravel roads cannot rely on normalized, consistent features found on urban roads such as curbs, as rural roads lack these features \cite{skorseth_gravel_nodate}.
+As such, many of the proposed models are rendered unserviceable, as they rely on curbs or painted lines to offer distinctions in data sets \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_nodate,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja_road_nodate}.}
 
-	Further limitations are imposed when considering data set size and processing speed.
+	{Further limitations are imposed when considering data set size and processing speed.
 Real time analysis of road surfaces dictates that trajectory updates to the vehicle in motion must have a rapid update rate, which prohibits the relatively lengthy collection of large data sets or any form of post-processing.
@@
 Principal component analysis on the height of each LiDAR data point was proposed to detect a straight, curb-less rural road with low computation time \cite{yadav_road_2018}.
-However a computational time of 24 minutes was required to process a 156 meter stretch of road, rendering real time detection impossible.
+However a computational time of 24 minutes was required to process a 156 meter stretch of road, rendering real time detection impossible.}
 
-	Classification of the surrounding environment is an integral part to the proposed work, as differentiation of gravel road surfaces and surrounding terrain is necessary to determine road boundaries.
-Terrain classification is the analysis of the ground surface in order to specify the ground surface type \cite{laible_3d_2012,laible_terrain_2013,laible_map_building,rasmussen_combining_2002,reymann_improving_2015,walas_terrain_2014,wietrzykowski_boosting_2014,wang_road_nodate}.
-Current literature proposes multiple methods of terrain classification using LiDAR and RGB cameras, and may be roughly broken into two areas of application.
+	% Current literature proposes multiple methods of terrain classification using LiDAR and RGB cameras, and may be roughly broken into two areas of application.
 Traversability or trafficability is the analysis of generally unstructured environments that includes obstacles or rough terrain types for autonomous vehicle path finding solutions  \cite{schilling_geometric_2017,ojeda_terrain_2006,coombs_driving_2000,stavens_self-supervised_nodate,belter_rough_2010,bartoszyk_terrain-aware_2017,noauthor_fusion_nodate,li_rugged_2019,wilson_terrain_2014,siva_robot_2019}.
@@
 
+	{Classification of the surrounding environment is an integral part to the proposed work, as differentiation of gravel road surfaces and surrounding terrain is necessary to determine road boundaries.
+Terrain classification is the analysis of the ground surface in order to specify the ground surface type \cite{laible_3d_2012,laible_terrain_2013,laible_map_building,rasmussen_combining_2002,reymann_improving_2015,walas_terrain_2014,wietrzykowski_boosting_2014,wang_road_nodate}.
 LiDAR and camera based real-time terrain classification was studied by projecting a 2D plane unto the point cloud data using a variant of Random Sample Consensus (RANSAC) called M-estimator Sample Consensus (MSAC) \cite{mijakovska_generating_2014} that is built for a more robust result \cite{laible_3d_2012,laible_map_building,laible_terrain_2013}.
@@
 Higher resolution LiDAR that produces a high density point cloud, such as the Velodyne VLP-32 that will be used in the proposed work \cite{vlp_32c}, may allow for better terrain classification with Gaussian characterization of return intensity and spatial data.
-While capable of real-time classification of terrain type, this method was not employed in road identification and used a low resolution LiDAR system.
+While capable of real-time classification of terrain type, this method was not employed in road identification and used a low resolution LiDAR system.}
 
-	Neural networks and similar machine learning tools may be used in terrain and road identification.
+	{Neural networks and similar machine learning tools may be used in terrain and road identification.
 Road classification with LiDAR spacial and remission features was completed with Neural Network classifier, Naive Bayes classifier, and Support Vector Machines (SVM) \cite{wang_road_nodate,wang_two-stage_2018}.
@@
 Histograms and averages for image hue, saturation, and color value, along with LiDAR intensity were used as points of classification for different terrain types.
-Although a 96\% true positive rate was reported for a rocky surface, the overall process was computationally expensive with 1.8 seconds per image.
+Although a 96\% true positive rate was reported for a rocky surface, the overall process was computationally expensive with 1.8 seconds per image.}
 
-	MLS and RANSAC were compared as methods to project a two dimensional plane unto a point cloud, facilitating examination of points by providing a frame of reference \cite{miller_method_nodate, gojcic_perfect_2019}.
+	{MLS and RANSAC were compared as methods to project a two dimensional plane unto a point cloud, facilitating examination of points by providing a frame of reference \cite{miller_method_nodate, gojcic_perfect_2019}.
 Method of least-squares analyses uses regression to approximate the best fit to a data set (\ref{eqn:MLS}).
 RANSAC is an iterative algorithm that is designed to handle large numbers of outliers in the sample data set \cite{derpanis_overview_nodate,yaniv_random_2010,fischler_random_1987,cantzler_random_nodate}.
-
-	{Testing LiDAR data for normality was completed to determine its usefulness as a feature.
-Shapiro-Wilk analysis \cite{shapiro_analysis_1965, royston_extension_1982} tests the null hypothesis that a sample vector comes from a normally distributed population, using the test statistic in equation \ref{eq:shapiro_wilks}.}
-
 	RDFs are a supervised machine learning method that operates on a set of decision trees \cite{ho_random_1995}.
@@
 Multiple decision trees are created with bootstrapped data sets - randomly selected data sub-sets from a training set.
-Classification may be determined by the majority results as opposed to the mean \cite{breiman_random_2001,ho_random_1995}.
-RDFs may be used to predict road surface area by detecting gravel terrain.
+Classification was determined by the majority results as opposed to the mean \cite{breiman_random_2001,ho_random_1995}.
+RDFs may be used to predict road surface area by detecting gravel terrain.}
 
@@
 
-	Required physical data was gathered by the Hexegon Chrysler Pacifica New Eagle DBW van (Figure \ref{fig:van}), built by AutonomouStuff, owned and operated by Ohio University.
+	{Required physical data was gathered by the Hexegon Chrysler Pacifica New Eagle DBW van (Figure \ref{fig:van}), built by AutonomouStuff, owned and operated by Ohio University.
 LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C (Figure \ref{fig:vlp32mount}).
 Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a high output data stream \cite{vlp_32c}.
-GPS data was recorded from the van's PwrPak7D-E2 GNSS \& INS enclosure manufactured by Novatel equipped with two GNSS-502 antennas manufactured by NavtechGPS.
+GPS data was recorded with a PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS) manufactured by Novatel, and two GNSS-502 antennas by NavtechGPS.
 Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using Robotic Operating System (ROS).
-Data gathering locations included lots and unmarked roads near Ohio University in Athens, Ohio (Figure \ref{fig:Combined_Roads}).
+Scanning LiDAR training data was collected on various roads and lots near Athens, Ohio.
+Scanning LiDAR data of gravel surfaces was gathered principally from an un-named gravel road near Armig Road (Figure \ref{fig:Armig_Road_Camera_View})and a gravel lot maintained by the Ohio University Civil Engineering Department near Blackburn Road.
+Additional gravel road data for testing purposes only was gathered on Sturbois Road.
+Scanning LiDAR data of chipseal surfaces was gathered principally from Bean Hollow Road.
+Additional chipseal road data for testing purposes only was gathered on later portions of both Bean Hollow Road (Figure \ref{fig:Bean_Cam_View}).
+Scanning LiDAR data of grassy surfaces and foliage was gathered from the sides of the previously stated roads, therefore no special data collection drive was made for those terrain types.}
 
@@
 		\end{subfigure}
-		\caption[Armig Road \& Bean Hollow Road]{Road view of Bean Hollow Road, an unmarked chipseal road (a). Road view of Armig Road, and unmarked gravel road (b).}
+		\caption[Armig Road \& Bean Hollow Road]{Typical examples of unmarked gravel and chipseal roads.
+Road view of Bean Hollow Road - an unmarked chipseal road (a).
+Road view of Armig Road - an unmarked gravel road (b).}
 		\label{fig:Combined_Roads}
@@
 
-	Scanning LiDAR training and verification data was extracted manually selecting appropriate consecutive points representing gravel, chipseal, foliage, and grass surfaces.
-Feature extraction was completed by performing mathematical functions on the gathered training data with reference to the projected plane.
+	{MATLAB was used to process the gathered rosbag files using the MATLAB ROSBAG library.
+Rosbags were directly loaded into the MATLAB environment.
+Point cloud topic was defined, and the point cloud messages were read from the topic and imported as a data structure.
+Cartesian coordinates and corresponding intensity values were extracted and saved to disk.
+GPS and IMU data was extracted in a similar manner.
+GPS and IMU topics were defined, and topic messages are read.
+Longitude, latitude, and altitude were extracted from the GPS messages, while roll, pitch, and yaw were extracted from the IMU messages.}
 
-	MATLAB's $TreeBagger$ function was used to create RDFs.
+	{Scanning LiDAR training and verification data was extracted by selecting appropriate consecutive points from a single LiDAR channel lying on gravel, chipseal, foliage, and grass surfaces.
+Number of points included in each training data set was not controlled.
+Foliage was defined as trees and bushes, while grass was defined as a trimmed lawn.
+Projection of a plane was necessary to provide a reference point for describing point variance.
+RANSAC and Method of Least Squares were tested for use as plane projection methods.
+Additionally, a point's perpendicular distance as well as range from the LiDAR origin (Figure \ref{fig:xy_vs_range}) was also considered, both of which has a computational advantage over the plane projection method in that they may be read directly from the Velodyne VLP-32 data stream.}
+
+	{RANSAC was implementing by a built-in MATLAB function, $pcfitplane$.
+Method of Least Squares was implemented by importing a MATLAB library \cite{noauthor_object-oriented_nodate}.
+Point clouds are individually imported into the MATLAB environment.
+RANSAC or Method of Least Squares may be used to project a reference plane unto the point cloud.
+Displaying the point cloud is completed using MATLAB's $pcshow$ function.
+Training data is selected by manually drawing an area around appropriate points of a desired terrain type using MATLAB's $drawrectangle$ function (Figure \ref{fig:Rectangle_ROI}).
+Indices for the rectangle are extracted, and MATLAB's function $findPointsInROI$ is used to extract any points within the drawn rectangle (Figure \ref{fig:ROI_Points}).
+Captured points are then saved to disk along with the projected plane's normal vector and distance from the origin $[a,b,c,d]$.}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=0.5\linewidth]{figures/Rectangle_ROI}
+			\caption[Region of Interest Selection]{MATLAB was used to select training data using the $drawrectangle$ function and extracting points that lay within the defined area.}
+			\label{fig:Rectangle_ROI}
+		\end{subfigure}
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=0.5\linewidth]{figures/ROI_Points}
+			\caption[Points in Region of Interest]{Selected points (Figure \ref{fig:Rectangle_ROI}) lying on an unmarked road.}
+			\label{fig:ROI_Points}
+		\end{subfigure}
+		\caption[Armig Road \& Bean Hollow Road]{Road view of Bean Hollow Road - an unmarked chipseal road (a). Road view of Armig Road - an unmarked gravel road (b).}
+		\label{fig:Combined_Things}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=0.5\linewidth]{figures/xy_vs_range}
+		\caption[XY vs Range vs Z Height]{Three references for feature extraction may be seen in this simplified graph.
+First is the Range shown by the magenta line from the LiDAR origin to the Point of Interest (PoI).
+Second is the height from  LiDAR origin to the PoI (cyan).
+Finally, distance from the projected plane (red) to the PoI (green).
+XY Distance (blue) and Range (magenta) from the sensor was used consider the effect of distance from LiDAR origin to the PoI.}
+		\label{fig:xy_vs_range}
+	\end{figure}
+
+	{Feature extraction was completed by performing mathematical functions on the gathered spatial and remission training data.
+Reference plane normal vector and distance from origin $[a,b,c,d]$ data derived from either RANSAC or Method of Least Squares methods was selected.
+Base-line features extracted include the following: Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an array comprising of the differences between consecutive points in the spatial or remission data array.
+Distance from the LiDAR point of origin was also considered by dividing the base-line features by either the average range or X-Y distances for a set of points, as LiDAR accuracy may decrease as distance increases.
+Sixty features were considered in total for this work, and multiple feature sub-sets were used to create a variety of Random Decision Forests.
+Plane-projection based algorithms were used considering all and top twenty features.
+Non-plane projection based algorithms were created considering only range and Z height features.
+For all RDFs, both spatial and remission features were used.}
+
+	{Random Decision Forests were created using MATLAB's $TreeBagger$ function, which creates an ensemble of bagged decision trees.
 Included features, tree depth, and number of trees were principal guiding factors in RDF creation.
-Features to be included in the training data set was determined by examination of feature usage in each RDF and validation accuracy, eliminating lesser-used features if it had negligible effect on the validation accuracy.
-Tree depth was determined by examination of the validation accuracy and out-of-bag error to determine a point of diminishing returns.
-RDF classification accuracy was verified (Table \ref{tab:Verification_Results}).
-Quadrant size inversely affects rate of classification, as smaller quadrants allow for greater resolution at the expense of more computational time.
+Tree depth was determined by examination of the validation accuracy to determine a point of diminishing returns.
+Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data.}
 
-	\begin{table}
-		\centering
-		\begin{tabular}{lc|c|c|c|c|}
-			\multicolumn{1}{c}{}       & \multicolumn{1}{c}{} & \multicolumn{4}{c}{Actual}                                                                                           \\
-			\multicolumn{1}{c}{}       & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Gravel} & \multicolumn{1}{c}{Chipseal} & \multicolumn{1}{c}{Foliage} & \multicolumn{1}{c}{Grass}  \\ 
-			\cline{3-6}
-			\multirow{4}{*}{\rotatebox[origin=c]{90}{Predicted}} 	& Gravel 	& 93\% 		& 27\% 		& 0\% 		& 35\% \\ 
-			\cline{3-6}
-			& Chipseal 	& 3\% 		& 67\% 		& 0\% 		& 9\% \\ 
-			\cline{3-6}
-			& Foliage 	& 0\% 		& 0\% 		& 100\% 	& 7\% \\ 
-			\cline{3-6}
-			& Grass 	& 4\% 		& 16\% 		& 0\% 		& 54\% \\
-			\cline{3-6}
-		\end{tabular}
-		\caption{Verification accuracy table for the principal RDF used for this work}
-		\label{tab:Verification_Results}
-	\end{table}
-
-	Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data.
-Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU.
+%		{Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU.
 LiDAR origin was found using GPS longitude, latitude, and altitude data.
@@
 GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors.
-Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}).
+Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}).}
+
+	{Validation error was used as the primary means to compare RDF performance.
+MATLAB was used to load verification data and test increasing sizes of RDF algorithms for accuracy by comparing the known data types against the RDF result.
+RANSAC versus MLS was tested using identical feature sets (Figures \ref{fig:all_mls_ransac_comp} \ref{fig:xyz_mls_ransac_comp}).
+Validation error plots were used to influence feature sets and depth in later iterations of RDFs.
+Number of trees in the RDF verses error allowed for determination of an early stopping point for algorithm growth (Figure \ref{fig:mls_zxy_early_stopping}) by choosing the point where the change in moving average did not exceed 10\% and was within 15\% of the minimum error.}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{figures/MLS_RAN_ALL}
+			\caption[]{}
+			\label{fig:all_mls_ransac_comp}
+		\end{subfigure}
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{figures/MLS_RAN_TT}
+			\caption[]{}
+			\label{fig:xyz_mls_ransac_comp}
+		\end{subfigure}
+		\label{fig:xyz_mls_comp_1}
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{figures/Range_Vali_Error}
+			\caption[]{}
+			\label{fig:range_mls_ransac_comp}
+		\end{subfigure}
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{figures/ZXY}
+			\caption[]{}
+			\label{fig:t20_mls_ransac_comp}
+		\end{subfigure}
+		\caption[RDF Validation Accuracy]{RDF validation accuracy directly compared.
+All features are considered (a); Top 20 features used in (a) only (b); Range was used without X, Y, and Z components broken out (c); Plane projection not considered in (d), reference plane set as the LiDAR origin point.}
+		\label{fig:xyz_mls_comp_2}
+	\end{figure}
+
+	\begin{figure}[H]
+			\centering
+			\includegraphics[width=0.75\linewidth]{figures/mls_zxy_early_stopping}
+			\caption[Early Stopping]{Diminishing returns on increasing algorithm size versus accuracy can be used to determine the early stopping point of algorithm growth.}
+			\label{fig:mls_zxy_early_stopping}
+	\end{figure}
+
+	% Quadrant size inversely affects rate of classification, as smaller quadrants allow for greater resolution at the expense of more computational time.
 
@@
 
-	Manually defined areas were projected unto consecutive LiDAR scans to score classification results.
+	{Road and side-of-road areas were manually defined and projected unto consecutive LiDAR scans to score classification results.
 Classification results that were sympatric to manually defined areas were examined for accuracy by scoring the exact terrain classification and true/false-positive road surface detection.
-Drop off in positive road surface classification rates in side-of-road areas of 50\% or greater in side-of-road areas would be indicative that the random decision forest classification algorithm was adequate at detecting the road edge.
+Drop off in positive road surface classification rates in side-of-road areas of 50\% or greater in side-of-road areas would be indicative of adequate road edge detection.}
 
@@
 
-	Scanning LiDAR, GPS, and IMU data was collected on closed lots and public roads near Athens Ohio using the Experimental Apparatus.
+	{Scanning LiDAR, GPS, and IMU data was collected on closed lots and public roads near Athens Ohio using the Experimental Apparatus.
 Point cloud data was manually defined by area as one of four basic terrain types: gravel, chipseal, foliage, or grass.
-Gravel and chipseal data was gathered from closed lots and unmarked public roads (Figures \ref{fig:Road_Examples}), while grass (well-trimmed lawn grass) and foliage (leafy trees and bushes) was gathered from areas surrounding public roads.
+Gravel and chipseal data was gathered from closed lots and unmarked public roads (Figures \ref{fig:Combined_Roads}), while grass (well-trimmed lawn grass) and foliage (leafy trees and bushes) was gathered from areas surrounding public roads.}
 
-	Testing consecutive scans was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
+	{Testing consecutive scans was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
 Physical distances between the GPS, IMU, and LiDAR were rectified by measuring the distance between the modules.
@@
 Consecutive scanning LiDAR data was compiled allowing for manually assigning areas representing road or side of road surfaces (Figure \ref{fig:Classified_PointCloud_1_ALL}).
-Side of road area were manually defined as an area roughly one meter outwards from the road edge for consideration of road edge detection.
+Side of road area were manually defined as an area roughly one meter outwards from the road edge for consideration of road edge detection.}
 
@@
 		\includegraphics[width=0.45\linewidth]{figures/both_classification_things_sturbois_curve_1}
-		\caption{Classified point cloud with projected manually defined areas}
+		\caption{Typical example of a classified point cloud with projected manually defined areas}
 		\label{fig:Classified_PointCloud_1_ALL}
@@
 
-	Quadrant size affects terrain classification accuracy at the expense of granularity of results and computational time.
+	{Quadrant size affects terrain classification accuracy at the expense of granularity of results and computational time.
 Smaller quadrants allow for greater resolution of terrain classification, however more computational time is required.
 Larger quadrants allow for swifter classification, however as consecutive scanning LiDAR points may include road and non-road surfaces, classification accuracy may suffer.
-Per-Quadrant classification rate data was gathered and plotted, the average rate being 2.97 quadrants per second.
-This was tested with a machine with 64 GB of DDR4 RAM and an Intel Core i7-770K @ 4.20GHz*8.
+Best per-Quadrant average classification rate was X.XX quadrants per second as tested on a machine with 64 GB of DDR4 RAM and a Ryzen 5900x @ 3.70GHz*24.}
 
@@
 
-	The proposed work addresses the problem of road surface detection on unmarked gravel and chipseal roads.
+	{The proposed work addresses the problem of road surface detection on unmarked gravel and chipseal roads.
 Current LiDAR and Camera based detection models are insufficient due to lack of distinct features typical of urban roads, such as painted line markings or curbs, or have excessive computational or storage requirements.
@@
 The final deliverable of the described work was a method of detecting physical unmarked gravel and chipseal roads by using a terrain classification approach to predicting road surface area.
-The impact of this work is that autonomous vehicles using LiDAR may be able to detect unmarked gravel and chipseal road surfaces, allowing autonomous operations on 1.5 million miles of previously undetected rural roads.
+The impact of this work is that autonomous vehicles using LiDAR may be able to detect unmarked gravel and chipseal road surfaces, allowing autonomous operations on 1.5 million miles of previously undetected rural roads.}
 
@@
 \end{equation}
-
-\begin{equation}
-	W = (\sum_{i=1}^{n} a_i x_{(i)})^2 / \sum_{i=1}^{n}(x_i - x_m)^2
-	\label{eq:shapiro_wilks}
-\end{equation}
-
-{where $x_i$ is the $i_{th}$ order statistic and $x_m$ is the sample mean \cite{shapiro_analysis_1965}. Coefficients $a_{i}$ are given by equation \ref{eq:shapiro_wilks_coeff}.}
-
-\begin{equation}
-	(a_1, a_2, ...,a_n) = m^{T} V^{-1} / C
-	\label{eq:shapiro_wilks_coeff}
-\end{equation}
-
-{where C is a vector norm in equation \ref{eq:shapiro_wilks_vector_norm} \cite{davis_covariance_1977}.}
-
-\begin{equation}
-	C = |mV^{-1}| = (m^{T} V^{-1} V^{-1} m)^{1/2}
-	\label{eq:shapiro_wilks_vector_norm}
-\end{equation}
-
-{where $m$ is the vector of expected values in the normal order statistics: $m = (m_1, m_2,...,m_n)^{T}$ and $V$ is the covariance matrix of the normal order statistics.
-Shapiro-Wilks was used to test the variance of LiDAR spatial and remission data for normality.}
+\centering
+{Method of Least Squares Planar Projection}
 
@@
 % use section* for acknowledgment
+\justifying
 \section*{Acknowledgment}
 
-The authors would like to thank the United States Dairy Farmers for their continuous cheese production. This work would not happen without them.
+	{The authors would like to thank the United States Dairy Farmers for their continuous cheese production. This work would not happen without them.}
 

### 2023-02-20 Huston_Paper.tex
Updates

@@
 Included features, tree depth, and number of trees were principal guiding factors in RDF creation.
-Tree depth was determined by examination of the validation accuracy to determine a point of diminishing returns.
-Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data.}
-
-%		{Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU.
-LiDAR origin was found using GPS longitude, latitude, and altitude data.
-Physical distances between the GPS, IMU, and LiDAR was rectitude by obtaining reference frames provided by AutonomouStuff.
-GPS coordinates were offset by the current orientation and converts the ground truth to the LiDAR frame.
-GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors.
-Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}).}
-
-	{Validation error was used as the primary means to compare RDF performance.
+Validation error was used as the primary means to compare RDF performance.
 MATLAB was used to load verification data and test increasing sizes of RDF algorithms for accuracy by comparing the known data types against the RDF result.
 RANSAC versus MLS was tested using identical feature sets (Figures \ref{fig:all_mls_ransac_comp} \ref{fig:xyz_mls_ransac_comp}).
-Validation error plots were used to influence feature sets and depth in later iterations of RDFs.
 Number of trees in the RDF verses error allowed for determination of an early stopping point for algorithm growth (Figure \ref{fig:mls_zxy_early_stopping}) by choosing the point where the change in moving average did not exceed 10\% and was within 15\% of the minimum error.}
@@
 	% Quadrant size inversely affects rate of classification, as smaller quadrants allow for greater resolution at the expense of more computational time.
+
+	{Testing consecutive scans required a method of compiling LiDAR data into a single point cloud and manually classifying areas for . Six stretches of unmarked roads was gathered for this work. }
+
+	%		{Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data.
+Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU.
+LiDAR origin was found using GPS longitude, latitude, and altitude data.
+Physical distances between the GPS, IMU, and LiDAR was rectitude by obtaining reference frames provided by AutonomouStuff.
+GPS coordinates were offset by the current orientation and converts the ground truth to the LiDAR frame.
+GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors.
+Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}).}
+
+	{Testing consecutive scans was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
+Physical distance between the GPS, IMU, and LiDAR were rectified by measuring the distance between the modules.
+Rotational reference frames between the GPS, IMU, and LiDAR were rectified by applying appropriate rotational matrices.
+Transformation matrices were created by considering the GPS position and the IMU's roll, pitch, and yaw data.
+Point cloud data was then classified using one of the generated RDF algorithms.
+Translation and rotation was applied to the terrain classification results with the closest matching transformation matrices (Figure \ref{fig:Classified_PointCloud_1_ALL}).
+Comparing averaged X, Y, and Z coordinates of classified quadrant (Figure \ref{fig:Classified_PointCloud_1_AVG}) allowed for easier comparison and scoring of the RDF classification to the manually classified areas.
+Consecutive scanning LiDAR data was compiled allowing for manually assigning areas representing road or side of road surfaces (Figure \ref{fig:Manually_Classified_Areas}).
+Side of road area were manually defined as an area roughly one meter outwards from the road edge for consideration of road edge detection.
+Per-Quadrant classification rate data was gathered and plotted, the average rate being 2.97 quadrants per second (Figure \ref{fig:Time_Hz_Rate}).
+This was tested with a machine with 64 GB of DDR4 RAM and a Ryzen 9 5900X 12 core x 24 thread CPU.}
 

### 2023-02-23 Huston_Paper.tex
updatez

--- 2023-02-20 Huston_Paper.tex
@@
 \usepackage{enumitem}
+\usepackage{attachfile}
+\usepackage{multirow}
+\usepackage{tabularray}
+
+% Table Stuff
+\newcommand{\spheading}[2][9.5em]{% \spheading[<width>]{<stuff>}
+	\rotatebox{90}{\parbox{#1}{\raggedright #2}}}
 
@@
 Training data is selected by manually drawing an area around appropriate points of a desired terrain type using MATLAB's $drawrectangle$ function (Figure \ref{fig:Rectangle_ROI}).
-Indices for the rectangle are extracted, and MATLAB's function $findPointsInROI$ is used to extract any points within the drawn rectangle (Figure \ref{fig:ROI_Points}).
+Indices for the rectangle are extracted, and MATLAB's function $findPointsInROI$ is used to extract sympatric points within the drawn rectangle (Figure \ref{fig:ROI_Points}).
 Captured points are then saved to disk along with the projected plane's normal vector and distance from the origin $[a,b,c,d]$.}
@@
 Reference plane normal vector and distance from origin $[a,b,c,d]$ data derived from either RANSAC or Method of Least Squares methods was selected.
-Base-line features extracted include the following: Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an array comprising of the differences between consecutive points in the spatial or remission data array.
+Base-line features extracted include the following: Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial or remission data array.
 Distance from the LiDAR point of origin was also considered by dividing the base-line features by either the average range or X-Y distances for a set of points, as LiDAR accuracy may decrease as distance increases.
 Sixty features were considered in total for this work, and multiple feature sub-sets were used to create a variety of Random Decision Forests.
-Plane-projection based algorithms were used considering all and top twenty features.
+Plane-projection based algorithms were created considering all features.
+Determination of feature usage allowed for creation of RDFs that used only the top twenty features.
 Non-plane projection based algorithms were created considering only range and Z height features.
-For all RDFs, both spatial and remission features were used.}
+For all RDFs, both spatial and remission features were used.
+}
 
-	{Random Decision Forests were created using MATLAB's $TreeBagger$ function, which creates an ensemble of bagged decision trees.
-Included features, tree depth, and number of trees were principal guiding factors in RDF creation.
-Tree depth was determined by examination of the validation accuracy to determine a point of diminishing returns.
-Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data.}
-
-%		{Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU.
-LiDAR origin was found using GPS longitude, latitude, and altitude data.
-Physical distances between the GPS, IMU, and LiDAR was rectitude by obtaining reference frames provided by AutonomouStuff.
-GPS coordinates were offset by the current orientation and converts the ground truth to the LiDAR frame.
-GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors.
-Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}).}
-
-	{Validation error was used as the primary means to compare RDF performance.
+	{Random Decision Forests were created using MATLAB's $TreeBagger$ function, which creates an ensemble of bagged decision trees using the input arguments of included features, tree depth, number of trees, and a cost array.
+RDFs were split into sub categories within each feature set to test varying tree depths (Figure \ref{fig:all_tree_comp}), and allowed to grow in number of trees until three hundred trees were created.
+Discouragement of false-positive identification for road surfaces (gravel and chipseal) was accomplished by introducing a cost array which would induce a bias against gravel and chipseal .
+Validation error was used as the primary means to compare RDF performance prior to testing with data outside the training and validation data sets.
 MATLAB was used to load verification data and test increasing sizes of RDF algorithms for accuracy by comparing the known data types against the RDF result.
-RANSAC versus MLS was tested using identical feature sets (Figures \ref{fig:all_mls_ransac_comp} \ref{fig:xyz_mls_ransac_comp}).
-Validation error plots were used to influence feature sets and depth in later iterations of RDFs.
-Number of trees in the RDF verses error allowed for determination of an early stopping point for algorithm growth (Figure \ref{fig:mls_zxy_early_stopping}) by choosing the point where the change in moving average did not exceed 10\% and was within 15\% of the minimum error.}
+Plotting number of trees verses error allowed for determination of an early stopping point for algorithm growth (Figure \ref{fig:mls_zxy_early_stopping}) by choosing the point where the change in error moving average did not exceed 10\% and was within 20\% of the minimum error.}
 
@@
 All features are considered (a); Top 20 features used in (a) only (b); Range was used without X, Y, and Z components broken out (c); Plane projection not considered in (d), reference plane set as the LiDAR origin point.}
-		\label{fig:xyz_mls_comp_2}
+		\label{fig:all_tree_comp}
 	\end{figure}
@@
 
-	\begin{figure}
+	{Testing consecutive scans required a method of compiling LiDAR data into a single point cloud and manually classifying areas for road detection accuracy, completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
+Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data.
+Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU.
+LiDAR origin was found using GPS longitude, latitude, and altitude data.
+Physical distances between the GPS, IMU, and LiDAR was rectitude by obtaining reference frames provided by AutonomouStuff.
+GPS coordinates were offset by the current orientation and converts the ground truth to the LiDAR frame.
+GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors.
+Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}).
+While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for scoring purposes.
+Six stretches of unmarked roads were examined for accuracy: three of chipseal and three of gravel.
+Road and side-of-road areas were manually defined and projected unto consecutive LiDAR scans to score classification results.
+Due to the fuzzy nature of the classification results, the side-of-road areas were roughly a standard lane width to the sides of the road to accommodate the inability to define a ridge road edge.}
+
+	\begin{figure}[H]
 		\centering
-		\includegraphics[width=0.7\linewidth]{figures/Compiled_PCD}
-		\caption[Compiled Point Cloud Data]{Compiled Point Cloud Data}
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=0.95\linewidth]{figures/Compiled_PCD}
+			\caption[C]{}
 			\label{fig:Compiled_PCD}
+		\end{subfigure}
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=0.95\linewidth]{figures/short_chip_redux}
+			\caption[]{}
+			\label{fig:Classified_Areas_PCD}
+		\end{subfigure}
+		\caption[Compiled and Classified Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR (a).
+Point cloud data was classified, and then scored using manually defined road and side-of-road areas (b).}
+		\label{fig:Combined_Classified_PCD}
 	\end{figure}
 
-	{Road and side-of-road areas were manually defined and projected unto consecutive LiDAR scans to score classification results.
-Classification results that were sympatric to manually defined areas were examined for accuracy by scoring the exact terrain classification and true/false-positive road surface detection.
-Drop off in positive road surface classification rates in side-of-road areas of 50\% or greater in side-of-road areas would be indicative of adequate road edge detection.}
+	{Classifying gathered consecutive LiDAR scans was accomplished by analyzing each individual 360$^{\circ}$ sweep of the discretized channels.
+Each channel was broken into quadrants.
+Quadrant size inversely affects rate of classification, as smaller quadrants allow for greater resolution at the expense of more computational time.
+Translation and rotation was applied to the terrain classification results with the closest matching transformation matrices.Classification results that were sympatric to manually defined areas were examined for accuracy by scoring the exact terrain classification and true/false-positive road surface detection.
+Drop off in positive road surface classification rates in side-of-road areas in side-of-road areas would be indicative of adequate road edge detection.
+Consecutive scanning LiDAR data was compiled allowing for manually assigning areas representing road or side of road surfaces (Figure \ref{fig:Classified_Areas_PCD}).}
+
+%	{ Comparing averaged X, Y, and Z coordinates of the classified quadrant allowed for easier comparison and scoring of the RDF classification to the manually classified areas. 
+% Insert sentence or two on rate of of classification, feat extract, etc.; reference figure with rates> This was tested with a machine with 64 GB of DDR4 RAM and a Ryzen 9 5900X 12 core x 24 thread CPU.}
 
@@
 	{Scanning LiDAR, GPS, and IMU data was collected on closed lots and public roads near Athens Ohio using the Experimental Apparatus.
-Point cloud data was manually defined by area as one of four basic terrain types: gravel, chipseal, foliage, or grass.
-Gravel and chipseal data was gathered from closed lots and unmarked public roads (Figures \ref{fig:Combined_Roads}), while grass (well-trimmed lawn grass) and foliage (leafy trees and bushes) was gathered from areas surrounding public roads.}
-
-	{Testing consecutive scans was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
+Gravel and chipseal data was gathered from closed lots and unmarked public roads (Figures \ref{fig:Combined_Roads}), while grass (well-trimmed lawn grass) and foliage (leafy trees and bushes) was gathered from areas surrounding public roads.
+Training data was generated using MATLAB to extract LiDAR data from rosbags.
+Classification algorithms using Random Decision Forests were generated using a variety of feature sets and reference points.
+Testing consecutive scans was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data.
 Physical distances between the GPS, IMU, and LiDAR were rectified by measuring the distance between the modules.
@@
 Translation and rotation was applied to the terrain classification results.
-Consecutive scanning LiDAR data was compiled allowing for manually assigning areas representing road or side of road surfaces (Figure \ref{fig:Classified_PointCloud_1_ALL}).
+Consecutive scanning LiDAR data was compiled allowing for manually assigning areas representing road or side of road surfaces (Figure \ref{fig:classified_point_clouds_all}).
 Side of road area were manually defined as an area roughly one meter outwards from the road edge for consideration of road edge detection.}
@@
 		\centering
-		\includegraphics[width=0.45\linewidth]{figures/both_classification_things_sturbois_curve_1}
-		\caption{Typical example of a classified point cloud with projected manually defined areas}
-		\label{fig:Classified_PointCloud_1_ALL}
+
+		\begin{subfigure}{0.495\textwidth}
+			\centering
+			\includegraphics[width=0.95\linewidth]{figures/curve_gravel}
+			\caption[]{}
+			\label{fig:curve_gravel_2}	
+		\end{subfigure}
+		\begin{subfigure}{0.495\textwidth}
+			\centering
+			\includegraphics[width=0.95\linewidth]{figures/curve_gravel_2}
+			\caption[]{}
+			\label{fig:curve_gravel}
+		\end{subfigure}
+		\begin{subfigure}{0.495\textwidth}
+			\centering
+			\includegraphics[width=0.95\linewidth]{figures/long_gravel_1}
+			\caption[]{}
+			\label{fig:long_gravel_1}	
+		\end{subfigure}
+		\begin{subfigure}{0.495\textwidth}
+			\centering
+			\includegraphics[width=0.95\linewidth]{figures/long_chip_redux}
+			\caption[]{}
+			\label{fig:long_chip_classified}
+		\end{subfigure}
+		\begin{subfigure}{0.495\textwidth}
+			\centering
+			\includegraphics[width=0.95\linewidth]{figures/long_chip_redux_2}
+			\caption[]{}
+			\label{fig:long_chip_2_classified}
+		\end{subfigure}
+		\begin{subfigure}{0.495\textwidth}
+			\centering
+			\includegraphics[width=0.95\linewidth]{figures/short_chip_redux}
+			\caption[]{}
+			\label{fig:short_chip_classified}
+		\end{subfigure}
+		\caption[Classified Point Clouds]{Classified point clouds with manually defined surface areas that allow for scoring of classification results.
+Gravel roads include an un-named gravel side-road near Armig Road (a) and Sturbois Road (b), (c).
+Chipseal roads include Willow Creek Road (d), (e), and Bean Hollow Road (f).
+The algorithm that produced these results used all features and reference plane projected with RANSAC.
+Chipseal roads are shown as black areas, gravel roads and driveways are shown as red areas, and side-of-roads are designated as orange.
+}
+		\label{fig:classified_point_clouds_all}
 	\end{figure}
 
-	\begin{table}
+	{Results were scored by comparing the location of the averaged classified quadrant's (x,y) coordinate to the area that the manually defined areas (Table \ref{tab:averaged_results} shows results for algorithm designation RAN ALL which produced the results in Figure \ref{fig:classified_point_clouds_all}.
+All results are shown in Table \ref{tab:all_results} in Appendix \ref{apdx:Appendix_2}).
+Ratio of road to non-road in manually defined road and side-of-road areas are shown in columns 3 and 4, with higher values indicating a greater presence of positive road identification.
+The quotient of the two ratios (col.
+5) allows for estimation of road edge detection ability.
+Scores for correct and incorrect road surface type is shown, indicating the algorithm's ability to correctly determine the road surface type (cols.
+6,7).
+Percentages of grass and foliage classification for both side-of-road and on road are given in cols.
+8 \& 9, indicating true/false-negative road surface identification capability.
+True/false-positive road surface identification percentages are indicated in cols 10-11.}
+
+	{Of the algorithms tested (Appendix \ref{apdx:Appendix_2}), the RDF that used RANSAC as the plane projection method and the full feature set seemed to perform the most consistently for true non-road surface identification.
+While it did struggle with gravel identification (32.72\% average accuracy) with a high confusion rate with grass (27.47\% false identification), there was a consistent decrease ($>20\%$) in false-positive road identification in side-of-road areas.
+}
+
+	{It was found that while none of the algorithms excelled in all areas, }
+
+	\begin{table}[H]
 		\centering
-		\begin{tabular}{lll}
-			& Road 		& Side-of-Road 	\\
-			Gravel   	& 79.3\% 	& 0\%			\\
-			Chipseal 	& 2.0\% 	& 0\%  			\\
-			Foliage  	& 11.2\% 	& 0\% 			\\
-			Grass    	& 7.5\% 	& 0\%
-		\end{tabular}
-		\caption[Sturbois Road Curve Example 1]{Accuracy for Sturbois Road and Side-of-Road analysis}
-		\label{tab:Sturbois_Curve_1_Road_Results}
+		\begin{tblr}{width=0.9\linewidth,colspec={|Q[1,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|}}
+			\hline
+			Algorithm & Average \% of True Gravel Type & Average \% of True Chipseal Type & Average \% of True Surface Type Identification & Average \% True Road Surface Detection & Average \% of True  Non-Road Surface Detection\\
+			\hline
+			MLS ALL & 27.35 & 61.50 & 46.59 & 82.90 & 55.01 \\
+			\hline
+			MLS TT  & 27.62 & 50.37 & 40.86 & 89.44 & 37.31 \\
+			\hline
+			ZXY     & 31.70 & 65.65 & 50.60 & 93.15 & 55.16 \\
+			\hline
+			RAN ALL & 32.72 & 48.42 & 42.38 & 56.25 & 66.50 \\
+			\hline
+			RANGE   & 56.92 & 39.97 & 51.61 & 78.53 & 47.92 \\
+			\hline
+			RAN TT  & 82.84 & 27.60 & 57.29 & 65.32 & 58.51 \\
+			\hline
+		\end{tblr}
+		\caption[Averaged Results]{Average results for all algorithms and all tested roads}
+		\label{tab:averaged_results}
 	\end{table}
@@
 
+\section{All Results}\label{apdx:Appendix_2}
+
+\begin{table}[]
+	\centering
+	\begin{tabular}{|>{\centering\arraybackslash}p{.25cm}|>{\centering\arraybackslash}p{.25cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.4cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|}
+		\hline
+		\spheading{Algorithm} & 
+		\spheading{Road Samp} & 
+		\spheading{Ratio of Road / Non-Road: On Road} & 
+		\spheading{Ratio of Road / Non-Road: Off Road} & 
+		\spheading{Ratio Score} & 
+		\spheading{True Road Surface Type Identification Score} & 
+		\spheading{False Road Surface Type Identification Score} & 
+		\spheading{False Non-Road CR Score} & 
+		\spheading{True Non-Road CR Score} &
+		\spheading{True Road Surface Identification} &
+		\spheading{False Road Surface Identification} \\
+		\hline
+		\multirow{6}{*}{\rotatebox[origin=c]{90}{MLS ALL}}
+		&		A		& 4.00   & 0.21 & 19.01  & 36.10 & 43.91 & 19.99 & 82.61 & 80.01  & 19.99 \\
+		&		B		& 3.89   & 2.36 & 1.65   & 23.93 & 55.61 & 20.45 & 29.75 & 79.55  & 20.45 \\
+		&		C		& 4.06   & 1.63 & 2.49   & 22.03 & 58.20 & 19.77 & 38.07 & 80.23  & 19.77 \\
+		&		D		& 2.22   & 0.44 & 5.09   & 58.80 & 22.25 & 18.96 & 69.67 & 81.04  & 18.96 \\
+		&		E		& 6.51   & 0.83 & 7.84   & 58.93 & 24.15 & 16.92 & 54.62 & 83.08  & 16.92 \\
+		&		F		& 26.20  & 0.81 & 32.44  & 79.74 & 13.73 & 6.54  & 55.32 & 93.46  & 6.54  \\
+		\hline
+		\multirow{6}{*}{\rotatebox[origin=c]{90}{MLS TT}}       
+		&		A		& 19.94  & 0.38 & 52.93  & 31.69 & 63.53 & 4.78  & 72.64 & 95.22  & 4.78  \\
+		&		B		& 19.52  & 4.69 & 4.16   & 25.24 & 69.89 & 4.87  & 17.56 & 95.13  & 4.87  \\
+		&		C		& 19.00  & 3.04 & 6.26   & 25.94 & 69.06 & 5.00  & 24.77 & 95.00  & 5.00  \\
+		&		D		& 2.09   & 1.27 & 1.65   & 8.94  & 45.08 & 45.98 & 44.10 & 54.02  & 45.98 \\
+		&		E		& 21.59  & 2.86 & 7.55   & 68.87 & 28.41 & 2.71  & 25.91 & 97.29  & 2.71  \\
+		&		F		& 100.00 & 1.57 & 63.56  & 84.50 & 15.50 & 0.00  & 38.86 & 100.00 & 0.00  \\
+		\hline
+		\multirow{6}{*}{\rotatebox[origin=c]{90}{RAN ALL}}     
+		&		A		& 11.42  & 0.23 & 50.35  & 41.39 & 50.56 & 8.05  & 81.51 & 91.95  & 8.05  \\
+		&		B		& 13.01  & 2.38 & 5.47   & 26.46 & 66.41 & 7.14  & 29.60 & 92.86  & 7.14  \\
+		&		C		& 14.61  & 1.72 & 8.49   & 27.27 & 66.33 & 6.41  & 36.76 & 93.59  & 6.41  \\
+		&		D		& 2.67   & 0.42 & 6.29   & 62.55 & 25.32 & 12.13 & 70.19 & 87.87  & 12.13 \\
+		&		E		& 10.74  & 0.76 & 14.05  & 63.03 & 29.61 & 7.36  & 56.68 & 92.64  & 7.36  \\
+		&		F		& 100.00 & 0.78 & 128.22 & 82.91 & 17.09 & 0.00  & 56.18 & 100.00 & 0.00  \\		
+		\hline
+		\multirow{6}{*}{\rotatebox[origin=c]{90}{RAN TT}}       
+		&		A		& 0.27   & 0.22 & 1.23   & 21.35 & 0.19  & 78.46 & 81.80 & 21.54  & 78.46 \\
+		&		B		& 0.70   & 0.54 & 1.29   & 38.21 & 2.87  & 58.92 & 64.92 & 41.08  & 58.92 \\
+		&		C		& 0.68   & 0.51 & 1.32   & 38.59 & 1.72  & 59.69 & 66.11 & 40.31  & 59.69 \\
+		&		D		& 1.17   & 0.52 & 2.27   & 43.30 & 29.72 & 26.99 & 65.98 & 73.01  & 26.99 \\
+		&		E		& 2.97   & 0.77 & 3.87   & 38.49 & 37.55 & 23.96 & 56.60 & 76.04  & 23.96 \\
+		&		F		& 2.05   & 0.57 & 3.58   & 74.32 & 11.20 & 14.47 & 63.60 & 85.53  & 14.47 \\
+		\hline
+		\multirow{6}{*}{\rotatebox[origin=c]{90}{RANGE}}  
+		&		A		& 4.00   & 0.21 & 19.01  & 36.10 & 43.91 & 19.99 & 82.61 & 80.01  & 19.99 \\
+		&		B		& 2.59   & 2.47 & 1.05   & 71.63 & 0.52  & 27.85 & 28.84 & 72.15  & 27.85 \\
+		&		C		& 1.75   & 2.26 & 0.77   & 63.05 & 0.55  & 36.41 & 30.66 & 63.59  & 36.41 \\
+		&		D		& 1.95   & 1.05 & 1.86   & 34.50 & 46.97 & 18.53 & 48.87 & 81.47  & 18.53 \\
+		&		E		& 6.54   & 1.59 & 4.12   & 29.32 & 53.86 & 16.82 & 38.62 & 83.18  & 16.82 \\
+		&		F		& 5.30   & 0.73 & 7.30   & 75.07 & 15.69 & 9.24  & 57.90 & 90.76  & 9.24  \\
+		\hline
+		\multirow{6}{*}{\rotatebox[origin=c]{90}{ZXY}}          
+		&		A		& 2.47   & 0.37 & 6.67   & 71.11 & 0.09  & 28.79 & 72.95 & 71.21  & 28.79 \\
+		&		B		& 13.36  & 1.78 & 7.53   & 89.99 & 3.05  & 6.96  & 36.03 & 93.04  & 6.96  \\
+		&		C		& 10.13  & 1.66 & 6.11   & 87.42 & 3.59  & 8.98  & 37.61 & 91.02  & 8.98  \\
+		&		D		& 0.38   & 0.29 & 1.28   & 3.24  & 14.44 & 82.33 & 77.28 & 17.67  & 82.33 \\
+		&		E		& 0.37   & 0.36 & 1.04   & 2.26  & 18.98 & 78.76 & 73.76 & 21.24  & 78.76 \\
+		&		F		& 43.63  & 0.87 & 49.99  & 89.73 & 8.03  & 2.24  & 53.40 & 97.76  & 2.24 \\
+		\hline                              
+	\end{tabular}
+	\caption[All Results]{All Results XD}
+	\label{tab:all_results}
+\end{table}
+
 % use section* for acknowledgment

### 2023-02-25 Huston_Paper.tex
Logic for simple ensemble

@@
 
-	{Of the algorithms tested (Appendix \ref{apdx:Appendix_2}), the RDF that used RANSAC as the plane projection method and the full feature set seemed to perform the most consistently for true non-road surface identification.
-While it did struggle with gravel identification (32.72\% average accuracy) with a high confusion rate with grass (27.47\% false identification), there was a consistent decrease ($>20\%$) in false-positive road identification in side-of-road areas.
+	{It was found that while none of the algorithms excelled in all areas, the results seem to indicate that implementing a multiple-model algorithm to create an ensemble would greatly increase road detection accuracy.
+Of the algorithms tested (Appendix \ref{apdx:Appendix_2}), the RDF that used RANSAC as the plane projection method and the full feature (designated RAN ALL) performed the most consistently for true non-road surface identification, with an average of 66.5\% (median of 65.5\%).
+While it did struggle with gravel identification (32.72\% average accuracy) with a high confusion rate with grass (27.47\% false identification), it may be supplemented with the RDF designated RANGE, which has the highest true-positive gravel identification rate of 82.8\%.
+A classification method was tested to verify this method based on RAN ALL and RANGE algorithms.
+RAN ALL would would designate non-road surfaces and the determination of road surface type was a combined effort of the two (Figure \ref{}), thus achieving a higher accuracy.
 }
-
-	{It was found that while none of the algorithms excelled in all areas, }
 
@@
 			\hline
-			Algorithm & Average \% of True Gravel Type & Average \% of True Chipseal Type & Average \% of True Surface Type Identification & Average \% True Road Surface Detection & Average \% of True  Non-Road Surface Detection\\
+			Algorithm Designation & Average \% of True Gravel Type Identification & Average \% of True Chipseal Type Identification & Average \% of True Surface Type Identification & Average \% True Road Surface Detection & Average \% of True  Non-Road Surface Detection\\
 			\hline
-			MLS ALL & 27.35 & 61.50 & 46.59 & 82.90 & 55.01 \\
+			MLS ALL & 35.354 & 68.581 & 46.587 & 82.896 & 55.006 \\
 			\hline
-			MLS TT  & 27.62 & 50.37 & 40.86 & 89.44 & 37.31 \\
+			MLS TT  & 37.833 & 55.162 & 40.865 & 89.444 & 37.308 \\
 			\hline
-			ZXY     & 31.70 & 65.65 & 50.60 & 93.15 & 55.16 \\
+			ZXY     & 35.697 & 73.211 & 50.601 & 93.153 & 55.156 \\
 			\hline
-			RAN ALL & 32.72 & 48.42 & 42.38 & 56.25 & 66.50 \\
+			RAN ALL & 36.119 & 54.169 & 42.376 & 56.252 & 66.500 \\
 			\hline
-			RANGE   & 56.92 & 39.97 & 51.61 & 78.53 & 47.92 \\
+			RANGE   & 61.822 & 44.758 & 51.610 & 78.526 & 47.918 \\
 			\hline
-			RAN TT  & 82.84 & 27.60 & 57.29 & 65.32 & 58.51 \\
+			RAN TT  & 54.508 & 27.617 & 57.293 & 65.323 & 58.506 \\
 			\hline
+%			Ens\_1  & 61.822 & 44.758 & 51.610 & 78.526 & 66.685 \\
+%			\hline
 		\end{tblr}
-		\caption[Averaged Results]{Average results for all algorithms and all tested roads}
+		\caption[Averaged Results]{Average results of all algorithms on tested roads. }
 		\label{tab:averaged_results}
 	\end{table}
+
+%	\begin{figure}[H]
+%		\centering
+%		\includegraphics[width=0.75\linewidth]{figures/RAN_ALL_ofroad_RANGE_onroad}
+%		\caption[Early Stopping]{Diminishing returns on increasing algorithm size versus accuracy can be used to determine the early stopping point of algorithm growth.}
+%		\label{fig:combo_result}
+%	\end{figure}
 

### 2023-04-27 Huston_Paper.tex
okay here we go

@@
 
-	\author{Rhett Huston, Jay Wilhelm}
+	\author{Rhett Huston, Dr. Jay Wilhelm}
 %	{
@@
 % The paper headers
-\markboth{Some Sort of Journal}%
-{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for IEEE Journals}
+\markboth{Journal of Autonomous Vehicles and Systems}%
+{}
 % The only time the second header will appear is for the odd numbered pages
@@
 % in the abstract or keywords.
-\begin{abstract}
+\begin{abstract}{
 
-	{Gravel and chipseal roads that lack standardized features such as curbs or painted lines present detection and navigation challenges to autonomous vehicles.
+	Gravel roads lack standardized features such as curbs or painted lines, presenting detection challenges to autonomous vehicles.
 Differentiation between road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation.
-Global Positioning Service (GPS) and high resolution maps cannot be relied upon for navigation of rural roads, as some may only be width of the vehicle and GPS may not be accurate enough.
-Normal Distribution Transform (NDT) based on scanning LiDAR data may be insufficient for navigating on gravel roads.
-This work will examine a method of classifying scanning LiDAR data spatial and remission features for explicit detection of unmarked gravel and chipseal road surfaces.
-Exploration of terrain classification using high resolution scanning LiDAR data of these road surfaces will allow for predicting road boundary location potentially enabling confident autonomous operations on gravel roads.
-The principal outcome of this work is a method for gravel road detection using LiDAR data for the purpose of predicting road boundary locations.
-Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.}
-
+Global Positioning Service (GPS) and high resolution maps may not be reliable for navigation of gravel roads, as some may only be width of the vehicle and GPS may not be accurate enough.
+Normal Distribution Transform (NDT) based on scanning LiDAR data may be insufficient for navigating on gravel roads as there may not be enough geometrically distinct features for reliable scan matching.
+This work will examine a method of classifying scanning LiDAR data spatial and remission features for explicit detection of unmarked gravel road surfaces.
+Exploration of terrain classification using high resolution scanning LiDAR data of these road surfaces may allow for predicting gravel road boundary locations potentially enabling confident autonomous operations on gravel roads.
+The principal outcome of this work was a method for gravel road detection using LiDAR data for the purpose of predicting road boundary locations.
+Random Decision Forests were tested for the purpose of scanning LiDAR data terrain classification.
+Manually discovered gravel and paved road surface based on classified results were compared to actual road bounds.
+<SUMMARY OF RESULTS HERE>.
+Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.
+}	
 \end{abstract}
@@
 	%	Autonomous vehicles have relied upon standardized road features such as curbs and painted lane lines or high definition maps paired with precision GPS for navigational solutions. 
+	% LiDAR has been used by autonomous vehicles for road edge detection purposes, however challenges arise when trajectory solutions are required on unpaved gravel and chipseal roads due to the lack of standardized road features such as curbs and painted lane lines.
 
@@
 
-	{LiDAR has been used by autonomous vehicles for road edge detection purposes, however challenges arise in environments with unpaved gravel and chipseal roads due to the lack of standardized road features such as curbs and painted lane lines.
-Gravel roads comprise 34.8\% \cite{road_stats_2} of all road surfaces in the United States.
+	{Gravel roads comprise 34.8\% \cite{road_stats_2} of all road surfaces in the United States.
 Some nations have predominately unpaved road networks, such as India, where 70.7\% of all roads by mile are categorized as unpaved \cite{malik_lal_2019}.
-Detection of unmarked roads may allow for expansion of current civilian autonomous vehicle capabilities as well as introduce opportunities for industries that operate in rural areas.
-Global Positioning Service (GPS) and high resolution maps cannot be relied upon for navigation of rural roads, as some gravel roads may only be width of the vehicle and GPS is not accurate enough or may not have a reliable signal for precision driving \cite{noauthor_gpsgov_nodate}.
+Detection of gravel roads may allow for expansion of current civilian autonomous vehicle capabilities as well as introduce opportunities for industries that operate in rural areas.
+Global Positioning Service (GPS) and high resolution maps cannot be relied upon for navigation of rural roads alone as GPS is not accurate enough or may not have a reliable signal for precision driving \cite{noauthor_gpsgov_nodate}, and gravel roads are not marked or mapped.
 Systems that rely on cameras may fail when analyzing gravel roads, as these lack visual cues such as painted lane markings \cite{crisman_scarf_1993} and may in some circumstances closely resemble surrounding terrain.
@@
 Using 2D projection of planes unto a 3D point cloud of a road surface is an alternative approach, but likewise requires curbs or painted lines for road boundary definition \cite{fernandes_road_2014, borkar_robust_2009-1, guo_lane_2015}.
-Instead, using a LiDAR forward approach may overcome difficulties in achieving real-time detection of unmarked roads by using terrain classification to discover road type, boundaries, and intersections.}
+Instead, using LiDAR may overcome difficulties in achieving real-time detection of unmarked roads by using terrain classification to discover road type, boundaries, and intersections.}
 
@@
 
-	{Detection of unmarked road surfaces may rely on the roughness properties without relying on distinct markings or topographical features.
-Surface roughness is a measurable property that may be used to characterize a gravel road, as they are typically consistent with gravel type used \cite{skorseth_gravel_nodate} which may be exploited in a search and compare function, and is distinct from grass and other ground types \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain}.
+	{Detection of unmarked road surfaces may utilize roughness properties without distinct markings or topographical features.
+Surface roughness is a measurable property that may be used to characterize a gravel road, as they are typically consistent with gravel type used \cite{skorseth_gravel_nodate} which may be exploited in a search and compare function, and is distinct from chipseal, grass, and other ground types \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain}.
 Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}.
-Implementing a terrain classification model using Random Decision Forests (RDF) using scanning LiDAR data to exploit surface roughness was studied for road surface detection in this study.}
-
-	{Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.
-Problem Statement: Determine a processing method for increasing road detection precision of unmarked gravel rural roads.}
+Implementing a terrain classification model using Random Decision Forests (RDF) using scanning LiDAR data to exploit surface roughness was studied for road surface detection in this study.
+Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.
+This work will attempt to determine a processing method for increasing road detection precision of unmarked gravel rural roads.}
 
@@
 
-	{MLS and RANSAC were compared as methods to project a two dimensional plane unto a point cloud, facilitating examination of points by providing a frame of reference \cite{miller_method_nodate, gojcic_perfect_2019}.
+%	{MLS and RANSAC were compared as methods to project a two dimensional plane unto a point cloud, facilitating examination of points by providing a frame of reference \cite{miller_method_nodate, gojcic_perfect_2019}.
 Method of least-squares analyses uses regression to approximate the best fit to a data set (\ref{eqn:MLS}).
-RANSAC is an iterative algorithm that is designed to handle large numbers of outliers in the sample data set \cite{derpanis_overview_nodate,yaniv_random_2010,fischler_random_1987,cantzler_random_nodate}.
-	RDFs are a supervised machine learning method that operates on a set of decision trees \cite{ho_random_1995}.
-REFs have been shown to be useful with handling LiDAR data sets \cite{breiman_random_2001}, and have been employed in terrain classification \cite{laible_3d_2012,laible_map_building,laible_terrain_2013,khan_high_2011,reymann_improving_2015,schilling_geometric_2017, wietrzykowski_context-aware_2019}.
+RANSAC is an iterative algorithm that is designed to handle large numbers of outliers in the sample data set \cite{derpanis_overview_nodate,yaniv_random_2010,fischler_random_1987,cantzler_random_nodate}.}
+
+	{Random Decision Forests are a supervised machine learning method that operates on a set of decision trees \cite{ho_random_1995}.
+RDFs have been shown to be useful with handling LiDAR data sets \cite{breiman_random_2001}, and have been employed in terrain classification \cite{laible_3d_2012,laible_map_building,laible_terrain_2013,khan_high_2011,reymann_improving_2015,schilling_geometric_2017, wietrzykowski_context-aware_2019}.
 Multiple decision trees are created with bootstrapped data sets - randomly selected data sub-sets from a training set.
-Classification was determined by the majority results as opposed to the mean \cite{breiman_random_2001,ho_random_1995}.
-RDFs may be used to predict road surface area by detecting gravel terrain.}
+Classification is determined by the majority results as opposed to the mean \cite{breiman_random_2001,ho_random_1995}.
+Bayesian Optimization may be used to tune the hyper parameters of a machine algorithm \cite{noauthor_bayesian_nodate, snoek_practical_2012}.}
+
+	% Summary of lit review in 1 sentance:
+	{Current methods for road detection may be insufficient for real-time detection of unmarked gravel and chipseal roads due to ambient conditions, lack of standardized features, or rely on post-processing large data sets, while terrain classification of scanning LiDAR data may overcome these difficulties.}
 
@@
 
-	{Required physical data was gathered by the Hexegon Chrysler Pacifica New Eagle DBW van (Figure \ref{fig:van}), built by AutonomouStuff, owned and operated by Ohio University.
+	{Required physical data was gathered by the Ohio University Autonomous Van van (Figure \ref{fig:van}), built by AutonomouStuff.
 LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C (Figure \ref{fig:vlp32mount}).
 Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a high output data stream \cite{vlp_32c}.
-GPS data was recorded with a PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS) manufactured by Novatel, and two GNSS-502 antennas by NavtechGPS.
+GPS data was recorded with a Novatel PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS), and two GNSS-502 antennas by NavtechGPS.
 Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using Robotic Operating System (ROS).
-Scanning LiDAR training data was collected on various roads and lots near Athens, Ohio.
-Scanning LiDAR data of gravel surfaces was gathered principally from an un-named gravel road near Armig Road (Figure \ref{fig:Armig_Road_Camera_View})and a gravel lot maintained by the Ohio University Civil Engineering Department near Blackburn Road.
-Additional gravel road data for testing purposes only was gathered on Sturbois Road.
-Scanning LiDAR data of chipseal surfaces was gathered principally from Bean Hollow Road.
-Additional chipseal road data for testing purposes only was gathered on later portions of both Bean Hollow Road (Figure \ref{fig:Bean_Cam_View}).
-Scanning LiDAR data of grassy surfaces and foliage was gathered from the sides of the previously stated roads, therefore no special data collection drive was made for those terrain types.}
+Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot that intersects Blackburn Road near Athens Ohio.
+Scanning LiDAR data of an asphalt surface was gathered from Blackburn Road near the intersecting gravel lot.
+Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot.
+Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection.
+Only three classes are considered in this work: "gravel", "asphalt", and "unknown".
+}
 
@@
 			\centering
-			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/Van}
+			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/van_on_van}
 			\caption[Sensor Van]{}
@@
 			\centering
-			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/vlp_32_mount_2}
+			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/LiDAR_on_van}
 			\caption[VLP 32 on Van]{}
@@
 			\centering
-			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/Bean_Hollow_Road_Camera}
-			\caption[Bean Hollow Road Camera View]{}
+			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/vlcsnap-2023-04-20-08h48m17s447}
+			\caption[Blackburn Road Camera View]{}
 			\label{fig:Bean_Cam_View}
@@
 		\end{subfigure}
-		\caption[Armig Road \& Bean Hollow Road]{Typical examples of unmarked gravel and chipseal roads.
-Road view of Bean Hollow Road - an unmarked chipseal road (a).
-Road view of Armig Road - an unmarked gravel road (b).}
+		\caption[Blackburn Road \& Gravel Parking Lot]{Road view of Blackburn Road - an unmarked asphalt road (a). Road view of the gravel driveway (b).}
 		\label{fig:Combined_Roads}
@@
 
-	{MATLAB was used to process the gathered rosbag files using the MATLAB ROSBAG library.
+	{MATLAB was used to process the gathered rosbag files using the their ROSBAG library.
 Rosbags were directly loaded into the MATLAB environment.
-Point cloud topic was defined, and the point cloud messages were read from the topic and imported as a data structure.
-Cartesian coordinates and corresponding intensity values were extracted and saved to disk.
-GPS and IMU data was extracted in a similar manner.
-GPS and IMU topics were defined, and topic messages are read.
-Longitude, latitude, and altitude were extracted from the GPS messages, while roll, pitch, and yaw were extracted from the IMU messages.}
+Point cloud, GPS, and IMU topics were defined, and the messages were read from the topic and imported.
+Cartesian coordinates, longitude, latitude, altitude, roll, pitch, and yaw were extracted from the messages.}
 
-	{Scanning LiDAR training and verification data was extracted by selecting appropriate consecutive points from a single LiDAR channel lying on gravel, chipseal, foliage, and grass surfaces.
-Number of points included in each training data set was not controlled.
-Foliage was defined as trees and bushes, while grass was defined as a trimmed lawn.
-Projection of a plane was necessary to provide a reference point for describing point variance.
-RANSAC and Method of Least Squares were tested for use as plane projection methods.
-Additionally, a point's perpendicular distance as well as range from the LiDAR origin (Figure \ref{fig:xy_vs_range}) was also considered, both of which has a computational advantage over the plane projection method in that they may be read directly from the Velodyne VLP-32 data stream.}
-
-	{RANSAC was implementing by a built-in MATLAB function, $pcfitplane$.
-Method of Least Squares was implemented by importing a MATLAB library \cite{noauthor_object-oriented_nodate}.
-Point clouds are individually imported into the MATLAB environment.
-RANSAC or Method of Least Squares may be used to project a reference plane unto the point cloud.
-Displaying the point cloud is completed using MATLAB's $pcshow$ function.
-Training data is selected by manually drawing an area around appropriate points of a desired terrain type using MATLAB's $drawrectangle$ function (Figure \ref{fig:Rectangle_ROI}).
-Indices for the rectangle are extracted, and MATLAB's function $findPointsInROI$ is used to extract sympatric points within the drawn rectangle (Figure \ref{fig:ROI_Points}).
-Captured points are then saved to disk along with the projected plane's normal vector and distance from the origin $[a,b,c,d]$.}
-
-	\begin{figure}[H]
-		\centering
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=0.5\linewidth]{figures/Rectangle_ROI}
-			\caption[Region of Interest Selection]{MATLAB was used to select training data using the $drawrectangle$ function and extracting points that lay within the defined area.}
-			\label{fig:Rectangle_ROI}
-		\end{subfigure}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=0.5\linewidth]{figures/ROI_Points}
-			\caption[Points in Region of Interest]{Selected points (Figure \ref{fig:Rectangle_ROI}) lying on an unmarked road.}
-			\label{fig:ROI_Points}
-		\end{subfigure}
-		\caption[Armig Road \& Bean Hollow Road]{Road view of Bean Hollow Road - an unmarked chipseal road (a). Road view of Armig Road - an unmarked gravel road (b).}
-		\label{fig:Combined_Things}
-	\end{figure}
-
-	\begin{figure}[H]
-		\centering
-		\includegraphics[width=0.5\linewidth]{figures/xy_vs_range}
-		\caption[XY vs Range vs Z Height]{Three references for feature extraction may be seen in this simplified graph.
-First is the Range shown by the magenta line from the LiDAR origin to the Point of Interest (PoI).
-Second is the height from  LiDAR origin to the PoI (cyan).
-Finally, distance from the projected plane (red) to the PoI (green).
-XY Distance (blue) and Range (magenta) from the sensor was used consider the effect of distance from LiDAR origin to the PoI.}
-		\label{fig:xy_vs_range}
-	\end{figure}
-
-	{Feature extraction was completed by performing mathematical functions on the gathered spatial and remission training data.
-Reference plane normal vector and distance from origin $[a,b,c,d]$ data derived from either RANSAC or Method of Least Squares methods was selected.
-Base-line features extracted include the following: Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial or remission data array.
-Distance from the LiDAR point of origin was also considered by dividing the base-line features by either the average range or X-Y distances for a set of points, as LiDAR accuracy may decrease as distance increases.
-Sixty features were considered in total for this work, and multiple feature sub-sets were used to create a variety of Random Decision Forests.
-Plane-projection based algorithms were created considering all features.
-Determination of feature usage allowed for creation of RDFs that used only the top twenty features.
-Non-plane projection based algorithms were created considering only range and Z height features.
-For all RDFs, both spatial and remission features were used.
-}
-
-	{Random Decision Forests were created using MATLAB's $TreeBagger$ function, which creates an ensemble of bagged decision trees using the input arguments of included features, tree depth, number of trees, and a cost array.
-RDFs were split into sub categories within each feature set to test varying tree depths (Figure \ref{fig:all_tree_comp}), and allowed to grow in number of trees until three hundred trees were created.
-Discouragement of false-positive identification for road surfaces (gravel and chipseal) was accomplished by introducing a cost array which would induce a bias against gravel and chipseal .
-Validation error was used as the primary means to compare RDF performance prior to testing with data outside the training and validation data sets.
-MATLAB was used to load verification data and test increasing sizes of RDF algorithms for accuracy by comparing the known data types against the RDF result.
-Plotting number of trees verses error allowed for determination of an early stopping point for algorithm growth (Figure \ref{fig:mls_zxy_early_stopping}) by choosing the point where the change in error moving average did not exceed 10\% and was within 20\% of the minimum error.}
-
-	\begin{figure}[H]
-		\centering
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=1\linewidth]{figures/MLS_RAN_ALL}
-			\caption[]{}
-			\label{fig:all_mls_ransac_comp}
-		\end{subfigure}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=1\linewidth]{figures/MLS_RAN_TT}
-			\caption[]{}
-			\label{fig:xyz_mls_ransac_comp}
-		\end{subfigure}
-		\label{fig:xyz_mls_comp_1}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=1\linewidth]{figures/Range_Vali_Error}
-			\caption[]{}
-			\label{fig:range_mls_ransac_comp}
-		\end{subfigure}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=1\linewidth]{figures/ZXY}
-			\caption[]{}
-			\label{fig:t20_mls_ransac_comp}
-		\end{subfigure}
-		\caption[RDF Validation Accuracy]{RDF validation accuracy directly compared.
-All features are considered (a); Top 20 features used in (a) only (b); Range was used without X, Y, and Z components broken out (c); Plane projection not considered in (d), reference plane set as the LiDAR origin point.}
-		\label{fig:all_tree_comp}
-	\end{figure}
-
-	\begin{figure}[H]
-			\centering
-			\includegraphics[width=0.75\linewidth]{figures/mls_zxy_early_stopping}
-			\caption[Early Stopping]{Diminishing returns on increasing algorithm size versus accuracy can be used to determine the early stopping point of algorithm growth.}
-			\label{fig:mls_zxy_early_stopping}
-	\end{figure}
-
-	% Quadrant size inversely affects rate of classification, as smaller quadrants allow for greater resolution at the expense of more computational time.
-
-	{Testing consecutive scans required a method of compiling LiDAR data into a single point cloud and manually classifying areas for road detection accuracy, completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
+	{Training data was extracted by manually defining gravel, asphalt, and unknown terrain areas on compiled point clouds.
+Compiling LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
 Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data.
@@
 While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for scoring purposes.
-Six stretches of unmarked roads were examined for accuracy: three of chipseal and three of gravel.
-Road and side-of-road areas were manually defined and projected unto consecutive LiDAR scans to score classification results.
-Due to the fuzzy nature of the classification results, the side-of-road areas were roughly a standard lane width to the sides of the road to accommodate the inability to define a ridge road edge.}
+Manually defined areas may then be overlayed unto the point cloud (Figure \ref{fig:Combined_Classified_PCD}).
+Individual scans were consecutively examined to extract per-channel scanning LiDAR training data and exported to a database if coincident with the manually defined areas.}
 
@@
 			\centering
-			\includegraphics[width=0.95\linewidth]{figures/short_chip_redux}
+			\includegraphics[width=0.95\linewidth]{figures/Manually_Defined_Area_Train_Dat_Grab_2}
 			\caption[]{}
@@
 		\end{subfigure}
-		\caption[Compiled and Classified Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR (a).
-Point cloud data was classified, and then scored using manually defined road and side-of-road areas (b).}
+		\caption[Compiled and Classified Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map (a).
+Manually defined areas inform where training data lies (b).}
 		\label{fig:Combined_Classified_PCD}
 	\end{figure}
+
+%	{RANSAC was implementing by a built-in MATLAB function, $pcfitplane$.
+Method of Least Squares was implemented by importing a MATLAB library \cite{noauthor_object-oriented_nodate}.
+Point clouds are individually imported into the MATLAB environment.
+RANSAC or Method of Least Squares may be used to project a reference plane unto the point cloud.
+Displaying the point cloud is completed using MATLAB's $pcshow$ function.
+Training data is selected by manually drawing an area around appropriate points of a desired terrain type using MATLAB's $drawrectangle$ function (Figure \ref{fig:Rectangle_ROI}).
+Indices for the rectangle are extracted, and MATLAB's function $findPointsInROI$ is used to extract sympatric points within the drawn rectangle.
+Captured points are then saved to disk along with the projected plane's normal vector and distance from the origin $[a,b,c,d]$.}
+
+%	\begin{figure}[H]
+%		\centering
+%		\begin{minipage}[c]{.45\linewidth}
+%			\includegraphics[width=.95\linewidth]{figures/Rectangle_ROI}
+%			\caption[Region of Interest Selection]{MATLAB was used to select training data using the $drawrectangle$ function and extracting points that lay within the defined area.}
+%			\label{fig:Rectangle_ROI}
+%		\end{minipage}
+%		\hfill
+%		\begin{minipage}[c]{0.45\linewidth}
+%			\includegraphics[width=.95\linewidth]{figures/xy_vs_range}
+%			\caption[XY vs Range vs Z Height]{Three references for feature extraction may be seen in this simplified graph.
+First is the Range shown by the magenta line from the LiDAR origin to the Point of Interest (PoI).
+Second is the height from  LiDAR origin to the PoI (cyan).
+Finally, distance from the projected plane (red) to the PoI (green).
+XY Distance (blue) and Range (magenta) from the sensor was used consider the effect of distance from LiDAR origin to the PoI.}
+%			\label{fig:xy_vs_range}
+%		\end{minipage}
+%	\end{figure}
+
+	{Feature extraction was completed by performing mathematical functions on the gathered spatial and remission training data (Figure \ref{fig:flowzchart}).
+Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial or remission data array.
+Features were non-dimensionalized as necessary by dividing by the average distance to the appropriate power to cancel units.
+Thirty percent of the training data was randomly selected and reserved to create a validation data set.}
+
+	{MATLAB's Classification Learner Application was used to generated Random Decision Forests.
+Decision tree depth and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.}
+
+	{Random Decision Forests were created using MATLAB's $TreeBagger$ function, which creates an ensemble of bagged decision trees using the input arguments of included features, tree depth, number of trees, and a cost array.
+RDFs were split into sub categories within each feature set to test varying tree depths (Figure \ref{fig:all_tree_comp}), and allowed to grow in number of trees until three hundred trees were created.
+Discouragement of false-positive identification for road surfaces (gravel and chipseal) was accomplished by introducing a cost array which would induce a bias against gravel and chipseal.
+Validation error was used as the primary means to compare RDF performance prior to testing with data outside the training and validation data sets.
+MATLAB was used to load verification data and test increasing sizes of RDF algorithms for accuracy by comparing the known data types against the RDF result.
+Plotting number of trees verses error allowed for determination of an early stopping point for algorithm growth (Figure \ref{fig:mls_zxy_early_stopping}) by choosing the point where the change in error moving average did not exceed 10\% and was within 20\% of the minimum error.}
+
+	\begin{figure}
+		\centering
+		\includegraphics[width=0.9\linewidth]{figures/flowz}
+		\caption[Project Flow]{Project flowchart}
+		\label{fig:flowzchart}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{figures/RAN_MLS_ALL}
+			\caption[]{}
+			\label{fig:all_mls_ransac_comp}
+		\end{subfigure}
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{figures/RAN_MLS_TT}
+			\caption[]{}
+			\label{fig:xyz_mls_ransac_comp}
+		\end{subfigure}
+		\label{fig:xyz_mls_comp_1}
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{figures/Range}
+			\caption[]{}
+			\label{fig:range_mls_ransac_comp}
+		\end{subfigure}
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{figures/zxy}
+			\caption[]{}
+			\label{fig:t20_mls_ransac_comp}
+		\end{subfigure}
+		\caption[RDF Validation Accuracy]{RDF validation accuracy directly compared.
+All features are considered (a); Top 20 features used in (a) only (b); Range was used without X, Y, and Z components broken out (c); Plane projection not considered in (d), reference plane set as the LiDAR origin point.}
+		\label{fig:all_tree_comp}
+	\end{figure}
+
+	\begin{figure}[H]
+			\centering
+			\includegraphics[width=0.75\linewidth]{figures/early_stopping_example}
+			\caption[Early Stopping]{Diminishing returns on increasing algorithm size versus accuracy can be used to determine the early stopping point of algorithm growth.}
+			\label{fig:mls_zxy_early_stopping}
+	\end{figure}
+
+	% Quadrant size inversely affects rate of classification, as smaller quadrants allow for greater resolution at the expense of more computational time.
+
+	{Testing consecutive scans required a method of compiling LiDAR data into a single point cloud and manually classifying areas for road detection accuracy.
+This was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
+Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data.
+Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU.
+LiDAR origin was found using GPS longitude, latitude, and altitude data.
+Physical distances between the GPS, IMU, and LiDAR was rectitude by obtaining reference frames provided by AutonomouStuff.
+GPS coordinates were offset by the current orientation and converts the ground truth to the LiDAR frame.
+GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors.
+Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}).
+While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for scoring purposes.
+Six stretches of unmarked roads were examined for accuracy: three of chipseal and three of gravel.
+Road and side-of-road areas were manually defined and projected unto consecutive LiDAR scans to score classification results.
+Due to the fuzzy nature of the classification results, the side-of-road areas were roughly a standard lane width to the sides of the road to accommodate the inability to define a rigid road edge.}
 
@@
 Quadrant size inversely affects rate of classification, as smaller quadrants allow for greater resolution at the expense of more computational time.
-Translation and rotation was applied to the terrain classification results with the closest matching transformation matrices.Classification results that were sympatric to manually defined areas were examined for accuracy by scoring the exact terrain classification and true/false-positive road surface detection.
+Translation and rotation was applied to the terrain classification results with the closest matching transformation matrices.
+Classification results that were sympatric to manually defined areas were examined for accuracy by scoring the exact terrain classification and true/false-positive road surface detection.
 Drop off in positive road surface classification rates in side-of-road areas in side-of-road areas would be indicative of adequate road edge detection.
@@
 
-	{Scanning LiDAR, GPS, and IMU data was collected on closed lots and public roads near Athens Ohio using the Experimental Apparatus.
+	{Scanning LiDAR, GPS, and IMU data was collected on closed lots and public roads near Athens Ohio.
 Gravel and chipseal data was gathered from closed lots and unmarked public roads (Figures \ref{fig:Combined_Roads}), while grass (well-trimmed lawn grass) and foliage (leafy trees and bushes) was gathered from areas surrounding public roads.
 Training data was generated using MATLAB to extract LiDAR data from rosbags.
-Classification algorithms using Random Decision Forests were generated using a variety of feature sets and reference points.
+Random Decision Forests classification algorithms were generated using a variety of feature sets and reference points.
 Testing consecutive scans was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data.
@@
 Transformation matrices were created by considering the GPS position and the IMU's roll, pitch, and yaw data.
-Point cloud data was then classified using the RDF derived in Objective I.
+For the initial results shown (Table \ref{tab:averaged_results}), LiDAR channels 2-4 were used (counting upwards from the lowest deflection channel).
 Translation and rotation was applied to the terrain classification results.
@@
 
-	{Results were scored by comparing the location of the averaged classified quadrant's (x,y) coordinate to the area that the manually defined areas (Table \ref{tab:averaged_results} shows results for algorithm designation RAN ALL which produced the results in Figure \ref{fig:classified_point_clouds_all}.
-All results are shown in Table \ref{tab:all_results} in Appendix \ref{apdx:Appendix_2}).
+	{Results were scored by comparing the location of the averaged classified quadrant's (x,y) coordinate to the area that the manually defined areas (Table \ref{tab:averaged_results}) shows results for algorithm designation RAN ALL which produced the results in Figure \ref{fig:classified_point_clouds_all}.
 Ratio of road to non-road in manually defined road and side-of-road areas are shown in columns 3 and 4, with higher values indicating a greater presence of positive road identification.
@@
 	{It was found that while none of the algorithms excelled in all areas, the results seem to indicate that implementing a multiple-model algorithm to create an ensemble would greatly increase road detection accuracy.
-Of the algorithms tested (Appendix \ref{apdx:Appendix_2}), the RDF that used RANSAC as the plane projection method and the full feature (designated RAN ALL) performed the most consistently for true non-road surface identification, with an average of 66.5\% (median of 65.5\%).
-While it did struggle with gravel identification (32.72\% average accuracy) with a high confusion rate with grass (27.47\% false identification), it may be supplemented with the RDF designated RANGE, which has the highest true-positive gravel identification rate of 82.8\%.
-A classification method was tested to verify this method based on RAN ALL and RANGE algorithms.
-RAN ALL would would designate non-road surfaces and the determination of road surface type was a combined effort of the two (Figure \ref{}), thus achieving a higher accuracy.
+Of the algorithms tested, the RDF that used RANSAC as the plane projection method and the full feature (designated RAN ALL) performed the most consistently for true non-road surface identification, with an average of 66.5\% (median of 65.5\%).
+While it did struggle with gravel identification (32.72\% average accuracy) with a high confusion rate with grass (27.47\% false identification), it may be supplemented in future work  with the RDF designated RANGE, which has the highest true-positive gravel identification rate of 82.8\%.
+Future work may also explore exploiting classification confidence scores by setting a base confidence threshold for identifying a road-surface terrain type.
+Preliminary work was completed to compare the confidence score of gravel classification against foliage and grass in order to over-ride the gravel classification, with the goal of decreasing road-detection false detection (Table \ref{tab:score_trim_method}).
+Gravel confidence score was compared to that of foliage and grass on Sturbois Road, which features mostly grassy center similar to that of Armig Road (Figure \ref{fig:Armig_Road_Camera_View}).
+Differences of less than 10\% between gravel and foliage as well as a difference of less than 10\% between gravel and grass results in a re-classification of grass or foliage.
+True-road identification accuracy was 38.5\%, nearly identical to the 40.3\% accuracy previously obtained.
+However side-of-road detection accuracy increased by 8.48\%.
 }
+
+	%{A classification method was tested to verify this method based on RAN ALL and RANGE algorithms.
+RAN ALL would would designate non-road surfaces and the determination of road surface type was a combined effort of the two (Figure \ref{}), thus achieving a higher accuracy.}
 
@@
 		\centering
-		\begin{tblr}{width=0.9\linewidth,colspec={|Q[1,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|}}
+		\begin{tblr}{width=0.9\linewidth,colspec={|Q[1,m,c]||Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|}}
 			\hline
@@
 
-	{Quadrant size affects terrain classification accuracy at the expense of granularity of results and computational time.
+	{Increasing the number of quadrants each channel was broken into increased terrain classification accuracy at the expense of granularity of results and computational time.
 Smaller quadrants allow for greater resolution of terrain classification, however more computational time is required.
 Larger quadrants allow for swifter classification, however as consecutive scanning LiDAR points may include road and non-road surfaces, classification accuracy may suffer.
-Best per-Quadrant average classification rate was X.XX quadrants per second as tested on a machine with 64 GB of DDR4 RAM and a Ryzen 5900x @ 3.70GHz*24.}
+Best per-Quadrant average classification rate with a quadrant size of 36 points was 0.1220 seconds per quadrant (includes recording and saving all results and diagnostic information to disk) as tested on a machine with 64 GB of DDR4 RAM and a Ryzen 5900x @ 3.70GHz*24.
+Decreasing quadrant sizes was tested for accuracy vs rate of classification by classifying a chipseal road sample.
+Results suggest that there is a point of diminishing returns when more than 60 quadrants (in this case 30 LiDAR points per quadrant) are used.
+Quadrant classification rate tends to decrease with increased numbers of quadrants, while the per channel 360 classification rate increases linearly.}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=0.95\linewidth]{figures/class_rate_2}
+			\caption[]{}
+			\label{fig:quad_vs_scan}
+		\end{subfigure}
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=0.95\linewidth]{figures/accs_v_size_2}
+			\caption[]{}
+			\label{fig:accs_v_size}
+		\end{subfigure}
+		\caption[Quadrant Num]{Number of quadrants was increased to determine the effect on accuracy and rate of classification.
+The algorithm used was the RDF designated RAN ALL.
+Quadrant classification rate tends to decrease with increased numbers of quadrants, while the per channel 360 classification rate increases linearly.
+The road used to derive these plots was a chipseal road.}
+		\label{fig:quadrant_size_comps}
+	\end{figure}
+
+	{Difficulties arise when attempting to classify gravel roads with a grassy center, as in the case of Armig Road (Figure \ref{fig:Armig_Road_Camera_View}).
+Discretizing LiDAR channels into smaller quadrants while also considering confidence scores may increase road detection capabilities.
+Compared to 100 quadrants, it was found that increasing the number of quadrants to 210 the true-road identification accuracy was decreased by 3.3\% while also increasing true-side-of-road detection by 3.9\% (Table \ref{tab:score_trim_method}).}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=0.95\linewidth]{figures/100_quad_score_test}
+			\caption[]{}
+			\label{fig:100_quad_size_score_trim}
+		\end{subfigure}
+		\begin{subfigure}{0.45\textwidth}
+			\centering
+			\includegraphics[width=0.95\linewidth]{figures/210_quad_score_test}
+			\caption[]{}
+			\label{fig:210_quad_size_score_trim}
+		\end{subfigure}
+		\caption[Comparing Quadrant Size and Score-based Re-classification]{Confidence scores of gravel, foliage, and grass were considered to potentially decrease side-of-road (SoR) false-positive road identification.Discretizing LiDAR channels affects the accuracy and classification rate.
+100 Quadrants (a) is compared to 210 quadrants (b): while the density of classification is increased, the difference between results is negligible (Table \ref{tab:score_trim_method}).}
+		\label{fig:score_trim_method}
+	\end{figure}
+
+%	\begin{figure}[H]
+%		\centering
+%		\begin{subfigure}{0.495\textwidth}
+%			\centering
+%			\includegraphics[width=0.95\linewidth]{figures/100_quad_score_test_2}
+%			\caption[]{}
+%			\label{fig:100_quad_score_test_2}	
+%		\end{subfigure}
+%		\begin{subfigure}{0.495\textwidth}
+%			\centering
+%			\includegraphics[width=0.95\linewidth]{figures/210_quad_score_test_2}
+%			\caption[]{}
+%			\label{fig:210_quad_score_test_2}
+%		\end{subfigure}
+%		\caption[Quadrant Limit Tests]{Discretizing LiDAR channels affects the accuracy and classification rate.
+100 Quadrants (a) is compared to 210 quadrants (b): while the density of classification is increased, the difference between results is negligible (Table \ref{tab:score_trim_method}).}
+%		\label{fig:score_trim_method}
+%	\end{figure}
+
+	\begin{table}[H]
+
+		\centering
+
+		\begin{tblr}{width=0.8\linewidth,colspec={|Q[1.0cm,m,c]||Q[1.0cm,m,c]|Q[1.0cm,m,c]|Q[1.5cm,m,c]|Q[1.5cm,m,c]|Q[1.5cm,m,c]|Q[1.5cm,m,c]|Q[1.5cm,m,c]|Q[1.5cm,m,c]|}}
+
+			\hline
+			\SetCell[r=2,c=1]{c}  
+			& \SetCell[r=1,c=2]{} Base (100Q) & 
+			& \SetCell[r=1,c=2]{} Conf. Score Adjusted (100Q) & 
+			& \SetCell[r=1,c=2]{} Conf. Score Adjusted (210Q) & 
+			& \SetCell[r=1,c=2]{} Channel Limited (100Q) & 										\\ \hline
+					 & Gravel & SoR    & Gravel & SoR    & Gravel & SoR		& Gravel & SoR 		\\ \hline
+			Gravel   & 38.594 & 33.787 & 36.657 & 24.847 & 32.466 & 20.497 	& 72.628 & 32.375 	\\ \hline
+			Chipseal & 1.719  & 0.101  & 1.813  & 0.552  & 2.701  & 1.017 	& 4.030  & 0.446 	\\ \hline
+			Foliage  & 14.766 & 21.198 & 3.525  & 9.264  & 3.152  & 7.271 	&  0.251 & 14.824 	\\ \hline
+			Grass    & 44.922 & 44.914 & 58.006 & 65.337 & 61.681 & 71.215	& 23.090 & 52.355  	\\ \hline
+
+		\end{tblr}
+
+		\caption[Confidence Score based Classification]{Confidence scores of gravel, foliage, and grass were considered to potentially decrease side-of-road (SoR) false-positive road identification (Figure \ref{fig:score_trim_method}).
+Distance from LiDAR is a factor that affects classification accuracy}
+		\label{tab:score_trim_method}
+
+	\end{table}
+
+	{Increasing distance from the LiDAR sensor may cause classification accuracy to decrease.
+Accuracy versus LiDAR channel was tested with channels 2-4 using the RAN ALL RDF without considering confidence scores to increase road-side detection accuracy (Table \ref{tab:accuracy_v_channel}) using the data gathered from Sturbois Road.
+Results indicate that accuracy decreases at increasing LiDAR ranges.
+Distance from LiDAR point of origin is considered in the feature extraction and the training data includes data from random LiDAR channels.
+These results may indicate that an ensemble of random decision forests to handle both long and short range data may be needed.
+Elimination of channel 4 (Table \ref{tab:accuracy_v_channel}) was completed to determine the increase in road surface detection using the data set from Sturbois Road.
+It was found that the road surface accuracy increased from 38.6\% to 76.7\% and side of road detection increased from 66.1\% to 67.2\% (Table \ref{tab:score_trim_method}).}
+
+	\begin{table}[H]
+
+		\centering
+
+		\begin{tblr}{width=0.5\linewidth,colspec={|Q[2,m,c]||Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|}}
+
+			\hline
+			\SetCell[r=2,c=1]{c}  
+			& \SetCell[r=1,c=2]{} Channel 2 & 
+			& \SetCell[r=1,c=2]{} Channel 3 &
+			& \SetCell[r=1,c=2]{} Channel 4 & \\ \hline
+				 & Gravel & SoR    & Gravel & SoR    & Gravel & SoR 	\\ \hline
+			Grav & 61.670 & 30.551 & 62.050 & 32.641 & 36.183 & 24.932 	\\ \hline
+			Chip & 6.250  & 0.449  & 11.123 & 1.953  & 1.392  & 0.933  	\\ \hline
+			Foli & 0.087  & 8.622  & 0.491  & 17.235 & 0.928  & 21.898 	\\ \hline
+			Gras & 31.993 & 60.379 & 26.336 & 48.171 & 61.498 & 52.236 	\\ \hline
+
+		\end{tblr}
+
+		\caption[Accuracy versus Channel]{Classification accuracy was tested for channels 2-4 using data from Sturbois Road. }
+		\label{tab:accuracy_v_channel}
+
+	\end{table}
+
+	% NEED: side-by-side of quadrant test & channel trim
 
@@
 
-	{The proposed work addresses the problem of road surface detection on unmarked gravel and chipseal roads.
-Current LiDAR and Camera based detection models are insufficient due to lack of distinct features typical of urban roads, such as painted line markings or curbs, or have excessive computational or storage requirements.
-The completed work balanced accuracy and efficiency by using less intensive analysis techniques of smaller point cloud data sets.
-The final deliverable of the described work was a method of detecting physical unmarked gravel and chipseal roads by using a terrain classification approach to predicting road surface area.
+	% Summary of Work & Closing Statements
+	{High definition scanning LiDAR and GPS data was gathered on unmarked chipseal and gravel roads around Athens Ohio.
+Properties describing surface roughness were extracted and used to train a number of Random Decision Forests.
+Terrain classification of consecutive scans compared to manually defined road and side-of-road surface areas yielded accuracy scores for gauging algorithm capability for road-surface detection.
+Use of multiple terrain classification algorithms was tested.
+Results indicate that a base accuracy rate of 36.1\%, 54.2\%, and 56.3\% for true positive detection rate for gravel, chipseal, and side-of-road surfaces.
+Number of consecutive points per scanning LiDAR channel quadrant influences terrain classification rate and accuracy, however accuracy increases were negligible.
+Distance from LiDAR point of origin influences terrain classification accuracy.
+Limiting classification by LiDAR channel indicated that a much higher classification accuracy may be achieved, accomplishing a peak gravel road detection accuracy of 72.3\%.
+This work addressed the problem of road surface detection on unmarked gravel and chipseal roads.
+Current LiDAR and Camera based detection models were insufficient due to lack of distinct features typical of urban roads, such as painted line markings or curbs, or have excessive computational or storage requirements.
+This work balanced accuracy and efficiency by using less intensive analysis techniques of smaller point cloud data sets.
+The described work was a method of detecting physical unmarked gravel and chipseal roads by using a terrain classification approach to predicting road surface area.
 The impact of this work is that autonomous vehicles using LiDAR may be able to detect unmarked gravel and chipseal road surfaces, allowing autonomous operations on 1.5 million miles of previously undetected rural roads.}
-
 % For peer review papers, you can put extra information on the cover
@@
 
-\section{All Results}\label{apdx:Appendix_2}
-
-\begin{table}[]
-	\centering
-	\begin{tabular}{|>{\centering\arraybackslash}p{.25cm}|>{\centering\arraybackslash}p{.25cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.4cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|>{\raggedleft\arraybackslash}p{.8cm}|}
-		\hline
-		\spheading{Algorithm} & 
-		\spheading{Road Samp} & 
-		\spheading{Ratio of Road / Non-Road: On Road} & 
-		\spheading{Ratio of Road / Non-Road: Off Road} & 
-		\spheading{Ratio Score} & 
-		\spheading{True Road Surface Type Identification Score} & 
-		\spheading{False Road Surface Type Identification Score} & 
-		\spheading{False Non-Road CR Score} & 
-		\spheading{True Non-Road CR Score} &
-		\spheading{True Road Surface Identification} &
-		\spheading{False Road Surface Identification} \\
-		\hline
-		\multirow{6}{*}{\rotatebox[origin=c]{90}{MLS ALL}}
-		&		A		& 4.00   & 0.21 & 19.01  & 36.10 & 43.91 & 19.99 & 82.61 & 80.01  & 19.99 \\
-		&		B		& 3.89   & 2.36 & 1.65   & 23.93 & 55.61 & 20.45 & 29.75 & 79.55  & 20.45 \\
-		&		C		& 4.06   & 1.63 & 2.49   & 22.03 & 58.20 & 19.77 & 38.07 & 80.23  & 19.77 \\
-		&		D		& 2.22   & 0.44 & 5.09   & 58.80 & 22.25 & 18.96 & 69.67 & 81.04  & 18.96 \\
-		&		E		& 6.51   & 0.83 & 7.84   & 58.93 & 24.15 & 16.92 & 54.62 & 83.08  & 16.92 \\
-		&		F		& 26.20  & 0.81 & 32.44  & 79.74 & 13.73 & 6.54  & 55.32 & 93.46  & 6.54  \\
-		\hline
-		\multirow{6}{*}{\rotatebox[origin=c]{90}{MLS TT}}       
-		&		A		& 19.94  & 0.38 & 52.93  & 31.69 & 63.53 & 4.78  & 72.64 & 95.22  & 4.78  \\
-		&		B		& 19.52  & 4.69 & 4.16   & 25.24 & 69.89 & 4.87  & 17.56 & 95.13  & 4.87  \\
-		&		C		& 19.00  & 3.04 & 6.26   & 25.94 & 69.06 & 5.00  & 24.77 & 95.00  & 5.00  \\
-		&		D		& 2.09   & 1.27 & 1.65   & 8.94  & 45.08 & 45.98 & 44.10 & 54.02  & 45.98 \\
-		&		E		& 21.59  & 2.86 & 7.55   & 68.87 & 28.41 & 2.71  & 25.91 & 97.29  & 2.71  \\
-		&		F		& 100.00 & 1.57 & 63.56  & 84.50 & 15.50 & 0.00  & 38.86 & 100.00 & 0.00  \\
-		\hline
-		\multirow{6}{*}{\rotatebox[origin=c]{90}{RAN ALL}}     
-		&		A		& 11.42  & 0.23 & 50.35  & 41.39 & 50.56 & 8.05  & 81.51 & 91.95  & 8.05  \\
-		&		B		& 13.01  & 2.38 & 5.47   & 26.46 & 66.41 & 7.14  & 29.60 & 92.86  & 7.14  \\
-		&		C		& 14.61  & 1.72 & 8.49   & 27.27 & 66.33 & 6.41  & 36.76 & 93.59  & 6.41  \\
-		&		D		& 2.67   & 0.42 & 6.29   & 62.55 & 25.32 & 12.13 & 70.19 & 87.87  & 12.13 \\
-		&		E		& 10.74  & 0.76 & 14.05  & 63.03 & 29.61 & 7.36  & 56.68 & 92.64  & 7.36  \\
-		&		F		& 100.00 & 0.78 & 128.22 & 82.91 & 17.09 & 0.00  & 56.18 & 100.00 & 0.00  \\		
-		\hline
-		\multirow{6}{*}{\rotatebox[origin=c]{90}{RAN TT}}       
-		&		A		& 0.27   & 0.22 & 1.23   & 21.35 & 0.19  & 78.46 & 81.80 & 21.54  & 78.46 \\
-		&		B		& 0.70   & 0.54 & 1.29   & 38.21 & 2.87  & 58.92 & 64.92 & 41.08  & 58.92 \\
-		&		C		& 0.68   & 0.51 & 1.32   & 38.59 & 1.72  & 59.69 & 66.11 & 40.31  & 59.69 \\
-		&		D		& 1.17   & 0.52 & 2.27   & 43.30 & 29.72 & 26.99 & 65.98 & 73.01  & 26.99 \\
-		&		E		& 2.97   & 0.77 & 3.87   & 38.49 & 37.55 & 23.96 & 56.60 & 76.04  & 23.96 \\
-		&		F		& 2.05   & 0.57 & 3.58   & 74.32 & 11.20 & 14.47 & 63.60 & 85.53  & 14.47 \\
-		\hline
-		\multirow{6}{*}{\rotatebox[origin=c]{90}{RANGE}}  
-		&		A		& 4.00   & 0.21 & 19.01  & 36.10 & 43.91 & 19.99 & 82.61 & 80.01  & 19.99 \\
-		&		B		& 2.59   & 2.47 & 1.05   & 71.63 & 0.52  & 27.85 & 28.84 & 72.15  & 27.85 \\
-		&		C		& 1.75   & 2.26 & 0.77   & 63.05 & 0.55  & 36.41 & 30.66 & 63.59  & 36.41 \\
-		&		D		& 1.95   & 1.05 & 1.86   & 34.50 & 46.97 & 18.53 & 48.87 & 81.47  & 18.53 \\
-		&		E		& 6.54   & 1.59 & 4.12   & 29.32 & 53.86 & 16.82 & 38.62 & 83.18  & 16.82 \\
-		&		F		& 5.30   & 0.73 & 7.30   & 75.07 & 15.69 & 9.24  & 57.90 & 90.76  & 9.24  \\
-		\hline
-		\multirow{6}{*}{\rotatebox[origin=c]{90}{ZXY}}          
-		&		A		& 2.47   & 0.37 & 6.67   & 71.11 & 0.09  & 28.79 & 72.95 & 71.21  & 28.79 \\
-		&		B		& 13.36  & 1.78 & 7.53   & 89.99 & 3.05  & 6.96  & 36.03 & 93.04  & 6.96  \\
-		&		C		& 10.13  & 1.66 & 6.11   & 87.42 & 3.59  & 8.98  & 37.61 & 91.02  & 8.98  \\
-		&		D		& 0.38   & 0.29 & 1.28   & 3.24  & 14.44 & 82.33 & 77.28 & 17.67  & 82.33 \\
-		&		E		& 0.37   & 0.36 & 1.04   & 2.26  & 18.98 & 78.76 & 73.76 & 21.24  & 78.76 \\
-		&		F		& 43.63  & 0.87 & 49.99  & 89.73 & 8.03  & 2.24  & 53.40 & 97.76  & 2.24 \\
-		\hline                              
-	\end{tabular}
-	\caption[All Results]{All Results XD}
-	\label{tab:all_results}
-\end{table}
-
 % use section* for acknowledgment

### 2023-04-29 Huston_Paper.tex
more progress

@@
 Manually discovered gravel and paved road surface based on classified results were compared to actual road bounds.
-<SUMMARY OF RESULTS HERE>.
+It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively.
+Classified point clouds were manually examined and gravel areas were projected and compared with actual gravel surface areas.
+Overlapping areas indicated a high level of unmarked gravel and asphalt road surface detection rate.
 Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.
@@
 Using 2D projection of planes unto a 3D point cloud of a road surface is an alternative approach, but likewise requires curbs or painted lines for road boundary definition \cite{fernandes_road_2014, borkar_robust_2009-1, guo_lane_2015}.
-Instead, using LiDAR may overcome difficulties in achieving real-time detection of unmarked roads by using terrain classification to discover road type, boundaries, and intersections.}
+Instead, using LiDAR may overcome difficulties in achieving detection of gravel roads by using terrain classification.}
 
@@
 	% Summary of lit review in 1 sentance:
-	{Current methods for road detection may be insufficient for real-time detection of unmarked gravel and chipseal roads due to ambient conditions, lack of standardized features, or rely on post-processing large data sets, while terrain classification of scanning LiDAR data may overcome these difficulties.}
+	{Current methods for road detection may be insufficient for real-time detection of gravel roads due to ambient conditions, lack of standardized features, or rely on post-processing large data sets, while terrain classification of scanning LiDAR data may overcome these difficulties.}
 
@@
 			\centering
-			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/armig_2}
+			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/gravel_lot_pic.png}
 			\caption[Armig Road Camera View]{}
@@
 
-	{Training data was extracted by manually defining gravel, asphalt, and unknown terrain areas on compiled point clouds.
+	{Scanning LiDAR training data was extracted from manually defined gravel, asphalt, and unknown terrain areas on compiled point clouds.
 Compiling LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
@@
 Manually defined areas may then be overlayed unto the point cloud (Figure \ref{fig:Combined_Classified_PCD}).
-Individual scans were consecutively examined to extract per-channel scanning LiDAR training data and exported to a database if coincident with the manually defined areas.}
+Six degree arcs from the first three channels of scanning LiDAR data that lay directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.}
 
@@
 
-	{MATLAB's Classification Learner Application was used to generated Random Decision Forests.
-Decision tree depth and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.}
-
-	{Random Decision Forests were created using MATLAB's $TreeBagger$ function, which creates an ensemble of bagged decision trees using the input arguments of included features, tree depth, number of trees, and a cost array.
-RDFs were split into sub categories within each feature set to test varying tree depths (Figure \ref{fig:all_tree_comp}), and allowed to grow in number of trees until three hundred trees were created.
-Discouragement of false-positive identification for road surfaces (gravel and chipseal) was accomplished by introducing a cost array which would induce a bias against gravel and chipseal.
-Validation error was used as the primary means to compare RDF performance prior to testing with data outside the training and validation data sets.
-MATLAB was used to load verification data and test increasing sizes of RDF algorithms for accuracy by comparing the known data types against the RDF result.
-Plotting number of trees verses error allowed for determination of an early stopping point for algorithm growth (Figure \ref{fig:mls_zxy_early_stopping}) by choosing the point where the change in error moving average did not exceed 10\% and was within 20\% of the minimum error.}
+	{MATLAB's Classification Learner Application was used to generate a Random Decision Forests for each considered scanning LiDAR channel.
+Decision tree depth, number of learners, and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.}
 
@@
 
-	\begin{figure}[H]
-		\centering
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=1\linewidth]{figures/RAN_MLS_ALL}
-			\caption[]{}
-			\label{fig:all_mls_ransac_comp}
-		\end{subfigure}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=1\linewidth]{figures/RAN_MLS_TT}
-			\caption[]{}
-			\label{fig:xyz_mls_ransac_comp}
-		\end{subfigure}
-		\label{fig:xyz_mls_comp_1}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=1\linewidth]{figures/Range}
-			\caption[]{}
-			\label{fig:range_mls_ransac_comp}
-		\end{subfigure}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=1\linewidth]{figures/zxy}
-			\caption[]{}
-			\label{fig:t20_mls_ransac_comp}
-		\end{subfigure}
-		\caption[RDF Validation Accuracy]{RDF validation accuracy directly compared.
-All features are considered (a); Top 20 features used in (a) only (b); Range was used without X, Y, and Z components broken out (c); Plane projection not considered in (d), reference plane set as the LiDAR origin point.}
-		\label{fig:all_tree_comp}
-	\end{figure}
-
-	\begin{figure}[H]
-			\centering
-			\includegraphics[width=0.75\linewidth]{figures/early_stopping_example}
-			\caption[Early Stopping]{Diminishing returns on increasing algorithm size versus accuracy can be used to determine the early stopping point of algorithm growth.}
-			\label{fig:mls_zxy_early_stopping}
-	\end{figure}
-
 	% Quadrant size inversely affects rate of classification, as smaller quadrants allow for greater resolution at the expense of more computational time.
 
-	{Testing consecutive scans required a method of compiling LiDAR data into a single point cloud and manually classifying areas for road detection accuracy.
+%	{Testing consecutive scans required a method of compiling LiDAR data into a single point cloud and manually classifying areas for road detection accuracy.
 This was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
@@
 
-	{Classifying gathered consecutive LiDAR scans was accomplished by analyzing each individual 360$^{\circ}$ sweep of the discretized channels.
-Each channel was broken into quadrants.
-Quadrant size inversely affects rate of classification, as smaller quadrants allow for greater resolution at the expense of more computational time.
-Translation and rotation was applied to the terrain classification results with the closest matching transformation matrices.
-Classification results that were sympatric to manually defined areas were examined for accuracy by scoring the exact terrain classification and true/false-positive road surface detection.
-Drop off in positive road surface classification rates in side-of-road areas in side-of-road areas would be indicative of adequate road edge detection.
-Consecutive scanning LiDAR data was compiled allowing for manually assigning areas representing road or side of road surfaces (Figure \ref{fig:Classified_Areas_PCD}).}
+	{Classifying consecutive LiDAR scans was accomplished by examination of nine arcs of interest in front and forty-five degree angles left and right from the front of the experimental apparatus.
+Features from each arc were extracted and the arc was classified.
+Averaged x, y, and z coordinates were taken of each arc and projected unto a manually defined truth area map (Figure \ref{fig:Classified_Areas_PCD}).
+Accuracy scores of each area was calculated, allowing evaluation of road detection performance.}
 
@@
 
-	{Scanning LiDAR, GPS, and IMU data was collected on closed lots and public roads near Athens Ohio.
-Gravel and chipseal data was gathered from closed lots and unmarked public roads (Figures \ref{fig:Combined_Roads}), while grass (well-trimmed lawn grass) and foliage (leafy trees and bushes) was gathered from areas surrounding public roads.
+	{Five drive-bys of Blackburn Road with three intercepting gravel parking lots and driveways were completed and scanning LiDAR, GPS, and IMU data was collected.
 Training data was generated using MATLAB to extract LiDAR data from rosbags.
-Random Decision Forests classification algorithms were generated using a variety of feature sets and reference points.
+Random Decision Forests classification algorithms were generated.
 Testing consecutive scans was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data.
@@
 Transformation matrices were created by considering the GPS position and the IMU's roll, pitch, and yaw data.
-For the initial results shown (Table \ref{tab:averaged_results}), LiDAR channels 2-4 were used (counting upwards from the lowest deflection channel).
-Translation and rotation was applied to the terrain classification results.
-Consecutive scanning LiDAR data was compiled allowing for manually assigning areas representing road or side of road surfaces (Figure \ref{fig:classified_point_clouds_all}).
-Side of road area were manually defined as an area roughly one meter outwards from the road edge for consideration of road edge detection.}
+Tree scanning LiDAR arcs from three channels in front of the vehicle were classified.
+Translation and rotation was applied to the averaged x, y, and z coordinates of each arc's terrain classification results.
+Consecutive scanning LiDAR data was projected on manually classified truth areas representing road road surfaces (Figure \ref{fig:area_percentages}) and percentages of terrain types per area were calculated.
+}
+
+	{Due to the side of the road was being re-grassed near area $3$, the algorithm sometimes struggled to differentiate the dirt from gravel, leading to some miss-classification of the side-of-road areas (Figure \ref{fig:pre-adjust} vs Figure \ref{fig:post-adjust}).
+Confidence scores may be exploited to adjust the final result by requiring a very high confidence for a gravel surface to be classified as such.
+All gravel surfaces that did not meet a 80\% confidence score was re-labeled as "unknown".
+Furthermore distances between the averaged coordinates of the arcs that were classified as gravel were compared and if the distance was below a threshold the arc was re-labeled as "unknown".
+This was done as distances between channels tended to certain distances and evenly spaced from each other when the data is on a flat surface, typical of road surfaces.
+With these slight adjustments, it was found that an average true-positive accuracy of all intercepting gravel surfaces was 67.54\%.
+If any asphalt miss-classification on the gravel areas is included, the average true-positive road detection accuracy increases to 76.97\%.
+It was found that an average true-positive accuracy of all asphalt surfaces was 87.05\%.
+}
 
@@
 			\centering
-
 			\begin{subfigure}{0.495\textwidth}
 					\centering
-			\includegraphics[width=0.95\linewidth]{figures/curve_gravel}
-			\caption[]{}
-			\label{fig:curve_gravel_2}	
+					\includegraphics[width=0.95\linewidth]{figures/pre-adjust}
+					\caption[Pre-Adjusted]{}
+					\label{fig:pre-adjust}	
 				\end{subfigure}
@@
 					\centering
-			\includegraphics[width=0.95\linewidth]{figures/curve_gravel_2}
-			\caption[]{}
-			\label{fig:curve_gravel}
+					\includegraphics[width=0.95\linewidth]{figures/post-adjust}
+					\caption[Post-Adjusted]{}
+					\label{fig:post-adjust}
 				\end{subfigure}
-		\begin{subfigure}{0.495\textwidth}
-			\centering
-			\includegraphics[width=0.95\linewidth]{figures/long_gravel_1}
-			\caption[]{}
-			\label{fig:long_gravel_1}	
-		\end{subfigure}
-		\begin{subfigure}{0.495\textwidth}
-			\centering
-			\includegraphics[width=0.95\linewidth]{figures/long_chip_redux}
-			\caption[]{}
-			\label{fig:long_chip_classified}
-		\end{subfigure}
-		\begin{subfigure}{0.495\textwidth}
-			\centering
-			\includegraphics[width=0.95\linewidth]{figures/long_chip_redux_2}
-			\caption[]{}
-			\label{fig:long_chip_2_classified}
-		\end{subfigure}
-		\begin{subfigure}{0.495\textwidth}
-			\centering
-			\includegraphics[width=0.95\linewidth]{figures/short_chip_redux}
-			\caption[]{}
-			\label{fig:short_chip_classified}
-		\end{subfigure}
-		\caption[Classified Point Clouds]{Classified point clouds with manually defined surface areas that allow for scoring of classification results.
-Gravel roads include an un-named gravel side-road near Armig Road (a) and Sturbois Road (b), (c).
-Chipseal roads include Willow Creek Road (d), (e), and Bean Hollow Road (f).
-The algorithm that produced these results used all features and reference plane projected with RANSAC.
-Chipseal roads are shown as black areas, gravel roads and driveways are shown as red areas, and side-of-roads are designated as orange.
-}
-		\label{fig:classified_point_clouds_all}
+				\label{fig:pre_post_adjust}
+				\caption[Post-Adjusted]{Raw classification result example (a).
+Classification results may be slightly altered by considering confidence scores and inter-arc distances to decrease miss-classification of gravel surfaces.
+(b).}
 	\end{figure}
-
-	{Results were scored by comparing the location of the averaged classified quadrant's (x,y) coordinate to the area that the manually defined areas (Table \ref{tab:averaged_results}) shows results for algorithm designation RAN ALL which produced the results in Figure \ref{fig:classified_point_clouds_all}.
-Ratio of road to non-road in manually defined road and side-of-road areas are shown in columns 3 and 4, with higher values indicating a greater presence of positive road identification.
-The quotient of the two ratios (col.
-5) allows for estimation of road edge detection ability.
-Scores for correct and incorrect road surface type is shown, indicating the algorithm's ability to correctly determine the road surface type (cols.
-6,7).
-Percentages of grass and foliage classification for both side-of-road and on road are given in cols.
-8 \& 9, indicating true/false-negative road surface identification capability.
-True/false-positive road surface identification percentages are indicated in cols 10-11.}
-
-	{It was found that while none of the algorithms excelled in all areas, the results seem to indicate that implementing a multiple-model algorithm to create an ensemble would greatly increase road detection accuracy.
-Of the algorithms tested, the RDF that used RANSAC as the plane projection method and the full feature (designated RAN ALL) performed the most consistently for true non-road surface identification, with an average of 66.5\% (median of 65.5\%).
-While it did struggle with gravel identification (32.72\% average accuracy) with a high confusion rate with grass (27.47\% false identification), it may be supplemented in future work  with the RDF designated RANGE, which has the highest true-positive gravel identification rate of 82.8\%.
-Future work may also explore exploiting classification confidence scores by setting a base confidence threshold for identifying a road-surface terrain type.
-Preliminary work was completed to compare the confidence score of gravel classification against foliage and grass in order to over-ride the gravel classification, with the goal of decreasing road-detection false detection (Table \ref{tab:score_trim_method}).
-Gravel confidence score was compared to that of foliage and grass on Sturbois Road, which features mostly grassy center similar to that of Armig Road (Figure \ref{fig:Armig_Road_Camera_View}).
-Differences of less than 10\% between gravel and foliage as well as a difference of less than 10\% between gravel and grass results in a re-classification of grass or foliage.
-True-road identification accuracy was 38.5\%, nearly identical to the 40.3\% accuracy previously obtained.
-However side-of-road detection accuracy increased by 8.48\%.
-}
-
-	%{A classification method was tested to verify this method based on RAN ALL and RANGE algorithms.
-RAN ALL would would designate non-road surfaces and the determination of road surface type was a combined effort of the two (Figure \ref{}), thus achieving a higher accuracy.}
-
-	\begin{table}[H]
-		\centering
-		\begin{tblr}{width=0.9\linewidth,colspec={|Q[1,m,c]||Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|}}
-			\hline
-			Algorithm Designation & Average \% of True Gravel Type Identification & Average \% of True Chipseal Type Identification & Average \% of True Surface Type Identification & Average \% True Road Surface Detection & Average \% of True  Non-Road Surface Detection\\
-			\hline
-			MLS ALL & 35.354 & 68.581 & 46.587 & 82.896 & 55.006 \\
-			\hline
-			MLS TT  & 37.833 & 55.162 & 40.865 & 89.444 & 37.308 \\
-			\hline
-			ZXY     & 35.697 & 73.211 & 50.601 & 93.153 & 55.156 \\
-			\hline
-			RAN ALL & 36.119 & 54.169 & 42.376 & 56.252 & 66.500 \\
-			\hline
-			RANGE   & 61.822 & 44.758 & 51.610 & 78.526 & 47.918 \\
-			\hline
-			RAN TT  & 54.508 & 27.617 & 57.293 & 65.323 & 58.506 \\
-			\hline
-%			Ens\_1  & 61.822 & 44.758 & 51.610 & 78.526 & 66.685 \\
-%			\hline
-		\end{tblr}
-		\caption[Averaged Results]{Average results of all algorithms on tested roads. }
-		\label{tab:averaged_results}
-	\end{table}
-
-%	\begin{figure}[H]
-%		\centering
-%		\includegraphics[width=0.75\linewidth]{figures/RAN_ALL_ofroad_RANGE_onroad}
-%		\caption[Early Stopping]{Diminishing returns on increasing algorithm size versus accuracy can be used to determine the early stopping point of algorithm growth.}
-%		\label{fig:combo_result}
-%	\end{figure}
-
-	{Increasing the number of quadrants each channel was broken into increased terrain classification accuracy at the expense of granularity of results and computational time.
-Smaller quadrants allow for greater resolution of terrain classification, however more computational time is required.
-Larger quadrants allow for swifter classification, however as consecutive scanning LiDAR points may include road and non-road surfaces, classification accuracy may suffer.
-Best per-Quadrant average classification rate with a quadrant size of 36 points was 0.1220 seconds per quadrant (includes recording and saving all results and diagnostic information to disk) as tested on a machine with 64 GB of DDR4 RAM and a Ryzen 5900x @ 3.70GHz*24.
-Decreasing quadrant sizes was tested for accuracy vs rate of classification by classifying a chipseal road sample.
-Results suggest that there is a point of diminishing returns when more than 60 quadrants (in this case 30 LiDAR points per quadrant) are used.
-Quadrant classification rate tends to decrease with increased numbers of quadrants, while the per channel 360 classification rate increases linearly.}
 
@@
 		\centering
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=0.95\linewidth]{figures/class_rate_2}
-			\caption[]{}
-			\label{fig:quad_vs_scan}
-		\end{subfigure}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=0.95\linewidth]{figures/accs_v_size_2}
-			\caption[]{}
-			\label{fig:accs_v_size}
-		\end{subfigure}
-		\caption[Quadrant Num]{Number of quadrants was increased to determine the effect on accuracy and rate of classification.
-The algorithm used was the RDF designated RAN ALL.
-Quadrant classification rate tends to decrease with increased numbers of quadrants, while the per channel 360 classification rate increases linearly.
-The road used to derive these plots was a chipseal road.}
-		\label{fig:quadrant_size_comps}
+		\includegraphics[width=0.9\linewidth]{figures/rm_db_4_percentages_big_anotated}
+		\caption[Area Percentage Scores]{Five runs of Blackburn Road with the intercepting gravel parking lot were classified.
+Percentages of each terrain type per area were calculated.
+Area $3$ is the entrance to the gravel parking lot from which training data was extracted.}
+		\label{fig:area_percentages}
 	\end{figure}
 
-	{Difficulties arise when attempting to classify gravel roads with a grassy center, as in the case of Armig Road (Figure \ref{fig:Armig_Road_Camera_View}).
-Discretizing LiDAR channels into smaller quadrants while also considering confidence scores may increase road detection capabilities.
-Compared to 100 quadrants, it was found that increasing the number of quadrants to 210 the true-road identification accuracy was decreased by 3.3\% while also increasing true-side-of-road detection by 3.9\% (Table \ref{tab:score_trim_method}).}
+	{Classified point clouds were manually examined for 'best guess' locations for gravel and asphalt surfaces.
+Two out of three positive gravel identifications must be made on each channel, be evenly spaced, and be relatively co-planar to other road surface guesses in order for a positive guess for an asphalt or gravel surface to be made.
+It was found that the three intercepting gravel surfaces was consistently discovered (Figure \ref{fig:rm_db_4_toc}).}
 
@@
 		\centering
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=0.95\linewidth]{figures/100_quad_score_test}
-			\caption[]{}
-			\label{fig:100_quad_size_score_trim}
-		\end{subfigure}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=0.95\linewidth]{figures/210_quad_score_test}
-			\caption[]{}
-			\label{fig:210_quad_size_score_trim}
-		\end{subfigure}
-		\caption[Comparing Quadrant Size and Score-based Re-classification]{Confidence scores of gravel, foliage, and grass were considered to potentially decrease side-of-road (SoR) false-positive road identification.Discretizing LiDAR channels affects the accuracy and classification rate.
-100 Quadrants (a) is compared to 210 quadrants (b): while the density of classification is increased, the difference between results is negligible (Table \ref{tab:score_trim_method}).}
-		\label{fig:score_trim_method}
+		\includegraphics[width=0.9\linewidth]{figures/rm_db_4_toc}
+		\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed using a rule set and compared to actual road-surface areas. }
+		\label{fig:rm_db_4_toc}
 	\end{figure}
-
-%	\begin{figure}[H]
-%		\centering
-%		\begin{subfigure}{0.495\textwidth}
-%			\centering
-%			\includegraphics[width=0.95\linewidth]{figures/100_quad_score_test_2}
-%			\caption[]{}
-%			\label{fig:100_quad_score_test_2}	
-%		\end{subfigure}
-%		\begin{subfigure}{0.495\textwidth}
-%			\centering
-%			\includegraphics[width=0.95\linewidth]{figures/210_quad_score_test_2}
-%			\caption[]{}
-%			\label{fig:210_quad_score_test_2}
-%		\end{subfigure}
-%		\caption[Quadrant Limit Tests]{Discretizing LiDAR channels affects the accuracy and classification rate.
-100 Quadrants (a) is compared to 210 quadrants (b): while the density of classification is increased, the difference between results is negligible (Table \ref{tab:score_trim_method}).}
-%		\label{fig:score_trim_method}
-%	\end{figure}
-
-	\begin{table}[H]
-
-		\centering
-
-		\begin{tblr}{width=0.8\linewidth,colspec={|Q[1.0cm,m,c]||Q[1.0cm,m,c]|Q[1.0cm,m,c]|Q[1.5cm,m,c]|Q[1.5cm,m,c]|Q[1.5cm,m,c]|Q[1.5cm,m,c]|Q[1.5cm,m,c]|Q[1.5cm,m,c]|}}
-
-			\hline
-			\SetCell[r=2,c=1]{c}  
-			& \SetCell[r=1,c=2]{} Base (100Q) & 
-			& \SetCell[r=1,c=2]{} Conf. Score Adjusted (100Q) & 
-			& \SetCell[r=1,c=2]{} Conf. Score Adjusted (210Q) & 
-			& \SetCell[r=1,c=2]{} Channel Limited (100Q) & 										\\ \hline
-					 & Gravel & SoR    & Gravel & SoR    & Gravel & SoR		& Gravel & SoR 		\\ \hline
-			Gravel   & 38.594 & 33.787 & 36.657 & 24.847 & 32.466 & 20.497 	& 72.628 & 32.375 	\\ \hline
-			Chipseal & 1.719  & 0.101  & 1.813  & 0.552  & 2.701  & 1.017 	& 4.030  & 0.446 	\\ \hline
-			Foliage  & 14.766 & 21.198 & 3.525  & 9.264  & 3.152  & 7.271 	&  0.251 & 14.824 	\\ \hline
-			Grass    & 44.922 & 44.914 & 58.006 & 65.337 & 61.681 & 71.215	& 23.090 & 52.355  	\\ \hline
-
-		\end{tblr}
-
-		\caption[Confidence Score based Classification]{Confidence scores of gravel, foliage, and grass were considered to potentially decrease side-of-road (SoR) false-positive road identification (Figure \ref{fig:score_trim_method}).
-Distance from LiDAR is a factor that affects classification accuracy}
-		\label{tab:score_trim_method}
-
-	\end{table}
-
-	{Increasing distance from the LiDAR sensor may cause classification accuracy to decrease.
-Accuracy versus LiDAR channel was tested with channels 2-4 using the RAN ALL RDF without considering confidence scores to increase road-side detection accuracy (Table \ref{tab:accuracy_v_channel}) using the data gathered from Sturbois Road.
-Results indicate that accuracy decreases at increasing LiDAR ranges.
-Distance from LiDAR point of origin is considered in the feature extraction and the training data includes data from random LiDAR channels.
-These results may indicate that an ensemble of random decision forests to handle both long and short range data may be needed.
-Elimination of channel 4 (Table \ref{tab:accuracy_v_channel}) was completed to determine the increase in road surface detection using the data set from Sturbois Road.
-It was found that the road surface accuracy increased from 38.6\% to 76.7\% and side of road detection increased from 66.1\% to 67.2\% (Table \ref{tab:score_trim_method}).}
-
-	\begin{table}[H]
-
-		\centering
-
-		\begin{tblr}{width=0.5\linewidth,colspec={|Q[2,m,c]||Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|Q[1.5,m,c]|}}
-
-			\hline
-			\SetCell[r=2,c=1]{c}  
-			& \SetCell[r=1,c=2]{} Channel 2 & 
-			& \SetCell[r=1,c=2]{} Channel 3 &
-			& \SetCell[r=1,c=2]{} Channel 4 & \\ \hline
-				 & Gravel & SoR    & Gravel & SoR    & Gravel & SoR 	\\ \hline
-			Grav & 61.670 & 30.551 & 62.050 & 32.641 & 36.183 & 24.932 	\\ \hline
-			Chip & 6.250  & 0.449  & 11.123 & 1.953  & 1.392  & 0.933  	\\ \hline
-			Foli & 0.087  & 8.622  & 0.491  & 17.235 & 0.928  & 21.898 	\\ \hline
-			Gras & 31.993 & 60.379 & 26.336 & 48.171 & 61.498 & 52.236 	\\ \hline
-
-		\end{tblr}
-
-		\caption[Accuracy versus Channel]{Classification accuracy was tested for channels 2-4 using data from Sturbois Road. }
-		\label{tab:accuracy_v_channel}
-
-	\end{table}
-
-	% NEED: side-by-side of quadrant test & channel trim
 
@@
 	% Summary of Work & Closing Statements
-	{High definition scanning LiDAR and GPS data was gathered on unmarked chipseal and gravel roads around Athens Ohio.
-Properties describing surface roughness were extracted and used to train a number of Random Decision Forests.
+	{High definition scanning LiDAR and GPS data was gathered on Blackburn Road and an intercepting gravel lot in Athens Ohio.
+Properties describing surface roughness and remission data were extracted and used to train Random Decision Forests.
 Terrain classification of consecutive scans compared to manually defined road and side-of-road surface areas yielded accuracy scores for gauging algorithm capability for road-surface detection.
-Use of multiple terrain classification algorithms was tested.
-Results indicate that a base accuracy rate of 36.1\%, 54.2\%, and 56.3\% for true positive detection rate for gravel, chipseal, and side-of-road surfaces.
-Number of consecutive points per scanning LiDAR channel quadrant influences terrain classification rate and accuracy, however accuracy increases were negligible.
-Distance from LiDAR point of origin influences terrain classification accuracy.
-Limiting classification by LiDAR channel indicated that a much higher classification accuracy may be achieved, accomplishing a peak gravel road detection accuracy of 72.3\%.
+It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively.
+Classified point clouds were manually examined and gravel areas were projected and compared with actual gravel surface areas.
+In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is a promising method of gravel road surface detection.
 This work addressed the problem of road surface detection on unmarked gravel and chipseal roads.
@@
 The impact of this work is that autonomous vehicles using LiDAR may be able to detect unmarked gravel and chipseal road surfaces, allowing autonomous operations on 1.5 million miles of previously undetected rural roads.}
+
 % For peer review papers, you can put extra information on the cover
@@
 \appendices
-\section{Equations}
+%\section{Equations}
 
-\begin{equation}\label{eqn:MLS}
-	\left[ {\begin{array}{cc}
-			\sum_{i=1}^{m} x_i z_i \\
-			\sum_{i=1}^{m} y_i z_i \\
-			\sum_{i=1}^{m} z_i \\
-
-	\end{array} } \right]
-	=
-	\left[ {\begin{array}{ccc}
-			\sum_{i=1}^{m} x_i^2 		& \sum_{i=1}^{m} x_i y_i 		& \sum_{i=1}^{m} x_i \\
-			\sum_{i=1}^{m} x_i y_i 		& \sum_{i=1}^{m} y_i^2 			& \sum_{i=1}^{m} y_i \\
-			\sum_{i=1}^{m} x_i 			& \sum_{i=1}^{m} y_i 			& \sum_{i=1}^{m} 1   \\
-	\end{array} } \right]
-	\left[ {\begin{array}{cc}
-			A\\
-			B\\
-			C\\
-	\end{array} } \right]
-\end{equation}
-\centering
-{Method of Least Squares Planar Projection}
+%\begin{equation}\label{eqn:MLS}
+%	\left[ {\begin{array}{cc}
+%			\sum_{i=1}^{m} x_i z_i \\
+%			\sum_{i=1}^{m} y_i z_i \\
+%			\sum_{i=1}^{m} z_i \\
+%			
+%	\end{array} } \right]
+%	=
+%	\left[ {\begin{array}{ccc}
+%			\sum_{i=1}^{m} x_i^2 		& \sum_{i=1}^{m} x_i y_i 		& \sum_{i=1}^{m} x_i \\
+%			\sum_{i=1}^{m} x_i y_i 		& \sum_{i=1}^{m} y_i^2 			& \sum_{i=1}^{m} y_i \\
+%			\sum_{i=1}^{m} x_i 			& \sum_{i=1}^{m} y_i 			& \sum_{i=1}^{m} 1   \\
+%	\end{array} } \right]
+%	\left[ {\begin{array}{cc}
+%			A\\
+%			B\\
+%			C\\
+%	\end{array} } \right]
+%\end{equation}
+%\centering
+%{Method of Least Squares Planar Projection}
 

### 2023-05-02 Huston_Paper.tex
removed redundant libraries
more updates:

@@
 \usepackage{dcolumn}  
-\usepackage{multirow} 
 \usepackage{ragged2e}
@@
 \usepackage{attachfile}
-\usepackage{multirow}
-\usepackage{tabularray}
+%\usepackage{tabularray}
 
@@
 
-	Gravel roads lack standardized features such as curbs or painted lines, presenting detection challenges to autonomous vehicles.
-Differentiation between road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation.
+	{Gravel roads lack standardized features such as curbs or painted lines, presenting detection challenges to autonomous vehicles.
 Global Positioning Service (GPS) and high resolution maps may not be reliable for navigation of gravel roads, as some may only be width of the vehicle and GPS may not be accurate enough.
@@
 The principal outcome of this work was a method for gravel road detection using LiDAR data for the purpose of predicting road boundary locations.
-Random Decision Forests were tested for the purpose of scanning LiDAR data terrain classification.
-Manually discovered gravel and paved road surface based on classified results were compared to actual road bounds.
+Random Decision Forests were trained using scanning LiDAR data terrain classification to detect unmarked gravel and asphalt surfaces.
 It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively.
-Classified point clouds were manually examined and gravel areas were projected and compared with actual gravel surface areas.
-Overlapping areas indicated a high level of unmarked gravel and asphalt road surface detection rate.
-Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.
+Overlapping results between manually projected and actual road surface areas resulted in 91.67\% intercepting gravel road detection accuracy.
+Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.}
 }	
@@
 Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.
-This work will attempt to determine a processing method for increasing road detection precision of unmarked gravel rural roads.}
+This work achieved 91.67\% intercepting gravel road detection accuracy with a 1.226 seconds per 360 degree scan classification rate.}
 
@@
 
-	{Feature extraction was completed by performing mathematical functions on the gathered spatial and remission training data (Figure \ref{fig:flowzchart}).
+	{Feature extraction was completed by performing mathematical functions on the gathered spatial and remission training data.
 Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial or remission data array.
-Features were non-dimensionalized as necessary by dividing by the average distance to the appropriate power to cancel units.
+Features were non-dimensionalized as necessary by dividing by the average range of the arc to LiDAR point of origin to the appropriate power.
 Thirty percent of the training data was randomly selected and reserved to create a validation data set.}
@@
 
-	\begin{figure}
-		\centering
-		\includegraphics[width=0.9\linewidth]{figures/flowz}
-		\caption[Project Flow]{Project flowchart}
-		\label{fig:flowzchart}
-	\end{figure}
+%   (Figure \ref{fig:flowzchart})
+%	\begin{figure}
+%		\centering
+%		\includegraphics[width=0.9\linewidth]{figures/flowz}
+%		\caption[Project Flow]{Project flowchart}
+%		\label{fig:flowzchart}
+%	\end{figure}
 
@@
 Transformation matrices were created by considering the GPS position and the IMU's roll, pitch, and yaw data.
-Tree scanning LiDAR arcs from three channels in front of the vehicle were classified.
+Three scanning LiDAR arcs from three channels in front of the vehicle were classified.
 Translation and rotation was applied to the averaged x, y, and z coordinates of each arc's terrain classification results.
@@
 	{Due to the side of the road was being re-grassed near area $3$, the algorithm sometimes struggled to differentiate the dirt from gravel, leading to some miss-classification of the side-of-road areas (Figure \ref{fig:pre-adjust} vs Figure \ref{fig:post-adjust}).
-Confidence scores may be exploited to adjust the final result by requiring a very high confidence for a gravel surface to be classified as such.
-All gravel surfaces that did not meet a 80\% confidence score was re-labeled as "unknown".
-Furthermore distances between the averaged coordinates of the arcs that were classified as gravel were compared and if the distance was below a threshold the arc was re-labeled as "unknown".
-This was done as distances between channels tended to certain distances and evenly spaced from each other when the data is on a flat surface, typical of road surfaces.
-With these slight adjustments, it was found that an average true-positive accuracy of all intercepting gravel surfaces was 67.54\%.
+Exploitation of confidence scores to adjust results by requiring a very high confidence for positive gravel surface classification was explored.
+All gravel surfaces that did not meet a 80\% confidence score was re-labeled as "unknown", however these adjustment had either a negligible or detrimental impact on the final results.
+Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 67.54\%.
 If any asphalt miss-classification on the gravel areas is included, the average true-positive road detection accuracy increases to 76.97\%.
 It was found that an average true-positive accuracy of all asphalt surfaces was 87.05\%.
-}
+Differentiation between gravel road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation.
+Side-of-road areas were established (Figure \ref{fig:prepostadjust}), and the overall average for "unknown" terrain type was found to be 77.49\%.}
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=0.9\linewidth]{figures/rm_db_4_percentages_big_anotated}
+		\caption[Area Percentage Scores]{Five runs of Blackburn Road with the intercepting gravel parking lot were classified.
+Shown is the results of one run.
+Percentages of each terrain type per area were calculated.
+Area $3$ is the gravel parking lot entrance from which training data was extracted.}
+		\label{fig:area_percentages}
+	\end{figure}
 
@@
 			\centering
-					\includegraphics[width=0.95\linewidth]{figures/pre-adjust}
+			\includegraphics[width=0.975\linewidth]{figures/side_of_road_nums_4_with_no_filts_2}
 			\caption[Pre-Adjusted]{}
@@
 			\centering
-					\includegraphics[width=0.95\linewidth]{figures/post-adjust}
+			\includegraphics[width=0.975\linewidth]{figures/side_of_road_nums_4_with_filts_2}
 			\caption[Post-Adjusted]{}
@@
 		\end{subfigure}
-				\label{fig:pre_post_adjust}
+		\label{fig:prepostadjust}
 		\caption[Post-Adjusted]{Raw classification result example (a).
-Classification results may be slightly altered by considering confidence scores and inter-arc distances to decrease miss-classification of gravel surfaces.
-(b).}
-	\end{figure}
-
-	\begin{figure}[H]
-		\centering
-		\includegraphics[width=0.9\linewidth]{figures/rm_db_4_percentages_big_anotated}
-		\caption[Area Percentage Scores]{Five runs of Blackburn Road with the intercepting gravel parking lot were classified.
-Percentages of each terrain type per area were calculated.
-Area $3$ is the entrance to the gravel parking lot from which training data was extracted.}
-		\label{fig:area_percentages}
+Results were slightly altered by considering confidence scores and inter-arc distances to decrease miss-classification of gravel surfaces, however the effect was negligible (b).}
 	\end{figure}
@@
 	{Classified point clouds were manually examined for 'best guess' locations for gravel and asphalt surfaces.
-Two out of three positive gravel identifications must be made on each channel, be evenly spaced, and be relatively co-planar to other road surface guesses in order for a positive guess for an asphalt or gravel surface to be made.
-It was found that the three intercepting gravel surfaces was consistently discovered (Figure \ref{fig:rm_db_4_toc}).}
+Two out of three positive gravel identifications must be made on each channel, be evenly spaced, and be relatively co-planar to other road surface guesses in order for a positive guess for an asphalt or gravel surface to be made (Figure \ref{fig:rm_db_4_toc}).
+Driveways $1$ and $2$ were detected in all five drive-bys, driveway $3$ was detected in four leading to an 91.67\% accuracy in intercepting gravel road detection.
+It was found that in three drive-bys there was a false positive driveway detection, however the location of the false-positive was not repeated.
+In all drive-bys the asphalt road was accurately detected.
+}
+
+	{Classification was completed using Ubuntu 18.04 and an AMD Ryzen 9 5900X 12 core 24 thread.
+Rate of classification for all nine areas of interest in a single scan was found to be an average of 1.226 seconds using a single core.
+Further optimization may increase this speed, however real-time classification is not a focus of this work.
+Using all cores when processing a ROSBAG file indicated that a rate of 55.81 feet per second or 38.05 miles per hour was accomplished.}
 
@@
 
+\section{Discussion}
+
 \section{Conclusion}
@@
 Properties describing surface roughness and remission data were extracted and used to train Random Decision Forests.
-Terrain classification of consecutive scans compared to manually defined road and side-of-road surface areas yielded accuracy scores for gauging algorithm capability for road-surface detection.
+Terrain classification of consecutive scans compared to manually defined road surface areas yielded accuracy scores for gauging algorithm capability for gravel surface detection.
 It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively.
 Classified point clouds were manually examined and gravel areas were projected and compared with actual gravel surface areas.
+Overlapping results between manually projected and actual road surface areas resulted in 91.67\% intercepting gravel road detection accuracy.
 In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is a promising method of gravel road surface detection.
@@
 
-	{The authors would like to thank the United States Dairy Farmers for their continuous cheese production. This work would not happen without them.}
+	{Special thanks to Dr. Jay Wilhelm for his guidance and Travis Moleski for his help in LiDAR map building.}
 

### 2023-05-02 Huston_Paper.tex
formatting + moar stuff

@@
 			\caption[Blackburn Road Camera View]{}
-			\label{fig:Bean_Cam_View}
+			\label{fig:Blackburn_Road_View}
 		\end{subfigure}
@@
 			\caption[Armig Road Camera View]{}
-			\label{fig:Armig_Road_Camera_View}
+			\label{fig:Gravel_Lot_View}
 		\end{subfigure}
@@
 
-%   (Figure \ref{fig:flowzchart})
-%	\begin{figure}
-%		\centering
-%		\includegraphics[width=0.9\linewidth]{figures/flowz}
-%		\caption[Project Flow]{Project flowchart}
-%		\label{fig:flowzchart}
-%	\end{figure}
-
-	% Quadrant size inversely affects rate of classification, as smaller quadrants allow for greater resolution at the expense of more computational time.
-
-%	{Testing consecutive scans required a method of compiling LiDAR data into a single point cloud and manually classifying areas for road detection accuracy.
-This was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
-Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data.
-Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU.
-LiDAR origin was found using GPS longitude, latitude, and altitude data.
-Physical distances between the GPS, IMU, and LiDAR was rectitude by obtaining reference frames provided by AutonomouStuff.
-GPS coordinates were offset by the current orientation and converts the ground truth to the LiDAR frame.
-GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors.
-Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}).
-While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for scoring purposes.
-Six stretches of unmarked roads were examined for accuracy: three of chipseal and three of gravel.
-Road and side-of-road areas were manually defined and projected unto consecutive LiDAR scans to score classification results.
-Due to the fuzzy nature of the classification results, the side-of-road areas were roughly a standard lane width to the sides of the road to accommodate the inability to define a rigid road edge.}
-
 	{Classifying consecutive LiDAR scans was accomplished by examination of nine arcs of interest in front and forty-five degree angles left and right from the front of the experimental apparatus.
@@
 Accuracy scores of each area was calculated, allowing evaluation of road detection performance.}
-
-%	{ Comparing averaged X, Y, and Z coordinates of the classified quadrant allowed for easier comparison and scoring of the RDF classification to the manually classified areas. 
-% Insert sentence or two on rate of of classification, feat extract, etc.; reference figure with rates> This was tested with a machine with 64 GB of DDR4 RAM and a Ryzen 9 5900X 12 core x 24 thread CPU.}
 
@@
 
-	{Due to the side of the road was being re-grassed near area $3$, the algorithm sometimes struggled to differentiate the dirt from gravel, leading to some miss-classification of the side-of-road areas (Figure \ref{fig:pre-adjust} vs Figure \ref{fig:post-adjust}).
-Exploitation of confidence scores to adjust results by requiring a very high confidence for positive gravel surface classification was explored.
-All gravel surfaces that did not meet a 80\% confidence score was re-labeled as "unknown", however these adjustment had either a negligible or detrimental impact on the final results.
-Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 67.54\%.
+	{Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 67.54\%.
 If any asphalt miss-classification on the gravel areas is included, the average true-positive road detection accuracy increases to 76.97\%.
@@
 
+%	\begin{figure}[H]
+%		\centering
+%		\begin{subfigure}{0.495\textwidth}
+%			\centering
+%			\includegraphics[width=0.975\linewidth]{figures/side_of_road_nums_4_with_no_filts_2}
+%			\caption[Pre-Adjusted]{}
+%			\label{fig:pre-adjust}	
+%		\end{subfigure}
+%		\begin{subfigure}{0.495\textwidth}
+%			\centering
+%			\includegraphics[width=0.975\linewidth]{figures/side_of_road_nums_4_with_filts_2}
+%			\caption[Post-Adjusted]{}
+%			\label{fig:post-adjust}
+%		\end{subfigure}
+%		\label{fig:prepostadjust}
+%		\caption[Post-Adjusted]{Raw classification result example (a).
+Results were slightly altered by considering confidence scores and inter-arc distances to decrease miss-classification of gravel surfaces, however the effect was negligible (b).}
+%	\end{figure}
+
 	\begin{figure}[H]
-		\centering
-		\begin{subfigure}{0.495\textwidth}
-			\centering
-			\includegraphics[width=0.975\linewidth]{figures/side_of_road_nums_4_with_no_filts_2}
-			\caption[Pre-Adjusted]{}
-			\label{fig:pre-adjust}	
-		\end{subfigure}
-		\begin{subfigure}{0.495\textwidth}
 			\centering
@@
 			\caption[Post-Adjusted]{}
-			\label{fig:post-adjust}
-		\end{subfigure}
 			\label{fig:prepostadjust}
-		\caption[Post-Adjusted]{Raw classification result example (a).
-Results were slightly altered by considering confidence scores and inter-arc distances to decrease miss-classification of gravel surfaces, however the effect was negligible (b).}
+		\caption[Area Scores]{Per-area scores were calculated, indicating algorithm efficiency of unmarked road detection. }
 	\end{figure}
@@
 \section{Discussion}
+
+	{Classification of real-world data outside of the training data set presents difficulties to the generated Random Decision Forest Classifiers generated in this work.
+Due to the side of the road was being re-grassed near area $3$ (Figure \ref{fig:area_percentages}, Figure \ref{fig:Blackburn_Road_View}), the algorithm struggled to differentiate the dirt from gravel, leading to some miss-classification of the side-of-road areas.
+Examination of the feature distribution shows a high amount of overlap (Figure \ref{}).
+Exploitation of confidence scores to adjust results by requiring a very high confidence for positive gravel surface classification was explored.
+All gravel surfaces that did not meet a 80\% confidence score was re-labeled as "unknown", however these adjustment had either a negligible or detrimental impact on the final results.
+Higher confidence score requirements resulted in "unknown" over-classification, while too low resulted in "gravel" over-classification.
+}
 

### 2023-05-02 Huston_Paper_asme.tex
formatting + moar stuff

--- 2023-05-02 Huston_Paper.tex
@@
-\documentclass[journal,onecolumn]{IEEEtran}
+\documentclass[balance,upint,subscriptcorrection,varvw,mathalfa=cal=boondoxo,colorlinks]{asmeconf}
 
-\hyphenation{op-tical net-works semi-conduc-tor}
-\usepackage[square,sort&compress,numbers]{natbib}
-\usepackage[margin=1in]{geometry}
-\usepackage{textcomp} 
-\usepackage{amssymb}  
-\usepackage{bm}       
-\usepackage{booktabs}
-\usepackage{dcolumn}  
-\usepackage{ragged2e}
-\usepackage{graphicx} 
-\usepackage{float}
-\usepackage{rotating}
-\usepackage{makecell}
-\usepackage{multirow}
-\usepackage[english]{babel}
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{listings}
-\usepackage{array}
-\usepackage[printonlyused]{acronym}
-\usepackage[framed, numbered]{matlab-prettifier}
-\usepackage[utf8]{inputenc}
-\usepackage[english]{babel}
-\usepackage{times}
-\usepackage{indentfirst}
-\usepackage{multicol}
-\usepackage{enumitem}
-\usepackage{attachfile}
-%\usepackage{tabularray}
+%\hyphenation{op-tical net-works semi-conduc-tor}
+%\usepackage[square,sort&compress,numbers]{natbib}
+%\usepackage{amssymb}  
+%\usepackage{bm}       
+%\usepackage{booktabs}
+%\usepackage{dcolumn}  
+%\usepackage{ragged2e}
+%\usepackage{graphicx} 
+%\usepackage{float}
+%\usepackage{rotating}
+%\usepackage{makecell}
+%\usepackage{multirow}
+%\usepackage[english]{babel}
+%\usepackage{subcaption}
+%\usepackage{array}
+%\usepackage[printonlyused]{acronym}
+%\usepackage[framed, numbered]{matlab-prettifier}
+%\usepackage[utf8]{inputenc}
+%\usepackage[english]{babel}
 
-% Table Stuff
-\newcommand{\spheading}[2][9.5em]{% \spheading[<width>]{<stuff>}
-	\rotatebox{90}{\parbox{#1}{\raggedright #2}}}
+%%%%%  pdf metadata  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\hypersetup{%
+pdfauthor={John H. Lienhard},
+pdftitle={ASME Conference Paper LaTeX Template},                  % <=== change to YOUR pdf file title
+pdfkeywords={ASME conference paper, LaTeX template, BibTeX style},% <=== change to YOUR pdf keywords
+pdfsubject = {Describes the asmeconf LaTeX template},			  % <=== change to YOUR subject
+%	pdfurl={https://ctan.org/pkg/asmeconf},% may delete
+pdflicenseurl={https://ctan.org/pkg/asmeconf},% may delete
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \begin{document}
+
+% Change these fields to the right content for your conference.
+% You can comment these out if for some reason you don't want a header.
+% Use title case (first letters capitalized), not all capitals
+
+\ConfName{Proceedings of the ASME 2023\linebreak International Mechanical Engineering Congress and Exposition}
+\ConfAcronym{IMECE2023}
+\ConfDate{October 29--November 2, 2023} % update 
+\ConfCity{New Orleans, LA} % update 
+\PaperNo{IMECE2023-XXXX}
+
+% Units of measure (e.g., cm) and other specialty lowercase terms in the title should be 
+%   enclosed in \NoCaseChange{...} to maintain lower case type
+%   LaTeX will automatically set the rest of the title in all capital letters.
+
+\title{Enhance Road Detection Data Processing of LiDAR Point Clouds to Specifically Identify Unmarked Gravel Rural Roads} % <=== replace with YOUR title
+%\title{Place Title Here: Place Subtitle After Colon}
+
+%   Put author names into the order you want. Use the same order for affiliations.
+%   \affil{#} tags the author's affiliation to the address in \SetAffiliation{#}.
+%   No space between last name and \affil{#}, separate names with commas.
 %
-	% paper title
-	% Titles are generally capitalized except for words such as a, an, and, as,
-	% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
-	% not capitalized unless they are the first or last word of the title.
-	% Linebreaks \\ can be used within to get better formatting as desired.
-	% Do not put math or special symbols in the title.
-	\title{Enhance Road Detection Data Processing of LiDAR Point Clouds to Specifically Identify Unmarked Gravel Rural Roads}
+%	For a sole author or a single affiliation for all authors, {#} may be left empty, as \affil{} and \SetAffiliation{} (but not with [grid] option!)
 %
+%   \CorrespondingAuthor{email} follows that author's affiliation, no spaces.  
+%   If multiple corresponding authors, put both email addresses in the same command and place after both authors.
 %
-	% author names and IEEE memberships
-	% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
-	% a structure at a ~ so this keeps an author's name from being broken across
-	% two lines.
-	% use \thanks{} to gain access to the first footnote area
-	% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-	% was not built to handle multiple paragraphs
-	%
+%   \JointFirstAuthor, if applicable, follows the affiliation of the relevant authors, no spaces.
 
-	\author{Rhett Huston, Dr. Jay Wilhelm}
-%	{
-%		and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
-%		\thanks{M. Shell was with the Department
-%			of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
-%			GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
-%		\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
-%		\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}}
+\SetAuthors{
+	Rhett Huston\affil{1}\JointFirstAuthor, 
+	Dr. Jay Wilhelm\affil{2}\JointFirstAuthor, 
+}
 
-	% note the % following the last \IEEEmembership and also \thanks - 
-	% these prevent an unwanted space from occurring between the last author name
-	% and the end of the author line. i.e., if you had this:
-	% 
-	% \author{....lastname \thanks{...} \thanks{...} }
-	%                     ^------------^------------^----Do not want these spaces!
-	%
-	% a space would be appended to the last name and could cause every name on that
-	% line to be shifted left slightly. This is one of those "LaTeX things". For
-	% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-	% "AB" then you have to do: "\textbf{A}\textbf{B}"
-	% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names. For what it is worth,
-% this is a minor point as most people would not even notice if the said evil
-% space somehow managed to creep in.
+\SetAffiliation{1}{Ohio University, Athens, Ohio}
+\SetAffiliation{2}{Ohio University, Athens, Ohio}
 
-% The paper headers
-\markboth{Journal of Autonomous Vehicles and Systems}%
-{}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
 \maketitle
 
-% As a general rule, do not put math, special symbols or citations
-% in the abstract or keywords.
+%%% Use this footnote for tracking various versions of your draft. Change text to suit your own needs. 
+%%% \date{..} calls the same command. 
+\versionfootnote{Documentation for \texttt{asmeconf.cls}: Version~\versionno, \today.}% <=== Delete before final submission.
+
+%%% Change these to your keywords.  Keywords are automatically printed at the end of the abstract.
+%%% This command MUST COME BEFORE the end of the abstract.
+%%% If you don't want keywords, leave the argument of \keywords{} empty (or use the abstract* environment)
+
+%	\keywords{ASME conference paper, \LaTeX\ Template, \hologo{BibTeX}}
+
+%%%%%  End of fields to be completed. Now write your paper. %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%  ABSTRACT  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%
+%% Abstract should be 200 words or less
 \begin{abstract}{
@@
 Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.}
+
 	}	
@@
 
-%	
-%	% Note that keywords are not normally used for peerreview papers.
-%	\begin{IEEEkeywords}
-%		IEEE, IEEEtran, journal, \LaTeX, paper, template.
-%	\end{IEEEkeywords}
-%	
-%	
-%	
-%
+%%%%%%%%%  NOMENCLATURE (OPTIONAL) %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%
+%% To change space between the symbols and  definitions, use \begin{nomenclature}[Xcm] where X is a number 
+	%% The unit cm can be replaced by any LaTeX unit of dimension: pt, in, ex, em, pc, etc.
+	%% Default is 2em.
+	%% \EntryHeading{..} produces an italicized subheading in the nomenclature list, e.g., \EntryHeading{Greek letters}
+
+	%%%%%%%%%  BODY OF PAPER %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-	%Real time description of surface noise based on incoming LiDAR data, then subsequent comparison to a standard, is a possible alternative that may allow for computationally efficient road surface detection.
-Overlaying camera imagery data with LiDAR allows for assigning incoming LiDAR data with color information, potentially increasing efficiency by narrowing the search area.
-
 	{Detection of unmarked road surfaces may utilize roughness properties without distinct markings or topographical features.
@@
 Implementing a terrain classification model using Random Decision Forests (RDF) using scanning LiDAR data to exploit surface roughness was studied for road surface detection in this study.
-Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.
-This work achieved 91.67\% intercepting gravel road detection accuracy with a 1.226 seconds per 360 degree scan classification rate.}
+Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.}
 
@@
 
+	\subsubsection{Overview}
+
+		{Followed methodology to accomplish this work is outlined in Figure \ref{fig:flowz_2}.
+The experimental apperatus for this work is the Ohio University Autonomous Van van (Figure \ref{fig:van}), which has a Velodyne VLP-32C scanning LiDAR sensor, a Novatel PwrPak 7D-E2 GNSS and INS system, and Mako cameras.
+Raw data was gathered using the Robotic Operating System (ROS) to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags.
+Scanning LiDAR data was sent from the sensor as a single 360 degree scan at 20 Hz.
+GNSS and IMU data was captured at a rate of 100 Hz.
+Rosbags were unpacked using MATLAB tools to separate data streams.
+Prior to training data extraction, manual dictation of gravel and asphalt areas is necessary.
+In order to do so, all scanning LiDAR data from a rosbag was compiled in order to create a single point cloud map.
+Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data.
+Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan (Figure \ref{fig:Combined_Classified_PCD}).
+Compiled point cloud maps may then be examined and compared to camera and satellite data to determine gravel and asphalt surface location.
+Manually defined 2D areas representing gravel and asphalt were projected unto the point cloud (Figure \ref{fig:Classified_Areas_PCD}).}
+
+		{A single fixed arc from three scanning LiDAR channels in front of the experimental apparatus were considered for classification.
+Future work may incorporate additional channels.
+Training data was extracted by examining each 360 scan and comparing the location of the three arcs and the manually projected 2D areas.
+If the arc resides within the manually defined surface area, raw spatial and intensity values from the arc were saved to a raw training database.
+Each channel had a separate training data base.
+Features that describe the raw spatial and intensity data were extracted by performing mathematical functions and saving the results to a feature data base.
+Thirty percent of the training data base was randomly selected and set aside as a validation data base.
+Training data was visually examined for clustering to verify that a Random Decision Forest would be able to be trained (Figure \ref{fig:training_data_cluster}).
+It was found that each class generally clustered to a certain range with some overlap, which indicates that a Random Decision Forest may be trained.
+MATLAB was used to train a Random Decision Forests for each of the three considered channels using Bayesian Optimization to tune tree depth and number of splits.
+Training continued until a minimum error was reached (Figure \ref{fig:c2_min_class_error}).}
+
+		{Validation of the }
+
+		%When the app performs hyperparameter tuning by using Bayesian optimization (see Optimization Options for a brief introduction), it chooses the set ofhyperparameter values that minimizes an upper confidence interval of the classification error objective model, rather than the set that minimizes the classification error.
+
+	\subsection{Detailed Methodology}
+
+	\begin{figure}
+		\centering
+		\includegraphics[width=\linewidth]{figures/flowz_2}
+		\caption[Project Flow]{High level view of project work}
+		\label{fig:flowz_2}
+	\end{figure}
+
+	\begin{figure}
+		\centering
+		\includegraphics[width=\linewidth]{figures/training_data_cluster}
+		\caption[Example Clustering]{Example of training data comparison between two features.}
+		\label{fig:training_data_cluster}
+	\end{figure}
+
+	\begin{figure}
+		\centering
+		\includegraphics[width=\linewidth]{figures/c2_min_class_error}
+		\caption[RDF Training Classification Error]{}
+		\label{fig:c2_min_class_error}
+	\end{figure}
+
 	{Required physical data was gathered by the Ohio University Autonomous Van van (Figure \ref{fig:van}), built by AutonomouStuff.
@@
 GPS data was recorded with a Novatel PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS), and two GNSS-502 antennas by NavtechGPS.
-Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using Robotic Operating System (ROS).
+Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using ROS.
 Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot that intersects Blackburn Road near Athens Ohio.
@@
 
-	\begin{figure}[H] 
+	\begin{figure} 
 		\centering
-		\begin{subfigure}{0.45\textwidth}
+		\begin{subfigure}{0.5\textwidth}
 			\centering
-			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/van_on_van}
+			\includegraphics[width=1\linewidth,height=4.60 cm,keepaspectratio]{figures/van_on_van}
 			\caption[Sensor Van]{}
@@
 		\end{subfigure}
-		\begin{subfigure}{0.45\textwidth}
+		\begin{subfigure}{0.5\textwidth}
 			\centering
-			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/LiDAR_on_van}
+			\includegraphics[width=1\linewidth,height=4.60 cm,keepaspectratio]{figures/LiDAR_on_van}
 			\caption[VLP 32 on Van]{}
@@
 
-	\begin{figure}[H]
+	\begin{figure}
 		\centering
@@
 Physical distances between the GPS, IMU, and LiDAR was rectitude by obtaining reference frames provided by AutonomouStuff.
-GPS coordinates were offset by the current orientation and converts the ground truth to the LiDAR frame.
+GPS coordinates were offset by the current orientation and converted the ground truth to the LiDAR frame.
 GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors.
@@
 }
+
+%%%%% Conclusions %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
+%%%%% Acknowledgments %%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
+\section*{Acknowledgments}
+Place any acknowledgments here.
 
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
+%%%  REFERENCES  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%
+%% Put your references into your .bib file in the usual way. Run latex once, bibtex once, then latex twice.
+%% The asmeconf.bst style allows @inproceedings and @proceedings to include: 
+%%		venue = {Location of Conference}, 
+%%		eventdate = {Month, days},
 
-\appendices
-%\section{Equations}
+\nocite{*}%% <=== Delete this line unless you want to typeset the entire contents of your .bib file!
 
-%\begin{equation}\label{eqn:MLS}
-%	\left[ {\begin{array}{cc}
-%			\sum_{i=1}^{m} x_i z_i \\
-%			\sum_{i=1}^{m} y_i z_i \\
-%			\sum_{i=1}^{m} z_i \\
-%			
-%	\end{array} } \right]
-%	=
-%	\left[ {\begin{array}{ccc}
-%			\sum_{i=1}^{m} x_i^2 		& \sum_{i=1}^{m} x_i y_i 		& \sum_{i=1}^{m} x_i \\
-%			\sum_{i=1}^{m} x_i y_i 		& \sum_{i=1}^{m} y_i^2 			& \sum_{i=1}^{m} y_i \\
-%			\sum_{i=1}^{m} x_i 			& \sum_{i=1}^{m} y_i 			& \sum_{i=1}^{m} 1   \\
-%	\end{array} } \right]
-%	\left[ {\begin{array}{cc}
-%			A\\
-%			B\\
-%			C\\
-%	\end{array} } \right]
-%\end{equation}
-%\centering
-%{Method of Least Squares Planar Projection}
-
-% you can choose not to have a title for an appendix
-%% if you want by leaving the argument blank
-%\section{}
-%Appendix two text goes here.
-
-% use section* for acknowledgment
-\justifying
-\section*{Acknowledgment}
-
-	{Special thanks to Dr. Jay Wilhelm for his guidance and Travis Moleski for his help in LiDAR map building.}
-
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
-\ifCLASSOPTIONcaptionsoff
-\newpage
-\fi
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-
-	\bibliographystyle{ieeetr}  
+\bibliographystyle{asmeconf}  %% .bst file following ASME conference format. Do not change.
 \bibliography{resources.bib}
 
-% that's all folks
+%%%  APPENDICES  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\appendix
+
 \end{document}

### 2023-05-03 Huston_Paper_asme.tex
methodolgy draft complete???

@@
 \hypersetup{%
-pdfauthor={John H. Lienhard},
-pdftitle={ASME Conference Paper LaTeX Template},                  % <=== change to YOUR pdf file title
-pdfkeywords={ASME conference paper, LaTeX template, BibTeX style},% <=== change to YOUR pdf keywords
-pdfsubject = {Describes the asmeconf LaTeX template},			  % <=== change to YOUR subject
-%	pdfurl={https://ctan.org/pkg/asmeconf},% may delete
-pdflicenseurl={https://ctan.org/pkg/asmeconf},% may delete
+pdfauthor={Rhett Huston},
+pdftitle={LiDAR Stuff},                  % <=== change to YOUR pdf file title
 }
@@
 	{Followed methodology to accomplish this work is outlined in Figure \ref{fig:flowz_2}.
-The experimental apperatus for this work is the Ohio University Autonomous Van van (Figure \ref{fig:van}), which has a Velodyne VLP-32C scanning LiDAR sensor, a Novatel PwrPak 7D-E2 GNSS and INS system, and Mako cameras.
+The experimental apparatus for this work was the Ohio University Autonomous Van van (Figure \ref{fig:van}), which has a Velodyne VLP-32C scanning LiDAR sensor, a Novatel PwrPak 7D-E2 GNSS and INS system, and Mako cameras.
 Raw data was gathered using the Robotic Operating System (ROS) to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags.
@@
 
-		{A single fixed arc from three scanning LiDAR channels in front of the experimental apparatus were considered for classification.
-Future work may incorporate additional channels.
-Training data was extracted by examining each 360 scan and comparing the location of the three arcs and the manually projected 2D areas.
+	{Training data was extracted by examining each 360 scan and comparing the location of a 6 degree arc directly in front of the area and the manually projected 2D areas.
 If the arc resides within the manually defined surface area, raw spatial and intensity values from the arc were saved to a raw training database.
-Each channel had a separate training data base.
+Seperate training data bases were used for each individual channel.
 Features that describe the raw spatial and intensity data were extracted by performing mathematical functions and saving the results to a feature data base.
@@
 MATLAB was used to train a Random Decision Forests for each of the three considered channels using Bayesian Optimization to tune tree depth and number of splits.
-Training continued until a minimum error was reached (Figure \ref{fig:c2_min_class_error}).}
+Training continued until a minimum error was reached (Figure \ref{fig:c2_min_class_error}).
+Validation of the RDF was then completed using the validation database (Figure \ref{fig:vali_conf_mat}).
+}
 
-		{Validation of the }
+	{Classification of five rosbags with data gathered from a section of road with three intercepting gravel driveways (one of the gravel driveways was the entrance to the source of the gravel training data).
+Three arcs from three channels were considered for classification - the centers of the arcs were directly in front of the vehicle, and 45 degrees left and right of center, creating nine total areas in a 3x3 grid in front of the van.
+Following the same process as aggregating point clouds, GNSS and IMU data informed the location and orientation of the arcs in each 360 degree scan.}
+
+	{Manually defined gravel, asphalt, and side-of-road areas were projected onto the averaged coordinates of all the classified arcs.
+Manually defined gravel, asphalt, and side-of-road areas were then examined for the number of type of classified arcs, yielding the accuracy scores for each area.
+Classified point clouds were then manually examined to determine if the location of gravel and asphalt surfaces may be identified.
+If consecutive points from each channel in any direction had two or more matching gravel or asphalt classification and was evenly spaced, a gravel or road surface area was projected.
+Guessed road surface areas were then compared to actual areas to determine the method's ability at unmarked gravel road discovery.}
 
 	%When the app performs hyperparameter tuning by using Bayesian optimization (see Optimization Options for a brief introduction), it chooses the set ofhyperparameter values that minimizes an upper confidence interval of the classification error objective model, rather than the set that minimizes the classification error.
-
-	\subsection{Detailed Methodology}
 
@@
 	\end{figure}
+
+	\subsection{Data Collection}
 
@@
 
+	\subsection{Training Data Extraction}
+
 	{MATLAB was used to process the gathered rosbag files using the their ROSBAG library.
@@
 Compiling LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
+Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data.
+Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan.
 Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data.
@@
 LiDAR origin was found using GPS longitude, latitude, and altitude data.
-Physical distances between the GPS, IMU, and LiDAR was rectitude by obtaining reference frames provided by AutonomouStuff.
+Physical distances between the GPS, IMU, and LiDAR sensors was rectified by obtaining reference frames provided by AutonomouStuff.
 GPS coordinates were offset by the current orientation and converted the ground truth to the LiDAR frame.
@@
 Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}).
-While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for scoring purposes.
+While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for this work.
 Manually defined areas may then be overlayed unto the point cloud (Figure \ref{fig:Combined_Classified_PCD}).
@@
 
+	\subsection{RDF Training and Testing}
+
 	{MATLAB's Classification Learner Application was used to generate a Random Decision Forests for each considered scanning LiDAR channel.
-Decision tree depth, number of learners, and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.}
+Decision tree depth, number of learners, and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.
+All arcs in a channel used the same RDF.
+}
 
-	{Classifying consecutive LiDAR scans was accomplished by examination of nine arcs of interest in front and forty-five degree angles left and right from the front of the experimental apparatus.
+	{Classifying consecutive LiDAR scans was accomplished by examination of the nine arcs of interest in front and forty-five degree angles left and right from the front of the experimental apparatus.
 Features from each arc were extracted and the arc was classified.

### 2023-05-03 asmeconf-template.tex
methodolgy draft complete???

% !TEX encoding = UTF-8 Unicode
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%
%%  This file is asmeconf-template.tex, a LaTeX template to format ASME Conference papers according to
%%  the requirements on ASME's conference web pages, and including hypertext support for the pdf.
%%
%%  This file is version 1.34 dated 2022/12/30
%%  
%%  As of version 1.11, this template defaults to ASME's newer conference guidelines first posted July 2019.
%% 			Those guidelines changed the requested author block formatting to be inline. 
%%			This LaTeX template continues to support the traditional grid format as a package option, [grid].
%%			Nomenclature now follows the abstract. Abstract text is set in italics.
%%
%%  Author: John H. Lienhard V
%%          Department of Mechanical Engineering
%%          Massachusetts Institute of Technology
%%          Cambridge, MA 02139-4307 USA
%%
%%  Class options include:
%%
%%          * Math options from M. Sharpe's newtxmath package: upright integrals [upint]; 
%%          *    [varvw] for a v and w that are better distinguished from Greek nu; and also 
%%          *    [subscriptcorrection, smallerops, varg, frenchmath, varbb, cmbraces, slantedGreek,...] 
%%			* 	 See newtx documentation for descriptions (at CTAN: http://ctan.org, v1.6 or higher is best).
%%
%%          * Options to: 	omit the ASME copyright footer [nofoot]; 
%%			*				use government employee copyright notice [govt];
%%			*				use government contractor copyright notice [contractor]
%%
%%          * An option to use the traditional grid arrangement of author names [grid].  
%%			*	 With this option, line breaks (\\) may be inserted into the address as needed. 
%%			*	 Author names that include commas should be enclosed in braces, e.g., {Joseph L. Smith, Jr.}.  
%%			*	 Authors may be grouped above a single affiliation using braces, e.g., {Henry Tudor, Catherine Parr}.
%%
%%          * An option to balance the heights of columns on the last page [balance]. 
%%          *    This option is NOT compatible with the [lineno] option.
%%
%%          * An option to include line numbers [lineno]. You must *run twice* for proper placement of the line numbers. 
%%			*	 The lineno package does not number titles, footnotes, captions, or tables.
%%          *    This option will disable balancing column height on final page if that option has been invoked.
%%          *    The lineno package won't always number the lines preceding displayed math in a paragraph because
%%          *    paragraph has not ended.  See that package's documentation for macros to address this problem, or
%%          *    just leave a blank line above the displayed equation while you are editing and then remove the 
%%          *    blank line and [lineno] option when you move to your final version.
%%
%%			* Options for PDF/A compliance. [pdf-a] will produce PDF/A-3u compliance with sRGB OutputIntent.
%%			*	 [pdfapart= 1 or 2 or 3] and [pdfaconformance= b or u] can enable levels 1b, 2b, 2u, and 3b.
%%			*    The most recent versions of LaTeX (2021 and later) are moving toward integrated support for pdf-a, 
%%			*    through \DeclareDocumentMetadata{..}. The asmeconf class supports these new features, which can 
%%			*	 replace the aforementioned class options. (An up-to-date LaTeX installation is required to use this.)
%%
%%          * Many options for calligraphic, script, and fraktur fonts from the mathalfa package; the
%%          *    example shown here is: [mathalfa=cal=boondoxo] to use a Boondox font for \mathcal.
%%          *    Some other options for cal are: dutchcal, zapfc, cm (default), euler,...
%%          *    frak (fraktur), bb (blackboard bold), scr (script) may also be chosen this way.
%%			*	 For details, refer to mathalfa documentation (at CTAN: http://ctan.org).
%%
%%          * Option to use superiors font from newtxtext for footnotes [nodefaultsups] and
%%          *    for slightly larger small capitals, [largesc], also from newtxtext.
%%
%%          * An option to allow hyphenation of the typewriter font [hyphenate], from inconsolata package.
%%          *    Hyphenation is normally suppressed for typewriter mode because it is often used for code.
%%			*	 To replace the default variable word spacing by monospacing, use the option [mono].
%%			*	 To get a zero without a slash, use [var0]
%%
%%          * Options (used by the babel package) to include passages in languages other than English (e.g., a translation 
%%			*    of the abstract). Languages are called as options, e.g. [french], [spanish], [greek], [russian], etc. 
%%			*    Language support is greatest when running LuaLaTeX with the [fontspec] option.
%%			*    See Appendix B for details.
%%
%%  The use of commands defined or modified by the asmeconf class is illustrated throughout this file. In particular, 
%%  ASME requires capitalized, sans-serif section headings, and as a result some care is needed when using macros 
%%  in section headings, as also illustrated below.
%%
%%  Use an up-to-date and complete installation of LaTeX, such as TeX Live 2020 or later. Errors may occur with old set-ups.
%%
 %=========================================================
%% 
%% LICENSE: 
%%
%% Copyright (c) 2022 John H. Lienhard
%%
%% Offered under the MIT license: https://ctan.org/license/mit 
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% New pdf management code (June 2021); with this, the class option [pdf-a] can be omitted.
%% This change to the LaTeX kernel is being phased-in by the LaTeX3 team. Can delete if it gives you trouble.
%% Under LuaLaTeX, choose pdfstandard=A-3b (and be cautious when loading extra fonts)

%\DocumentMetadata{%
%	pdfstandard=A-3b,% A-2b, A-2u, A-3b, or A-3u
%	pdfversion=1.7,
%	lang=en-US,
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Class options are described above. Change these options as desired. 
%%		If you are not using the language options, remove them (together with Appendices B and C)
%%	 	Remove the [colorlinks] option before *final* submission to ASME, to get black text for printing,
%%		but keep that option for other uses.

\documentclass[balance,upint,subscriptcorrection,varvw,mathalfa=cal=boondoxo,spanish,french,vietnamese,russian,greek,pdf-a,colorlinks]{asmeconf}

%%%%%  pdf metadata  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\hypersetup{%
	pdfauthor={John H. Lienhard},
	pdftitle={ASME Conference Paper LaTeX Template},                  % <=== change to YOUR pdf file title
	pdfkeywords={ASME conference paper, LaTeX template, BibTeX style},% <=== change to YOUR pdf keywords
	pdfsubject = {Describes the asmeconf LaTeX template},			  % <=== change to YOUR subject
%	pdfurl={https://ctan.org/pkg/asmeconf},% may delete
	pdflicenseurl={https://ctan.org/pkg/asmeconf},% may delete
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

% Change these fields to the right content for your conference.
% You can comment these out if for some reason you don't want a header.
% Use title case (first letters capitalized), not all capitals

\ConfName{Proceedings of the ASME 2023\linebreak International Mechanical Engineering Congress and Exposition}
\ConfAcronym{IMECE2023}
\ConfDate{October 29--November 2, 2023} % update 
\ConfCity{New Orleans, LA} % update 
\PaperNo{IMECE2023-XXXX}

% Units of measure (e.g., cm) and other specialty lowercase terms in the title should be 
%   enclosed in \NoCaseChange{...} to maintain lower case type
%   LaTeX will automatically set the rest of the title in all capital letters.

\title{A LATEX Template for ASME Conference Papers: \NoCaseChange{\texttt{asmeconf.cls}}} % <=== replace with YOUR title
%\title{Place Title Here: Place Subtitle After Colon}

%   Put author names into the order you want. Use the same order for affiliations.
%   \affil{#} tags the author's affiliation to the address in \SetAffiliation{#}.
%   No space between last name and \affil{#}, separate names with commas.
%
%	For a sole author or a single affiliation for all authors, {#} may be left empty, as \affil{} and \SetAffiliation{} (but not with [grid] option!)
%
%   \CorrespondingAuthor{email} follows that author's affiliation, no spaces.  
%   If multiple corresponding authors, put both email addresses in the same command and place after both authors.
%
%   \JointFirstAuthor, if applicable, follows the affiliation of the relevant authors, no spaces.

\SetAuthors{%
	John H.\ Lienhard\affil{1}\JointFirstAuthor\CorrespondingAuthor{}, 
	Luis Hern\'andez\affil{2}\JointFirstAuthor, 
	Maria Silva\affil{3}, 
	Henry Tudor\affil{4},  
	Catherine~Parr\affil{4}\CorrespondingAuthor{lienhard@mit.edu, kate@thepalace.gov}
	}

\SetAffiliation{1}{Massachusetts Institute of Technology, Cambridge, MA }
\SetAffiliation{2}{Institution or Company Name, City, State}
\SetAffiliation{3}{Institution or Company Name, City, Province, Canada}
\SetAffiliation{4}{Hampton Court Palace, Richmond, England}
%	Note: Luis and Maria are not real people.  Henry and Catherine have been dead for >450 years.

%	To switch from inline author names to gridded names, use the [grid] option.

\maketitle

%%% Use this footnote for tracking various versions of your draft. Change text to suit your own needs. 
%%% \date{..} calls the same command. 
\versionfootnote{Documentation for \texttt{asmeconf.cls}: Version~\versionno, \today.}% <=== Delete before final submission.

%%% Change these to your keywords.  Keywords are automatically printed at the end of the abstract.
%%% This command MUST COME BEFORE the end of the abstract.
%%% If you don't want keywords, leave the argument of \keywords{} empty (or use the abstract* environment)

\keywords{ASME conference paper, \LaTeX\ Template, \hologo{BibTeX}}

%%%%%  End of fields to be completed. Now write your paper. %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%  ABSTRACT  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% Abstract should be 200 words or less
\begin{abstract}

This paper is an example of and a  {\upshape\LaTeX} template for typesetting ASME conference papers using the {\upshape\texttt{asmeconf}} class. This  {\upshape\LaTeX} template follows ASME guidelines for margins, fonts, headings, captions, and reference formats as of 2022. The class is intended to be used with the {\upshape\texttt{asmeconf.bst} \hologo{BibTeX}} style for reference formatting, which is part of this distribution. The template produces pdfs that contain hyperlinks, bookmarks, and metadata; and references can include the DOI and URL fields. Links may be colored, for online use, or black, for publication. The class enables inline author names, following ASME's current style, but can also produce the traditional grid style. Options include line numbering, final column balancing, various math options, government copyright, and archivability (PDF/A). In addition, section headers may contain mathematics, references, citations, and footnotes. The class is compatible with {\upshape\hologo{pdfLaTeX}} or {\upshape\hologo{LuaLaTeX}}.
\end{abstract}

%%%%%%%%%  NOMENCLATURE (OPTIONAL) %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% To change space between the symbols and  definitions, use \begin{nomenclature}[Xcm] where X is a number 
%% The unit cm can be replaced by any LaTeX unit of dimension: pt, in, ex, em, pc, etc.
%% Default is 2em.
%% \EntryHeading{..} produces an italicized subheading in the nomenclature list, e.g., \EntryHeading{Greek letters}

\begin{nomenclature}
\EntryHeading{Roman letters}
\entry{$k$}{Thermal conductivity [W m$^{-1}$ K$^{-1}$]}
\entry{$\vec{q}$}{Heat flux vector [W m$^{-2}$]}

\EntryHeading{Greek letters}
\entry{$\alpha$}{Thermal diffusivity [m$^2$ s$^{-1}$]}
\entry{$\nu$}{Kinematic viscosity [m$^2$ s$^{-1}$]}

\EntryHeading{Dimensionless groups}
\entry{Pr}{Prandtl number, $\nu/\alpha$}
\entry{Sc}{Schmidt number, $\nu/\mathcal{D}_{1,2}$}

\EntryHeading{Superscripts and subscripts}
\entry{b}{bulk value}
\entry{$\infty$}{free stream value}
\end{nomenclature}

%%%%%%%%%  BODY OF PAPER %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
The \texttt{\hrefurl{https://ctan.org/pkg/asmeconf}{asmeconf}} class typesets papers with margins, fonts, headings, captions, and reference formats that follow the specifications for conference papers of the American Society of Mechanical Engineers (ASME). In contrast to older ASME \LaTeX\ templates, \texttt{asmeconf} will set internal and external hyperlinks, and the pdf file will contain bookmarks and metadata. Many other useful features have been incorporated. This class is not a publication of ASME, but the author has published in ASME conferences since 1983.

The \texttt{.tex} file may be written using standard \LaTeX\ commands, although some specific initial commands are needed to format the blocks containing the author[s], title, and abstract.  This class loads a number of other packages, all of which are contained in up-to-date versions of \hrefurl{https://www.tug.org/texlive/}{\TeX\ Live}, \hrefurl{http://www.tug.org/mactex/}{Mac\TeX}, and similar platforms. If you get an error message about a missing package, you may download it at no cost from CTAN (\hrefurl{https://ctan.org}{ctan.org}).

\subsection{Essential Initial Commands}

To begin, fill in the fields to be completed at top of the \texttt{asmeconf-template.tex} file. These fields include the headers for your conference and your paper number. Specified metadata will be placed into the pdf file itself. 
The title should be placed into \verb|\title{..}|.

Put author names into the \verb|\SetAuthors{name, name,...}| command in the desired order; follow the syntax illustrated \texttt{asmeconf-template.tex} file. Put each distinct address sequentially into a separate \verb|\SetAffiliation{n}{address}|, where $n = 1,2,\ldots$ Tag each author with an affiliation by putting \verb|\affil{n}| after that author's name inside the \verb|\SetAuthors{..| command.

Keep author addresses short.  List the author institution, and the City, State (US authors), City, Province, Canada (Canadian authors), or City, Country (other international authors).

One author (or more) may be designated as the corresponding author by placing \verb|\CorrespondingAuthor{email}|  after \verb|\affil{n}|. Two or more authors may be joint first authors by putting \verb|\JointFirstAuthor| after \verb|\affil{n}|.

After setting up the headers, authors,  and title, issue the \verb|\maketitle| command.

The abstract text must be placed into \verb|\begin{abstract}| \ldots \verb|\end{abstract}|. The abstract will automatically be italicized. Keywords may be included using the \verb|\keywords{..}| command. The \texttt{keyword} command \textit{must} be issued before the abstract environment.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Referring to Citations, Figures, and Equations}

Citations are automatically numbered \cite{ning2002}. They should be inserted in the text using a \verb|\cite{ref}| command~\cite{gibson2008,stevens1999}. The citations will be automatically sorted and compressed if they are given in a set \cite{stevens1999,ning2002,gibson2008,wions2005,smith2002,watson1982}. 
A specific reference may be named with an abbreviation, as in Ref.~\cite{watson1982}.
See the \texttt{asmeconf-sample.bib} file and Sect.~\ref{sec:references} for examples of entering references.

For ASME conference papers, the labels Equation and Figure should be abbreviated when they do not start a sentence, as in  Eq.~\eqref{eqn:dw} and Fig.~\ref{fig:1}. Figure~\ref{fig:1} is spelled out when it starts a sentence. Equation~\eqref{eqn:dw} is spelled out when it starts a sentence.

Equations are typeset in the usual way and will be automatically numbered.  The class file loads the \texttt{amsmath} and \texttt{mathtools} packages. Further, the \texttt{newtxmath} package used for the math fonts includes many additional features (see Sect.~\ref{sec:moremath}).
\begin{equation}\label{eqn:fourier}
\vec{q} = -k\nabla T
\end{equation}

ASME prefers SI units. (U.S.\ style units may follow in parentheses.) Be sure to put all symbols into the nomenclature list, including their units.

%%%%%%%%%%%%% begin figure %%%%%%%%%%%%%%%%%

%% captions go below figures

\begin{figure}
\centering\includegraphics[width=0.7\linewidth]{sample-figure-1.pdf}
\caption{Caption with math, eqn.~\eqref{eqn:fourier}: $z = (r,\phi)$ \cite{Lienhard2019b}}\label{fig:1}
\end{figure}

%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use title case for subsections and subsubsections (first letter of words capitalized)

\section{Section Headings and Captions}

ASME requires that section headings and captions be set in an uppercase, sans serif font.  The class will do this automatically.  You can place \verb|\cite{..}|, \verb|\ref{..}|, \verb|\label{..}|, and mathematics into headings and captions directly, as you would in the main text. Do not enclose them braces, e.g.\ \verb|{\cite{..}}|, which will cause errors. You can place \verb|\footnote{..}| into headings, but not into captions.\footnote{See \texttt{tex-stackexchange} for various approaches to footnotes in captions, if they seem necessary. For footnotes in tables, use the \texttt{tablefootnote} package.}\footnote{Sequential footnotes are automatically separated by a comma.}

Text in section headings and captions will not be capitalized if enclosed in a \verb|\NoCaseChange{..}| command.

Sections may either be numbered or left unnumbered.

Simple mathematical expressions can be used in either captions or section headings. For a section heading that includes more complicated math (and macros), you may use the optional argument of \verb|\section[..]{..}| to create a pdf bookmark without losing characters or producing warnings or errors. See the \texttt{asmeconf-template.tex} source file for examples of this procedure. These bookmarks should usually be text expressions, although some math is supported.

To override the \texttt{sansbold} math version in captions, put \verb|\NoCaseChange{\mathversion{normal}}| in the caption.

\subsection{Subsection and Sub-subsection Headings}

Subsections and sub-subsection headings should be entered in title case, with the first letter of primary words capitalized. Sub-subsections (i.e., paragraphs) are never numbered.

%%%%%%%%%%%%%%% begin simple table %%%%%%%%%%%%%%%%%%%%%%%%%%

%% Captions go above tables
%%
%% Note that placement of figures and tables can managed with the [!tbhp] options. See: https://latexref.xyz/dev/latex2e.html#Floats

\begin{table}[t]
\caption[Table]{A simple table}\label{tab:1}
\centering{%
\begin{tabular}{llr}
\toprule
Experiment & $u$ [m/s] & $T$ [\textdegree C] \\
\midrule
Run 11 & 12.5 & 103.4 \\
Run 12 & 24   & 68.3 \\
\bottomrule
\end{tabular}
}
\end{table}

%%%%%%%%%%%%%%%% end table  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%% begin more complicated table %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{table}[t]
\caption{Table with more complicated columns}\label{tab:2}%
\centering{%
\begin{tabular}{!{\hspace*{0.5cm}} >{\raggedright\hangindent=1em} p{3cm} d{3.3} @{\hspace*{1cm}} d{3.3} !{\hspace*{0.5cm}}}
\toprule
Experiment & \multicolumn{1}{c@{\hspace*{1cm}}}{$u$ [m/s]} & \multicolumn{1}{c!{\hspace*{0.5cm}}}{$T$ [\textdegree C]} \\
\midrule
The first test we ran this morning   & 124.3     &   68.3   \\
The second test we ran this morning  &  82.50    &  103.46  \\
Our competitor's test                &  72.321   &  141.384 \\
\bottomrule
\end{tabular}
}
\end{table}

%%%%%%%%%%%%%%%% end table  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%% begin two column table %%%%%%%%%%%%%%%%%% 
\begin{table*}
\caption{A table spanning two columns}\label{tab:3}%
\centering{%
\begin{tabular*}{0.8\textwidth}{@{\hspace*{1.5em}}@{\extracolsep{\fill}}ccc!{\hspace*{3.em}}ccc@{\hspace*{1.5em}}}
\toprule
\multicolumn{1}{@{\hspace*{1.5em}}c}{$x$\rule{0pt}{11pt}} &
\multicolumn{1}{c}{$\textrm{erf}(x)$} &
\multicolumn{1}{c!{\hspace*{3.em}}}{$\textrm{erfc}(x)$} &
\multicolumn{1}{c}{$x$} &
\multicolumn{1}{c}{$\textrm{erf}(x)$} &
\multicolumn{1}{c@{\hspace*{1.5em}}}{$\textrm{erfc}(x)$} \\ \midrule
0.00 & 0.00000 & 1.00000 & 1.10 & 0.88021 & 0.11980\rule{0pt}{11pt} \\
0.05 & 0.05637 & 0.94363 & 1.20 & 0.91031 & 0.08969 \\
0.10 & 0.11246 & 0.88754 & 1.30 & 0.93401 & 0.06599 \\
0.15 & 0.16800 & 0.83200 & 1.40 & 0.95229 & 0.04771 \\
0.20 & 0.22270 & 0.77730 & 1.50 & 0.96611 & 0.03389 \\
0.30 & 0.32863 & 0.67137 & 1.60 & 0.97635 & 0.02365 \\
0.40 & 0.42839 & 0.57161 & 1.70 & 0.98379 & 0.01621 \\
0.50 & 0.52050 & 0.47950 & 1.80 & 0.98909 & 0.01091 \\
0.60 & 0.60386 & 0.39614 & 1.82\makebox[0pt][l]{14} & 0.99000 & 0.01000 \\
0.70 & 0.67780 & 0.32220 & 1.90 & 0.99279 & 0.00721 \\
0.80 & 0.74210 & 0.25790 & 2.00 & 0.99532 & 0.00468 \\
0.90 & 0.79691 & 0.20309 & 2.50 & 0.99959 & 0.00041 \\
1.00 & 0.84270 & 0.15730 & 3.00 & 0.99998 & 0.00002 \\[2pt]
\bottomrule\end{tabular*}
}
\end{table*}

%%%%%%%%%%%%%%%%% end two column table  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tables and Figures}

Table \ref{tab:1} is an example of a simple table. Table captions should be placed above tables.
The class loads the \texttt{booktabs} package (used for horizontal rules in Tables \ref{tab:1} and \ref{tab:2}), and the \texttt{array} and \texttt{dcolumn} packages which provide extended capabilities for columns in the \texttt{tabular} environment (see Table \ref{tab:2}).  Table \ref{tab:3} is an example of a table that spans two columns. Two column tables (and figures) will always float to the top of a later page.

Figure captions go below figures. Figure~\ref{fig:2} is an example of a figure that spans two columns and includes subfigures. The text in figures (and tables) should be no smaller than 6~point type. Images in figures are handled by the standard \texttt{graphicx} package.

Landscape figures and tables may be produced at full-page size by putting \verb|\usepackage[figuresright]{rotating}| in your \texttt{.tex} file's preamble and using the \texttt{sidewaystable*} and \texttt{sidewaysfigure*} environments~\cite{fairbairns}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Reference Formatting with \NoCaseChange{\texttt{asmeconf.bst}}\footnote{To prevent capitalization of text in a section heading or caption, such as an SI unit, enclose it in a \texttt{\textbackslash NoCaseChange} command. As of the July 2022 release of \LaTeX, commands used in a heading or caption may be protected globally by putting this in the preamble: \texttt{\textbackslash AddToNoCaseChangeList\{\textbackslash MyCommand\}}.}}\label{sec:references}

The {\upshape\texttt{asmeconf.bst}} \hologo{BibTeX}  style follows the reference styles shown on ASME's conference web site in  2022.\footnote{\texttt{asmeconf.bst} is intended as a replacement for the old \texttt{asmems4.bst}, which does not follow ASME's current reference formats or support DOI and URL.}
Examples for these and many other cases are given in the \texttt{asmeconf-sample.bib} file, which is part of this distribution. Citations and references are managed by the standard \texttt{natbib} package.  Nevertheless, a few comments are necessary.

%% sub-subsections should *not* be numbered according to ASME's style

\subsubsection*{DOI, URL, and eprint} Include DOI numbers when they are available.  URL's may alternatively be given. ASME requests that URLs point to a document's abstract.

Basic support for \texttt{eprint} numbers is also included, generating a url at the end of the citation. The \texttt{archive} type may be specified using the macros \texttt{arxiv, google\-books, hdl, jstore, oclc}, or \texttt{pubmed} (e.g., \texttt{archive=hdl},  \textit{without} braces). Both \texttt{eprint} and \texttt{archive} fields \textit{must} be given. Other root urls may be invoked using \verb|archive = {https://another.url.org/}|.

\subsubsection*{Online Sources} A bibliography entry \verb|@online{..| is included for citation of online sources, such as web pages. A \texttt{url} or \texttt{eprint} with \texttt{archive} must be included. See the examples of use in the \texttt{asmeconf-sample.bib} file.

\subsubsection*{Date Accessed} The \verb|urldate={..}| field may be used to provide the date on which a given url was accessed. By default, the text printed will be \texttt{Accessed `date',}. The word ``Accessed'' may be changed using the \verb|urltype={..}| field.

\subsubsection*{Conference Location and Date} To specify the city and date of a conference, you can use \verb|venue={..}| and \verb|eventdate={..}| with the entries \verb|@inproceeedings{..| and \verb|@proceeedings{..|.

\subsubsection*{Capitalization of Titles} ASME's bibliography style requires that document titles be in title case. The first letters of principal words are capitalized. Do this in the \texttt{.bib} file.

%%%%%%%%%%%%%%%%%  begin two column figure  %%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure*}
\begin{subfigure}[t]{0.5\textwidth} % subfigure is basically the same as minipage
\vbox{
\vspace*{1.7em}
\centering{
  \includegraphics{sample-figure-2a.pdf}
}%
\vspace*{1.7em}
}%
\subcaption{Interior region\label{fig:interior-region}}
\end{subfigure}%
%%%%%%%%%%%%% no spaces or line breaks between these two subfigures
\begin{subfigure}[t]{0.5\textwidth}
\centering{%
\includegraphics{sample-figure-2b.pdf}
}%
\subcaption{Exterior region\label{fig:exterior-region}}
\end{subfigure}
\caption{A figure with two subfigures  \cite{lienhard2023}\label{fig:2}}
\end{figure*}

%%%%%%%%%%%%%%%%%%%  end two column figure  %%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%  MORE ON MATH   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Here is an example of managing complicated math in a section or subsection heading: 
%%    the optional argument to \section will provide the pdf bookmark
%%    without losing characters or producing warnings/errors.
%%
%% In this heading, letter u is forced to be upright with \mathrm{u}
%%
\section[More on math: u\cdot\omega=0]{More on math: $\vec{\mathrm{u}}\cdot\vec{\omega}=0$}\label{sec:moremath}

In most cases, the need for a wide equation can be eliminated by using one of the multiline equation environments defined by 
\texttt{amsmath}, such as \texttt{align}, \texttt{split}, or \texttt{multline}~\cite{amsmath}. The following equation is set with the 
\texttt{multline} environment:
\begin{multline}\label{eqn:energy}
\frac{\partial}{\partial t}\left[\rho\bigl(e + \lvert\vec{u}\rvert^2\big/2\bigr)\right]  + \nabla\cdot\left[\rho\bigl(h + \lvert\vec{u}\rvert^2\big/2 \bigr)\vec{u}\right] \\
 ={}-\nabla \cdot \vec{q} +  \rho \vec{u}\cdot\vec{g}+ \frac{\partial}{\partial x_j}\bigl(d_{ji}u_i\bigr) + \dot{Q}_v
\end{multline}
An example using \texttt{align} appears in Appendix~\ref{appendix:a}.

An alternative solution may be to set large equations into two-column-wide tables or figures. An experimental package for setting equations that span two columns, \texttt{asmewide.sty}, can be loaded as well, but that code may require hand-fitting around figures, tables, and page breaks. See the examples in~\cite{lienhard2022}.

Math italics are used for Roman and Greek letters by default.  If you want an upright letter in math, you can use the relevant math alphabet, e.g., \verb|\mathrm, \mathbf, \mathsf|:
\begin{equation}\label{eqn:dw}
\vec{F} = m \vec{a} \quad\textrm{or}\quad \vec{\mathrm{F}} = m \vec{\mathrm{a}} \quad\textrm{or}\quad \mathbf{F} = m \mathbf{a} \quad\textrm{or}\quad \vec{\mathsf{F}} = m \vec{\mathsf{a}}
\end{equation}
To get additional symbols in bold math, use the \verb|\bm{..}| macro from the \texttt{bm} package (which is loaded by the class) or, for longer passages, use \verb|{\mathversion{bold}|\ldots\texttt{\}}.

The class file also provides upright sans-serif Greek letters with \verb|\sfalpha| and similar expressions (e.g., $\sfalpha, \sfbeta, \sfgamma, \sfdelta$ \ldots $\bm{\sfalpha, \sfbeta, \sfgamma, \sfdelta \ldots}$), in case they are needed (but note that the \verb|newtxmath| options \verb|frenchmath| and \verb|slantedGreek| also affect how Greek letters are presented).

\subsection{The \texttt{newtxmath} and \texttt{mathalpha} Packages~\cite{sharpe1,sharpe2}} The \texttt{newtxmath} package~\cite{sharpe1}, loaded by default, includes many options for mathematics, most of which can be called as options to \verb|\documentclass|. For example, the \texttt{upint} option selects upright integral signs (rather than slanted integral signs):
\begin{quote}
\verb|\documentclass[upint]{asmeconf}|. 
\end{quote}  
The option \verb|subscriptcorrection| improves the spacing of math subscripts. These math options are discussed further in the \texttt{asmeconf-template.tex} file.

In addition, many options for calligraphic, fraktur, and script fonts are available as options to the \texttt{mathalfa} package, which is also loaded. These may be invoked, for example, as 
\begin{center}
\verb|\documentclass[mathalfa=cal=boondoxo]{asmeconf}| 
\end{center}
which selects a Boondox font for \verb|\mathcal|, as in $A \in \mathcal{P}(A)$. To find all the font options, refer to the \texttt{mathalfa} package documentation \cite{sharpe2}.

The \texttt{asmeconf} class is designed to be used with \texttt{newtxmath} and does not support the \texttt{unicode-math} package.

%%%%%%%%%%%%%%%  ADDITIONAL PACKAGE OPTIONS  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Additional Options for \NoCaseChange{\texttt{asmeconf.cls}}}
The class accepts a number of options in addition to those already described. These options are discussed next.

\subsection{Colored Hyperlinks}
ASME requires that all text be \textbf{in black} when the paper is submitted for publication.  For other uses, authors may
obtain colored hyperlinks with the [\texttt{colorlinks}] option.

\subsection{Final Column Balancing} The option \texttt{[balance]} invokes the the \texttt{flushend} package~\cite{tolusis}.
This package will attempt to give equal height to the two columns on the last page. The performance of this package is sometimes inconsistent (with odd page layout or, very rarely, errors), so use this option with caution.

\subsection{Line Numbers} The option \texttt{[lineno]} invokes the the \texttt{lineno} package~\cite{bottcher}. This option will produce line numbers in the margins. You must run \LaTeX\ \textit{twice} for proper placement of the numbers. Tables, captions, and footnotes will not be numbered.  Line numbers can be helpful for review and editing, but should not be used in your final manuscript. See the documentation of the \texttt{lineno} package for further commands to control line numbering.

The \texttt{lineno} package is not compatible with the \texttt{flushend} package that makes final short columns the same height. Balancing is automatically disabled when this option is called.

\subsection{Grid-Style Author Block} The option \texttt{[grid]} invokes ASME's grid-style arrangement of author names. Author names are recognized by the commas that separate them. (To include a comma in a name, enclose the name in braces.) Line breaks (\verb|\\|) may be inserted into the address of \verb|\SetAffiliation{n}{address}| as needed.

Note that ASME interprets the author order in the grid style by reading names from left-to-right in the top row, then left-to-right in each subsequent row.

\subsection{Changing the Copyright Footer} The option \texttt{[nofoot]} will omit the ASME copyright from the page footer. The option \texttt{[govt]} will produce a copyright notice for authors who are employees of the U.\ S.\ Government.  The option \texttt{[contractor]} will produce a copyright
notice for authors who are employed by a U.\ S.\ Government contractor.

The footers are generated with the \texttt{fancyhdr} package~\cite{oostrum} and can be changed using the commands of that package. Only the default arrangement matches ASME's style, however.

\subsection{Archivability:~PDF/A} Compliance with PDF/A standards can be enabled using the option \texttt{[pdf-a]} 
when running with \hologo{pdfLaTeX}. The default setting is for PDF/A-3u with sRGB OutputIntent (\texttt{sRGB.icc}). 
If levels 1b, 2b, 2u, or 3b are preferred, use the options \texttt{[pdfapart=1 or 2 or 3]} and 
\texttt{[pdfaconformance=b or u]}. Note that accessible conformance~(\texttt{a}) is not currently possible with \LaTeX.

As of late 2021, the \LaTeX 3 team is phasing in native support for PDF/A in the \LaTeX\ kernel, which will make these
class options unnecessary when using up-to-date installations.

\subsection{Superiors Font} The \texttt{newtxtext} package includes a superiors font (both numbers and letters) for use in footnote markers and text superscripts. To enable this font, use the option \texttt{[nodefaultsups]}.

\subsection{Typewriter Font Options} This font is the sans-serif \texttt{inconsolata}. By default, the word spacing is variable, but option \texttt{[mono]} ends this behavior. A slashed zero is the default; option \texttt{[var0]} removes the slash. Option \texttt{[hyphenate]} enables hyphenation. (The hyphenation option is not available if the \texttt{[fontspec]} option is used.)

\subsection{Support for Other Languages}  This package can be adapted to incorporate (or entirely use) languages other than English. See Appendix \ref{appendix:b} for details.

%%%%% Conclusions %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}
Provide a brief conclusion (3-4 lines).

%%%%% Acknowledgments %%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Acknowledgments}
Place any acknowledgments here.

%%%  REFERENCES  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% Put your references into your .bib file in the usual way. Run latex once, bibtex once, then latex twice.
%% The asmeconf.bst style allows @inproceedings and @proceedings to include: 
%%		venue = {Location of Conference}, 
%%		eventdate = {Month, days},

\nocite{*}%% <=== Delete this line unless you want to typeset the entire contents of your .bib file!

\bibliographystyle{asmeconf}  %% .bst file following ASME conference format. Do not change.
\bibliography{asmeconf-sample}%% <=== change this to name of your bib file

%%%  APPENDICES  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix

%% Note that appendices will be "numbered" A, B, C, ... etc. Use \section, not \section*
%% Equations will be numbered sequentially following those in the paper. Do not reset the equation counter.

%% Here we use the optional argument to control the pdf bookmark and prevent errors.
\section[The Vector Product A\times B]{The vector product $\vec{A}\times\vec{B}$}\label{appendix:a}

This brief illustration of an appendix shows the numbering of the appendix and equations. Equations are numbered
consecutively, following those in the paper. Consider $\rho \neq \textrm{fn}(p)$:
\begin{align}
\frac{d\Gamma}{dt} &{}= \frac{d}{dt} \int_{\mathcal{C}} \mathbf{u} \cdot d\mathbf{r}\\
				   &{}= \int_{\mathcal{C}} \frac{D\mathbf{u}}{Dt} \cdot d\mathbf{r} + \underbrace{\int_{\mathcal{C}} \mathbf{u}\cdot d\biggl( \frac{d\mathbf{r}}{dt}\Biggr)}_{=\, 0} \\[-2pt]
                   &{}= \iint_{\mathcal{S}} \nabla \times \frac{D\mathbf{u}}{Dt}  \cdot d\mathbf{A}\\
                   &{}= \iint_{\mathcal{S}}  \nabla p \times \nabla \left( \frac{1}{\rho}\right) \cdot d\mathbf{A}
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Multilingual Support}\label{appendix:b}

ASME publishes in English, but the \texttt{babel} package is loaded for 
users who may wish to include other languages. For example, an author might wish to include an appendix that provides the 
abstract in another language.

When more than one language option is included in \verb|\documentclass[..]{asmeconf}|, English will be 
assumed to be the main language of the document. (To choose a different main language, set \texttt{[main=..]}).
If no language options are given, the package defaults to English.  As examples, a passage in French is 
shown in \selectlanguage{french}\appendixname~\ref{app:fourier}\selectlanguage{english}, followed by 
\ifpdftex abstracts in Spanish, Greek, Russian, and Vietnamese.\else abstracts in other languages.\fi

The input encoding can be utf-8, as for these glyphs:
%% If you have trouble with the next line, your file may not be saved in utf-8 format. You can delete that line to resolve the issue.
\typeout{If you have trouble with the next line, your file may not be saved in utf-8 format. You can delete that line to resolve the issue. Under LuaLaTeX, you can load the [fontspec] option if you have the relevant systems fonts installed}%
àáâäæãåā  èéęëêēė  îïíīįì ôöòóœøōõ ûüùúū çćč ł ñń ßśš ÿ žźż.

Fonts similar to Times/Helvetica are automatically used when the Greek, Vietnamese, or selected cyrillic-alphabet languages are called as options under {\upshape\hologo{pdfLaTeX}}. Using {\upshape\hologo{LuaLaTeX}} with the \texttt{[fontspec]} option, many additional scripts are available; see the supplemental notes for such usage~\cite{lienhard2021}. Possibilities include Arabic, Bengali, Chinese, Devanagari (e.g., for Hindi), Hangul (for Korean), Kana (for Japanese), and Tamil. \textit{The {\upshape\texttt{[fontspec]}} option requires a \LaTeX\ installation dated October 2020 or later.}

The bibliography style, \texttt{asmeconf.bst}, is designed in English and aimed at \hologo{BibTeX}.  
% Multilingual bibliographies can be supported using \texttt{BibLaTeX}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{selectlanguage}{french}%
\section{Discours Préliminaire de Fourier}\label{app:fourier}
Les causes primordiales ne nous sont point con­nues; mais elles sont assujetties à des lois simples et constantes, que l'on peut découvrir par l'obser­vation, et dont l'étude est l'objet de la philosophie naturelle.

La chale ur pénètre, comme la gravité, toutes les substances de l'univers, ses rayons occupent toutes les parties de l'espace. Le but de notre ouvrage est d'exposer les lois mathématiques que suit cet élé­ment. Cette théorie formera désormais une des branches les plus importantes de la physique gé­nérale~\cite{fourier1822}. 
\end{selectlanguage}%

\begin{selectlanguage}{spanish}%
\begin{abstract*}
Este es el resumen del artículo. Escribimos en español. Se describen el problema, los métodos y los resultados. También se incluyen referencias.
\end{abstract*}
\end{selectlanguage}% edited by Aarón Montoya-Moraga

%% If you have trouble with the following passages, you may deleting this stuff and remove the associated language option from \documentclass[..].
\typeout{If you have trouble with the following passages, your file may not be saved in utf-8 format, or your LaTeX format may be old, or you may not have the assumed font installed. You can delete those lines to resolve the issue.}

%% Examples of abstracts in other languages. The first three are intended for pdflatex, not lualatex.
\ifpdftex
    \begin{selectlanguage}{greek}%
    \begin{abstract*}
    Αυτή είναι η περίληψη του άρθρου. Χρησιμοποιούμε την ελληνική γλώσσα. Περιγράφεται το πρόβλημα, οι μέθοδοι και τα αποτελέσματα. Περιλαμβάνονται επίσης αναφορές.
    \end{abstract*}
    \end{selectlanguage}% Edited by George Barbastathis

    \begin{selectlanguage}{russian}
    \begin{abstract*}
    Это резюме статьи. Пишем по русски. Описаны проблема, методы и результаты. Библиография также включена.%
    \end{abstract*}
    \end{selectlanguage}% edited by Steven Gerasimoff

    \begin{selectlanguage}{vietnamese}
    \begin{abstract*}
    Đây là phần tóm tắt của bài báo khoa học. Chúng tôi viết bằng tiếng Việt. Vấn đề, các phương pháp và các kết quả được mô tả trong phần này. Tài liệu tham khảo cũng được bao gồm.
    \end{abstract*}
    \end{selectlanguage}% Checked and edited by Nguyen Le and Thao Nguyen
\fi

\iffontspecloaded %%% requires lualatex
%
    \begin{selectlanguage}{greek}%
    \begin{abstract*}
    Αυτή είναι η περίληψη του άρθρου. Χρησιμοποιούμε την ελληνική γλώσσα. Περιγράφεται το πρόβλημα, οι μέθοδοι και τα αποτελέσματα. Περιλαμβάνονται επίσης αναφορές.
    \end{abstract*}
    \end{selectlanguage}% Edited by George Barbastathis   
%    
    \begin{selectlanguage}{russian}
    \begin{abstract*}
    Это резюме статьи. Пишем по русски. Описаны проблема, методы и результаты. Библиография также включена.%
    \end{abstract*}
    \end{selectlanguage}% edited by Steven Gerasimoff
%
    \begin{selectlanguage}{vietnamese}
    \begin{abstract*}
    Đây là phần tóm tắt của bài báo khoa học. Chúng tôi viết bằng tiếng Việt. Vấn đề, các phương pháp và các kết quả được mô tả trong phần này. Tài liệu tham khảo cũng được bao gồm.
    \end{abstract*}
    \end{selectlanguage}% Checked and edited by Nguyen Le and Thao Nguyen
%
%    \begin{selectlanguage}{japanese}% Use class option [japanese] if you uncomment this passage.
%    \begin{abstract*}
%    % 論文の要約です。日本語で記述します。問題、方法、および結果について説明します。また、参考文献も含めます。           
%    この論文の日本語での要約は以下のとおりです。問題、方法、および結果が説明されています。参考資料も添付してあります。
%    \end{abstract*}
%    \end{selectlanguage}% Edited by Keiji Yano and Yoshiki Okamoto
%
    \begin{selectlanguage}{korean}
    \begin{abstract*}
    이것은 한국어로 쓰인 논문의 초록입니다. 문제, 방법 및 결과가 설명되어 있습니다. 참조도 포함됩니다.
    \end{abstract*}
    \end{selectlanguage}% Edited by Hyung Won Chung.
%
    \begin{selectlanguage}{chinese-simplified}
    \begin{abstract*}
    这是文章的摘要。我们用中文书写，描述了问题，方法和结果，还包括了参考文献。
    \end{abstract*}
    \end{selectlanguage}% edited by Zi Hao Foo
%
\fi

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2023-05-03 Huston_Paper_asme.tex
what is bayesian

@@
 Three arcs from three channels were considered for classification - the centers of the arcs were directly in front of the vehicle, and 45 degrees left and right of center, creating nine total areas in a 3x3 grid in front of the van.
-Following the same process as aggregating point clouds, GNSS and IMU data informed the location and orientation of the arcs in each 360 degree scan.}
+Following the same process as aggregating point clouds, GNSS and IMU data informed the location and orientation of the arcs in each 360 degree scan.
+All arcs were then classified as "asphalt", "gravel", or "unknown", and the averaged coordinates of each arc was aggregated into a single, classified point cloud.}
 
-	{Manually defined gravel, asphalt, and side-of-road areas were projected onto the averaged coordinates of all the classified arcs.
-Manually defined gravel, asphalt, and side-of-road areas were then examined for the number of type of classified arcs, yielding the accuracy scores for each area.
+	{Manually defined gravel, asphalt, and side-of-road areas were projected onto the classified point cloud.
+Manually defined gravel, asphalt, and side-of-road areas were then used to determine the accuracy of the terrain classification algorithm.
 Classified point clouds were then manually examined to determine if the location of gravel and asphalt surfaces may be identified.

### 2023-05-03 Huston_Paper_asme.tex
hey

@@
 
-	%When the app performs hyperparameter tuning by using Bayesian optimization (see Optimization Options for a brief introduction), it chooses the set ofhyperparameter values that minimizes an upper confidence interval of the classification error objective model, rather than the set that minimizes the classification error.
-
 	\begin{figure}
@@
 		\centering
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/vlcsnap-2023-04-20-08h48m17s447}
-			\caption[Blackburn Road Camera View]{}
+		\includegraphics[width=0.9\linewidth]{figures/vlcsnap-2023-04-20-08h48m17s447}
+		\caption[Blackburn Road Camera View]{Road view of Blackburn Road - an unmarked asphalt road with three intercepting gravel driveways}
 		\label{fig:Blackburn_Road_View}
-		\end{subfigure}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/gravel_lot_pic.png}
-			\caption[Armig Road Camera View]{}
-			\label{fig:Gravel_Lot_View}
-		\end{subfigure}
-		\caption[Blackburn Road \& Gravel Parking Lot]{Road view of Blackburn Road - an unmarked asphalt road (a). Road view of the gravel driveway (b).}
-		\label{fig:Combined_Roads}
 	\end{figure}
@@
 
+	% When the app performs hyperparameter tuning by using Bayesian optimization (see Optimization Options for a brief introduction), it chooses the set ofhyperparameter values that minimizes an upper confidence interval of the classification error objective model, rather than the set that minimizes the classification error.
+
 	\subsection{RDF Training and Testing}
@@
 Decision tree depth, number of learners, and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.
+Bayesian Optimization automates the manual hyper-parameter tuning process creating a surrogate model based on .
 All arcs in a channel used the same RDF.

### 2023-05-10 Huston_Paper.tex
05102023

@@
 Random Decision Forests were trained using scanning LiDAR data terrain classification to detect unmarked gravel and asphalt surfaces.
-It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively.
+It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively at an estimated rate of 13.19 ms per 360 degree scan.
 Overlapping results between manually projected and actual road surface areas resulted in 91.67\% intercepting gravel road detection accuracy.
 Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.}
+
 	}	
@@
 Supervised Classification Applied to Road Following (SCARF) is an algorithm that detects road surfaces based on color differences between the road and the surrounding terrain \cite{crisman_scarf_1993}.
-It was found that difficulties arise when detecting road surfaces in less colored environments \cite{crisman_scarf_1993,manz_detection_2011}.}
+It was found that difficulties arise when detecting road surfaces in less colored environments \cite{crisman_scarf_1993,manz_detection_2011}.
+Edmond et al.
+propose a method to detect terrain type through implementation of vehicle vibration, however this method is limited to what the vehicle is currently driving over and thus cannot detect intercepting gravel roads \cite{dupont_online_2008}.}
 
@@
 Computation efficiency of road surface detection is mentioned only by a few studies, such as the one completed by Yadav et al \cite{yadav_road_2018}.
-Principal component analysis on the height of each LiDAR data point was proposed to detect a straight, curb-less rural road with low computation time \cite{yadav_road_2018}.
+Principal component analysis on the height of each LiDAR data point was proposed to detect a straight, curb-less gravel road with low computation time \cite{yadav_road_2018}.
 However a computational time of 24 minutes was required to process a 156 meter stretch of road, rendering real time detection impossible.}
-
-	% Current literature proposes multiple methods of terrain classification using LiDAR and RGB cameras, and may be roughly broken into two areas of application.
-Traversability or trafficability is the analysis of generally unstructured environments that includes obstacles or rough terrain types for autonomous vehicle path finding solutions  \cite{schilling_geometric_2017,ojeda_terrain_2006,coombs_driving_2000,stavens_self-supervised_nodate,belter_rough_2010,bartoszyk_terrain-aware_2017,noauthor_fusion_nodate,li_rugged_2019,wilson_terrain_2014,siva_robot_2019}.
-While this research may be useful for detection of obstacles on the road for the proposed work, it is assumed that the roads that will be studied will be free of obstacles.
 
@@
 
-	{Neural networks and similar machine learning tools may be used in terrain and road identification.
+	{Neural Networks and similar machine learning tools may be used in terrain and road identification.
 Road classification with LiDAR spacial and remission features was completed with Neural Network classifier, Naive Bayes classifier, and Support Vector Machines (SVM) \cite{wang_road_nodate,wang_two-stage_2018}.
@@
 
-%	{MLS and RANSAC were compared as methods to project a two dimensional plane unto a point cloud, facilitating examination of points by providing a frame of reference \cite{miller_method_nodate, gojcic_perfect_2019}.
-Method of least-squares analyses uses regression to approximate the best fit to a data set (\ref{eqn:MLS}).
-RANSAC is an iterative algorithm that is designed to handle large numbers of outliers in the sample data set \cite{derpanis_overview_nodate,yaniv_random_2010,fischler_random_1987,cantzler_random_nodate}.}
-
 	{Random Decision Forests are a supervised machine learning method that operates on a set of decision trees \cite{ho_random_1995}.
@@
 Classification is determined by the majority results as opposed to the mean \cite{breiman_random_2001,ho_random_1995}.
-Bayesian Optimization may be used to tune the hyper parameters of a machine algorithm \cite{noauthor_bayesian_nodate, snoek_practical_2012}.}
+Bayesian Optimization may be used to tune the hyper parameters of a machine algorithm \cite{noauthor_bayesian_nodate, snoek_practical_2012}.
+Bayesian Optimization initializes a surrogate function to represent the objective function (in this case minimum total classification error) with randomly selected hyper-parameters.
+Additional points on the objective function are chosen using an acquisition function to determine the utility of hyper-parameters.}
 
-	% Summary of lit review in 1 sentance:
 	{Current methods for road detection may be insufficient for real-time detection of gravel roads due to ambient conditions, lack of standardized features, or rely on post-processing large data sets, while terrain classification of scanning LiDAR data may overcome these difficulties.}
+
+	% Current literature proposes multiple methods of terrain classification using LiDAR and RGB cameras, and may be roughly broken into two areas of application.
+Traversability or trafficability is the analysis of generally unstructured environments that includes obstacles or rough terrain types for autonomous vehicle path finding solutions  \cite{schilling_geometric_2017,ojeda_terrain_2006,coombs_driving_2000,stavens_self-supervised_nodate,belter_rough_2010,bartoszyk_terrain-aware_2017,noauthor_fusion_nodate,li_rugged_2019,wilson_terrain_2014,siva_robot_2019}.
+While this research may be useful for detection of obstacles on the road for the proposed work, it is assumed that the roads that will be studied will be free of obstacles.
 
@@
 
-	{Required physical data was gathered by the Ohio University Autonomous Van van (Figure \ref{fig:van}), built by AutonomouStuff.
-LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C (Figure \ref{fig:vlp32mount}).
+		\subsection{Overview}
+
+			{Followed methodology to accomplish this work is outlined in Fig.
+\ref{fig:flowz_2}.
+Training data was gathered by the Ohio University Autonomous Van van (Fig.
+\ref{fig:Experimental_Apperatus}).
+MATLAB was used to extract and parse LiDAR and GPS data.
+Scanning LiDAR data was aggregated into a large point cloud by using GNSS and IMU data to determine each LiDAR scan's position and orientation in a dead-reckoning process.
+Aggregated point cloud data was manually examined and compared to satellite and camera images to determine and project road boundaries.
+Training data was extracted from a single arc per 360 degree scanning LiDAR channel if the arc was coincident to a manually projected 2D area.
+Random Decision Forests were trained and verified using MATLAB's Classifier App.
+Scanning LiDAR data was consecutively classified and compared to manually defined truth areas.
+True-positive accuracy for gravel, asphalt, and non-road surfaces was found.
+Classified point cloud data was examined without reference to truth areas in order to manually project possible road surfaces.
+Manually projected road surface areas were compared to truth areas to determine the accuracy of asphalt road and intercepting gravel driveway detection.}
+
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.9\linewidth]{figures/flowz_2}
+				\caption[Project Flow]{High level view of project work}
+				\label{fig:flowz_2}
+			\end{figure}
+
+	%	\begin{figure}
+		%		\centering
+		%		\includegraphics[width=\linewidth]{figures/test_conf_mat}
+		%		\caption[Validation Confusion Matrix]{Confusion matrix from validation process.}
+		%		\label{fig:vali_conf_mat}
+		%	\end{figure}
+
+		\subsection{Data Collection}
+
+%	{The experimental apparatus for this work was the Ohio University Autonomous Van van (Fig.
+\ref{fig:Experimental_Apperatus}), which has a Velodyne VLP-32C scanning LiDAR sensor, Novatel PwrPak 7D-E2 GNSS and INS system, and Mako cameras.
+Raw data was gathered using the Robotic Operating System (ROS) to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags.
+Rosbags were unpacked using MATLAB tools to separate data streams.
+Prior to creating a training database, manual dictation of gravel and asphalt areas of the gathered point cloud data is necessary.
+In order to do so, all scanning LiDAR data from a rosbag was aggregated in order to create a single point cloud map.
+Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data.
+Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan (Fig.
+\ref{fig:road_areas_annotated_22}).
+Compiled point cloud maps was examined and compared to camera and satellite data to determine gravel and asphalt surface location.
+Manually defined 2D areas representing gravel and asphalt were projected unto the point cloud (Fig.
+\ref{fig:road_areas_annotated_22}).}
+
+			{Required physical data was gathered by the Ohio University Autonomous Van (Fig.
+\ref{fig:Experimental_Apperatus}).
+LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C.
 Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a high output data stream \cite{vlp_32c}.
 GPS data was recorded with a Novatel PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS), and two GNSS-502 antennas by NavtechGPS.
-Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using Robotic Operating System (ROS).
+Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using ROS.
 Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot that intersects Blackburn Road near Athens Ohio.
@@
 Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot.
-Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection.
-Only three classes are considered in this work: "gravel", "asphalt", and "unknown".
-}
+Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection, therefore grass is labeled as "unknown" in this work.
+Raw data was gathered using the Robotic Operating System (ROS) to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags.
+Rosbags were unpacked using MATLAB tools to separate data streams.
+Point cloud, GPS, and IMU topics were defined, and Cartesian coordinates, longitude, latitude, altitude, roll, pitch, and yaw messages were read from the appropriate topics and imported}
+
+	%	\begin{figure} 
+		%		\centering
+		%		\begin{subfigure}{0.5\textwidth}
+			%			\centering
+			%			\includegraphics[width=1\linewidth,height=4.60 cm,keepaspectratio]{figures/van_on_van}
+			%			\caption[Sensor Van]{}
+			%			\label{fig:van}
+			%		\end{subfigure}
+		%		\begin{subfigure}{0.5\textwidth}
+			%			\centering
+			%			\includegraphics[width=1\linewidth,height=4.60 cm,keepaspectratio]{figures/LiDAR_on_van}
+			%			\caption[VLP 32 on Van]{}
+			%			\label{fig:vlp32mount}
+			%		\end{subfigure}
+		%		\caption[Experimental Apparatus]{Experimental Apparatus (a). Velodyne VLP-32 Scanning LiDAR is mounted on top of the vehicle (b).}
+		%		\label{fig:Experimental_Apperatus}
+		%	\end{figure}
 
@@
 				\centering
-		\begin{subfigure}{0.45\textwidth}
+				\includegraphics[width=0.75\linewidth]{figures/van_on_van}
+				\caption[Experimental Apparatus]{Experimental Apparatus. Velodyne VLP-32 Scanning LiDAR is mounted on top of the vehicle.}
+				\label{fig:Experimental_Apperatus}
+			\end{figure}
+
+		\subsection{Training Data Extraction}
+
+			{Scanning LiDAR training data was extracted from manually defined gravel, asphalt, and unknown terrain areas on compiled point clouds.
+Compiling LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
+Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data.
+Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan.
+Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data.
+Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU.
+LiDAR origin was found using GPS longitude, latitude, and altitude data.
+Physical distances between the GPS, IMU, and LiDAR sensors was rectified by obtaining reference frames provided by AutonomouStuff.
+GPS coordinates were offset by the current orientation and converted the ground truth to the LiDAR frame.
+GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors.
+Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Fig.
+\ref{fig:road_areas_annotated_22}).
+While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for this work.}
+
+			{Manually defined areas may then be overlayed unto the point cloud (Fig.
+\ref{fig:road_areas_annotated_22}).
+Six degree arcs from the first three channels of scanning LiDAR data that lay directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.
+Training data was extracted from a single arc from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area.
+Raw spatial and intensity values from the coincident arc were saved to a raw training database.
+Separate training databases were maintained for each individual channel.
+Features that describe the raw spatial and intensity data were extracted.
+Thirty percent of the training data base was randomly selected and set aside as a validation data base.
+Training data was visually examined for clustering to verify that a Random Decision Forest would be able to be trained (Fig.
+\ref{fig:training_data_cluster}).
+It was found that each class generally clustered to a certain range with some overlap, which indicates that a Random Decision Forest may be trained.}
+
+			\begin{figure}[H]
 				\centering
-			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/van_on_van}
-			\caption[Sensor Van]{}
-			\label{fig:van}
-		\end{subfigure}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/LiDAR_on_van}
-			\caption[VLP 32 on Van]{}
-			\label{fig:vlp32mount}
-		\end{subfigure}
-		\caption[Experimental Apparatus]{Experimental Apparatus (a). Velodyne VLP-32 Scanning LiDAR is mounted on top of the vehicle (b).}
-		\label{fig:Experimental_Apperatus}
+				\includegraphics[width=0.75\linewidth]{figures/vlcsnap-2023-04-20-08h48m17s447}
+				\caption[Blackburn Road Camera View]{Camera view of Blackburn Road - an unmarked asphalt road with three intercepting gravel driveways from which asphalt data was gathered.}
+				\label{fig:Blackburn_Road_View}
 			\end{figure}
@@
 				\centering
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/vlcsnap-2023-04-20-08h48m17s447}
-			\caption[Blackburn Road Camera View]{}
-			\label{fig:Blackburn_Road_View}
-		\end{subfigure}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/gravel_lot_pic.png}
-			\caption[Armig Road Camera View]{}
-			\label{fig:Gravel_Lot_View}
-		\end{subfigure}
-		\caption[Blackburn Road \& Gravel Parking Lot]{Road view of Blackburn Road - an unmarked asphalt road (a). Road view of the gravel driveway (b).}
-		\label{fig:Combined_Roads}
+				\includegraphics[width=0.75\linewidth]{figures/gravel_training_lot}
+				\caption[Gravel Training Lot]{Camera view of the gravel lot from which gravel training data was gathered.}
+				\label{fig:gravel_training_lot}
 			\end{figure}
-
-	{MATLAB was used to process the gathered rosbag files using the their ROSBAG library.
-Rosbags were directly loaded into the MATLAB environment.
-Point cloud, GPS, and IMU topics were defined, and the messages were read from the topic and imported.
-Cartesian coordinates, longitude, latitude, altitude, roll, pitch, and yaw were extracted from the messages.}
-
-	{Scanning LiDAR training data was extracted from manually defined gravel, asphalt, and unknown terrain areas on compiled point clouds.
-Compiling LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
-Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data.
-Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU.
-LiDAR origin was found using GPS longitude, latitude, and altitude data.
-Physical distances between the GPS, IMU, and LiDAR was rectitude by obtaining reference frames provided by AutonomouStuff.
-GPS coordinates were offset by the current orientation and converts the ground truth to the LiDAR frame.
-GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors.
-Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}).
-While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for scoring purposes.
-Manually defined areas may then be overlayed unto the point cloud (Figure \ref{fig:Combined_Classified_PCD}).
-Six degree arcs from the first three channels of scanning LiDAR data that lay directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.}
 
@@
 				\centering
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=0.95\linewidth]{figures/Compiled_PCD}
-			\caption[C]{}
-			\label{fig:Compiled_PCD}
-		\end{subfigure}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=0.95\linewidth]{figures/Manually_Defined_Area_Train_Dat_Grab_2}
-			\caption[]{}
-			\label{fig:Classified_Areas_PCD}
-		\end{subfigure}
-		\caption[Compiled and Classified Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map (a).
-Manually defined areas inform where training data lies (b).}
-		\label{fig:Combined_Classified_PCD}
+				\includegraphics[width=0.75\linewidth]{figures/road_areas_annotated_22}
+				\caption[Compiled and Manually Classified Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.
+Projected manually defined areas define terrain surface type.
+Area $3$ is the gravel parking lot entrance from which training data was extracted.}
+				\label{fig:road_areas_annotated_22}
 			\end{figure}
 
-%	{RANSAC was implementing by a built-in MATLAB function, $pcfitplane$.
+			\begin{figure}[H]
+			\centering
+			\includegraphics[width=0.75\linewidth]{figures/training_data_cluster}
+			\caption[Example Clustering]{Example of training data comparison between two features.}
+			\label{fig:training_data_cluster}
+			\end{figure}
+
+			%{RANSAC was implementing by a built-in MATLAB function, $pcfitplane$.
 Method of Least Squares was implemented by importing a MATLAB library \cite{noauthor_object-oriented_nodate}.
@@
 
+	% When the app performs hyperparameter tuning by using Bayesian optimization (see Optimization Options for a brief introduction), it chooses the set ofhyperparameter values that minimizes an upper confidence interval of the classification error objective model, rather than the set that minimizes the classification error.
+
+		\subsection{RDF Training and Testing}
+
+			%MATLAB was used to train a Random Decision Forests for each of the three considered channels using Bayesian Optimization to tune tree depth and number of splits.
+Training continued until a minimum error was reached (Fig.
+\ref{fig:c2_min_class_error}).
+Validation of the RDF was then completed using the validation database.
+
 			{MATLAB's Classification Learner Application was used to generate a Random Decision Forests for each considered scanning LiDAR channel.
-Decision tree depth, number of learners, and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.}
+Decision tree depth, number of learners, and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.
+Bayesian Optimization automates the manual hyper-parameter tuning process by creating a surrogate model representing the objective function, in this case out-of-bag error.
+Initialization of the objective function is completed by training a number of RDFs using randomly selected hyper-parameters.
+Out-of-bag error is then found and used to create a series of Gaussian curves.
+Additional hyper-parameters are chosen by an acquisition function to determine their utility in minimizing out-of-bag error.
+Best hyper-parameters are chosen after a set number of iterations and used to create a final RDF.
+Confusion matrixes were generating showing the out-of-bag error to determine the training error (Fig.
+\ref{fig:out_of_bag_err_conf_mat}).
+Validation of the RDF was completed using the reserved thirty percent of the training data (Fig.
+\ref{fig:vali_err_conf_mat}).
+This process was completed for each of the three considered Scanning LiDAR channels used for classification.}
 
-	{Classifying consecutive LiDAR scans was accomplished by examination of nine arcs of interest in front and forty-five degree angles left and right from the front of the experimental apparatus.
-Features from each arc were extracted and the arc was classified.
-Averaged x, y, and z coordinates were taken of each arc and projected unto a manually defined truth area map (Figure \ref{fig:Classified_Areas_PCD}).
-Accuracy scores of each area was calculated, allowing evaluation of road detection performance.}
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.75\linewidth]{figures/c2_min_class_error}
+				\caption[RDF Training Classification Error]{Classification error during Bayesian Optimization tuning.}
+				\label{fig:c2_min_class_error}
+			\end{figure}
+
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.75\linewidth]{figures/out_of_bag_err_conf_mat}
+				\caption[Out-of-Bag Error]{During training the RDF may be tested using the out-of-bag error. Confusion matrices are used to indicate the final RDF iteration training error.}
+				\label{fig:out_of_bag_err_conf_mat}
+			\end{figure}
+
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.75\linewidth]{figures/vali_err_conf_mat}
+				\caption[Validation Error]{Confusion matrix with the validation training data set.}
+				\label{fig:vali_err_conf_mat}
+			\end{figure}
+
+			{Classifying consecutive LiDAR scans was accomplished by examination of the nine arcs of interest in front and forty-five degree angles left and right from the front of the experimental apparatus (Figure \ref{fig:area_example}).
+Features from each arc were extracted and the arc was classified (Fig.
+\ref{fig:raw_classification_results}).
+Manually defined gravel, asphalt, and side-of-road areas were projected onto the classified point cloud then used to determine the accuracy of the terrain classification algorithm.
+Averaged x, y, and z coordinates were taken of each arc and projected unto a manually defined truth area map (Fig.
+\ref{fig:rm_db_4_toc}).
+Accuracy scores of each area was calculated, allowing evaluation of road detection performance.
+Classified point clouds were visually inspected to determine if the manual detection of a road surface may be accomplished.
+If consecutive points from each channel in any direction had two or more matching gravel or asphalt classification and was evenly spaced, a gravel or road surface area was projected.
+Guessed road surface areas were then compared to actual areas to determine the method's ability at unmarked gravel road discovery, indicating the method's efficiency at detecting intercepting gravel roads.
+}
+
+				\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.25\linewidth]{figures/area_example}
+				\caption[Areas to Classify]{Three arcs from three channels were classified per 360 degree scan. }
+				\label{fig:area_example}
+			\end{figure}
+
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.75\linewidth]{figures/raw_classification_results}
+				\caption[Classified Point Cloud]{Classified Point Cloud.}
+				\label{fig:raw_classification_results}
+			\end{figure}
+
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.9\linewidth]{figures/rm_db_4_ToC_2}
+				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed and compared to actual road-surface areas. }
+				\label{fig:rm_db_4_toc}
+			\end{figure}
 
@@
 Translation and rotation was applied to the averaged x, y, and z coordinates of each arc's terrain classification results.
-Consecutive scanning LiDAR data was projected on manually classified truth areas representing road road surfaces (Figure \ref{fig:area_percentages}) and percentages of terrain types per area were calculated.
+Consecutive scanning LiDAR data was projected on manually classified truth areas representing road road surfaces (Figure \ref{fig:prepostadjust}) and percentages of terrain types per area were calculated.}
+
+		\begin{table}[t]
+			\caption[Table]{Average Results\label{tab:1}}
+			\centering{%
+				\begin{tabular}{lrl}
+					\toprule
+					Gravel Areas \% 						& 74.97 	& \% Gravel \\
+					Asphalt Area \%							& 87.05   	& \% Asphalt \\
+					Side-of-Road \% 						& 77.47 	& \% Unknown\\
+					Asphalt Road Detection 					& 100.00 	& \% True Pos. Detection Rate \\
+					Gravel Driveway Detection 				& 93.33 	& \% True Pos. Detection Rate \\
+					Misidentified Gravel Road Detection 	& 0.60 		&  Avg. False Pos. Detection Rate per Run\\
+					\bottomrule
+				\end{tabular}
 			}
+		\end{table}
 
-	{Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 67.54\%.
+		{Classification of real-world data outside of the training data set presents difficulties to the generated Random Decision Forest Classifiers generated in this work.
+Due to the side of the road was being re-grassed near area $3$ (Fig.
+\ref{fig:Blackburn_Road_View}, \ref{fig:road_areas_annotated_22}), the algorithm struggled to differentiate the dirt from gravel, leading to some miss-classification of the side-of-road areas.
+Examination of the feature distribution shows a high amount of overlap (Fig.
+\ref{fig:dirt_v_gravel}).
+Exploitation of confidence scores to adjust results by requiring a very high confidence for positive gravel surface classification was explored.
+All gravel surfaces that did not meet a 80\% confidence score was re-labeled as "unknown", however these adjustment had either a negligible or detrimental impact on the final results.
+Higher confidence score requirements resulted in "unknown" over-classification, while too low resulted in "gravel" over-classification.
+Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 67.54\%.
 If any asphalt miss-classification on the gravel areas is included, the average true-positive road detection accuracy increases to 76.97\%.
@@
 Differentiation between gravel road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation.
-Side-of-road areas were established (Figure \ref{fig:prepostadjust}), and the overall average for "unknown" terrain type was found to be 77.49\%.}
+Side-of-road areas were established (Fig.
+\ref{fig:prepostadjust}), and the overall average for "unknown" terrain type was found to be 77.49\%.}
 
@@
 			\centering
-		\includegraphics[width=0.9\linewidth]{figures/rm_db_4_percentages_big_anotated}
-		\caption[Area Percentage Scores]{Five runs of Blackburn Road with the intercepting gravel parking lot were classified.
-Shown is the results of one run.
-Percentages of each terrain type per area were calculated.
-Area $3$ is the gravel parking lot entrance from which training data was extracted.}
-		\label{fig:area_percentages}
+			\includegraphics[width=0.75\linewidth]{figures/dirt_v_gravel}
+			\caption[Re-grassed Dirt vs Gravel]{Side-of-road detection accuracy was diminished due to overlapping features between dirt and gravel. }
+			\label{fig:dirt_v_gravel}
 		\end{figure}
-
-%	\begin{figure}[H]
-%		\centering
-%		\begin{subfigure}{0.495\textwidth}
-%			\centering
-%			\includegraphics[width=0.975\linewidth]{figures/side_of_road_nums_4_with_no_filts_2}
-%			\caption[Pre-Adjusted]{}
-%			\label{fig:pre-adjust}	
-%		\end{subfigure}
-%		\begin{subfigure}{0.495\textwidth}
-%			\centering
-%			\includegraphics[width=0.975\linewidth]{figures/side_of_road_nums_4_with_filts_2}
-%			\caption[Post-Adjusted]{}
-%			\label{fig:post-adjust}
-%		\end{subfigure}
-%		\label{fig:prepostadjust}
-%		\caption[Post-Adjusted]{Raw classification result example (a).
-Results were slightly altered by considering confidence scores and inter-arc distances to decrease miss-classification of gravel surfaces, however the effect was negligible (b).}
-%	\end{figure}
 
@@
 			\centering
-			\includegraphics[width=0.975\linewidth]{figures/side_of_road_nums_4_with_filts_2}
-			\caption[Post-Adjusted]{}
+			\includegraphics[width=0.75\linewidth]{figures/side_of_road_nums_4_with_filts_2_2}
+			\caption[Area Scores]{Per-area scores were calculated, indicating algorithm unmarked road detection accuracy. }
 			\label{fig:prepostadjust}
-		\caption[Area Scores]{Per-area scores were calculated, indicating algorithm efficiency of unmarked road detection. }
 		\end{figure}
@@
 		{Classified point clouds were manually examined for 'best guess' locations for gravel and asphalt surfaces.
-Two out of three positive gravel identifications must be made on each channel, be evenly spaced, and be relatively co-planar to other road surface guesses in order for a positive guess for an asphalt or gravel surface to be made (Figure \ref{fig:rm_db_4_toc}).
+Two out of three positive gravel identifications must be made on each channel, be evenly spaced, and be relatively co-planar to other road surface guesses in order for a positive guess for an asphalt or gravel surface to be made (Fig.
+\ref{fig:rm_db_4_toc}).
 Driveways $1$ and $2$ were detected in all five drive-bys, driveway $3$ was detected in four leading to an 91.67\% accuracy in intercepting gravel road detection.
@@
 		{Classification was completed using Ubuntu 18.04 and an AMD Ryzen 9 5900X 12 core 24 thread.
-Rate of classification for all nine areas of interest in a single scan was found to be an average of 1.226 seconds using a single core.
+Rate of classification for all nine areas of interest in a single scan was found to be an average of 35.19 ms using a single core (Fig.
+\ref{fig:per_arc_classification_time}) or 316.68 ms per 360 degree scan (Fig.
+\ref{fig:per_scan_classification_rate}).
 Further optimization may increase this speed, however real-time classification is not a focus of this work.
-Using all cores when processing a ROSBAG file indicated that a rate of 55.81 feet per second or 38.05 miles per hour was accomplished.}
+When utilizing all 24 cores during this indicates that a rate of 13.19 ms per 360 scan.
+Extrapolating from the distance traveled during a drive-by, a hypothetical 71.39 feet per second or 48.68 miles per hour may be possible.}
 
@@
 			\centering
-		\includegraphics[width=0.9\linewidth]{figures/rm_db_4_toc}
-		\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed using a rule set and compared to actual road-surface areas. }
-		\label{fig:rm_db_4_toc}
+			\includegraphics[width=0.75\linewidth]{figures/per_arc_classification_time}
+			\caption[Per-Arc Time]{Per-arc classification time}
+			\label{fig:per_arc_classification_time}
 		\end{figure}
 
-\section{Discussion}
+		\begin{figure}[H]
+			\centering
+			\includegraphics[width=0.75\linewidth]{figures/per_scan_classification_rate}
+			\caption[Per-Scan Time]{Per-scan classification time}
+			\label{fig:per_scan_classification_rate}
+		\end{figure}
 
-	{Classification of real-world data outside of the training data set presents difficulties to the generated Random Decision Forest Classifiers generated in this work.
-Due to the side of the road was being re-grassed near area $3$ (Figure \ref{fig:area_percentages}, Figure \ref{fig:Blackburn_Road_View}), the algorithm struggled to differentiate the dirt from gravel, leading to some miss-classification of the side-of-road areas.
-Examination of the feature distribution shows a high amount of overlap (Figure \ref{}).
-Exploitation of confidence scores to adjust results by requiring a very high confidence for positive gravel surface classification was explored.
-All gravel surfaces that did not meet a 80\% confidence score was re-labeled as "unknown", however these adjustment had either a negligible or detrimental impact on the final results.
-Higher confidence score requirements resulted in "unknown" over-classification, while too low resulted in "gravel" over-classification.
-}
+	%\section{Discussion}
+
+	%%%%% Conclusions %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Terrain classification of consecutive scans compared to manually defined road surface areas yielded accuracy scores for gauging algorithm capability for gravel surface detection.
-It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively.
+It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively at an estimated rate of 13.19 ms per 360 degree scan.
 Classified point clouds were manually examined and gravel areas were projected and compared with actual gravel surface areas.

### 2023-05-10 Huston_Paper.tex
0510202023_2

@@
 Aggregated point cloud data was manually examined and compared to satellite and camera images to determine and project road boundaries.
-Training data was extracted from a single arc per 360 degree scanning LiDAR channel if the arc was coincident to a manually projected 2D area.
+Training data was extracted from a single arc per 360 degree scanning LiDAR channel if the arc was coincident to a manually projected 2D training data area.
 Random Decision Forests were trained and verified using MATLAB's Classifier App.
-Scanning LiDAR data was consecutively classified and compared to manually defined truth areas.
-True-positive accuracy for gravel, asphalt, and non-road surfaces was found.
+Scanning LiDAR data was consecutively classified and compared to manually defined truth areas yielded the true-positive accuracy for gravel, asphalt, and non-road surfaces.
 Classified point cloud data was examined without reference to truth areas in order to manually project possible road surfaces.
@@
 Rosbags were unpacked using MATLAB tools to separate data streams.
-Point cloud, GPS, and IMU topics were defined, and Cartesian coordinates, longitude, latitude, altitude, roll, pitch, and yaw messages were read from the appropriate topics and imported}
+Point cloud, GPS, and IMU topics were defined, and point cloud Cartesian coordinates with corresponding intensities, longitude, latitude, altitude, roll, pitch, and yaw messages were read from the appropriate topics and imported}
 
@@
 
-			{Scanning LiDAR training data was extracted from manually defined gravel, asphalt, and unknown terrain areas on compiled point clouds.
-Compiling LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
+			{Scanning LiDAR training data requires extraction from manually defined gravel, asphalt, and "unknown" terrain areas on aggregated point clouds as manual extraction from consecutive scans is extremely time consuming.
+Aggregating LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
 Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data.
@@
 
-			{Feature extraction was completed by performing mathematical functions on the gathered spatial and remission training data.
-Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial or remission data array.
-Features were non-dimensionalized as necessary by dividing by the average range of the arc to LiDAR point of origin to the appropriate power.
+			{Feature extraction was completed by performing mathematical functions on the gathered spatial training data.
+Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial data array.
+Spatial features were non-dimensionalized as necessary by dividing by the average range of the arc to LiDAR point of origin to the appropriate power.
+Remission features include Standard Deviation ($\sigma$), Mean, Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), however as remission is a ratio of minimum to maximum detectable reflected intensity there was no need to render the features non-dimensionalized.
 Thirty percent of the training data was randomly selected and reserved to create a validation data set.}
@@
 Features from each arc were extracted and the arc was classified (Fig.
-\ref{fig:raw_classification_results}).
-Manually defined gravel, asphalt, and side-of-road areas were projected onto the classified point cloud then used to determine the accuracy of the terrain classification algorithm.
+\ref{fig:raw_classification_results} shows the average spatial coordinates of each classified arc).
+Previously-created manually defined gravel, asphalt, and side-of-road truth areas were projected onto the classified point cloud for the determination of classification accuracy.
 Averaged x, y, and z coordinates were taken of each arc and projected unto a manually defined truth area map (Fig.
 \ref{fig:rm_db_4_toc}).
-Accuracy scores of each area was calculated, allowing evaluation of road detection performance.
+Accuracy scores was calculated based on the distribution of classified points with each area, allowing evaluation of unmarked road detection performance.
 Classified point clouds were visually inspected to determine if the manual detection of a road surface may be accomplished.
@@
 Driveways $1$ and $2$ were detected in all five drive-bys, driveway $3$ was detected in four leading to an 91.67\% accuracy in intercepting gravel road detection.
-It was found that in three drive-bys there was a false positive driveway detection, however the location of the false-positive was not repeated.
+It was found that in three drive-bys there was a false positive driveway detection, however the location of the false-positives were not repeated in any of the other drivebys.
 In all drive-bys the asphalt road was accurately detected.

### 2023-05-18 Huston_Paper.tex
05182023

@@
 Detection of gravel roads may allow for expansion of current civilian autonomous vehicle capabilities as well as introduce opportunities for industries that operate in rural areas.
-Global Positioning Service (GPS) and high resolution maps cannot be relied upon for navigation of rural roads alone as GPS is not accurate enough or may not have a reliable signal for precision driving \cite{noauthor_gpsgov_nodate}, and gravel roads are not marked or mapped.
+Global Positioning Service (GPS) and high resolution maps should not be relied upon for navigation of rural roads alone as GPS is not accurate enough or may not have a reliable signal for precision driving \cite{noauthor_gpsgov_nodate}, and gravel roads are not marked or mapped.
 Systems that rely on cameras may fail when analyzing gravel roads, as these lack visual cues such as painted lane markings \cite{crisman_scarf_1993} and may in some circumstances closely resemble surrounding terrain.
@@
 Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}.
-Implementing a terrain classification model using Random Decision Forests (RDF) using scanning LiDAR data to exploit surface roughness was studied for road surface detection in this study.
+Implementing a Random Decision Forests (RDF) terrain classification model using scanning LiDAR data to exploit surface roughness was studied for road surface detection in this study.
 Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.}
@@
 	{Current LiDAR based road surface detection models may be categorized into two methodologies.
-Firstly is the analysis of road surface properties, which considers aspects such as topology and surface roughness.
+First is the analysis of road surface properties, which considers aspects such as topology and surface roughness.
 Detection based on road surface smoothness was studied by segmenting the point cloud into candidate road surface regions and searching for elevation jumps \cite{liu_new_2013}.
@@
 
-			{Followed methodology to accomplish this work is outlined in Fig.
-\ref{fig:flowz_2}.
+			{Detection of unmarked gravel road surfaces using terrain classification requires building a data base of terrain features, training an RDF, and testing the algorithm for accuracy of unmarked road detection (outline of work in Fig.
+\ref{fig:flowz_2}).
 Training data was gathered by the Ohio University Autonomous Van van (Fig.
@@
 LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C.
-Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a high output data stream \cite{vlp_32c}.
+Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a minimum 0.33$^{\circ}$ vertical angular resolution \cite{vlp_32c} and 0.2$^{\circ}$ horizontal angular resolution.
 GPS data was recorded with a Novatel PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS), and two GNSS-502 antennas by NavtechGPS.
-Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using ROS.
+Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using the Robot Operating System (ROS).
 Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot that intersects Blackburn Road near Athens Ohio.
@@
 
-			{Scanning LiDAR training data requires extraction from manually defined gravel, asphalt, and "unknown" terrain areas on aggregated point clouds as manual extraction from consecutive scans is extremely time consuming.
+			{Scanning LiDAR training data requires extraction from manually defined gravel, asphalt, and "unknown" terrain areas on aggregated point clouds.
 Aggregating LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
@@
 
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.75\linewidth]{figures/blackburn_road}
+				\caption[Blackburn Road Camera View]{Camera view of Blackburn Road - an unmarked asphalt road with three intercepting gravel driveways from which asphalt data was gathered.}
+				\label{fig:Blackburn_Road_View}
+			\end{figure}
+
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.75\linewidth]{figures/gravel_lot_pic}
+				\caption[Gravel Training Lot]{Camera view of the gravel lot from which gravel training data was gathered.}
+				\label{fig:gravel_training_lot}
+			\end{figure}
+
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.75\linewidth]{figures/pc_example}
+				\caption[Aggregated Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.}
+			\end{figure}
+
 			{Manually defined areas may then be overlayed unto the point cloud (Fig.
 \ref{fig:road_areas_annotated_22}).
-Six degree arcs from the first three channels of scanning LiDAR data that lay directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.
+Six degree arcs (Fig.
+\cite{fig:area_example}) from the first three channels of scanning LiDAR data that lay directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.
 Training data was extracted from a single arc from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area.
@@
 Training data was visually examined for clustering to verify that a Random Decision Forest would be able to be trained (Fig.
-\ref{fig:training_data_cluster}).
+\ref{fig:training_data_cluster_2}).
 It was found that each class generally clustered to a certain range with some overlap, which indicates that a Random Decision Forest may be trained.}
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/vlcsnap-2023-04-20-08h48m17s447}
-				\caption[Blackburn Road Camera View]{Camera view of Blackburn Road - an unmarked asphalt road with three intercepting gravel driveways from which asphalt data was gathered.}
-				\label{fig:Blackburn_Road_View}
-			\end{figure}
-
-			\begin{figure}[H]
-				\centering
-				\includegraphics[width=0.75\linewidth]{figures/gravel_training_lot}
-				\caption[Gravel Training Lot]{Camera view of the gravel lot from which gravel training data was gathered.}
-				\label{fig:gravel_training_lot}
+				\includegraphics[width=0.25\linewidth]{figures/area_example}
+				\caption[Areas to Classify]{Three arcs from three channels were classified per 360 degree scan. }
+				\label{fig:area_example}
 			\end{figure}
@@
 				\includegraphics[width=0.75\linewidth]{figures/road_areas_annotated_22}
-				\caption[Compiled and Manually Classified Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.
-Projected manually defined areas define terrain surface type.
-Area $3$ is the gravel parking lot entrance from which training data was extracted.}
+				\caption[Manually Classified Point Cloud Data]{ Projected manually defined areas define terrain surface type.
+Three gravel driveways (annotated 1-3) intercept an asphalt road, area $3$ is the gravel parking lot entrance from which training data was extracted.}
 				\label{fig:road_areas_annotated_22}
@@
 			\centering
-			\includegraphics[width=0.75\linewidth]{figures/training_data_cluster}
+			\includegraphics[width=0.9\linewidth]{figures/training_data_cluster_3}
 			\caption[Example Clustering]{Example of training data comparison between two features.}
-			\label{fig:training_data_cluster}
+			\label{fig:training_data_cluster_2}
 			\end{figure}
@@
 
-			{Feature extraction was completed by performing mathematical functions on the gathered spatial training data.
+			{LiDAR returns spatial data represented by Cartesian coordinates with units in meters and remission data represented by a dimensionless ratio of minimum to maximum detected brightness.
+Feature extraction was completed by performing mathematical functions on the gathered spatial training data.
 Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial data array.
@@
 Decision tree depth, number of learners, and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.
-Bayesian Optimization automates the manual hyper-parameter tuning process by creating a surrogate model representing the objective function, in this case out-of-bag error.
+Bayesian Optimization automates the manual hyper-parameter tuning process by creating a surrogate model representing the objective function, in this the upper confidence interval of the classification error objective model (Fig.
+\ref{fig:c2_min_class_error}).
 Initialization of the objective function is completed by training a number of RDFs using randomly selected hyper-parameters.
@@
 Additional hyper-parameters are chosen by an acquisition function to determine their utility in minimizing out-of-bag error.
-Best hyper-parameters are chosen after a set number of iterations and used to create a final RDF.
-Confusion matrixes were generating showing the out-of-bag error to determine the training error (Fig.
-\ref{fig:out_of_bag_err_conf_mat}).
-Validation of the RDF was completed using the reserved thirty percent of the training data (Fig.
-\ref{fig:vali_err_conf_mat}).
-This process was completed for each of the three considered Scanning LiDAR channels used for classification.}
+After a set number of iterations, best hyper-parameters are chosen based minimum upper confidence interval of the classification error objective model.
+Hyperparameter tuning may lead to over-fitting to the training data base.
+Testing for over-fitting was accomplished by comparing the out-of-bag training error to the validation error (Fig.
+\ref{fig:train_vs_valid_overfit_test2}).
+Confusion matrices provide classification accuracy information on training and validation data sets by comparing actual and predicted classes (Fig.
+\ref{fig:out_of_bag_err_conf_mat}, \ref{fig:vali_err_conf_mat}).
+This process was completed for each of the three considered scanning LiDAR channels used for classification.}
 
@@
 				\includegraphics[width=0.75\linewidth]{figures/c2_min_class_error}
-				\caption[RDF Training Classification Error]{Classification error during Bayesian Optimization tuning.}
+				\caption[RDF Training Classification Error]{Classification error during Bayesian Optimization tuning.
+Best hyper-parameters are chosen based minimum upper confidence interval of the classification error objective model, not minimum classification error.}
 				\label{fig:c2_min_class_error}
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/out_of_bag_err_conf_mat}
+				\includegraphics[width=0.75\linewidth]{figures/train_vs_valid_overfit_test2}
+				\caption[Training vs Validation Error]{Hyperparameter tuning may lead to over-fitting.
+Testing out-of-bag versus validation error demonstrates that the model is not over-fitting to the training data base if the validation error does not diverge from the out-of-bag error.}
+				\label{fig:train_vs_valid_overfit_test2}
+			\end{figure}
+
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.75\linewidth]{figures/chan_2c_conf_OOB_mat222}
 				\caption[Out-of-Bag Error]{During training the RDF may be tested using the out-of-bag error. Confusion matrices are used to indicate the final RDF iteration training error.}
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/vali_err_conf_mat}
+				\includegraphics[width=0.75\linewidth]{figures/chan_2c_conf_VALIDATION_mat222}
 				\caption[Validation Error]{Confusion matrix with the validation training data set.}
@@
 				\centering
-				\includegraphics[width=0.25\linewidth]{figures/area_example}
-				\caption[Areas to Classify]{Three arcs from three channels were classified per 360 degree scan. }
-				\label{fig:area_example}
-			\end{figure}
-
-			\begin{figure}[H]
-				\centering
-				\includegraphics[width=0.75\linewidth]{figures/raw_classification_results}
+				\includegraphics[width=0.9\linewidth]{figures/raw_classification_results_v3}
 				\caption[Classified Point Cloud]{Classified Point Cloud.}
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/rm_db_4_ToC_2}
+				\includegraphics[width=0.9\linewidth]{figures/rm_db_4_ToC_22}
 				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed and compared to actual road-surface areas. }
@@
 			\centering
-			\includegraphics[width=0.75\linewidth]{figures/dirt_v_gravel}
+			\includegraphics[width=0.75\linewidth]{figures/dirt_v_gravel2}
 			\caption[Re-grassed Dirt vs Gravel]{Side-of-road detection accuracy was diminished due to overlapping features between dirt and gravel. }
@@
 Terrain classification of consecutive scans compared to manually defined road surface areas yielded accuracy scores for gauging algorithm capability for gravel surface detection.
-It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively at an estimated rate of 13.19 ms per 360 degree scan.
-Classified point clouds were manually examined and gravel areas were projected and compared with actual gravel surface areas.
-Overlapping results between manually projected and actual road surface areas resulted in 91.67\% intercepting gravel road detection accuracy.
-In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is a promising method of gravel road surface detection.
+It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively at an estimated rate of 13.19 milliseconds per 360 degree scan.
+Classified point clouds were examined and gravel areas were projected and compared with actual gravel surface areas.
+Overlapping results between detected and actual road surface areas resulted in 91.67\% intercepting gravel road detection accuracy.
+In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is promising.
 This work addressed the problem of road surface detection on unmarked gravel and chipseal roads.
-Current LiDAR and Camera based detection models were insufficient due to lack of distinct features typical of urban roads, such as painted line markings or curbs, or have excessive computational or storage requirements.
-This work balanced accuracy and efficiency by using less intensive analysis techniques of smaller point cloud data sets.
 The described work was a method of detecting physical unmarked gravel and chipseal roads by using a terrain classification approach to predicting road surface area.

### 2023-05-24 Huston_Paper.tex
Finished Draft

@@
 
+\renewcommand{\footnoterule}{\vspace*{1.5ex}\noindent\rule{\linewidth}{0.4pt}\vspace*{1ex}}
+\setlength{\skip\footins}{1.5ex}
+
 \begin{document}
@@
 
-	\author{Rhett Huston, Dr. Jay Wilhelm}
+	\author{Rhett G.
+Huston\textsuperscript{\textdagger}\thanks{\textsuperscript{\textdagger} Graduate Research Assistant, Mechanical Engineering, Ohio University, Athens, Ohio, 45701, Student Member}, Jay P.
+Wilhelm\textsuperscript{\textdaggerdbl}\thanks{\textsuperscript{\textdaggerdbl} Associate Professor, Mechanical Engineering, Ohio University, Athens, Ohio 45701, Senior Member}}
+
 %	{
@@
 Random Decision Forests were trained using scanning LiDAR data terrain classification to detect unmarked gravel and asphalt surfaces.
-It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively at an estimated rate of 13.19 ms per 360 degree scan.
+It was found that a true-positive accuracy for gravel and asphalt surfaces was 69.15\% and 78.07\% respectively at an estimated rate of 13.19 ms per 360 degree scan.
 Overlapping results between manually projected and actual road surface areas resulted in 91.67\% intercepting gravel road detection accuracy.
+Results indicate that using a Random Decision Forest to classify scanning LiDAR data is a feasible method for unmarked gravel and asphalt road detection.
 Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.}
@@
 Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}.
-Implementing a Random Decision Forests (RDF) terrain classification model using scanning LiDAR data to exploit surface roughness was studied for road surface detection in this study.
+Implementing a Random Decision Forests (RDF) terrain classification model using scanning LiDAR data to exploit surface roughness was studied to determine the method's feasibility in road surface detection.
 Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.}
@@
 Supervised Classification Applied to Road Following (SCARF) is an algorithm that detects road surfaces based on color differences between the road and the surrounding terrain \cite{crisman_scarf_1993}.
-It was found that difficulties arise when detecting road surfaces in less colored environments \cite{crisman_scarf_1993,manz_detection_2011}.
-Edmond et al.
-propose a method to detect terrain type through implementation of vehicle vibration, however this method is limited to what the vehicle is currently driving over and thus cannot detect intercepting gravel roads \cite{dupont_online_2008}.}
+It was found that difficulties arise when detecting road surfaces in less colored environments \cite{crisman_scarf_1993,manz_detection_2011}.}
 
@@
 It was found that characterizing the LiDAR intensity as Gaussian distributed noise yielded poor results with only a 49.5\% true positive rate, however a low-resolution LiDAR was used.
-Higher resolution LiDAR that produces a high density point cloud, such as the Velodyne VLP-32 that will be used in the proposed work \cite{vlp_32c}, may allow for better terrain classification with Gaussian characterization of return intensity and spatial data.
-While capable of real-time classification of terrain type, this method was not employed in road identification and used a low resolution LiDAR system.}
+Higher resolution LiDAR that produces a high density point cloud, such as the Velodyne VLP-32 that will be used in the proposed work \cite{vlp_32c}, may allow for better terrain classification.
+While capable of real-time classification of terrain type, this method was not employed in road identification and used a low resolution LiDAR system.
+Edmond et al.
+propose a method to detect terrain type through implementation of vehicle vibration, however this method is limited to what the vehicle is currently driving over and thus cannot detect intercepting gravel roads \cite{dupont_online_2008}.
+Filitchkin et al.
+propose a method to detect terrain type using a downward orientated camera for use in a small robotic dog, however this method was not extended to explore terrain surrounding the robot \cite{filitchkin_feature_based_2012}.
+Weska et al.
+propose a method of terrain classification through arieal photography, however the implementation on a road vehicle was not explored \cite{weszka_comparative_1976}.
+Obstacle detection using a single axis LiDAR was studied \cite{manduchi_obstacle_2005}, however this method cannot detect road surfaces, only detect obstacles in roads.
+Terrain classification using cameras for use in off-road robots was studied \cite{walch_offroad_2022}, however the method focuses on terrain traversability, furthermore ambient conditions may lower classification accuracy \cite{levi_3d_2012_light}.
+}
 
@@
 
-			{Detection of unmarked gravel road surfaces using terrain classification requires building a data base of terrain features, training an RDF, and testing the algorithm for accuracy of unmarked road detection (outline of work in Fig.
+			{Detection of unmarked gravel road surfaces using terrain classification requires collecting data, training an RDF, and testing the algorithm for accuracy of unmarked road detection (outline of work in Fig.
 \ref{fig:flowz_2}).
@@
 LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C.
-Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a minimum 0.33$^{\circ}$ vertical angular resolution \cite{vlp_32c} and 0.2$^{\circ}$ horizontal angular resolution.
+Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a minimum 0.33$^{\circ}$ vertical angular resolution and 0.2$^{\circ}$ horizontal angular resolution \cite{vlp_32c}.
 GPS data was recorded with a Novatel PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS), and two GNSS-502 antennas by NavtechGPS.
 Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using the Robot Operating System (ROS).
-Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot that intersects Blackburn Road near Athens Ohio.
-Scanning LiDAR data of an asphalt surface was gathered from Blackburn Road near the intersecting gravel lot.
+Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot (Fig.
+\ref{fig:gravel_training_lot}) near Athens Ohio.
+Scanning LiDAR data of an asphalt surface was gathered from Blackburn Road (Fig.
+\ref{fig:Blackburn_Road_View}) near the intersecting gravel lot.
 Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot.
@@
 
-	%	\begin{figure} 
-		%		\centering
-		%		\begin{subfigure}{0.5\textwidth}
-			%			\centering
-			%			\includegraphics[width=1\linewidth,height=4.60 cm,keepaspectratio]{figures/van_on_van}
-			%			\caption[Sensor Van]{}
-			%			\label{fig:van}
-			%		\end{subfigure}
-		%		\begin{subfigure}{0.5\textwidth}
-			%			\centering
-			%			\includegraphics[width=1\linewidth,height=4.60 cm,keepaspectratio]{figures/LiDAR_on_van}
-			%			\caption[VLP 32 on Van]{}
-			%			\label{fig:vlp32mount}
-			%		\end{subfigure}
-		%		\caption[Experimental Apparatus]{Experimental Apparatus (a). Velodyne VLP-32 Scanning LiDAR is mounted on top of the vehicle (b).}
-		%		\label{fig:Experimental_Apperatus}
-		%	\end{figure}
-
 			\begin{figure}[H]
@@
 				\label{fig:Experimental_Apperatus}
+			\end{figure}
+
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.75\linewidth]{figures/blackburn_road}
+				\caption[Blackburn Road Camera View]{Camera view of Blackburn Road - an unmarked asphalt road with three intercepting gravel driveways from which asphalt data was gathered.}
+				\label{fig:Blackburn_Road_View}
+			\end{figure}
+
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.75\linewidth]{figures/gravel_lot_pic}
+				\caption[Gravel Training Lot]{Camera view of the gravel lot from which gravel training data was gathered.}
+				\label{fig:gravel_training_lot}
 			\end{figure}
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/blackburn_road}
-				\caption[Blackburn Road Camera View]{Camera view of Blackburn Road - an unmarked asphalt road with three intercepting gravel driveways from which asphalt data was gathered.}
-				\label{fig:Blackburn_Road_View}
-			\end{figure}
-
-			\begin{figure}[H]
-				\centering
-				\includegraphics[width=0.75\linewidth]{figures/gravel_lot_pic}
-				\caption[Gravel Training Lot]{Camera view of the gravel lot from which gravel training data was gathered.}
-				\label{fig:gravel_training_lot}
-			\end{figure}
-
-			\begin{figure}[H]
-				\centering
-				\includegraphics[width=0.75\linewidth]{figures/pc_example}
+				\includegraphics[width=0.75\linewidth]{figures/classified_point_cloud_example_redux}
 				\caption[Aggregated Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.}
@@
 
-			{Manually defined areas may then be overlayed unto the point cloud (Fig.
-\ref{fig:road_areas_annotated_22}).
+			{Manually defined areas may then be overlayed unto the point cloud representing training areas (Fig.
+\ref{fig:test_vs_train_areas}).
 Six degree arcs (Fig.
-\cite{fig:area_example}) from the first three channels of scanning LiDAR data that lay directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.
+\ref{fig:area_example}) from the first three channels of scanning LiDAR data that lay directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.
 Training data was extracted from a single arc from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area.
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/road_areas_annotated_22}
-				\caption[Manually Classified Point Cloud Data]{ Projected manually defined areas define terrain surface type.
+				\includegraphics[width=0.75\linewidth]{figures/test_vs_train_areas_hatch}
+				\caption[Training vs Testing Areas]{Training areas were kept separate from the testing areas.}
+				\label{fig:test_vs_train_areas}
+			\end{figure}
+
+%			\begin{figure}[H]
+%				\centering
+%				\includegraphics[width=0.75\linewidth]{figures/road_areas_annotated_2}
+%				\caption[Manually Classified Point Cloud Data]{ Projected manually defined areas define terrain surface type.
 Three gravel driveways (annotated 1-3) intercept an asphalt road, area $3$ is the gravel parking lot entrance from which training data was extracted.}
-				\label{fig:road_areas_annotated_22}
-			\end{figure}
+%				\label{fig:road_areas_annotated_22}
+%			\end{figure}
 
@@
 			\centering
-			\includegraphics[width=0.9\linewidth]{figures/training_data_cluster_3}
+			\includegraphics[width=0.75\linewidth]{figures/training_data_cluster_3}
 			\caption[Example Clustering]{Example of training data comparison between two features.}
@@
 			\end{figure}
-
-			%{RANSAC was implementing by a built-in MATLAB function, $pcfitplane$.
-Method of Least Squares was implemented by importing a MATLAB library \cite{noauthor_object-oriented_nodate}.
-Point clouds are individually imported into the MATLAB environment.
-RANSAC or Method of Least Squares may be used to project a reference plane unto the point cloud.
-Displaying the point cloud is completed using MATLAB's $pcshow$ function.
-Training data is selected by manually drawing an area around appropriate points of a desired terrain type using MATLAB's $drawrectangle$ function (Figure \ref{fig:Rectangle_ROI}).
-Indices for the rectangle are extracted, and MATLAB's function $findPointsInROI$ is used to extract sympatric points within the drawn rectangle.
-Captured points are then saved to disk along with the projected plane's normal vector and distance from the origin $[a,b,c,d]$.}
-
-			%	\begin{figure}[H]
-				%		\centering
-				%		\begin{minipage}[c]{.45\linewidth}
-					%			\includegraphics[width=.95\linewidth]{figures/Rectangle_ROI}
-					%			\caption[Region of Interest Selection]{MATLAB was used to select training data using the $drawrectangle$ function and extracting points that lay within the defined area.}
-					%			\label{fig:Rectangle_ROI}
-					%		\end{minipage}
-				%		\hfill
-				%		\begin{minipage}[c]{0.45\linewidth}
-					%			\includegraphics[width=.95\linewidth]{figures/xy_vs_range}
-					%			\caption[XY vs Range vs Z Height]{Three references for feature extraction may be seen in this simplified graph.
-First is the Range shown by the magenta line from the LiDAR origin to the Point of Interest (PoI).
-Second is the height from  LiDAR origin to the PoI (cyan).
-Finally, distance from the projected plane (red) to the PoI (green).
-XY Distance (blue) and Range (magenta) from the sensor was used consider the effect of distance from LiDAR origin to the PoI.}
-					%			\label{fig:xy_vs_range}
-					%		\end{minipage}
-				%	\end{figure}
 
@@
 Spatial features were non-dimensionalized as necessary by dividing by the average range of the arc to LiDAR point of origin to the appropriate power.
-Remission features include Standard Deviation ($\sigma$), Mean, Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), however as remission is a ratio of minimum to maximum detectable reflected intensity there was no need to render the features non-dimensionalized.
+Remission features include Standard Deviation ($\sigma$), Mean, Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$).
+As remission is a ratio of minimum to maximum detectable reflected intensity there was no need to render the features non-dimensionalized.
 Thirty percent of the training data was randomly selected and reserved to create a validation data set.}
@@
 Out-of-bag error is then found and used to create a series of Gaussian curves.
-Additional hyper-parameters are chosen by an acquisition function to determine their utility in minimizing out-of-bag error.
+Additional hyper-parameter values are chosen by an acquisition function to determine their utility in minimizing out-of-bag error.
 After a set number of iterations, best hyper-parameters are chosen based minimum upper confidence interval of the classification error objective model.
@@
 \ref{fig:train_vs_valid_overfit_test2}).
-Confusion matrices provide classification accuracy information on training and validation data sets by comparing actual and predicted classes (Fig.
-\ref{fig:out_of_bag_err_conf_mat}, \ref{fig:vali_err_conf_mat}).
+Confusion matrices provide classification accuracy information on validation database by classifying each data set in the validation database and comparing actual and predicted classes, resulting in a matrix that visualizes the accuracy for each terrain type as well as which terrain types are more likely to be confused for a different class (Fig.
+\ref{fig:vali_err_conf_mat}).
 This process was completed for each of the three considered scanning LiDAR channels used for classification.}
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/train_vs_valid_overfit_test2}
+				\includegraphics[width=0.75\linewidth]{figures/train_vs_valid_overfit_test3}
 				\caption[Training vs Validation Error]{Hyperparameter tuning may lead to over-fitting.
@@
 
-			\begin{figure}[H]
-				\centering
-				\includegraphics[width=0.75\linewidth]{figures/chan_2c_conf_OOB_mat222}
-				\caption[Out-of-Bag Error]{During training the RDF may be tested using the out-of-bag error. Confusion matrices are used to indicate the final RDF iteration training error.}
-				\label{fig:out_of_bag_err_conf_mat}
-			\end{figure}
+%			\begin{figure}[H]
+%				\centering
+%				\includegraphics[width=0.75\linewidth]{figures/chan_2c_conf_OOB_mat222}
+%				\caption[Out-of-Bag Error]{During training the RDF may be tested using the out-of-bag error. Confusion matrices are used to indicate the final RDF iteration training error.}
+%				\label{fig:out_of_bag_err_conf_mat}
+%			\end{figure}
 
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/chan_2c_conf_VALIDATION_mat222}
+				\includegraphics[width=0.55\linewidth]{figures/chan_2c_conf_VALIDATION_mat222}
 				\caption[Validation Error]{Confusion matrix with the validation training data set.}
@@
 
-			{Classifying consecutive LiDAR scans was accomplished by examination of the nine arcs of interest in front and forty-five degree angles left and right from the front of the experimental apparatus (Figure \ref{fig:area_example}).
+			{Classifying consecutive LiDAR scans was accomplished by examination of the nine arcs of interest in front and forty-five degree angles left and right from the front of the experimental apparatus (Fig.
+\ref{fig:area_example}).
 Features from each arc were extracted and the arc was classified (Fig.
 \ref{fig:raw_classification_results} shows the average spatial coordinates of each classified arc).
-Previously-created manually defined gravel, asphalt, and side-of-road truth areas were projected onto the classified point cloud for the determination of classification accuracy.
+Previously-created manually defined gravel, asphalt, and side-of-road truth areas (denoted as Testing areas in Fig.
+\ref{fig:test_vs_train_areas})were projected onto the classified point cloud for the determination of classification accuracy .
 Averaged x, y, and z coordinates were taken of each arc and projected unto a manually defined truth area map (Fig.
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/raw_classification_results_v3}
+				\includegraphics[width=0.9\linewidth]{figures/classified_point_cloud_example_redux}
 				\caption[Classified Point Cloud]{Classified Point Cloud.}
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/rm_db_4_ToC_22}
+				\includegraphics[width=0.9\linewidth]{figures/rm_db_4_overlap_reredux}
 				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed and compared to actual road-surface areas. }
@@
 		\begin{table}[t]
-			\caption[Table]{Average Results\label{tab:1}}
+
 			\centering{%
@@
 			}
+			\caption[Table]{Average Results\label{tab:1}}
 		\end{table}
@@
 		{Classification of real-world data outside of the training data set presents difficulties to the generated Random Decision Forest Classifiers generated in this work.
-Due to the side of the road was being re-grassed near area $3$ (Fig.
-\ref{fig:Blackburn_Road_View}, \ref{fig:road_areas_annotated_22}), the algorithm struggled to differentiate the dirt from gravel, leading to some miss-classification of the side-of-road areas.
+Due to the side of the road was being re-grassed near a portion of the asphalt road, the terrain classification algorithm struggled to differentiate the dirt from gravel, leading to some miss-classification of the side-of-road areas.
 Examination of the feature distribution shows a high amount of overlap (Fig.
@@
 Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 67.54\%.
-If any asphalt miss-classification on the gravel areas is included, the average true-positive road detection accuracy increases to 76.97\%.
-It was found that an average true-positive accuracy of all asphalt surfaces was 87.05\%.
+If any asphalt miss-classification on the gravel areas is included, the average true-positive road detection accuracy increases to 79.18\%.
+It was found that an average true-positive accuracy of all asphalt surfaces was 78.06\%.
 Differentiation between gravel road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation.
 Side-of-road areas were established (Fig.
-\ref{fig:prepostadjust}), and the overall average for "unknown" terrain type was found to be 77.49\%.}
+\ref{fig:prepostadjust}), and the overall average for "unknown" terrain type was found to be 77.94\%.}
 
@@
 			\centering
-			\includegraphics[width=0.75\linewidth]{figures/side_of_road_nums_4_with_filts_2_2}
+			\includegraphics[width=0.75\linewidth]{figures/rm_db_1_area_score}
 			\caption[Area Scores]{Per-area scores were calculated, indicating algorithm unmarked road detection accuracy. }
@@
 
-	%\section{Discussion}
-
 	%%%%% Conclusions %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Properties describing surface roughness and remission data were extracted and used to train Random Decision Forests.
-Terrain classification of consecutive scans compared to manually defined road surface areas yielded accuracy scores for gauging algorithm capability for gravel surface detection.
+Terrain classification of consecutive scans compared to manually detected road surface areas yielded accuracy scores for gauging algorithm capability for gravel surface detection.
 It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively at an estimated rate of 13.19 milliseconds per 360 degree scan.
@@
 Overlapping results between detected and actual road surface areas resulted in 91.67\% intercepting gravel road detection accuracy.
-In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is promising.
-This work addressed the problem of road surface detection on unmarked gravel and chipseal roads.
+In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is promising for future exploration of additional roads and road surfaces.
+Work as described addressed the problem of road surface detection on unmarked gravel and chipseal roads.
 The described work was a method of detecting physical unmarked gravel and chipseal roads by using a terrain classification approach to predicting road surface area.
-The impact of this work is that autonomous vehicles using LiDAR may be able to detect unmarked gravel and chipseal road surfaces, allowing autonomous operations on 1.5 million miles of previously undetected rural roads.}
+The impact of this work is that autonomous vehicles using the developed method may be able to detect unmarked gravel and chipseal road surfaces, allowing autonomous operations on 1.5 million miles of previously undetected rural roads.}
 
@@
 
-	{Special thanks to Dr. Jay Wilhelm for his guidance and Travis Moleski for his help in LiDAR map building.}
+	{The authors would like to acknowledge the Ohio Department of Transportation, United States Department of Transportation, and the Drive Ohio Automated Driving Systems Project for making this work possible.}
 

### 2023-05-25 Huston_Paper.tex
More Updates

@@
 Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Fig.
-\ref{fig:road_areas_annotated_22}).
+\ref{fig:pc_example}).
 While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for this work.}
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/classified_point_cloud_example_redux}
+				\includegraphics[width=0.75\linewidth]{figures/pc_example}
 				\caption[Aggregated Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.}
+				\label{fig:pc_example}
 			\end{figure}
@@
 \ref{fig:train_vs_valid_overfit_test2}).
-Confusion matrices provide classification accuracy information on validation database by classifying each data set in the validation database and comparing actual and predicted classes, resulting in a matrix that visualizes the accuracy for each terrain type as well as which terrain types are more likely to be confused for a different class (Fig.
+Confusion matrices provide validation database classification accuracy information by classifying each data set in the validation database and comparing actual and predicted classes, resulting in a matrix that visualizes the accuracy for each terrain type as well as which terrain types are more likely to be confused for a different class (Fig.
 \ref{fig:vali_err_conf_mat}).
@@
 Previously-created manually defined gravel, asphalt, and side-of-road truth areas (denoted as Testing areas in Fig.
-\ref{fig:test_vs_train_areas})were projected onto the classified point cloud for the determination of classification accuracy .
+\ref{fig:test_vs_train_areas}) were projected onto the classified point cloud for the determination of classification accuracy.
 Averaged x, y, and z coordinates were taken of each arc and projected unto a manually defined truth area map (Fig.
@@
 Higher confidence score requirements resulted in "unknown" over-classification, while too low resulted in "gravel" over-classification.
-Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 67.54\%.
+Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 67.54\% (Fig.
+\ref{fig:prepostadjust}).
 If any asphalt miss-classification on the gravel areas is included, the average true-positive road detection accuracy increases to 79.18\%.
@@
 Differentiation between gravel road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation.
-Side-of-road areas were established (Fig.
-\ref{fig:prepostadjust}), and the overall average for "unknown" terrain type was found to be 77.94\%.}
+Overall average distribution for "unknown" terrain class within side-of-road areas was found to be 77.94\%.}
 
@@
 		% Summary of Work & Closing Statements
-		{High definition scanning LiDAR and GPS data was gathered on Blackburn Road and an intercepting gravel lot in Athens Ohio.
+		{High definition scanning LiDAR and GPS data was gathered on Blackburn Road and intercepting gravel lots in Athens Ohio.
 Properties describing surface roughness and remission data were extracted and used to train Random Decision Forests.

### 2023-06-15 Huston_Paper.tex
hey

@@
 Random Decision Forests were trained using scanning LiDAR data terrain classification to detect unmarked gravel and asphalt surfaces.
-It was found that a true-positive accuracy for gravel and asphalt surfaces was 69.15\% and 78.07\% respectively at an estimated rate of 13.19 ms per 360 degree scan.
-Overlapping results between manually projected and actual road surface areas resulted in 91.67\% intercepting gravel road detection accuracy.
+It was found that a true-positive accuracy for gravel and asphalt surfaces was 74.97\% and 84.05\% respectively at an estimated rate of 13.19 ms per 360 degree scan.
+Overlapping results between manually projected and actual road surface areas resulted in 93.33\% intercepting gravel road detection accuracy.
 Results indicate that using a Random Decision Forest to classify scanning LiDAR data is a feasible method for unmarked gravel and asphalt road detection.
@@
 Detection based on road surface smoothness was studied by segmenting the point cloud into candidate road surface regions and searching for elevation jumps \cite{liu_new_2013}.
-Secondly is the detection of roadside curbs, which relies upon the height difference between the road surface and the curb for road edge detection.
+Second is the detection of roadside curbs, which relies upon the height difference between the road surface and the curb for road edge detection.
 LiDAR point density at a curb's upper and lower edges can be used to indicate road edges \cite{ibrahim_curb-based_2012}.
@@
 Comparison with neighboring cells increases classification type accuracy, as adjacent cell are likely to have the same terrain type \cite{haselich_terrain_2011,zhao_fusion_2014}.
-It was found that CRF was better suited for classification.	Random Forest (RF) classification showed good results for identification of analysis of LiDAR and camera data \cite{breiman_random_2001}, and was used to predict cell terrain into one of five types, including gravel and grass.
-It was found that characterizing the LiDAR intensity as Gaussian distributed noise yielded poor results with only a 49.5\% true positive rate, however a low-resolution LiDAR was used.
+It was found that CRF was better suited for classification.}
+
+	{Random Decision Forest (RDF) classification showed good results for identification of analysis of LiDAR and camera data \cite{breiman_random_2001}, and was used to predict cell terrain into one of five types, including gravel and grass.
+It was found that characterizing the LiDAR intensity as Gaussian distributed noise yielded poor results with only a 49.5\% true positive rate, however a low-resolution LiDAR was used \cite{rauscher_comparison_2016}.
 Higher resolution LiDAR that produces a high density point cloud, such as the Velodyne VLP-32 that will be used in the proposed work \cite{vlp_32c}, may allow for better terrain classification.
@@
 Edmond et al.
-propose a method to detect terrain type through implementation of vehicle vibration, however this method is limited to what the vehicle is currently driving over and thus cannot detect intercepting gravel roads \cite{dupont_online_2008}.
+proposed a method to detect terrain type through implementation of vehicle vibration, however this method is limited to what the vehicle is currently driving over and thus cannot detect intercepting gravel roads \cite{dupont_online_2008}.
 Filitchkin et al.
-propose a method to detect terrain type using a downward orientated camera for use in a small robotic dog, however this method was not extended to explore terrain surrounding the robot \cite{filitchkin_feature_based_2012}.
+proposed a method to detect terrain type using a downward orientated camera for use in a small robotic dog; however, this method was not extended to explore terrain surrounding the robot \cite{filitchkin_feature_based_2012}.
 Weska et al.
-propose a method of terrain classification through arieal photography, however the implementation on a road vehicle was not explored \cite{weszka_comparative_1976}.
-Obstacle detection using a single axis LiDAR was studied \cite{manduchi_obstacle_2005}, however this method cannot detect road surfaces, only detect obstacles in roads.
-Terrain classification using cameras for use in off-road robots was studied \cite{walch_offroad_2022}, however the method focuses on terrain traversability, furthermore ambient conditions may lower classification accuracy \cite{levi_3d_2012_light}.
+proposed a method of terrain classification through aerial photography; however, the implementation on a road vehicle was not explored \cite{weszka_comparative_1976}.
+Obstacle detection using a single axis LiDAR was studied \cite{manduchi_obstacle_2005}; however, this method could not detect road surfaces, only obstacles in roads.
+Terrain classification using cameras for use in off-road robots was studied \cite{walch_offroad_2022}; however, the method focuses on terrain traversability, and ambient conditions may lower classification accuracy \cite{levi_3d_2012_light}.
 }
@@
 Road classification with LiDAR spacial and remission features was completed with Neural Network classifier, Naive Bayes classifier, and Support Vector Machines (SVM) \cite{wang_road_nodate,wang_two-stage_2018}.
-Gravel roads were successfully identified with a 98.5\% true positive identification rate, however data collected required post processing.
-MATLAB's Neural Network tools were used in autonomous road identification using LiDAR and Camera data \cite{rasmussen_combining_2002}, however post processing was required, with unknown computational time.
+Gravel roads were successfully identified with a 98.5\% true positive identification rate; however, data collected required post processing.
+MATLAB's Neural Network tools were used in autonomous road identification using LiDAR and Camera data \cite{rasmussen_combining_2002}; however, post processing was required, with unknown computational time.
 Terrain classification was completed using Support Vector Machine (SVM), a classification algorithm where a line is drawn between two different categories to differentiate between them \cite{breiman_random_2001}.
@@
 
-	{Random Decision Forests are a supervised machine learning method that operates on a set of decision trees \cite{ho_random_1995}.
+	{RDFs are a supervised machine learning method that operates on a set of decision trees \cite{ho_random_1995}.
 RDFs have been shown to be useful with handling LiDAR data sets \cite{breiman_random_2001}, and have been employed in terrain classification \cite{laible_3d_2012,laible_map_building,laible_terrain_2013,khan_high_2011,reymann_improving_2015,schilling_geometric_2017, wietrzykowski_context-aware_2019}.
@@
 \ref{fig:flowz_2}).
-Training data was gathered by the Ohio University Autonomous Van van (Fig.
+Training data was gathered by the Ohio University Autonomous Van (Fig.
 \ref{fig:Experimental_Apperatus}).
@@
 Random Decision Forests were trained and verified using MATLAB's Classifier App.
-Scanning LiDAR data was consecutively classified and compared to manually defined truth areas yielded the true-positive accuracy for gravel, asphalt, and non-road surfaces.
+Scanning LiDAR data was consecutively classified and compared to manually defined truth areas, yielding the true-positive accuracy for gravel, asphalt, and non-road surfaces.
 Classified point cloud data was examined without reference to truth areas in order to manually project possible road surfaces.
@@
 Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot.
-Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection, therefore grass is labeled as "unknown" in this work.
+Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection; therefore, grass is labeled as \textit{unknown} in this work.
 Raw data was gathered using the Robotic Operating System (ROS) to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags.
 Rosbags were unpacked using MATLAB tools to separate data streams.
-Point cloud, GPS, and IMU topics were defined, and point cloud Cartesian coordinates with corresponding intensities, longitude, latitude, altitude, roll, pitch, and yaw messages were read from the appropriate topics and imported}
+Point cloud, GPS, and IMU topics were defined, and point cloud Cartesian coordinates with corresponding intensities, longitude, latitude, altitude, roll, pitch, and yaw messages were read from the appropriate topics and imported.}
 
@@
 
-			{Scanning LiDAR training data requires extraction from manually defined gravel, asphalt, and "unknown" terrain areas on aggregated point clouds.
+			{Scanning LiDAR training data requires extraction from manually defined \textit{gravel}, \textit{asphalt}, and \textit{unknown} terrain areas on aggregated point clouds.
 Aggregating LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
@@
 LiDAR origin was found using GPS longitude, latitude, and altitude data.
-Physical distances between the GPS, IMU, and LiDAR sensors was rectified by obtaining reference frames provided by AutonomouStuff.
+Physical distances between the GPS, IMU, and LiDAR sensors were rectified by obtaining reference frames provided by AutonomouStuff.
 GPS coordinates were offset by the current orientation and converted the ground truth to the LiDAR frame.
 GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors.
-Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Fig.
+Consecutive point cloud scans were then translated and rotated with derived transformation matrices (Fig.
 \ref{fig:pc_example}).
-While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for this work.}
+While not as robust as more sophisticated methods of point cloud aggregation, such as NDT or ICP scan matching, using this method proved to be adequate over shorter distances.}
 
@@
 
-			{Manually defined areas may then be overlayed unto the point cloud representing training areas (Fig.
+			{Manually defined areas were then overlayed onto the point cloud, representing training areas (Fig.
 \ref{fig:test_vs_train_areas}).
 Six degree arcs (Fig.
-\ref{fig:area_example}) from the first three channels of scanning LiDAR data that lay directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.
+\ref{fig:area_example}) from the first three channels of scanning LiDAR data directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.
 Training data was extracted from a single arc from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area.
@@
 Thirty percent of the training data base was randomly selected and set aside as a validation data base.
-Training data was visually examined for clustering to verify that a Random Decision Forest would be able to be trained (Fig.
+Training data was visually examined for clustering to verify that a RDF would be able to be trained (Fig.
 \ref{fig:training_data_cluster_2}).
-It was found that each class generally clustered to a certain range with some overlap, which indicates that a Random Decision Forest may be trained.}
+It was found that each class generally clustered to a certain range with some overlap, which indicates that a RDF may be trained.}
 
@@
 			{LiDAR returns spatial data represented by Cartesian coordinates with units in meters and remission data represented by a dimensionless ratio of minimum to maximum detected brightness.
-Feature extraction was completed by performing mathematical functions on the gathered spatial training data.
-Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial data array.
-Spatial features were non-dimensionalized as necessary by dividing by the average range of the arc to LiDAR point of origin to the appropriate power.
+Feature extraction was completed by performing mathematical functions on the gathered spatial training data, yielding the Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial data array.
+Spatial features were non-dimensionalized as necessary by dividing the average range of the arc to LiDAR point of origin to the appropriate power.
 Remission features include Standard Deviation ($\sigma$), Mean, Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$).
-As remission is a ratio of minimum to maximum detectable reflected intensity there was no need to render the features non-dimensionalized.
+As remission is a ratio of minimum to maximum detectable reflected intensity, there was no need to render the features non-dimensionalized.
 Thirty percent of the training data was randomly selected and reserved to create a validation data set.}
@@
 
-			{MATLAB's Classification Learner Application was used to generate a Random Decision Forests for each considered scanning LiDAR channel.
+			{MATLAB's Classification Learner Application was used to generate an RDF for each considered scanning LiDAR channel.
 Decision tree depth, number of learners, and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.
-Bayesian Optimization automates the manual hyper-parameter tuning process by creating a surrogate model representing the objective function, in this the upper confidence interval of the classification error objective model (Fig.
+Bayesian Optimization automates the manual hyper-parameter tuning process by creating a surrogate model representing the objective function, in this case the upper confidence interval of the classification error objective model (Fig.
 \ref{fig:c2_min_class_error}).
@@
 Additional hyper-parameter values are chosen by an acquisition function to determine their utility in minimizing out-of-bag error.
-After a set number of iterations, best hyper-parameters are chosen based minimum upper confidence interval of the classification error objective model.
+After a set number of iterations, best hyper-parameters are chosen based on minimum upper confidence interval of the classification error objective model.
 Hyperparameter tuning may lead to over-fitting to the training data base.
@@
 				\caption[RDF Training Classification Error]{Classification error during Bayesian Optimization tuning.
-Best hyper-parameters are chosen based minimum upper confidence interval of the classification error objective model, not minimum classification error.}
+Best hyper-parameters are chosen based on minimum upper confidence interval of the classification error objective model, not minimum classification error.}
 				\label{fig:c2_min_class_error}
@@
 				\includegraphics[width=0.75\linewidth]{figures/train_vs_valid_overfit_test3}
-				\caption[Training vs Validation Error]{Hyperparameter tuning may lead to over-fitting.
+				\caption[Training vs Validation Error]{Hyper-parameter tuning may lead to over-fitting.
 Testing out-of-bag versus validation error demonstrates that the model is not over-fitting to the training data base if the validation error does not diverge from the out-of-bag error.}
@@
 \ref{fig:test_vs_train_areas}) were projected onto the classified point cloud for the determination of classification accuracy.
-Averaged x, y, and z coordinates were taken of each arc and projected unto a manually defined truth area map (Fig.
+Averaged x, y, and z coordinates were taken of each arc, and projected unto a manually defined truth area map (Fig.
 \ref{fig:rm_db_4_toc}).
@@
 If consecutive points from each channel in any direction had two or more matching gravel or asphalt classification and was evenly spaced, a gravel or road surface area was projected.
-Guessed road surface areas were then compared to actual areas to determine the method's ability at unmarked gravel road discovery, indicating the method's efficiency at detecting intercepting gravel roads.
+Guessed road surface areas were then compared to actual areas to determine the method's ability at unmarked gravel road discovery, indicating the method's efficiency at detecting intersecting gravel roads.
 }
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/classified_point_cloud_example_redux}
+				\includegraphics[width=0.9\linewidth]{figures/range_classified_pc_db_6}
 				\caption[Classified Point Cloud]{Classified Point Cloud.}
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/rm_db_4_overlap_reredux}
-				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed and compared to actual road-surface areas. }
+				\includegraphics[width=0.9\linewidth]{figures/cross_channel_trend}
+				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed based on trends between channels and compared to actual road-surface areas.}
 				\label{fig:rm_db_4_toc}
@@
 
-		{Five drive-bys of Blackburn Road with three intercepting gravel parking lots and driveways were completed and scanning LiDAR, GPS, and IMU data was collected.
-Training data was generated using MATLAB to extract LiDAR data from rosbags.
-Random Decision Forests classification algorithms were generated.
+		{Five drive-bys of Blackburn Road, which has three intersecting gravel parking lots and driveways, were completed, and scanning LiDAR, GPS, and IMU data was collected.
+Training data was generated using MATLAB to extract LiDAR data from rosbags, and RDF classification algorithms were generated.
 Testing consecutive scans was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data.
@@
 		{Classification of real-world data outside of the training data set presents difficulties to the generated Random Decision Forest Classifiers generated in this work.
-Due to the side of the road was being re-grassed near a portion of the asphalt road, the terrain classification algorithm struggled to differentiate the dirt from gravel, leading to some miss-classification of the side-of-road areas.
+Due to the side of the road being re-grassed near a portion of the asphalt road, the terrain classification algorithm struggled to differentiate the dirt from gravel, leading to some misclassification of the side-of-road areas.
 Examination of the feature distribution shows a high amount of overlap (Fig.
@@
 Exploitation of confidence scores to adjust results by requiring a very high confidence for positive gravel surface classification was explored.
-All gravel surfaces that did not meet a 80\% confidence score was re-labeled as "unknown", however these adjustment had either a negligible or detrimental impact on the final results.
-Higher confidence score requirements resulted in "unknown" over-classification, while too low resulted in "gravel" over-classification.
-Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 67.54\% (Fig.
+All gravel surfaces that did not meet a 80\% confidence score was re-labeled as \textit{unknown}, however these adjustment had either a negligible or detrimental impact on the final results.
+Higher confidence score requirements resulted in \textit{unknown} over-classification, while too low resulted in \textit{gravel} over-classification.
+Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 74.97\% (Fig.
 \ref{fig:prepostadjust}).
 If any asphalt miss-classification on the gravel areas is included, the average true-positive road detection accuracy increases to 79.18\%.
-It was found that an average true-positive accuracy of all asphalt surfaces was 78.06\%.
+It was found that an average true-positive accuracy of all asphalt surfaces was 87.05\%.
+Due to recent construction, discovered asphalt accuracy was artificially lowered due to the area in front of a gravel driveway being contaminated with gravel.
+It was found that an average of 18.55\% of asphalt areas were classified as gravel because of contamination.
 Differentiation between gravel road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation.
-Overall average distribution for "unknown" terrain class within side-of-road areas was found to be 77.94\%.}
+Overall average distribution for \textit{unknown} terrain class within side-of-road areas was found to be 77.94\%, which is lower than expected due to the side of the road being re-grassed.
+It is acknowledged by the authors that the classification algorithm must yield a result that most closely matches one of the trained classes, which led to lower accuracy scores in these circumstances.
+It is difficult to determine if these results are comparable to similar research, as found research classifies larger data sets.
+For example, work completed by Laible et.
+al.
+\cite{levi_3d_2012_light} used $\sim1200$ data points, compared to the $\sim28$ data points per arc in this work.
+Overall discovered accuracy levels may be sufficient for the purpose of identifying asphalt and gravel roads by considering trends between classified arcs.}
 
@@
 			\centering
-			\includegraphics[width=0.75\linewidth]{figures/rm_db_1_area_score}
-			\caption[Area Scores]{Per-area scores were calculated, indicating algorithm unmarked road detection accuracy. }
+			\includegraphics[width=0.75\linewidth]{figures/range_actual_rm_db_4_area_score}
+			\caption[Area Scores]{Per-area scores were calculated, indicating algorithm unmarked road detection accuracy.
+Note that the asphalt area in front of the gravel driveway has significant \textit{gravel} classification due to construction.
+These results are included with the overall asphalt classification score}
 			\label{fig:prepostadjust}
@@
 Rate of classification for all nine areas of interest in a single scan was found to be an average of 35.19 ms using a single core (Fig.
-\ref{fig:per_arc_classification_time}) or 316.68 ms per 360 degree scan (Fig.
-\ref{fig:per_scan_classification_rate}).
-Further optimization may increase this speed, however real-time classification is not a focus of this work.
-When utilizing all 24 cores during this indicates that a rate of 13.19 ms per 360 scan.
-Extrapolating from the distance traveled during a drive-by, a hypothetical 71.39 feet per second or 48.68 miles per hour may be possible.}
+\ref{fig:per_arc_classification_time}) or 316.68 ms per 360 degree scan.
+When utilizing all 24 cores during classification a rate of 13.19 ms per 360 scan.
+Extrapolating from the distance traveled during a drive-by, a hypothetical 71.39 feet per second or 48.68 miles per hour may be possible.
+Further optimization may increase this speed, however implementing real-time classification was not a focus of this work.}
 
@@
 
-		\begin{figure}[H]
-			\centering
-			\includegraphics[width=0.75\linewidth]{figures/per_scan_classification_rate}
-			\caption[Per-Scan Time]{Per-scan classification time}
-			\label{fig:per_scan_classification_rate}
-		\end{figure}
+% 		(Fig. \ref{fig:per_scan_classification_rate})		
+%		\begin{figure}[H]
+%			\centering
+%			\includegraphics[width=0.75\linewidth]{figures/per_scan_classification_rate}
+%			\caption[Per-Scan Time]{Per-scan classification time}
+%			\label{fig:per_scan_classification_rate}
+%		\end{figure}
 
@@
 		% Summary of Work & Closing Statements
-		{High definition scanning LiDAR and GPS data was gathered on Blackburn Road and intercepting gravel lots in Athens Ohio.
+		{High definition scanning LiDAR and GPS data was gathered on Blackburn Road and intersecting gravel lots in Athens Ohio.
 Properties describing surface roughness and remission data were extracted and used to train Random Decision Forests.
@@
 Classified point clouds were examined and gravel areas were projected and compared with actual gravel surface areas.
-Overlapping results between detected and actual road surface areas resulted in 91.67\% intercepting gravel road detection accuracy.
+Overlapping results between detected and actual road surface areas resulted in 91.67\% intersecting gravel road detection accuracy.
 In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is promising for future exploration of additional roads and road surfaces.

### 2023-07-07 Huston_Paper.tex
Autonomous Section Update

@@
 		{Gravel roads lack standardized features such as curbs or painted lines, presenting detection challenges to autonomous vehicles.
-Global Positioning Service (GPS) and high resolution maps may not be reliable for navigation of gravel roads, as some may only be width of the vehicle and GPS may not be accurate enough.
-Normal Distribution Transform (NDT) based on scanning LiDAR data may be insufficient for navigating on gravel roads as there may not be enough geometrically distinct features for reliable scan matching.
-This work will examine a method of classifying scanning LiDAR data spatial and remission features for explicit detection of unmarked gravel road surfaces.
+Global Positioning Service (GPS) and high resolution maps may not be reliable for navigation of gravel roads, as some roads may only be width of the vehicle and GPS may not be accurate enough.
+Normal Distribution Transform (NDT) LiDAR scan matching may be insufficient for navigating on gravel roads as there may not be enough geometrically distinct features for reliable scan matching.
+Completed work will examine a method of classifying scanning LiDAR spatial and remission data features for explicit detection of unmarked gravel road surfaces.
 Exploration of terrain classification using high resolution scanning LiDAR data of these road surfaces may allow for predicting gravel road boundary locations potentially enabling confident autonomous operations on gravel roads.
-The principal outcome of this work was a method for gravel road detection using LiDAR data for the purpose of predicting road boundary locations.
+The principal outcome of this work was a method for gravel road terrain detection using LiDAR data for the purpose of predicting potential road boundary locations.
 Random Decision Forests were trained using scanning LiDAR data terrain classification to detect unmarked gravel and asphalt surfaces.
-It was found that a true-positive accuracy for gravel and asphalt surfaces was 74.97\% and 84.05\% respectively at an estimated rate of 13.19 ms per 360 degree scan.
+It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively at an estimated rate of 13.19 ms per 360 degree scan.
 Overlapping results between manually projected and actual road surface areas resulted in 93.33\% intercepting gravel road detection accuracy.
-Results indicate that using a Random Decision Forest to classify scanning LiDAR data is a feasible method for unmarked gravel and asphalt road detection.
+Automated post-process examination of classification results yielded an  true-positive gravel road detection rate of 71.67\%.
 Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.}
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/flowz_2}
+				\includegraphics[width=0.9\linewidth]{figures/flowz_7}
 				\caption[Project Flow]{High level view of project work}
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/gravel_lot_pic}
+				\includegraphics[width=0.75\linewidth]{figures/gravel_training_lot}
 				\caption[Gravel Training Lot]{Camera view of the gravel lot from which gravel training data was gathered.}
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/pc_example}
+				\includegraphics[width=0.75\linewidth]{figures/combined_pcd_example}
 				\caption[Aggregated Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.}
@@
 
+		\subsection{Surface Classification and Road Detection}
+
 			{Classifying consecutive LiDAR scans was accomplished by examination of the nine arcs of interest in front and forty-five degree angles left and right from the front of the experimental apparatus (Fig.
@@
 Averaged x, y, and z coordinates were taken of each arc, and projected unto a manually defined truth area map (Fig.
-\ref{fig:rm_db_4_toc}).
+\ref{fig:visual_v_auto}).
 Accuracy scores was calculated based on the distribution of classified points with each area, allowing evaluation of unmarked road detection performance.
@@
 Guessed road surface areas were then compared to actual areas to determine the method's ability at unmarked gravel road discovery, indicating the method's efficiency at detecting intersecting gravel roads.
+Automated detection of unmarked road was explored in order to determine if visual detection results may be repeated.
+Class percentages, standard deviation of height, standard deviation of average [x,y] distance to the LiDAR point of origin, and distance trends between classified arcs among other features were exploited to determine if the area best describes a gravel, asphalt, or unknown surface (Fig.
+\ref{fig:visual_v_auto}).
+Results were compiled into a single classified map and compared to truth areas, then scored based off the number of true/false positive road detection.
 }
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/range_classified_pc_db_6}
+				\includegraphics[width=0.9\linewidth]{figures/db1range.png}
 				\caption[Classified Point Cloud]{Classified Point Cloud.}
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/cross_channel_trend}
-				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed based on trends between channels and compared to actual road-surface areas.}
-				\label{fig:rm_db_4_toc}
+				\includegraphics[width=0.9\linewidth]{figures/visual_v_auto}
+				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed based on trends between channels and compared to actual road-surface areas (left).
+Autonomous projection of road surface areas following developed methods (right).}
+				\label{fig:visual_v_auto}
 			\end{figure}
@@
 Two out of three positive gravel identifications must be made on each channel, be evenly spaced, and be relatively co-planar to other road surface guesses in order for a positive guess for an asphalt or gravel surface to be made (Fig.
-\ref{fig:rm_db_4_toc}).
+\ref{fig:visual_v_auto}).
 Driveways $1$ and $2$ were detected in all five drive-bys, driveway $3$ was detected in four leading to an 91.67\% accuracy in intercepting gravel road detection.
@@
 In all drive-bys the asphalt road was accurately detected.
-}
+It is acknowledged that the author's a-priori knowledge of the gravel drive locations may have introduced a bias, however this method proved useful for developing an autonomous method.
+Trends across classified arcs were autonomously exploited to determine the best guess at area terrain type.
+Guessed areas were aggregated into a larger map then compared to truth areas (Fig.
+\ref{fig:guess_grav_intersect}).
+Using autonomous methods to project road surfaces, it was found that a true-positive intersecting gravel road detection rate of 71.66\% with no false positives.
+It was found that the true positive rate may be increased to 100\% at the expense of increasing the number of false positives with a 1.8 average number of false positives per drive-by.}
 
-		{Classification was completed using Ubuntu 18.04 and an AMD Ryzen 9 5900X 12 core 24 thread.
+		\begin{figure}[H]
+			\centering
+			\includegraphics[width=0.95\linewidth]{figures/db1_conf_elim_auto_guess_mca}
+			\caption[Aggregated Automated Road Surface Detection]{Aggregated automated road surface detection was compared to truth areas. }
+			\label{fig:guess_grav_intersect}
+		\end{figure}
+
+		{Classification was completed using Ubuntu 18.04 and an AMD Ryzen 9 5900X 12 core 24 thread CPU.
 Rate of classification for all nine areas of interest in a single scan was found to be an average of 35.19 ms using a single core (Fig.
@@
 Classified point clouds were examined and gravel areas were projected and compared with actual gravel surface areas.
-Overlapping results between detected and actual road surface areas resulted in 91.67\% intersecting gravel road detection accuracy.
+Overlapping results between detected and actual road surface areas resulted in a minimum  71.66\% intersecting gravel road autonomous detection accuracy.
 In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is promising for future exploration of additional roads and road surfaces.

### 2023-08-14 Huston_Paper.tex
hey stop looking at my computer

@@
 Classified point clouds were examined and gravel areas were projected and compared with actual gravel surface areas.
-Overlapping results between detected and actual road surface areas resulted in a minimum  71.66\% intersecting gravel road autonomous detection accuracy.
+Overlapping results between autonomously detected and actual road surface areas resulted in a minimum  71.66\% intersecting gravel road detection accuracy.
 In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is promising for future exploration of additional roads and road surfaces.

### 2023-08-15 Huston_Paper.tex
start on asme feedback

@@
 Automated post-process examination of classification results yielded an  true-positive gravel road detection rate of 71.67\%.
-Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.}
+Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on previously undetected roads.}
 
@@
 	{Further limitations are imposed when considering data set size and processing speed.
-Real time analysis of road surfaces dictates that trajectory updates to the vehicle in motion must have a rapid update rate, which prohibits the relatively lengthy collection of large data sets or any form of post-processing.
+Real time trajectory updates to the vehicle in motion requires low latency road surface analysis, denying the use of methods that require post-processing due to high computational loads.
+Real time analysis of road surfaces dictates that trajectory updates to the vehicle in motion must have low latency, which prohibits the relatively lengthy collection of large data sets or any form of post-processing.
 Proposed methods \cite{yadav_extraction_2017,yadav_road_2018,yadav_rural_2018,yadav_pole-shaped_2015,miyazaki_line-based_2014,yang_semi-automated_2013,liu_new_2013,qiu_fast_2016} do not indicate the minimum number of points required for road surface detection, however large data sets numbering many millions of points are used in those studies.
@@
 Road classification with LiDAR spacial and remission features was completed with Neural Network classifier, Naive Bayes classifier, and Support Vector Machines (SVM) \cite{wang_road_nodate,wang_two-stage_2018}.
-Gravel roads were successfully identified with a 98.5\% true positive identification rate; however, data collected required post processing.
-MATLAB's Neural Network tools were used in autonomous road identification using LiDAR and Camera data \cite{rasmussen_combining_2002}; however, post processing was required, with unknown computational time.
+Gravel surfaces were successfully identified with a 98.5\% true positive identification rate; however, the method was only focused on classifying road surface type directly behind the vehicle, and was not tested for predicting road surface areas.
+MATLAB's Neural Network tools were used in autonomous road identification using LiDAR and Camera data \cite{rasmussen_combining_2002}; however, cameras may not be relied upon for ambient conditions may affect classification accuracy \cite{levi_3d_2012_light}, furthermore computational time was not mentioned.
 Terrain classification was completed using Support Vector Machine (SVM), a classification algorithm where a line is drawn between two different categories to differentiate between them \cite{breiman_random_2001}.
@@
 Classification is determined by the majority results as opposed to the mean \cite{breiman_random_2001,ho_random_1995}.
+While RDFs were used in terrain classification, they have not been, to the author's best knowledge, been employed in predicting road surface areas based on terrain classification results while also only using extremely small input data sets typical of real-time mobile scanning.
 Bayesian Optimization may be used to tune the hyper parameters of a machine algorithm \cite{noauthor_bayesian_nodate, snoek_practical_2012}.
@@
 Scanning LiDAR data was consecutively classified and compared to manually defined truth areas, yielding the true-positive accuracy for gravel, asphalt, and non-road surfaces.
-Classified point cloud data was examined without reference to truth areas in order to manually project possible road surfaces.
-Manually projected road surface areas were compared to truth areas to determine the accuracy of asphalt road and intercepting gravel driveway detection.}
+Classified point cloud data was manually examined in order to manually project possible road surfaces.
+Manually projected road surface areas were compared to truth areas to determine the accuracy of asphalt road and intersecting gravel driveway detection.}
 
@@
 			{Required physical data was gathered by the Ohio University Autonomous Van (Fig.
-\ref{fig:Experimental_Apperatus}).
+\ref{fig:Experimental_Apperatus})  on a sunny, cloudless day with no road obstacles or debris at a speed below 15 miles per hour.
+It is acknowledged that weather or non-ideal road conditions may affect results and is an area of future study.
 LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C.
@@
 Scanning LiDAR data of an asphalt surface was gathered from Blackburn Road (Fig.
-\ref{fig:Blackburn_Road_View}) near the intersecting gravel lot.
+\ref{fig:Blackburn_Road_View}) which provides access to the training gravel lot.
 Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot.
 Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection; therefore, grass is labeled as \textit{unknown} in this work.
+It is acknowledged by the authors that the algorithm must classify as one of these three terrain types.
+Due to the potentially unlimited number of terrain types,
+
 				Raw data was gathered using the Robotic Operating System (ROS) to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags.
@@
 \ref{fig:pc_example}).
-While not as robust as more sophisticated methods of point cloud aggregation, such as NDT or ICP scan matching, using this method proved to be adequate over shorter distances.}
+While not as robust as more sophisticated methods of point cloud aggregation due to IMU drift, such as NDT or ICP scan matching, using this method proved to be adequate over shorter distances.}
 
@@
 Six degree arcs (Fig.
-\ref{fig:area_example}) from the first three channels of scanning LiDAR data directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.
-Training data was extracted from a single arc from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area.
+\ref{fig:area_example}) from the first three channels of scanning LiDAR data centered directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.
+Later, nine arcs will be classified for the prediction of intersecting roads.
+Arc length was chosen based on approximate vehicle width (~6 degrees), and the location of each arc was fixed relative to the center line (length-wise) of the vehicle.
+Training data was extracted from a single arc, with approximately 28 LiDAR points per arc, from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area.
 Raw spatial and intensity values from the coincident arc were saved to a raw training database.
@@
 
-		{Five drive-bys of Blackburn Road, which has three intersecting gravel parking lots and driveways, were completed, and scanning LiDAR, GPS, and IMU data was collected.
+		{Training and verification LiDAR, GPS, and IMU data collection was completed using the Experimental Apparatus.
+Gravel data was gathered from the Redmen Lodge gravel parking lot, while grass data was gathered beside the lot.
+Asphalt data was gathered Blackburn Road.
+Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training; however, for this study the algorithm that produced the final results used multiple thousands of data sets.
+Results were then compared to algorithms trained with smaller data sets.
 Training data was generated using MATLAB to extract LiDAR data from rosbags, and RDF classification algorithms were generated.
@@
 It is acknowledged by the authors that the classification algorithm must yield a result that most closely matches one of the trained classes, which led to lower accuracy scores in these circumstances.
-It is difficult to determine if these results are comparable to similar research, as found research classifies larger data sets.
-For example, work completed by Laible et.
-al.
-\cite{levi_3d_2012_light} used $\sim1200$ data points, compared to the $\sim28$ data points per arc in this work.
 Overall discovered accuracy levels may be sufficient for the purpose of identifying asphalt and gravel roads by considering trends between classified arcs.}
@@
 
+		{It is difficult to determine if these results are comparable to similar research, as found research classifies larger LiDAR data sets.
+For example, work completed by Laible et.
+al.
+\cite{levi_3d_2012_light} used $\sim1200$ data points, compared to the $\sim28$ data points per arc in this work.
+Another example is the work completed by Wang et.
+al.
+\cite{wang_road_nodate}, which found a true positive gravel terrain classification rate of 98.5\%; however, the method again used significantly more data from the LiDAR sensor compared to this work, exploiting all LiDAR channels that fall within a 1.3 meter region of interest.
+To the best of the author's knowledge, this study employed a relatively unique approach for detecting road surface area and type.}
+
 % 		(Fig. \ref{fig:per_scan_classification_rate})		
@@
 In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is promising for future exploration of additional roads and road surfaces.
-Work as described addressed the problem of road surface detection on unmarked gravel and chipseal roads.
-The described work was a method of detecting physical unmarked gravel and chipseal roads by using a terrain classification approach to predicting road surface area.
-The impact of this work is that autonomous vehicles using the developed method may be able to detect unmarked gravel and chipseal road surfaces, allowing autonomous operations on 1.5 million miles of previously undetected rural roads.}
+Work as described addressed the problem of road surface detection on unmarked gravel and asphalt roads.
+The described work was a method of detecting physical unmarked gravel and asphalt roads by using a terrain classification approach to predicting road surface area.
+The impact of this work is that autonomous vehicles using the developed method may be able to detect unmarked gravel and asphalt road surfaces, allowing autonomous operations on previously undetected rural roads.}
 

### 2023-08-16 Huston_Paper.tex
I SEE YOU

@@
 It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively at an estimated rate of 13.19 ms per 360 degree scan.
-Overlapping results between manually projected and actual road surface areas resulted in 93.33\% intercepting gravel road detection accuracy.
+Overlapping results between manually projected and actual road surface areas resulted in 93.33\% intersecting gravel road detection accuracy.
 Automated post-process examination of classification results yielded an  true-positive gravel road detection rate of 71.67\%.
@@
 Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}.
-Implementing a Random Decision Forests (RDF) terrain classification model using scanning LiDAR data to exploit surface roughness was studied to determine the method's feasibility in road surface detection.
+Real-time detection of road surfaces limits the available computational time required for low-latency vehicle trajectory updates, prohibiting the collection of larger point LiDAR point clouds.
+Implementing a Random Decision Forests (RDF) terrain classification model using arcs from a single 360 degree scan of LiDAR data to exploit surface properties was studied to determine the method's feasibility in road surface detection.
+While an established method for terrain classification, to the author's best knowledge RDFs in found literature were not employed for the detection of road surface area and type using a small data input within the context of an unstructured environment, making the approach used in this work is a novel one.
 Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.}
@@
 Edmond et al.
-proposed a method to detect terrain type through implementation of vehicle vibration, however this method is limited to what the vehicle is currently driving over and thus cannot detect intercepting gravel roads \cite{dupont_online_2008}.
+proposed a method to detect terrain type through implementation of vehicle vibration, however this method is limited to what the vehicle is currently driving over and thus cannot detect intersecting gravel roads \cite{dupont_online_2008}.
 Filitchkin et al.
@@
 
+%		{Detection of unmarked gravel road surfaces using terrain classification requires building a data base of terrain features, training an RDF, and testing the algorithm for accuracy of unmarked road detection (outline of work in Figure \ref{fig:flowz_7}).
+Raw data was gathered using the experimental apparatus, which has a Velodyne VLP-32C scanning LiDAR sensor, and Novatel PwrPak 7D-E2 GNSS and INS system.
+Robotic Operating System (ROS) was used to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags, which were unpacked using MATLAB tools to separate data streams.
+Prior to creating a training database, manual dictation of gravel and asphalt areas of the gathered point cloud data is necessary for automatic training data extraction.
+In order to do so, all scanning LiDAR data from a rosbag was aggregated in order to create a single point cloud map.
+Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data.
+Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan.
+Compiled point cloud maps were examined and compared to camera and satellite data to determine gravel and asphalt surface location.
+Manually defined 2D areas representing gravel and asphalt were projected unto the point cloud \textit{[Manually Define Road \& Side-of-Road]} (Section \ref{sec:training_and_verification_data_selection_process}).}
+%		
+%		{Training data was extracted from a single arc from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area (Figure \ref{fig:test_vs_train_areas}).
+Raw spatial and intensity values from the coincident arc were saved to a raw training database \textit{[Training Data Extraction]}.
+Separate training data bases were maintained for each individual channel.
+Features that describe the raw spatial and intensity data were extracted.
+Thirty percent of the training data base was randomly selected and set aside as a validation data base.
+Training data was visually examined for clustering to verify that a Random Decision Forest would be able to be trained (Section \ref{sec:Feat_Extract}).
+It was found that each class generally clustered to a certain range with some overlap, which indicates that a Random Decision Forest may be trained.
+MATLAB was used to train a Random Decision Forests for each of the three considered channels using Bayesian Optimization to tune tree depth and number of splits \textit{[RDF Training \& Verification]} (Section \ref{sec:random-decision-forest-creation}).
+Training continued for thirty iterations and the algorithm that produced the minimum upper confidence interval for out-of-bag error was selected.
+Validation of the RDF was then completed using the validation database (Section \ref{sec:random-decision-forest-verification}).}
+%		
+%		{Scanning LiDAR data from five datasets representing five drives on a stretch of Blackburn Road was classified \textit{[Classify Scanning LiDAR Data]}.
+Following the same process as aggregating point clouds, GNSS and IMU data informed the location and orientation of the arcs in each 360 degree scan.
+Three arcs from three channels with sizes determined by approximate vehicle width (~7 ft, Figure \ref{fig:area_example}) were then classified as "asphalt", "gravel", or "unknown", and the averaged coordinates of each arc was aggregated into a single, classified point cloud (Figure \ref{fig:raw_classification_results}).}
+
 		{Detection of unmarked gravel road surfaces using terrain classification requires collecting data, training an RDF, and testing the algorithm for accuracy of unmarked road detection (outline of work in Fig.
-\ref{fig:flowz_2}).
+\ref{fig:flowz_7}).
 Training data was gathered by the Ohio University Autonomous Van (Fig.
@@
 Scanning LiDAR data was consecutively classified and compared to manually defined truth areas, yielding the true-positive accuracy for gravel, asphalt, and non-road surfaces.
-Classified point cloud data was manually examined in order to manually project possible road surfaces.
-Manually projected road surface areas were compared to truth areas to determine the accuracy of asphalt road and intersecting gravel driveway detection.}
+Classified point cloud data was manually examined in order to manually project possible road surfaces using manual and automated methods.
+Projected road surface areas were compared to truth areas to determine the accuracy of asphalt road and intersecting gravel driveway detection.}
 
@@
 				\caption[Project Flow]{High level view of project work}
-				\label{fig:flowz_2}
+				\label{fig:flowz_7}
 			\end{figure}
@@
 GPS data was recorded with a Novatel PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS), and two GNSS-502 antennas by NavtechGPS.
-Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using the Robot Operating System (ROS).
+GNSS was assisted with Real Time Kinematic correction data from the Ohio Continuously Operating Reference Station (CORS) and Ohio Real Time Network (RTN) \cite{noauthor_cors_nodate}, yielding in our tests sub 2cm location accuracy level.
+IMU specifications indicate $0.8 ^{\circ}/hr$ bias instability, $0.06 ^{\circ}/\sqrt{hr}$ angular random walk, and $0.025 m/s/\sqrt{hr}$ velocity random walk, in addition to a Extended Kalman Filter \cite{noauthor_pwrpak7_nodate}.
+Because the IMU was only used to orientate the individual LiDAR scans, IMU drift was not considered a factor that would effect the classification results themselves.
+As the data collection took place within an hour period, the IMU drift was considered a negligible factor when scoring the results.
+Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using the Robot Operating System (ROS) to be subsequently examined and categorized later using MATLAB.
 Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot (Fig.
@@
 Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection; therefore, grass is labeled as \textit{unknown} in this work.
-It is acknowledged by the authors that the algorithm must classify as one of these three terrain types.
-Due to the potentially unlimited number of terrain types,
-
-				Raw data was gathered using the Robotic Operating System (ROS) to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags.
-Rosbags were unpacked using MATLAB tools to separate data streams.
-Point cloud, GPS, and IMU topics were defined, and point cloud Cartesian coordinates with corresponding intensities, longitude, latitude, altitude, roll, pitch, and yaw messages were read from the appropriate topics and imported.}
+While the number of classes is low, this is done to lower computational load.
+It was the expectation for this work that terrain that did not match the asphalt or gravel classes would be classified as unknown, and this would be later confirmed in results.}
 
@@
 				\includegraphics[width=0.75\linewidth]{figures/blackburn_road}
-				\caption[Blackburn Road Camera View]{Camera view of Blackburn Road - an unmarked asphalt road with three intercepting gravel driveways from which asphalt data was gathered.}
+				\caption[Blackburn Road Camera View]{Camera view of Blackburn Road - an unmarked asphalt road with three intersecting gravel driveways from which asphalt data was gathered.}
 				\label{fig:Blackburn_Road_View}
@@
 \ref{fig:pc_example}).
-While not as robust as more sophisticated methods of point cloud aggregation due to IMU drift, such as NDT or ICP scan matching, using this method proved to be adequate over shorter distances.}
+While not as robust as more sophisticated methods of point cloud aggregation due to IMU drift, such as NDT or ICP scan matching, using this method proved to be adequate over shorter distances given the reasons previously mentioned.}
 
@@
 \ref{fig:area_example}) from the first three channels of scanning LiDAR data centered directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.
-Later, nine arcs will be classified for the prediction of intersecting roads.
-Arc length was chosen based on approximate vehicle width (~6 degrees), and the location of each arc was fixed relative to the center line (length-wise) of the vehicle.
+Later, nine arcs (three arcs from three channels) will be classified for the prediction of intersecting roads.
+Arc length was chosen based on approximate vehicle width (set to 6 degrees), and the location of each arc was fixed relative to the center line (length-wise) of the vehicle.
 Training data was extracted from a single arc, with approximately 28 LiDAR points per arc, from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area.
+As arc is fixed to 6 degrees, the number of data points is constant, thus the only factor contributing to point density would be potential elevation changes between the vehicle and arc.
+This study found that the training data had a standard deviation was distinct between LiDAR channels and simular between terrain classes in each LiDAR channel, therefore for this work LiDAR sparsity was considered a negligible issue.
 Raw spatial and intensity values from the coincident arc were saved to a raw training database.
@@
 \ref{fig:training_data_cluster_2}).
-It was found that each class generally clustered to a certain range with some overlap, which indicates that a RDF may be trained.}
+It was found that each class generally clustered to a certain range with some overlap, which indicates that a RDF may be trained.
+}
 
@@
 				\centering
-				\includegraphics[width=0.25\linewidth]{figures/area_example}
-				\caption[Areas to Classify]{Three arcs from three channels were classified per 360 degree scan. }
+				\includegraphics[width=0.25\linewidth]{figures/nine_arcs_example}
+				\caption[Areas to Classify]{Three arcs from three channels were classified per 360 degree scan.
+2C, 3C, and 4C were used exclusively for gathering training data, while the rest were included during testing.
+\textit{Note: Vehicle not to scale.}}
 				\label{fig:area_example}
@@
 
-			{Classifying consecutive LiDAR scans was accomplished by examination of the nine arcs of interest in front and forty-five degree angles left and right from the front of the experimental apparatus (Fig.
+%		It is acknowledged by the authors that the algorithm must classify as one of these three terrain types.
+
+			{Classifying consecutive LiDAR scans was accomplished by examination of nine arcs of interest (three arcs from three separate LiDAR channels) in front and forty-five degree angles left and right from the front of the experimental apparatus (Fig.
 \ref{fig:area_example}).
@@
 
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.9\linewidth]{figures/auto_guess_v_truth}
+				\caption[Projected Guess over Time vs Truth]{Aggregated automated road projection based on classification results.}
+				\label{fig:auto_guess_v_truth}
+			\end{figure}
+
 	\section{Results}
@@
 					Asphalt Road Detection 					& 100.00 	& \% True Pos. Detection Rate \\
-					Gravel Driveway Detection 				& 93.33 	& \% True Pos. Detection Rate \\
-					Misidentified Gravel Road Detection 	& 0.60 		&  Avg. False Pos. Detection Rate per Run\\
+					Gravel Driveway Detection 				& 71.67 	& \% True Pos. Detection Rate \\
+					Misidentified Gravel Road Detection 	& 0 		&  Avg. False Pos. Detection Rate per Run\\
 					\bottomrule
@@
 Higher confidence score requirements resulted in \textit{unknown} over-classification, while too low resulted in \textit{gravel} over-classification.
-Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 74.97\% (Fig.
+Overall it was found that an average true-positive accuracy of all intersecting gravel surfaces was 74.97\% (Fig.
 \ref{fig:prepostadjust}).
@@
 \ref{fig:visual_v_auto}).
-Driveways $1$ and $2$ were detected in all five drive-bys, driveway $3$ was detected in four leading to an 91.67\% accuracy in intercepting gravel road detection.
+Driveways $1$ and $2$ were detected in all five drive-bys, driveway $3$ was detected in four leading to an 91.67\% accuracy in intersecting gravel road detection.
 It was found that in three drive-bys there was a false positive driveway detection, however the location of the false-positives were not repeated in any of the other drivebys.
@@
 
-		{It is difficult to determine if these results are comparable to similar research, as found research classifies larger LiDAR data sets.
+		{It is difficult to determine if these results are comparable to similar research, as found research classifies using distinctly different sensor configurations,  algorithm input requirements, or have different objectives.
 For example, work completed by Laible et.
 al.
-\cite{levi_3d_2012_light} used $\sim1200$ data points, compared to the $\sim28$ data points per arc in this work.
-Another example is the work completed by Wang et.
+\cite{levi_3d_2012_light} used a camera as well as LiDAR, compared to the LiDAR-only focus for this work.
+Work completed by Wang et.
 al.
 \cite{wang_road_nodate}, which found a true positive gravel terrain classification rate of 98.5\%; however, the method again used significantly more data from the LiDAR sensor compared to this work, exploiting all LiDAR channels that fall within a 1.3 meter region of interest.
-To the best of the author's knowledge, this study employed a relatively unique approach for detecting road surface area and type.}
+There is a general lack of open source LiDAR data for the express purpose of comparing road detection and terrain classification methods in an unstructured, rural environment.
+This is in stark contrast to open source data for urban environments.
+Furthermore, there is to the best of the author's knowledge any comparison between road detection methods or employment of a third-party LiDAR data of unmarked, unstructured gravel and asphalt roads.
+To the best of the author's knowledge, this study employed a novel approach for detecting road surface area and type.}
 
@@
 The described work was a method of detecting physical unmarked gravel and asphalt roads by using a terrain classification approach to predicting road surface area.
-The impact of this work is that autonomous vehicles using the developed method may be able to detect unmarked gravel and asphalt road surfaces, allowing autonomous operations on previously undetected rural roads.}
+The impact of this work is that autonomous vehicles using the developed method may be able to detect unmarked road surfaces and differentiate between gravel and asphalt, allowing autonomous operations on previously undetected rural roads.}
 

### 2023-08-16 Huston_Paper.tex
Small Updates

@@
 Classification is determined by the majority results as opposed to the mean \cite{breiman_random_2001,ho_random_1995}.
-While RDFs were used in terrain classification, they have not been, to the author's best knowledge, been employed in predicting road surface areas based on terrain classification results while also only using extremely small input data sets typical of real-time mobile scanning.
+While RDFs were used in terrain classification in multiple instances in found literature, to the author's best knowledge they have not been employed in predicting road surface areas and type while also only using extremely small input data sets typical of real-time mobile scanning.
 Bayesian Optimization may be used to tune the hyper parameters of a machine algorithm \cite{noauthor_bayesian_nodate, snoek_practical_2012}.
@@
 As arc is fixed to 6 degrees, the number of data points is constant, thus the only factor contributing to point density would be potential elevation changes between the vehicle and arc.
-This study found that the training data had a standard deviation was distinct between LiDAR channels and simular between terrain classes in each LiDAR channel, therefore for this work LiDAR sparsity was considered a negligible issue.
+This study found that the training data sparsity had a standard deviation that was distinct between LiDAR channels and similar between terrain classes in each LiDAR channel, therefore for this work LiDAR sparsity was considered a negligible issue.
 Raw spatial and intensity values from the coincident arc were saved to a raw training database.
@@
 \ref{fig:visual_v_auto}).
+While a simple model, this will serve as a proof-of-concept for future development.
 Results were compiled into a single classified map and compared to truth areas, then scored based off the number of true/false positive road detection.
@@
 This is in stark contrast to open source data for urban environments.
-Furthermore, there is to the best of the author's knowledge any comparison between road detection methods or employment of a third-party LiDAR data of unmarked, unstructured gravel and asphalt roads.
+Furthermore, to the best of the author's knowledge there is no comparison between road detection methods using third-party LiDAR data of unmarked, unstructured gravel and asphalt roads in found literature.
 To the best of the author's knowledge, this study employed a novel approach for detecting road surface area and type.}

### 2023-08-17 Huston_Paper.tex
081620203

@@
 
-			{Manually defined areas were then overlayed onto the point cloud, representing training areas (Fig.
+			{Arcs of 6 degrees from the first three channels of scanning LiDAR data centered directly in front of the vehicle (Fig.
+\ref{fig:area_example}) were extracted and exported to a database if coincident with the manually defined areas (Fig.
 \ref{fig:test_vs_train_areas}).
-Six degree arcs (Fig.
-\ref{fig:area_example}) from the first three channels of scanning LiDAR data centered directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.
 Later, nine arcs (three arcs from three channels) will be classified for the prediction of intersecting roads.
@@
 Asphalt data was gathered Blackburn Road.
-Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training; however, for this study the algorithm that produced the final results used multiple thousands of data sets.
-Results were then compared to algorithms trained with smaller data sets.
+Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training; however, for this study the algorithm that produced the final results used multiple thousands of data sets per class.
+Results were compared to algorithms trained with smaller data sets, OH YEAH WHERE'S THE PROOF HONEY?.
 Training data was generated using MATLAB to extract LiDAR data from rosbags, and RDF classification algorithms were generated.

### 2023-08-17 Huston_Paper.tex
Added resources

@@
 Because the IMU was only used to orientate the individual LiDAR scans, IMU drift was not considered a factor that would effect the classification results themselves.
-As the data collection took place within an hour period, the IMU drift was considered a negligible factor when scoring the results.
 Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using the Robot Operating System (ROS) to be subsequently examined and categorized later using MATLAB.
@@
 			{Arcs of 6 degrees from the first three channels of scanning LiDAR data centered directly in front of the vehicle (Fig.
-\ref{fig:area_example}) were extracted and exported to a database if coincident with the manually defined areas (Fig.
+\ref{fig:nine_arcs_example}) were extracted and exported to a database if coincident with the manually defined areas (Fig.
 \ref{fig:test_vs_train_areas}).
@@
 \textit{Note: Vehicle not to scale.}}
-				\label{fig:area_example}
+				\label{fig:nine_arcs_example}
 			\end{figure}
@@
 			{Classifying consecutive LiDAR scans was accomplished by examination of nine arcs of interest (three arcs from three separate LiDAR channels) in front and forty-five degree angles left and right from the front of the experimental apparatus (Fig.
-\ref{fig:area_example}).
+\ref{fig:nine_arcs_example}).
 Features from each arc were extracted and the arc was classified (Fig.
@@
 Accuracy scores was calculated based on the distribution of classified points with each area, allowing evaluation of unmarked road detection performance.
+Due to the arc locations (Fig.
+\ref{fig:nine_arcs_example}), IMU drift was considered a negligible factor for this work as IMU drift would need to be extreme in order to impact these results, in which case it would be easily found visually.
 Classified point clouds were visually inspected to determine if the manual detection of a road surface may be accomplished.

### 2023-08-18 Huston_Paper.tex
Added diff

@@
 				\includegraphics[width=0.75\linewidth]{figures/combined_pcd_example}
-				\caption[Aggregated Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.}
+				\caption[Aggregated Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.
+Intensity informs point color in this example image.}
 				\label{fig:pc_example}
@@
 				\includegraphics[width=0.75\linewidth]{figures/test_vs_train_areas_hatch}
-				\caption[Training vs Testing Areas]{Training areas were kept separate from the testing areas.}
+				\caption[Training vs Testing Areas]{Training areas were kept separate from the testing areas, shown is a close-up, top-down view of the road of interest. }
 				\label{fig:test_vs_train_areas}
@@
 
-			\begin{figure}[H]
-				\centering
-				\includegraphics[width=0.9\linewidth]{figures/auto_guess_v_truth}
-				\caption[Projected Guess over Time vs Truth]{Aggregated automated road projection based on classification results.}
-				\label{fig:auto_guess_v_truth}
-			\end{figure}
+%			\begin{figure}[H]
+%				\centering
+%				\includegraphics[width=0.9\linewidth]{figures/auto_guess_v_truth}
+%				\caption[Projected Guess over Time vs Truth]{Aggregated automated road projection based on classification results.}
+%				\label{fig:auto_guess_v_truth}
+%			\end{figure}
 
@@
 Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training; however, for this study the algorithm that produced the final results used multiple thousands of data sets per class.
-Results were compared to algorithms trained with smaller data sets, OH YEAH WHERE'S THE PROOF HONEY?.
 Training data was generated using MATLAB to extract LiDAR data from rosbags, and RDF classification algorithms were generated.
@@
 It is acknowledged by the authors that the classification algorithm must yield a result that most closely matches one of the trained classes, which led to lower accuracy scores in these circumstances.
+Results were compared to algorithms trained with smaller data sets in order to determine if the training data was sufficiently large for this specific scenario.
+Using 25\% and 50\% of total available training data yielded poor results compared to shown, however using 75\% of all available training data yielded results comparable to the shown results, indicating that the training data set is sufficient for this specific scenario.
 Overall discovered accuracy levels may be sufficient for the purpose of identifying asphalt and gravel roads by considering trends between classified arcs.}

### 2023-12-15 Huston_Paper.tex
round 2 FIGHT

--- 2023-08-17 Huston_Paper.tex
@@
 Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}.
-Real-time detection of road surfaces limits the available computational time required for low-latency vehicle trajectory updates, prohibiting the collection of larger point LiDAR point clouds.
-Implementing a Random Decision Forests (RDF) terrain classification model using arcs from a single 360 degree scan of LiDAR data to exploit surface properties was studied to determine the method's feasibility in road surface detection.
-While an established method for terrain classification, to the author's best knowledge RDFs in found literature were not employed for the detection of road surface area and type using a small data input within the context of an unstructured environment, making the approach used in this work is a novel one.
+Implementing a Random Decision Forests (RDF) terrain classification model using subsets of a single 360 degree scan of LiDAR data to exploit surface properties was studied to determine the method's feasibility in road surface detection, without using a-priori knowledge of the road, additional sensors, nor memory of previous classifications.
+While an established method for terrain classification, to the author's best knowledge RDFs in found literature were not employed for the detection of road surface area and type using a small data input and the other aforementioned limitations within the context of an unstructured environment, making the approach used in this work is a novel one.
 Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.}
+
+%	Real-time detection of road surfaces limits the available computational time required for low-latency vehicle trajectory updates, prohibiting the collection of larger point LiDAR point clouds.
 
@@
 Histograms and averages for image hue, saturation, and color value, along with LiDAR intensity were used as points of classification for different terrain types.
-Although a 96\% true positive rate was reported for a rocky surface, the overall process was computationally expensive with 1.8 seconds per image.}
+Although a 96\% true positive rate was reported for a rocky surface, the overall process was computationally expensive with 1.8 seconds per image.
+Fernandez-Delgado et al found that RDFs out-performed other machine learning algorithms, including NNs, when classifying 121 data sets taken from the UCI data base \cite{fernandez_delgado_we_nodate}.
+Therefore while both Neural Networks and RDFs have been used to classify scanning LiDAR data, a RDF classifier was chosen.
+Furthermore, RDFs are less complex to train, able to be trained with smaller data sets, and are generally easier to interpret \cite{nawar_comparison_2017, rosbach_neural_nodate, rosbach_neural_nodate, ahmad_trees_2017}.}
 
@@
 Classification is determined by the majority results as opposed to the mean \cite{breiman_random_2001,ho_random_1995}.
-While RDFs were used in terrain classification in multiple instances in found literature, to the author's best knowledge they have not been employed in predicting road surface areas and type while also only using extremely small input data sets typical of real-time mobile scanning.
+While RDFs were used in terrain classification in multiple instances in found literature, to the author's best knowledge they have not been employed in predicting road surface areas and type while also only using extremely small input data sets of mobile scanning LiDAR.
 Bayesian Optimization may be used to tune the hyper parameters of a machine algorithm \cite{noauthor_bayesian_nodate, snoek_practical_2012}.
-Bayesian Optimization initializes a surrogate function to represent the objective function (in this case minimum total classification error) with randomly selected hyper-parameters.
-Additional points on the objective function are chosen using an acquisition function to determine the utility of hyper-parameters.}
+}
 
-	{Current methods for road detection may be insufficient for real-time detection of gravel roads due to ambient conditions, lack of standardized features, or rely on post-processing large data sets, while terrain classification of scanning LiDAR data may overcome these difficulties.}
+	{Current methods for road detection may be insufficient for detection of gravel roads due to ambient conditions, lack of standardized features, or rely on post-processing large data sets, while terrain classification of scanning LiDAR data may overcome these difficulties.}
 
@@
 				\caption[Project Flow]{High level view of project work}
-				\label{fig:flowz_7}
+				\label{fig:flowz_8}
 			\end{figure}
@@
 
-			{Arcs of 6 degrees from the first three channels of scanning LiDAR data centered directly in front of the vehicle (Fig.
-\ref{fig:nine_arcs_example}) were extracted and exported to a database if coincident with the manually defined areas (Fig.
-\ref{fig:test_vs_train_areas}).
-Later, nine arcs (three arcs from three channels) will be classified for the prediction of intersecting roads.
+			{Discretizing of scanning LiDAR data is required designating fixed sub-sets of data to be considered for training data extraction and future classification.
+For this work, arcs of six degrees centered in front of the vehicle were designated, each arc only containing points from one of the thirty two scanning LiDAR channels.
+For this work, only the closest three channels were considered in order to minimize the affect of LiDAR sparsity  \ref{fig:nine_arcs_example}.
+Later, nine arcs (three arcs from the three channels) will be classified for the prediction of intersecting roads \ref{fig:nine_arcs_example}.
 Arc length was chosen based on approximate vehicle width (set to 6 degrees), and the location of each arc was fixed relative to the center line (length-wise) of the vehicle.
@@
 2C, 3C, and 4C were used exclusively for gathering training data, while the rest were included during testing.
-\textit{Note: Vehicle not to scale.}}
+}
 				\label{fig:nine_arcs_example}
@@
 			\centering
-			\includegraphics[width=0.75\linewidth]{figures/training_data_cluster_3}
+			\includegraphics[width=0.75\linewidth]{figures/training_data_cluster_5}
 			\caption[Example Clustering]{Example of training data comparison between two features.}
@@
 			{LiDAR returns spatial data represented by Cartesian coordinates with units in meters and remission data represented by a dimensionless ratio of minimum to maximum detected brightness.
+No pre-processing of LiDAR data as extracted from the rosbags was required, however the Cartesian coordinates were converted to Polar by using the Pythagorean Theorem to determine range ($sqrt(x^{2} + y^{2} + z^{2})$) and simple trigonometry to determine the azimuth ($atan2(y,x)$).
 Feature extraction was completed by performing mathematical functions on the gathered spatial training data, yielding the Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial data array.
@@
 As remission is a ratio of minimum to maximum detectable reflected intensity, there was no need to render the features non-dimensionalized.
+Feature selection closely mimics those suggested in literature \cite{noauthor_pdf_nodate, reymann_improving_2015}.
+Future work may be completed to introduce additional features.
 Thirty percent of the training data was randomly selected and reserved to create a validation data set.}
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/c2_min_class_error}
+				\includegraphics[width=0.75\linewidth]{figures/c2_min_class_error_2}
 				\caption[RDF Training Classification Error]{Classification error during Bayesian Optimization tuning.
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/train_vs_valid_overfit_test3}
+				\includegraphics[width=0.75\linewidth]{figures/train_vs_valid_overfit_test_2}
 				\caption[Training vs Validation Error]{Hyper-parameter tuning may lead to over-fitting.
@@
 				\centering
-				\includegraphics[width=0.55\linewidth]{figures/chan_2c_conf_VALIDATION_mat222}
+				\includegraphics[width=0.75\linewidth]{figures/chan_2c_conf_VALIDATION_mat222}
 				\caption[Validation Error]{Confusion matrix with the validation training data set.}
@@
 Averaged x, y, and z coordinates were taken of each arc, and projected unto a manually defined truth area map (Fig.
-\ref{fig:visual_v_auto}).
+\ref{fig:rm_db_6_overlap_2}).
 Accuracy scores was calculated based on the distribution of classified points with each area, allowing evaluation of unmarked road detection performance.
 Due to the arc locations (Fig.
-\ref{fig:nine_arcs_example}), IMU drift was considered a negligible factor for this work as IMU drift would need to be extreme in order to impact these results, in which case it would be easily found visually.
+\ref{fig:nine_arcs_example}), IMU drift was considered a negligible factor for this work as IMU drift would need to be extreme in order to impact these results, in which case the results would be easily recognized as having experienced significant drift.
 Classified point clouds were visually inspected to determine if the manual detection of a road surface may be accomplished.
@@
 Class percentages, standard deviation of height, standard deviation of average [x,y] distance to the LiDAR point of origin, and distance trends between classified arcs among other features were exploited to determine if the area best describes a gravel, asphalt, or unknown surface (Fig.
-\ref{fig:visual_v_auto}).
+\ref{fig:auto_area_guess}).
 While a simple model, this will serve as a proof-of-concept for future development.
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/db1range.png}
+				\includegraphics[width=0.75\linewidth]{figures/db1range_3.png}
 				\caption[Classified Point Cloud]{Classified Point Cloud.}
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/visual_v_auto}
+				\includegraphics[width=0.75\linewidth]{figures/rm_db_6_overlap_2}
 				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed based on trends between channels and compared to actual road-surface areas (left).
 Autonomous projection of road surface areas following developed methods (right).}
-				\label{fig:visual_v_auto}
+				\label{fig:rm_db_6_overlap_2}
 			\end{figure}
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/auto_guess_v_truth}
+				\includegraphics[width=0.75\linewidth]{figures/auto_area_guess_4}
+				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed based on trends between channels and compared to actual road-surface areas (left).
+Autonomous projection of road surface areas following developed methods (right).}
+				\label{fig:auto_area_guess}
+			\end{figure}
+
+			\begin{figure}[H]
+				\centering
+				\includegraphics[width=0.75\linewidth]{figures/auto_guess_v_truth_2}
 				\caption[Projected Guess over Time vs Truth]{Aggregated automated road projection based on classification results.}
@@
 			\centering
-			\includegraphics[width=0.75\linewidth]{figures/dirt_v_gravel2}
+			\includegraphics[width=0.75\linewidth]{figures/dirt_v_gravel_3}
 			\caption[Re-grassed Dirt vs Gravel]{Side-of-road detection accuracy was diminished due to overlapping features between dirt and gravel. }
@@
 			\centering
-			\includegraphics[width=0.75\linewidth]{figures/range_actual_rm_db_4_area_score}
+			\includegraphics[width=0.75\linewidth]{figures/db_4_area_scores_2}
 			\caption[Area Scores]{Per-area scores were calculated, indicating algorithm unmarked road detection accuracy.
@@
 Two out of three positive gravel identifications must be made on each channel, be evenly spaced, and be relatively co-planar to other road surface guesses in order for a positive guess for an asphalt or gravel surface to be made (Fig.
-\ref{fig:visual_v_auto}).
+\ref{fig:auto_area_guess}).
 Driveways $1$ and $2$ were detected in all five drive-bys, driveway $3$ was detected in four leading to an 91.67\% accuracy in intersecting gravel road detection.
@@
 		{Classification was completed using Ubuntu 18.04 and an AMD Ryzen 9 5900X 12 core 24 thread CPU.
-Rate of classification for all nine areas of interest in a single scan was found to be an average of 35.19 ms using a single core (Fig.
-\ref{fig:per_arc_classification_time}) or 316.68 ms per 360 degree scan.
+Rate of classification for all nine areas of interest in a single scan was found to be an average of 35.19 ms using a single core or 316.68 ms per 360 degree scan.
 When utilizing all 24 cores during classification a rate of 13.19 ms per 360 scan.
 Extrapolating from the distance traveled during a drive-by, a hypothetical 71.39 feet per second or 48.68 miles per hour may be possible.
-Further optimization may increase this speed, however implementing real-time classification was not a focus of this work.}
+Future work may include a path of development for testing real-time performance.
+}
 
-		\begin{figure}[H]
-			\centering
-			\includegraphics[width=0.75\linewidth]{figures/per_arc_classification_time}
-			\caption[Per-Arc Time]{Per-arc classification time}
-			\label{fig:per_arc_classification_time}
-		\end{figure}
+%		Further optimization may increase this speed, however implementing real-time classification was not a focus of this work.
+%		\begin{figure}[H]
+%			\centering
+%			\includegraphics[width=0.75\linewidth]{figures/per_arc_classification_time}
+%			\caption[Per-Arc Time]{Per-arc classification time}
+%			\label{fig:per_arc_classification_time}
+%		\end{figure}
 

### 2023-12-18 Huston_Paper.tex
post review 1

@@
 LiDAR point cloud processing systems depend upon distinct geometric features, most commonly curbs \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_nodate,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja_road_nodate}, however as curbs are not installed on rural gravel roads \cite{skorseth_gravel_nodate} this method breaks down.
-Normal Distribution Transform (NDT) scan matching analytically compares two point cloud data sets to track current position, however this model relies on distinct geometric features, and suffers when distinguishing between terrain types making it insufficient for determining road boundaries \cite{biber_normal_2003}.
+Normal Distribution Transform (NDT) scan matching analytically compares two point cloud data sets to track current position, however this model relies on distinct geometric features, and suffers when distinguishing between terrain types making it insufficient for determining road boundaries and type \cite{biber_normal_2003}.
 Using 2D projection of planes unto a 3D point cloud of a road surface is an alternative approach, but likewise requires curbs or painted lines for road boundary definition \cite{fernandes_road_2014, borkar_robust_2009-1, guo_lane_2015}.
-Instead, using LiDAR may overcome difficulties in achieving detection of gravel roads by using terrain classification.}
+Instead, using LiDAR may overcome difficulties in achieving detection of gravel roads by using terrain classification to predict road surface location and type.}
 
@@
 Implementing a Random Decision Forests (RDF) terrain classification model using subsets of a single 360 degree scan of LiDAR data to exploit surface properties was studied to determine the method's feasibility in road surface detection, without using a-priori knowledge of the road, additional sensors, nor memory of previous classifications.
-While an established method for terrain classification, to the author's best knowledge RDFs in found literature were not employed for the detection of road surface area and type using a small data input and the other aforementioned limitations within the context of an unstructured environment, making the approach used in this work is a novel one.
-Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.}
+While an established method for terrain classification, to the author's best knowledge RDFs in found literature were not employed for the detection of road surface area and type using small portions of a single LiDAR scan and the other aforementioned limitations within the context of an unstructured environment, making the approach used in this work is a novel one.
+Evaluation of the detection model will rely principally upon accuracy of projected road location and surface type, as a moving vehicle must obtain reliable information of the road surface location and type to inform trajectory updates, or else risk suffering an accident.}
 
@@
 
-	{Real time processing of rural gravel roads cannot rely on normalized, consistent features found on urban roads such as curbs, as rural roads lack these features \cite{skorseth_gravel_nodate}.
+	{Processing rural gravel roads cannot rely on normalized, consistent features found on urban roads such as curbs, as rural roads lack these features \cite{skorseth_gravel_nodate}.
 As such, many of the proposed models are rendered unserviceable, as they rely on curbs or painted lines to offer distinctions in data sets \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_nodate,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja_road_nodate}.}
 
-	{Further limitations are imposed when considering data set size and processing speed.
+%	{Further limitations are imposed when considering data set size and processing speed.
 Real time trajectory updates to the vehicle in motion requires low latency road surface analysis, denying the use of methods that require post-processing due to high computational loads.
@@
 However a computational time of 24 minutes was required to process a 156 meter stretch of road, rendering real time detection impossible.}
+%	
+	{Further limitations are imposed when considering data set size, as a vehicle must generate trajectory solutions based on what the sensors currently see at the vehicle's current location.
+Post-processing aggregated data for the purpose of road surface detection presumes that a vehicle is able to navigate the area without the use of any sensors.
+Proposed methods \cite{yadav_extraction_2017,yadav_road_2018,yadav_rural_2018,yadav_pole-shaped_2015,miyazaki_line-based_2014,yang_semi-automated_2013,liu_new_2013,qiu_fast_2016} do not indicate the minimum number of points required for road surface detection, however those studies post-processed data sets with many millions of points.}
 
-	{Classification of the surrounding environment is an integral part to the proposed work, as differentiation of gravel road surfaces and surrounding terrain is necessary to determine road boundaries.
+	{Classification of the surrounding environment is an integral part to the proposed work, as differentiation of gravel road surfaces and surrounding terrain is necessary to determine road boundaries and type.
 Terrain classification is the analysis of the ground surface in order to specify the ground surface type \cite{laible_3d_2012,laible_terrain_2013,laible_map_building,rasmussen_combining_2002,reymann_improving_2015,walas_terrain_2014,wietrzykowski_boosting_2014,wang_road_nodate}.
@@
 Comparison with neighboring cells increases classification type accuracy, as adjacent cell are likely to have the same terrain type \cite{haselich_terrain_2011,zhao_fusion_2014}.
-It was found that CRF was better suited for classification.}
+It was found that CRF was better suited for classification, however this work did not use the method to study road projection and used cameras which may not be reliable for terrain classification \cite{levi_3d_2012_light}.}
 
-	{Random Decision Forest (RDF) classification showed good results for identification of analysis of LiDAR and camera data \cite{breiman_random_2001}, and was used to predict cell terrain into one of five types, including gravel and grass.
-It was found that characterizing the LiDAR intensity as Gaussian distributed noise yielded poor results with only a 49.5\% true positive rate, however a low-resolution LiDAR was used \cite{rauscher_comparison_2016}.
+	{Random Decision Forest (RDF) classification showed results for identification of analysis of LiDAR and camera data \cite{breiman_random_2001}, and was used to predict cell terrain into one of five types, including gravel and grass.
+It was found that characterizing the LiDAR intensity as Gaussian distributed noise yielded poor results with only a 49.5\% true positive rate, however a low-resolution LiDAR was used \cite{rauscher_comparison_2016} and this method was not employed in road identification.
 Higher resolution LiDAR that produces a high density point cloud, such as the Velodyne VLP-32 that will be used in the proposed work \cite{vlp_32c}, may allow for better terrain classification.
-While capable of real-time classification of terrain type, this method was not employed in road identification and used a low resolution LiDAR system.
 Edmond et al.
-proposed a method to detect terrain type through implementation of vehicle vibration, however this method is limited to what the vehicle is currently driving over and thus cannot detect intersecting gravel roads \cite{dupont_online_2008}.
+proposed a method to detect terrain type through implementation of vehicle vibration, however this method is limited to what the vehicle is currently driving over and thus cannot detect future unmarked road surfaces \cite{dupont_online_2008}.
 Filitchkin et al.
@@
 Road classification with LiDAR spacial and remission features was completed with Neural Network classifier, Naive Bayes classifier, and Support Vector Machines (SVM) \cite{wang_road_nodate,wang_two-stage_2018}.
-Gravel surfaces were successfully identified with a 98.5\% true positive identification rate; however, the method was only focused on classifying road surface type directly behind the vehicle, and was not tested for predicting road surface areas.
-MATLAB's Neural Network tools were used in autonomous road identification using LiDAR and Camera data \cite{rasmussen_combining_2002}; however, cameras may not be relied upon for ambient conditions may affect classification accuracy \cite{levi_3d_2012_light}, furthermore computational time was not mentioned.
-Terrain classification was completed using Support Vector Machine (SVM), a classification algorithm where a line is drawn between two different categories to differentiate between them \cite{breiman_random_2001}.
+Gravel surfaces were successfully identified with a 98.5\% true positive identification rate; however, the method was only focused on classifying road surface type directly behind the vehicle, and was not tested for informing vehicle trajectory based on forward-projected road surface areas.
+MATLAB's Neural Network tools were used in autonomous road identification using LiDAR and Camera data \cite{rasmussen_combining_2002}; however, cameras may not be relied upon for ambient conditions may affect classification accuracy \cite{levi_3d_2012_light}.
+Terrain classification was completed using Support Vector Machine (SVM), a classification algorithm where a line is drawn between two different categories to differentiate between them \cite{wietrzykowski_boosting_2014}.
 Histograms and averages for image hue, saturation, and color value, along with LiDAR intensity were used as points of classification for different terrain types.
-Although a 96\% true positive rate was reported for a rocky surface, the overall process was computationally expensive with 1.8 seconds per image.
+Although a 96\% true positive rate was reported for a rocky surface, the process was not tested for road detection.
+Lei et al suggest a method using an SVM for the detection of unstructured dirt and gravel roads using a 2D LiDAR and a camera \cite{lei_detection_2021}, however the method does not use a 3D scanning LiDAR and cameras may be insufficient for reliable road detection \cite{levi_3d_2012_light}.
 Fernandez-Delgado et al found that RDFs out-performed other machine learning algorithms, including NNs, when classifying 121 data sets taken from the UCI data base \cite{fernandez_delgado_we_nodate}.
@@
 Classification is determined by the majority results as opposed to the mean \cite{breiman_random_2001,ho_random_1995}.
-While RDFs were used in terrain classification in multiple instances in found literature, to the author's best knowledge they have not been employed in predicting road surface areas and type while also only using extremely small input data sets of mobile scanning LiDAR.
+While RDFs were used in terrain classification in multiple instances in found literature, to the author's best knowledge they have not been employed in predicting road surface areas and type while also only using small portions of a single LiDAR scan.
 Bayesian Optimization may be used to tune the hyper parameters of a machine algorithm \cite{noauthor_bayesian_nodate, snoek_practical_2012}.
@@
 
-	{Current methods for road detection may be insufficient for detection of gravel roads due to ambient conditions, lack of standardized features, or rely on post-processing large data sets, while terrain classification of scanning LiDAR data may overcome these difficulties.}
+	{Current methods for road detection may be insufficient for detection of gravel roads due to ambient conditions, lack of standardized features, or rely on post-processing large data sets.
+Developing a terrain classification algorithm to analyze scanning LiDAR data in order to project road surface area and type may overcome these difficulties.}
 
@@
 		{Detection of unmarked gravel road surfaces using terrain classification requires collecting data, training an RDF, and testing the algorithm for accuracy of unmarked road detection (outline of work in Fig.
-\ref{fig:flowz_7}).
+\ref{fig:flowz_8}).
 Training data was gathered by the Ohio University Autonomous Van (Fig.
@@
 MATLAB was used to extract and parse LiDAR and GPS data.
-Scanning LiDAR data was aggregated into a large point cloud by using GNSS and IMU data to determine each LiDAR scan's position and orientation in a dead-reckoning process.
+Scanning LiDAR data was aggregated into a large point cloud by using GNSS and IMU data to determine each LiDAR scan's position and orientation.
 Aggregated point cloud data was manually examined and compared to satellite and camera images to determine and project road boundaries.
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/flowz_7}
+				\includegraphics[width=0.9\linewidth]{figures/flowz_8}
 				\caption[Project Flow]{High level view of project work}
@@
 			{Required physical data was gathered by the Ohio University Autonomous Van (Fig.
-\ref{fig:Experimental_Apperatus})  on a sunny, cloudless day with no road obstacles or debris at a speed below 15 miles per hour.
+\ref{fig:Experimental_Apperatus}) on a single sunny, cloudless day with no road obstacles or debris at a speed below 15 miles per hour.
+Data collection was at mid-day, which minimized the risk of shadows from the road-side trees interfering with the data collection.
+Vehicle speed was limited to maximize the amount of data collected and to minimize any effect of IMU drift.
+It was assumed that vehicle speed did not affect the gathered data and is considered outside the scope of this work.
 It is acknowledged that weather or non-ideal road conditions may affect results and is an area of future study.
@@
 IMU specifications indicate $0.8 ^{\circ}/hr$ bias instability, $0.06 ^{\circ}/\sqrt{hr}$ angular random walk, and $0.025 m/s/\sqrt{hr}$ velocity random walk, in addition to a Extended Kalman Filter \cite{noauthor_pwrpak7_nodate}.
-Because the IMU was only used to orientate the individual LiDAR scans, IMU drift was not considered a factor that would effect the classification results themselves.
+Because the IMU was only used to orientate the individual LiDAR scans, IMU drift was not considered a factor that would affect the classification results themselves.
+It is assumed that the GNSS was not affected by the noise generated by the other electronic equipment as the antennas are located outside the vehicle, and is considered outside the scope of this work.
 Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using the Robot Operating System (ROS) to be subsequently examined and categorized later using MATLAB.
+Training data collection location was chosen for the clear distinction between gravel and paved surfaces, thus minimizing the risk of class confusion during the training process.
 Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot (Fig.
-\ref{fig:gravel_training_lot}) near Athens Ohio.
+\ref{fig:gravel_training_lot}) near Athens Ohio, which was freshly re-surfaced and re-graded.
+While there are a variety of gravel sizes, differentiating between multiple gravel sizes is outside the scope of this work.
+Furthermore, gravel roads are specified to have a consistent type and size \cite{skorseth_gravel_nodate}, therefore it is assumed that a single source for gravel data will be sufficient for demonstrating the ability of this work's method for gravel road detection.
+Two additional driveways intersected the main asphalt road, and while they had the same gravel type they differed from the training lot.
+Condition of one was severely segregated, however the other was in excellent condition.
+Road degradation and debris were not considered and were outside the scope of this work.
 Scanning LiDAR data of an asphalt surface was gathered from Blackburn Road (Fig.
 \ref{fig:Blackburn_Road_View}) which provides access to the training gravel lot.
+Due to the aforementioned re-surfacing and re-grading of the gravel lot, some gravel was in front of the lot entrance on the road.
+This data was excluded from the training process however was included in the testing phase.
+It was expected that there would be class confusion between asphalt and gravel in this specific area, which would be the case.
 Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot.
-Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection; therefore, grass is labeled as \textit{unknown} in this work.
-While the number of classes is low, this is done to lower computational load.
-It was the expectation for this work that terrain that did not match the asphalt or gravel classes would be classified as unknown, and this would be later confirmed in results.}
+Although this work does not evaluate the detection accuracy of grass specifically, differentiation between a road and non-road surface is necessary for evaluating road-surface detection; therefore, grass is labeled as \textit{unknown} in this work.}
+
+%				 While the number of classes is low, this is done to lower computational load.
+It was the expectation for this work that terrain that did not match the asphalt or gravel classes would be classified as unknown, and this would be later confirmed in results.
 
@@
 
-			{Discretizing of scanning LiDAR data is required designating fixed sub-sets of data to be considered for training data extraction and future classification.
+			{Discretization of scanning LiDAR data is required for training data extraction and future classification.
 For this work, arcs of six degrees centered in front of the vehicle were designated, each arc only containing points from one of the thirty two scanning LiDAR channels.
-For this work, only the closest three channels were considered in order to minimize the affect of LiDAR sparsity  \ref{fig:nine_arcs_example}.
-Later, nine arcs (three arcs from the three channels) will be classified for the prediction of intersecting roads \ref{fig:nine_arcs_example}.
+For this work, only the closest three channels were considered in order to minimize the affect of LiDAR sparsity (Fig \ref{fig:nine_arcs_example}).
+Later, nine arcs (three arcs from the three channels) will be classified for the prediction of intersecting roads (Fig \ref{fig:nine_arcs_example}).
 Arc length was chosen based on approximate vehicle width (set to 6 degrees), and the location of each arc was fixed relative to the center line (length-wise) of the vehicle.
@@
 Future work may be completed to introduce additional features.
-Thirty percent of the training data was randomly selected and reserved to create a validation data set.}
+Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training, however after thirty percent of the training data was randomly selected and reserved to create a validation data set, approximately two thousand feature sets per class were used to train.}
 
@@
 			{MATLAB's Classification Learner Application was used to generate an RDF for each considered scanning LiDAR channel.
+RDFs were found in literature to out-perform NNs for classification purposes \cite{fernandez_delgado_we_nodate} while also being generally less complex to train, able to be trained with smaller data sets, and easier to interpret \cite{nawar_comparison_2017, rosbach_neural_nodate, rosbach_neural_nodate, ahmad_trees_2017}.
 Decision tree depth, number of learners, and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.
+One advantage of RDFs is the relatively low number of tunable hyper-parameters, leading to a more rapid development \cite{nawar_comparison_2017, rosbach_neural_nodate, rosbach_neural_nodate, ahmad_trees_2017}.
+Listed are the ones available using MATLAB's Classification Learner Application that were able to be tuned using the Bayesian Optimization function.
 Bayesian Optimization automates the manual hyper-parameter tuning process by creating a surrogate model representing the objective function, in this case the upper confidence interval of the classification error objective model (Fig.
@@
 				\includegraphics[width=0.75\linewidth]{figures/chan_2c_conf_VALIDATION_mat222}
-				\caption[Validation Error]{Confusion matrix with the validation training data set.}
+				\caption[Validation Error]{Confusion matrix with the validation data set.}
 				\label{fig:vali_err_conf_mat}
@@
 If consecutive points from each channel in any direction had two or more matching gravel or asphalt classification and was evenly spaced, a gravel or road surface area was projected.
-Guessed road surface areas were then compared to actual areas to determine the method's ability at unmarked gravel road discovery, indicating the method's efficiency at detecting intersecting gravel roads.
+Guessed road surface areas were then compared to actual areas to determine the method's ability at detecting unmarked asphalt and gravel roads.
 Automated detection of unmarked road was explored in order to determine if visual detection results may be repeated.
@@
 Asphalt data was gathered Blackburn Road.
-Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training; however, for this study the algorithm that produced the final results used multiple thousands of data sets per class.
+Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training; however, for this study the algorithm that produced the final results used two thousand points of reference per class.
 Training data was generated using MATLAB to extract LiDAR data from rosbags, and RDF classification algorithms were generated.
@@
 Translation and rotation was applied to the averaged x, y, and z coordinates of each arc's terrain classification results.
-Consecutive scanning LiDAR data was projected on manually classified truth areas representing road road surfaces (Figure \ref{fig:prepostadjust}) and percentages of terrain types per area were calculated.}
+Consecutive scanning LiDAR data was projected on manually classified truth areas representing road surfaces (Figure \ref{fig:prepostadjust}) and percentages of terrain types per area were calculated.}
 
@@
 Due to the side of the road being re-grassed near a portion of the asphalt road, the terrain classification algorithm struggled to differentiate the dirt from gravel, leading to some misclassification of the side-of-road areas.
-Examination of the feature distribution shows a high amount of overlap (Fig.
-\ref{fig:dirt_v_gravel}).
+Examination of the feature distribution shows a high amount of overlap, Fig.
+\ref{fig:dirt_v_gravel} shows an example of this however all features suffered from class confusion.
 Exploitation of confidence scores to adjust results by requiring a very high confidence for positive gravel surface classification was explored.
 All gravel surfaces that did not meet a 80\% confidence score was re-labeled as \textit{unknown}, however these adjustment had either a negligible or detrimental impact on the final results.
-Higher confidence score requirements resulted in \textit{unknown} over-classification, while too low resulted in \textit{gravel} over-classification.
+Higher confidence score requirements resulted in \textit{unknown} over-classification, while too low resulted in \textit{gravel} over-classification (Fig.
+\ref{fig:precision_recall_curve_averaged} shows a typical precision-recall curve.).
 Overall it was found that an average true-positive accuracy of all intersecting gravel surfaces was 74.97\% (Fig.
@@
 			\centering
+			\includegraphics[width=0.75\linewidth]{figures/precision_recall_curve_averaged}
+			\caption[PR Curve]{Precision-Recall curve of the averaged \textit{gravel} vs \textit{unknown} classification results from all test runs. }
+			\label{fig:precision_recall_curve_averaged}
+		\end{figure}
+
+		\begin{figure}[H]
+			\centering
 			\includegraphics[width=0.75\linewidth]{figures/db_4_area_scores_2}
@@
 Note that the asphalt area in front of the gravel driveway has significant \textit{gravel} classification due to construction.
-These results are included with the overall asphalt classification score}
+These results are included with the overall asphalt classification score.}
 			\label{fig:prepostadjust}
@@
 Future work may include a path of development for testing real-time performance.
+Future work may include a real-time implementation of the described methods into the vehicle platform, however this is outside the scope of this current work.
 }
@@
 \cite{wang_road_nodate}, which found a true positive gravel terrain classification rate of 98.5\%; however, the method again used significantly more data from the LiDAR sensor compared to this work, exploiting all LiDAR channels that fall within a 1.3 meter region of interest.
-There is a general lack of open source LiDAR data for the express purpose of comparing road detection and terrain classification methods in an unstructured, rural environment.
-This is in stark contrast to open source data for urban environments.
+There is a general lack of open source LiDAR data for the express purpose of comparing road detection and terrain classification methods in an unstructured, rural environment, in stark contrast to open source data for urban environments.
 Furthermore, to the best of the author's knowledge there is no comparison between road detection methods using third-party LiDAR data of unmarked, unstructured gravel and asphalt roads in found literature.
 To the best of the author's knowledge, this study employed a novel approach for detecting road surface area and type.}
+
+%		{Future work may include a real-time implementation of the described methods into the vehicle platform. }
 

### 2024-07-10 Huston_Paper.tex
Oh yeah this exists

--- 2023-08-18 Huston_Paper.tex
@@
 \ref{fig:test_vs_train_areas}).
+Training data area and number of LiDAR points within the area has no bearing on the amount of training data used.
 Later, nine arcs (three arcs from three channels) will be classified for the prediction of intersecting roads.
@@
 Training data was extracted from a single arc, with approximately 28 LiDAR points per arc, from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area.
-As arc is fixed to 6 degrees, the number of data points is constant, thus the only factor contributing to point density would be potential elevation changes between the vehicle and arc.
+As each arc is fixed to 6 degrees, the number of data points is constant, thus the only factor contributing to point density would be potential elevation changes between the vehicle and arc.
 This study found that the training data sparsity had a standard deviation that was distinct between LiDAR channels and similar between terrain classes in each LiDAR channel, therefore for this work LiDAR sparsity was considered a negligible issue.
@@
 				\includegraphics[width=0.75\linewidth]{figures/test_vs_train_areas_hatch}
-				\caption[Training vs Testing Areas]{Training areas were kept separate from the testing areas, shown is a close-up, top-down view of the road of interest. }
+				\caption[Training vs Testing Areas]{Training areas were kept separate from the testing areas, shown is a top-down view of the road of interest along with most of the training and testing area.
+Some of the training area is off-image.
+Training data area and number of LiDAR points within the area has no bearing on the amount of training data used.}
 				\label{fig:test_vs_train_areas}
@@
 \ref{fig:dirt_v_gravel}).
+Higher LiDAR point density from a higher resolution LiDAR in conjunction with more descriptive features may yield better results, however for this was outside the scope of this work.
 Exploitation of confidence scores to adjust results by requiring a very high confidence for positive gravel surface classification was explored.

### 2024-07-10 Huston_Paper_07092024.tex
Oh yeah this exists

--- 2023-12-18 Huston_Paper.tex
@@
 Lei et al suggest a method using an SVM for the detection of unstructured dirt and gravel roads using a 2D LiDAR and a camera \cite{lei_detection_2021}, however the method does not use a 3D scanning LiDAR and cameras may be insufficient for reliable road detection \cite{levi_3d_2012_light}.
-Fernandez-Delgado et al found that RDFs out-performed other machine learning algorithms, including NNs, when classifying 121 data sets taken from the UCI data base \cite{fernandez_delgado_we_nodate}.
+Fernandez-Delgado et al found that RDFs out-performed other machine learning algorithms, including NNs, when classifying 121 data sets taken from the UCI data base \cite{fernandez2014we}.
 Therefore while both Neural Networks and RDFs have been used to classify scanning LiDAR data, a RDF classifier was chosen.
-Furthermore, RDFs are less complex to train, able to be trained with smaller data sets, and are generally easier to interpret \cite{nawar_comparison_2017, rosbach_neural_nodate, rosbach_neural_nodate, ahmad_trees_2017}.}
+Furthermore, RDFs are less complex to train, able to be trained with smaller data sets, and are generally easier to interpret \cite{nawar_comparison_2017, bernatz2020comparison, ahmad2017trees}.}
 
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/van_on_van}
+				\includegraphics[width=0.75\linewidth]{figures/smallVan}
 				\caption[Experimental Apparatus]{Experimental Apparatus. Velodyne VLP-32 Scanning LiDAR is mounted on top of the vehicle.}
@@
 				\includegraphics[width=0.75\linewidth]{figures/combined_pcd_example}
-				\caption[Aggregated Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.}
+				\caption[Aggregated Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.
+See Area 1 in Fig.
+\ref{fig:test_vs_train_areas}.}
 				\label{fig:pc_example}
@@
 Arc length was chosen based on approximate vehicle width (set to 6 degrees), and the location of each arc was fixed relative to the center line (length-wise) of the vehicle.
-Training data was extracted from a single arc, with approximately 28 LiDAR points per arc, from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area.
+Training data was extracted from a single arc, with approximately 28 LiDAR points per arc, from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area (Fig.
+\ref{fig:test_vs_train_areas}).
+As the training data is independent of the size of each designated training area and the number of LiDAR points in that area, it is not considered in this work.
 As arc is fixed to 6 degrees, the number of data points is constant, thus the only factor contributing to point density would be potential elevation changes between the vehicle and arc.
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/test_vs_train_areas_hatch}
-				\caption[Training vs Testing Areas]{Training areas were kept separate from the testing areas.}
+				\includegraphics[width=0.95\linewidth]{figures/test_vs_train_areas_hatch_annotated}
+				\caption[Training vs Testing Areas]{Training areas were kept separate from the testing areas.
+While not all are shown, the majority shown here is representative of the bulk of the areas.
+Area 1: Refers to area encompassing plots found in Figs.
+\ref{fig:pc_example}, \ref{fig:raw_classification_results}, \ref{fig:auto_guess_v_truth}, \ref{fig:guess_grav_intersect}.
+Area 2: Refer to area encompassing plots found in Figs.
+\ref{fig:rm_db_6_overlap_2}, \ref{fig:auto_area_guess}, \ref{fig:prepostadjust}
+				}
 				\label{fig:test_vs_train_areas}
@@
 As remission is a ratio of minimum to maximum detectable reflected intensity, there was no need to render the features non-dimensionalized.
-Feature selection closely mimics those suggested in literature \cite{noauthor_pdf_nodate, reymann_improving_2015}.
+Feature selection closely mimics those suggested in literature \cite{reymann_improving_2015}.
 Future work may be completed to introduce additional features.
-Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training, however after thirty percent of the training data was randomly selected and reserved to create a validation data set, approximately two thousand feature sets per class were used to train.}
+Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training, however after thirty percent of the training data was randomly selected and reserved to create a validation data set, approximately two thousand samples of \textit{gravel}, \textit{asphalt}, and \textit{unknown} were used to train the algorithm.}
 
-	% When the app performs hyperparameter tuning by using Bayesian optimization (see Optimization Options for a brief introduction), it chooses the set ofhyperparameter values that minimizes an upper confidence interval of the classification error objective model, rather than the set that minimizes the classification error.
+	% When the app performs hyperparameter tuning by using Bayesian optimization (see Optimization Options for a brief introduction), it chooses the set of hyperparameter values that minimizes an upper confidence interval of the classification error objective model, rather than the set that minimizes the classification error.
 
@@
 
-			%MATLAB was used to train a Random Decision Forests for each of the three considered channels using Bayesian Optimization to tune tree depth and number of splits.
-Training continued until a minimum error was reached (Fig.
-\ref{fig:c2_min_class_error}).
-Validation of the RDF was then completed using the validation database.
-
 			{MATLAB's Classification Learner Application was used to generate an RDF for each considered scanning LiDAR channel.
-RDFs were found in literature to out-perform NNs for classification purposes \cite{fernandez_delgado_we_nodate} while also being generally less complex to train, able to be trained with smaller data sets, and easier to interpret \cite{nawar_comparison_2017, rosbach_neural_nodate, rosbach_neural_nodate, ahmad_trees_2017}.
+RDFs were found in literature to out-perform NNs for classification purposes \cite{fernandez2014we} while also being generally less complex to train, able to be trained with smaller data sets, and easier to interpret \cite{nawar_comparison_2017, bernatz2020comparison, ahmad2017trees}.
 Decision tree depth, number of learners, and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.
-One advantage of RDFs is the relatively low number of tunable hyper-parameters, leading to a more rapid development \cite{nawar_comparison_2017, rosbach_neural_nodate, rosbach_neural_nodate, ahmad_trees_2017}.
+One advantage of RDFs is the relatively low number of tunable hyper-parameters, leading to a more rapid development \cite{nawar_comparison_2017, bernatz2020comparison, ahmad2017trees}.
 Listed are the ones available using MATLAB's Classification Learner Application that were able to be tuned using the Bayesian Optimization function.
@@
 				\includegraphics[width=0.75\linewidth]{figures/db1range_3.png}
-				\caption[Classified Point Cloud]{Classified Point Cloud.}
+				\caption[Classified Point Cloud]{Classified Point Cloud. See Area 1 in Fig. \ref{fig:test_vs_train_areas}.}
 				\label{fig:raw_classification_results}
@@
 				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed based on trends between channels and compared to actual road-surface areas (left).
-Autonomous projection of road surface areas following developed methods (right).}
+Autonomous projection of road surface areas following developed methods (right).
+See Area 2.
+in Fig.
+\ref{fig:test_vs_train_areas}.}
 				\label{fig:rm_db_6_overlap_2}
@@
 				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed based on trends between channels and compared to actual road-surface areas (left).
-Autonomous projection of road surface areas following developed methods (right).}
+Autonomous projection of road surface areas following developed methods (right).
+See Area 2.
+in Fig.
+\ref{fig:test_vs_train_areas}.}
 				\label{fig:auto_area_guess}
@@
 				\includegraphics[width=0.75\linewidth]{figures/auto_guess_v_truth_2}
-				\caption[Projected Guess over Time vs Truth]{Aggregated automated road projection based on classification results.}
+				\caption[Projected Guess over Time vs Truth]{Aggregated automated road projection based on classification results. See Area 1 in Fig. \ref{fig:test_vs_train_areas}.}
 				\label{fig:auto_guess_v_truth}
@@
 		{Training and verification LiDAR, GPS, and IMU data collection was completed using the Experimental Apparatus.
-Gravel data was gathered from the Redmen Lodge gravel parking lot, while grass data was gathered beside the lot.
+Gravel data was gathered from the Redmen Lodge gravel parking lot, while \textit{unknown} data was gathered beside the lot.
 Asphalt data was gathered Blackburn Road.
@@
 					Gravel Areas \% 						& 74.97 	& \% Gravel \\
+					\cline{1-3}
 					Asphalt Area \%							& 87.05   	& \% Asphalt \\
+					\cline{1-3}
 					Side-of-Road \% 						& 77.47 	& \% Unknown\\
+					\specialrule{1pt}{0pt}{0pt}
 					Asphalt Road Detection 					& 100.00 	& \% True Pos. Detection Rate \\
+					\cline{1-3}
 					Gravel Driveway Detection 				& 71.67 	& \% True Pos. Detection Rate \\
-					Misidentified Gravel Road Detection 	& 0 		&  Avg. False Pos. Detection Rate per Run\\
+					\cline{1-3}
+					Misidentified Gravel Road Detection 	& 0 		& \% False Pos. Detection Rate\\
 					\bottomrule
@@
 			}
-			\caption[Table]{Average Results\label{tab:1}}
+			\caption[Table]{Overall results for presented work.
+74.97\% of all samples were classified as \textit{gravel} inside all \textit{gravel} areas, with an overall 71.67\% true positive \textit{gravel} road detection rate and a 0\% false positive \textit{gravel} road detection rate.\label{tab:1}}
 		\end{table}
@@
 \ref{fig:precision_recall_curve_averaged} shows a typical precision-recall curve.).
-Overall it was found that an average true-positive accuracy of all intersecting gravel surfaces was 74.97\% (Fig.
+Overall it was found that an average true-positive accuracy of all intersecting gravel surfaces was 74.97\% (Tab.
+\ref{tab:1}, Fig.
 \ref{fig:prepostadjust}).
@@
 Due to recent construction, discovered asphalt accuracy was artificially lowered due to the area in front of a gravel driveway being contaminated with gravel.
-It was found that an average of 18.55\% of asphalt areas were classified as gravel because of contamination.
-Differentiation between gravel road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation.
-Overall average distribution for \textit{unknown} terrain class within side-of-road areas was found to be 77.94\%, which is lower than expected due to the side of the road being re-grassed.
+It was found that an average of 18.55\% of asphalt areas were classified as gravel because of contamination.}
+
+		{Differentiation between gravel road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation.
+Overall average distribution for \textit{unknown} terrain class within side-of-road areas was found to be 77.94\% (Tab.
+\ref{tab:1}), which is lower than expected due to the side of the road being re-grassed.
 It is acknowledged by the authors that the classification algorithm must yield a result that most closely matches one of the trained classes, which led to lower accuracy scores in these circumstances.
+It is also acknowledged by the authors that using a higher resolution LiDAR scanner in conjunction with a more comprehensive and descriptive feature set may be used to aid in differentiating between dirt and gravel, however this future study is outside the scope of this work.
 Overall discovered accuracy levels may be sufficient for the purpose of identifying asphalt and gravel roads by considering trends between classified arcs.}
@@
 			\centering
-			\includegraphics[width=0.75\linewidth]{figures/precision_recall_curve_averaged}
-			\caption[PR Curve]{Precision-Recall curve of the averaged \textit{gravel} vs \textit{unknown} classification results from all test runs. }
+			\includegraphics[width=0.75\linewidth]{figures/precision_recall_curve_averaged_marked}
+			\caption[PR Curve]{Precision-Recall curve of the averaged \textit{gravel} vs \textit{unknown} classification results from all test runs. The red circle indicates the presented method's location.}
 			\label{fig:precision_recall_curve_averaged}
@@
 Note that the asphalt area in front of the gravel driveway has significant \textit{gravel} classification due to construction.
-These results are included with the overall asphalt classification score.}
+These results are included with the overall asphalt classification score.
+See Area 2.
+in Fig.
+\ref{fig:test_vs_train_areas}.}
 			\label{fig:prepostadjust}
@@
 \ref{fig:guess_grav_intersect}).
-Using autonomous methods to project road surfaces, it was found that a true-positive intersecting gravel road detection rate of 71.66\% with no false positives.
+Using autonomous methods to project road surfaces, it was found that a true-positive intersecting gravel road detection rate of 71.66\% with no false positives (Tab.
+\ref{tab:1}).
 It was found that the true positive rate may be increased to 100\% at the expense of increasing the number of false positives with a 1.8 average number of false positives per drive-by.}
@@
 			\includegraphics[width=0.95\linewidth]{figures/db1_conf_elim_auto_guess_mca}
-			\caption[Aggregated Automated Road Surface Detection]{Aggregated automated road surface detection was compared to truth areas. }
+			\caption[Aggregated Automated Road Surface Detection]{Aggregated automated road surface detection was compared to truth areas. See Area 1 in Fig. \ref{fig:test_vs_train_areas}.}
 			\label{fig:guess_grav_intersect}
@@
 al.
-\cite{wang_road_nodate}, which found a true positive gravel terrain classification rate of 98.5\%; however, the method again used significantly more data from the LiDAR sensor compared to this work, exploiting all LiDAR channels that fall within a 1.3 meter region of interest.
+\cite{wang_road_nodate}, which found a true positive gravel terrain classification rate of 98.5\%; however, the method again used significantly more data from the LiDAR sensor compared to this work, exploiting all LiDAR channels that fall within a 1.3 meter region of interest leading to a higher point density.
 There is a general lack of open source LiDAR data for the express purpose of comparing road detection and terrain classification methods in an unstructured, rural environment, in stark contrast to open source data for urban environments.

### 2024-07-10 previous_builds/Huston_Paper_04262024_diff_backup.tex
Oh yeah this exists

--- 2023-12-18 Huston_Paper.tex
@@
 Future work may be completed to introduce additional features.
-Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training, however after thirty percent of the training data was randomly selected and reserved to create a validation data set, approximately two thousand feature sets per class were used to train.}
+Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training, however after thirty percent of the training data was randomly selected and reserved to create a validation data set, approximately two thousand samples of \textit{gravel}, \textit{asphalt}, and \textit{grass} were used to train the algorithm.}
 
@@
 It is acknowledged by the authors that the classification algorithm must yield a result that most closely matches one of the trained classes, which led to lower accuracy scores in these circumstances.
+It is also acknowledged by the authors that using a higher resolution LiDAR scanner in conjunction with a more comprehensive and descriptive feature set may be used to aid in differentiating between dirt and gravel, however this future study is outside the scope of this work.
 Overall discovered accuracy levels may be sufficient for the purpose of identifying asphalt and gravel roads by considering trends between classified arcs.}

### 2024-07-10 previous_builds/Huston_Paper_asme.tex
Oh yeah this exists

--- 2023-05-03 Huston_Paper_asme.tex
@@
 pdfauthor={Rhett Huston},
-pdftitle={LiDAR Stuff},                  % <=== change to YOUR pdf file title
+pdftitle={Enhance Road Detection Data Processing of LiDAR Point Clouds to Specifically Identify Unmarked Gravel Rural Roads},                  % <=== change to YOUR pdf file title
 }
@@
 
-	%	{MLS and RANSAC were compared as methods to project a two dimensional plane unto a point cloud, facilitating examination of points by providing a frame of reference \cite{miller_method_nodate, gojcic_perfect_2019}.
-Method of least-squares analyses uses regression to approximate the best fit to a data set (\ref{eqn:MLS}).
-RANSAC is an iterative algorithm that is designed to handle large numbers of outliers in the sample data set \cite{derpanis_overview_nodate,yaniv_random_2010,fischler_random_1987,cantzler_random_nodate}.}
-
 	{Random Decision Forests are a supervised machine learning method that operates on a set of decision trees \cite{ho_random_1995}.
@@
 
-	\subsubsection{Overview}
+	\subsection{Overview}
 
-	{Followed methodology to accomplish this work is outlined in Figure \ref{fig:flowz_2}.
-The experimental apparatus for this work was the Ohio University Autonomous Van van (Figure \ref{fig:van}), which has a Velodyne VLP-32C scanning LiDAR sensor, a Novatel PwrPak 7D-E2 GNSS and INS system, and Mako cameras.
+	{Followed methodology to accomplish this work is outlined in Fig.
+\ref{fig:flowz_2}.
+The experimental apparatus for this work was the Ohio University Autonomous Van van (Fig.
+\ref{fig:Experimental_Apperatus}), which has a Velodyne VLP-32C scanning LiDAR sensor, Novatel PwrPak 7D-E2 GNSS and INS system, and Mako cameras.
 Raw data was gathered using the Robotic Operating System (ROS) to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags.
-Scanning LiDAR data was sent from the sensor as a single 360 degree scan at 20 Hz.
-GNSS and IMU data was captured at a rate of 100 Hz.
 Rosbags were unpacked using MATLAB tools to separate data streams.
-Prior to training data extraction, manual dictation of gravel and asphalt areas is necessary.
-In order to do so, all scanning LiDAR data from a rosbag was compiled in order to create a single point cloud map.
+Prior to creating a training database, manual dictation of gravel and asphalt areas of the gathered point cloud data is necessary.
+In order to do so, all scanning LiDAR data from a rosbag was aggregated in order to create a single point cloud map.
 Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data.
-Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan (Figure \ref{fig:Combined_Classified_PCD}).
-Compiled point cloud maps may then be examined and compared to camera and satellite data to determine gravel and asphalt surface location.
-Manually defined 2D areas representing gravel and asphalt were projected unto the point cloud (Figure \ref{fig:Classified_Areas_PCD}).}
+Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan (Fig.
+\ref{fig:road_areas_annotated_22}).
+Compiled point cloud maps was examined and compared to camera and satellite data to determine gravel and asphalt surface location.
+Manually defined 2D areas representing gravel and asphalt were projected unto the point cloud (Fig.
+\ref{fig:road_areas_annotated_22}).}
 
-	{Training data was extracted by examining each 360 scan and comparing the location of a 6 degree arc directly in front of the area and the manually projected 2D areas.
-If the arc resides within the manually defined surface area, raw spatial and intensity values from the arc were saved to a raw training database.
-Seperate training data bases were used for each individual channel.
-Features that describe the raw spatial and intensity data were extracted by performing mathematical functions and saving the results to a feature data base.
+	{Training data was extracted from a single arc from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area.
+Raw spatial and intensity values from the coincident arc were saved to a raw training database.
+Separate training data bases were maintained for each individual channel.
+Features that describe the raw spatial and intensity data were extracted.
 Thirty percent of the training data base was randomly selected and set aside as a validation data base.
-Training data was visually examined for clustering to verify that a Random Decision Forest would be able to be trained (Figure \ref{fig:training_data_cluster}).
+Training data was visually examined for clustering to verify that a Random Decision Forest would be able to be trained (Fig.
+\ref{fig:training_data_cluster}).
 It was found that each class generally clustered to a certain range with some overlap, which indicates that a Random Decision Forest may be trained.
 MATLAB was used to train a Random Decision Forests for each of the three considered channels using Bayesian Optimization to tune tree depth and number of splits.
-Training continued until a minimum error was reached (Figure \ref{fig:c2_min_class_error}).
-Validation of the RDF was then completed using the validation database (Figure \ref{fig:vali_conf_mat}).
-}
+Training continued until a minimum error was reached (Fig.
+\ref{fig:c2_min_class_error}).
+Validation of the RDF was then completed using the validation database.}
 
@@
 
-	{Manually defined gravel, asphalt, and side-of-road areas were projected onto the classified point cloud.
-Manually defined gravel, asphalt, and side-of-road areas were then used to determine the accuracy of the terrain classification algorithm.
-Classified point clouds were then manually examined to determine if the location of gravel and asphalt surfaces may be identified.
+	{Manually defined gravel, asphalt, and side-of-road areas were projected onto the classified point cloud then used to determine the accuracy of the terrain classification algorithm.
+Classified point clouds were manually examined to determine if the location of gravel and asphalt surfaces may be identified visually.
 If consecutive points from each channel in any direction had two or more matching gravel or asphalt classification and was evenly spaced, a gravel or road surface area was projected.
-Guessed road surface areas were then compared to actual areas to determine the method's ability at unmarked gravel road discovery.}
+Guessed road surface areas were then compared to actual areas to determine the method's ability at unmarked gravel road discovery, indicating the method's efficiency at detecting intercepting gravel roads.}
 
@@
 		\includegraphics[width=\linewidth]{figures/c2_min_class_error}
-		\caption[RDF Training Classification Error]{}
+		\caption[RDF Training Classification Error]{Classification error during Bayesian Optimization tuning.}
 		\label{fig:c2_min_class_error}
@@
 
+%	\begin{figure}
+%		\centering
+%		\includegraphics[width=\linewidth]{figures/test_conf_mat}
+%		\caption[Validation Confusion Matrix]{Confusion matrix from validation process.}
+%		\label{fig:vali_conf_mat}
+%	\end{figure}
+
 	\subsection{Data Collection}
 
-	{Required physical data was gathered by the Ohio University Autonomous Van van (Figure \ref{fig:van}), built by AutonomouStuff.
-LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C (Figure \ref{fig:vlp32mount}).
+	{Required physical data was gathered by the Ohio University Autonomous Van van (Fig.
+\ref{fig:Experimental_Apperatus}), built by AutonomouStuff.
+LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C.
 Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a high output data stream \cite{vlp_32c}.
@@
 Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot.
-Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection.
-Only three classes are considered in this work: "gravel", "asphalt", and "unknown".
-}
+Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection, therefore grass is labeled as "unknown" in this work.}
+
+%	\begin{figure} 
+%		\centering
+%		\begin{subfigure}{0.5\textwidth}
+%			\centering
+%			\includegraphics[width=1\linewidth,height=4.60 cm,keepaspectratio]{figures/van_on_van}
+%			\caption[Sensor Van]{}
+%			\label{fig:van}
+%		\end{subfigure}
+%		\begin{subfigure}{0.5\textwidth}
+%			\centering
+%			\includegraphics[width=1\linewidth,height=4.60 cm,keepaspectratio]{figures/LiDAR_on_van}
+%			\caption[VLP 32 on Van]{}
+%			\label{fig:vlp32mount}
+%		\end{subfigure}
+%		\caption[Experimental Apparatus]{Experimental Apparatus (a). Velodyne VLP-32 Scanning LiDAR is mounted on top of the vehicle (b).}
+%		\label{fig:Experimental_Apperatus}
+%	\end{figure}
 
@@
 		\centering
-		\begin{subfigure}{0.5\textwidth}
-			\centering
-			\includegraphics[width=1\linewidth,height=4.60 cm,keepaspectratio]{figures/van_on_van}
-			\caption[Sensor Van]{}
-			\label{fig:van}
-		\end{subfigure}
-		\begin{subfigure}{0.5\textwidth}
-			\centering
-			\includegraphics[width=1\linewidth,height=4.60 cm,keepaspectratio]{figures/LiDAR_on_van}
-			\caption[VLP 32 on Van]{}
-			\label{fig:vlp32mount}
-		\end{subfigure}
-		\caption[Experimental Apparatus]{Experimental Apparatus (a). Velodyne VLP-32 Scanning LiDAR is mounted on top of the vehicle (b).}
+		\includegraphics[width=0.9\linewidth]{figures/van_on_van}
+		\caption[Experimental Apparatus]{Experimental Apparatus. Velodyne VLP-32 Scanning LiDAR is mounted on top of the vehicle.}
 		\label{fig:Experimental_Apperatus}
@@
 
-	{MATLAB was used to process the gathered rosbag files using the their ROSBAG library.
+	{MATLAB was used to process the gathered rosbag file.
 Rosbags were directly loaded into the MATLAB environment.
-Point cloud, GPS, and IMU topics were defined, and the messages were read from the topic and imported.
-Cartesian coordinates, longitude, latitude, altitude, roll, pitch, and yaw were extracted from the messages.}
+Point cloud, GPS, and IMU topics were defined, and Cartesian coordinates, longitude, latitude, altitude, roll, pitch, and yaw messages were read from the appropriate topics and imported.
+}
 
@@
 GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors.
-Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}).
+Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Fig.
+\ref{fig:road_areas_annotated_22}).
 While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for this work.
-Manually defined areas may then be overlayed unto the point cloud (Figure \ref{fig:Combined_Classified_PCD}).
+Manually defined areas may then be overlayed unto the point cloud (Fig.
+\ref{fig:road_areas_annotated_22}).
 Six degree arcs from the first three channels of scanning LiDAR data that lay directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.}
+
+%	\begin{figure}
+%		\centering
+%		\begin{subfigure}{0.45\textwidth}
+%			\centering
+%			\includegraphics[width=0.95\linewidth]{figures/Compiled_PCD}
+%			\caption[C]{}
+%			\label{fig:Compiled_PCD}
+%		\end{subfigure}
+%		\begin{subfigure}{0.45\textwidth}
+%			\centering
+%			\includegraphics[width=0.95\linewidth]{figures/Manually_Defined_Area_Train_Dat_Grab_2}
+%			\caption[]{}
+%			\label{fig:Classified_Areas_PCD}
+%		\end{subfigure}
+%		\caption[Compiled and Classified Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map (a).
+Manually defined areas inform where training data lies (b).}
+%		\label{fig:Combined_Classified_PCD}
+%	\end{figure}
 
@@
 		\centering
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=0.95\linewidth]{figures/Compiled_PCD}
-			\caption[C]{}
-			\label{fig:Compiled_PCD}
-		\end{subfigure}
-		\begin{subfigure}{0.45\textwidth}
-			\centering
-			\includegraphics[width=0.95\linewidth]{figures/Manually_Defined_Area_Train_Dat_Grab_2}
-			\caption[]{}
-			\label{fig:Classified_Areas_PCD}
-		\end{subfigure}
-		\caption[Compiled and Classified Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map (a).
-Manually defined areas inform where training data lies (b).}
-		\label{fig:Combined_Classified_PCD}
+		\includegraphics[width=0.95\linewidth]{figures/road_areas_annotated_22}
+		\caption[Compiled and Manually Classified Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.
+Projected manually defined areas define terrain surface type.
+Five runs of Blackburn Road with three intercepting gravel parking lot were classified.
+Area $3$ is the gravel parking lot entrance from which training data was extracted.}
+		\label{fig:road_areas_annotated_22}
 	\end{figure}
@@
 Decision tree depth, number of learners, and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization.
-Bayesian Optimization automates the manual hyper-parameter tuning process creating a surrogate model based on .
-All arcs in a channel used the same RDF.
-}
+Bayesian Optimization automates the manual hyper-parameter tuning process by creating a surrogate model representing the objective function, in this case out-of-bag error.
+Initialization of the objective function is completed by training a number of RDFs using randomly selected hyper-parameters.
+Out-of-bag error is then found and used to create a series of Gaussian curves.
+Additional hyper-parameters are chosen by an acquisition function to determine their utility in minimizing out-of-bag error.
+Best hyper-parameters are chosen after a set number of iterations and used to create a final RDF.
+Validation of the RDF was completed using the reserved thirty percent of the training data.
+This process was completed for each of the three considered Scanning LiDAR channels used for classification.}
 
@@
 Features from each arc were extracted and the arc was classified.
-Averaged x, y, and z coordinates were taken of each arc and projected unto a manually defined truth area map (Figure \ref{fig:Classified_Areas_PCD}).
-Accuracy scores of each area was calculated, allowing evaluation of road detection performance.}
+Averaged x, y, and z coordinates were taken of each arc and projected unto a manually defined truth area map (Fig.
+\ref{fig:road_areas_annotated_22}).
+Accuracy scores of each area was calculated, allowing evaluation of road detection performance.
+Classified point clouds were visually inspected to determine if the manual detection of a road surface may be accomplished.
+Manually detected road surface areas were compared to actual areas, yielding a road detection accuracy score.
+}
 
@@
 Translation and rotation was applied to the averaged x, y, and z coordinates of each arc's terrain classification results.
-Consecutive scanning LiDAR data was projected on manually classified truth areas representing road road surfaces (Figure \ref{fig:area_percentages}) and percentages of terrain types per area were calculated.
-}
+Consecutive scanning LiDAR data was projected on manually classified truth areas representing road road surfaces (Figure \ref{fig:prepostadjust}) and percentages of terrain types per area were calculated.}
 
-	{Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 67.54\%.
+	{Classification of real-world data outside of the training data set presents difficulties to the generated Random Decision Forest Classifiers generated in this work.
+Due to the side of the road was being re-grassed near area $3$ (Fig.
+\ref{fig:Blackburn_Road_View}, \ref{fig:road_areas_annotated_22}), the algorithm struggled to differentiate the dirt from gravel, leading to some miss-classification of the side-of-road areas.
+Examination of the feature distribution shows a high amount of overlap (Fig.
+\ref{fig:rm_db_4_toc}).
+Exploitation of confidence scores to adjust results by requiring a very high confidence for positive gravel surface classification was explored.
+All gravel surfaces that did not meet a 80\% confidence score was re-labeled as "unknown", however these adjustment had either a negligible or detrimental impact on the final results.
+Higher confidence score requirements resulted in "unknown" over-classification, while too low resulted in "gravel" over-classification.
+Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 67.54\%.
 If any asphalt miss-classification on the gravel areas is included, the average true-positive road detection accuracy increases to 76.97\%.
@@
 		\centering
-		\includegraphics[width=0.9\linewidth]{figures/rm_db_4_percentages_big_anotated}
-		\caption[Area Percentage Scores]{Five runs of Blackburn Road with the intercepting gravel parking lot were classified.
-Shown is the results of one run.
-Percentages of each terrain type per area were calculated.
-Area $3$ is the gravel parking lot entrance from which training data was extracted.}
-		\label{fig:area_percentages}
-	\end{figure}
-
-	%	\begin{figure}[H]
-		%		\centering
-		%		\begin{subfigure}{0.495\textwidth}
-			%			\centering
-			%			\includegraphics[width=0.975\linewidth]{figures/side_of_road_nums_4_with_no_filts_2}
-			%			\caption[Pre-Adjusted]{}
-			%			\label{fig:pre-adjust}	
-			%		\end{subfigure}
-		%		\begin{subfigure}{0.495\textwidth}
-			%			\centering
-			%			\includegraphics[width=0.975\linewidth]{figures/side_of_road_nums_4_with_filts_2}
-			%			\caption[Post-Adjusted]{}
-			%			\label{fig:post-adjust}
-			%		\end{subfigure}
-		%		\label{fig:prepostadjust}
-		%		\caption[Post-Adjusted]{Raw classification result example (a).
-Results were slightly altered by considering confidence scores and inter-arc distances to decrease miss-classification of gravel surfaces, however the effect was negligible (b).}
-		%	\end{figure}
-
-	\begin{figure}
-		\centering
-		\includegraphics[width=0.975\linewidth]{figures/side_of_road_nums_4_with_filts_2}
-		\caption[Post-Adjusted]{}
+		\includegraphics[width=0.9\linewidth]{figures/side_of_road_nums_4_with_filts_2_2}
+		\caption[Area Scores]{Per-area scores were calculated, indicating algorithm unmarked road detection accuracy. }
 		\label{fig:prepostadjust}
-		\caption[Area Scores]{Per-area scores were calculated, indicating algorithm efficiency of unmarked road detection. }
 	\end{figure}
@@
 	{Classified point clouds were manually examined for 'best guess' locations for gravel and asphalt surfaces.
-Two out of three positive gravel identifications must be made on each channel, be evenly spaced, and be relatively co-planar to other road surface guesses in order for a positive guess for an asphalt or gravel surface to be made (Figure \ref{fig:rm_db_4_toc}).
+Two out of three positive gravel identifications must be made on each channel, be evenly spaced, and be relatively co-planar to other road surface guesses in order for a positive guess for an asphalt or gravel surface to be made (Fig.
+\ref{fig:rm_db_4_toc}).
 Driveways $1$ and $2$ were detected in all five drive-bys, driveway $3$ was detected in four leading to an 91.67\% accuracy in intercepting gravel road detection.
@@
 		\centering
-		\includegraphics[width=0.9\linewidth]{figures/rm_db_4_toc}
-		\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed using a rule set and compared to actual road-surface areas. }
+		\includegraphics[width=0.9\linewidth]{figures/rm_db_4_ToC_2}
+		\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed and compared to actual road-surface areas. }
 		\label{fig:rm_db_4_toc}
@@
 
-\section{Discussion}
-
-	{Classification of real-world data outside of the training data set presents difficulties to the generated Random Decision Forest Classifiers generated in this work.
-Due to the side of the road was being re-grassed near area $3$ (Figure \ref{fig:area_percentages}, Figure \ref{fig:Blackburn_Road_View}), the algorithm struggled to differentiate the dirt from gravel, leading to some miss-classification of the side-of-road areas.
-Examination of the feature distribution shows a high amount of overlap (Figure \ref{}).
-Exploitation of confidence scores to adjust results by requiring a very high confidence for positive gravel surface classification was explored.
-All gravel surfaces that did not meet a 80\% confidence score was re-labeled as "unknown", however these adjustment had either a negligible or detrimental impact on the final results.
-Higher confidence score requirements resulted in "unknown" over-classification, while too low resulted in "gravel" over-classification.
-}
+%\section{Discussion}
 
@@
 
-\section*{Acknowledgments}
-Place any acknowledgments here.
+%\section*{Acknowledgments}
+%Place any acknowledgments here.
 
@@
 
-\nocite{*}%% <=== Delete this line unless you want to typeset the entire contents of your .bib file!
-
+%\nocite{*}
+\bibliography{resources.bib} 
 \bibliographystyle{asmeconf}  %% .bst file following ASME conference format. Do not change.
-\bibliography{resources.bib}
 

### 2024-07-10 previous_builds/Huston_Paper_old.tex
Oh yeah this exists

--- 2024-07-10 Huston_Paper.tex
@@
 		{Gravel roads lack standardized features such as curbs or painted lines, presenting detection challenges to autonomous vehicles.
-Global Positioning Service (GPS) and high resolution maps may not be reliable for navigation of gravel roads, as some roads may only be width of the vehicle and GPS may not be accurate enough.
-Normal Distribution Transform (NDT) LiDAR scan matching may be insufficient for navigating on gravel roads as there may not be enough geometrically distinct features for reliable scan matching.
-Completed work will examine a method of classifying scanning LiDAR spatial and remission data features for explicit detection of unmarked gravel road surfaces.
+Global Positioning Service (GPS) and high resolution maps may not be reliable for navigation of gravel roads, as some may only be width of the vehicle and GPS may not be accurate enough.
+Normal Distribution Transform (NDT) based on scanning LiDAR data may be insufficient for navigating on gravel roads as there may not be enough geometrically distinct features for reliable scan matching.
+This work will examine a method of classifying scanning LiDAR data spatial and remission features for explicit detection of unmarked gravel road surfaces.
 Exploration of terrain classification using high resolution scanning LiDAR data of these road surfaces may allow for predicting gravel road boundary locations potentially enabling confident autonomous operations on gravel roads.
-The principal outcome of this work was a method for gravel road terrain detection using LiDAR data for the purpose of predicting potential road boundary locations.
+The principal outcome of this work was a method for gravel road detection using LiDAR data for the purpose of predicting road boundary locations.
 Random Decision Forests were trained using scanning LiDAR data terrain classification to detect unmarked gravel and asphalt surfaces.
-It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively at an estimated rate of 13.19 ms per 360 degree scan.
-Overlapping results between manually projected and actual road surface areas resulted in 93.33\% intersecting gravel road detection accuracy.
-Automated post-process examination of classification results yielded an  true-positive gravel road detection rate of 71.67\%.
-Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on previously undetected roads.}
+It was found that a true-positive accuracy for gravel and asphalt surfaces was 74.97\% and 84.05\% respectively at an estimated rate of 13.19 ms per 360 degree scan.
+Overlapping results between manually projected and actual road surface areas resulted in 93.33\% intercepting gravel road detection accuracy.
+Results indicate that using a Random Decision Forest to classify scanning LiDAR data is a feasible method for unmarked gravel and asphalt road detection.
+Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.}
 
@@
 Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}.
-Real-time detection of road surfaces limits the available computational time required for low-latency vehicle trajectory updates, prohibiting the collection of larger point LiDAR point clouds.
-Implementing a Random Decision Forests (RDF) terrain classification model using arcs from a single 360 degree scan of LiDAR data to exploit surface properties was studied to determine the method's feasibility in road surface detection.
-While an established method for terrain classification, to the author's best knowledge RDFs in found literature were not employed for the detection of road surface area and type using a small data input within the context of an unstructured environment, making the approach used in this work is a novel one.
+Implementing a Random Decision Forests (RDF) terrain classification model using scanning LiDAR data to exploit surface roughness was studied to determine the method's feasibility in road surface detection.
 Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.}
@@
 	{Further limitations are imposed when considering data set size and processing speed.
-Real time trajectory updates to the vehicle in motion requires low latency road surface analysis, denying the use of methods that require post-processing due to high computational loads.
-Real time analysis of road surfaces dictates that trajectory updates to the vehicle in motion must have low latency, which prohibits the relatively lengthy collection of large data sets or any form of post-processing.
+Real time analysis of road surfaces dictates that trajectory updates to the vehicle in motion must have a rapid update rate, which prohibits the relatively lengthy collection of large data sets or any form of post-processing.
 Proposed methods \cite{yadav_extraction_2017,yadav_road_2018,yadav_rural_2018,yadav_pole-shaped_2015,miyazaki_line-based_2014,yang_semi-automated_2013,liu_new_2013,qiu_fast_2016} do not indicate the minimum number of points required for road surface detection, however large data sets numbering many millions of points are used in those studies.
@@
 Edmond et al.
-proposed a method to detect terrain type through implementation of vehicle vibration, however this method is limited to what the vehicle is currently driving over and thus cannot detect intersecting gravel roads \cite{dupont_online_2008}.
+proposed a method to detect terrain type through implementation of vehicle vibration, however this method is limited to what the vehicle is currently driving over and thus cannot detect intercepting gravel roads \cite{dupont_online_2008}.
 Filitchkin et al.
@@
 Road classification with LiDAR spacial and remission features was completed with Neural Network classifier, Naive Bayes classifier, and Support Vector Machines (SVM) \cite{wang_road_nodate,wang_two-stage_2018}.
-Gravel surfaces were successfully identified with a 98.5\% true positive identification rate; however, the method was only focused on classifying road surface type directly behind the vehicle, and was not tested for predicting road surface areas.
-MATLAB's Neural Network tools were used in autonomous road identification using LiDAR and Camera data \cite{rasmussen_combining_2002}; however, cameras may not be relied upon for ambient conditions may affect classification accuracy \cite{levi_3d_2012_light}, furthermore computational time was not mentioned.
+Gravel roads were successfully identified with a 98.5\% true positive identification rate; however, data collected required post processing.
+MATLAB's Neural Network tools were used in autonomous road identification using LiDAR and Camera data \cite{rasmussen_combining_2002}; however, post processing was required, with unknown computational time.
 Terrain classification was completed using Support Vector Machine (SVM), a classification algorithm where a line is drawn between two different categories to differentiate between them \cite{breiman_random_2001}.
@@
 Classification is determined by the majority results as opposed to the mean \cite{breiman_random_2001,ho_random_1995}.
-While RDFs were used in terrain classification in multiple instances in found literature, to the author's best knowledge they have not been employed in predicting road surface areas and type while also only using extremely small input data sets typical of real-time mobile scanning.
 Bayesian Optimization may be used to tune the hyper parameters of a machine algorithm \cite{noauthor_bayesian_nodate, snoek_practical_2012}.
@@
 
-%		{Detection of unmarked gravel road surfaces using terrain classification requires building a data base of terrain features, training an RDF, and testing the algorithm for accuracy of unmarked road detection (outline of work in Figure \ref{fig:flowz_7}).
-Raw data was gathered using the experimental apparatus, which has a Velodyne VLP-32C scanning LiDAR sensor, and Novatel PwrPak 7D-E2 GNSS and INS system.
-Robotic Operating System (ROS) was used to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags, which were unpacked using MATLAB tools to separate data streams.
-Prior to creating a training database, manual dictation of gravel and asphalt areas of the gathered point cloud data is necessary for automatic training data extraction.
-In order to do so, all scanning LiDAR data from a rosbag was aggregated in order to create a single point cloud map.
-Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data.
-Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan.
-Compiled point cloud maps were examined and compared to camera and satellite data to determine gravel and asphalt surface location.
-Manually defined 2D areas representing gravel and asphalt were projected unto the point cloud \textit{[Manually Define Road \& Side-of-Road]} (Section \ref{sec:training_and_verification_data_selection_process}).}
-%		
-%		{Training data was extracted from a single arc from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area (Figure \ref{fig:test_vs_train_areas}).
-Raw spatial and intensity values from the coincident arc were saved to a raw training database \textit{[Training Data Extraction]}.
-Separate training data bases were maintained for each individual channel.
-Features that describe the raw spatial and intensity data were extracted.
-Thirty percent of the training data base was randomly selected and set aside as a validation data base.
-Training data was visually examined for clustering to verify that a Random Decision Forest would be able to be trained (Section \ref{sec:Feat_Extract}).
-It was found that each class generally clustered to a certain range with some overlap, which indicates that a Random Decision Forest may be trained.
-MATLAB was used to train a Random Decision Forests for each of the three considered channels using Bayesian Optimization to tune tree depth and number of splits \textit{[RDF Training \& Verification]} (Section \ref{sec:random-decision-forest-creation}).
-Training continued for thirty iterations and the algorithm that produced the minimum upper confidence interval for out-of-bag error was selected.
-Validation of the RDF was then completed using the validation database (Section \ref{sec:random-decision-forest-verification}).}
-%		
-%		{Scanning LiDAR data from five datasets representing five drives on a stretch of Blackburn Road was classified \textit{[Classify Scanning LiDAR Data]}.
-Following the same process as aggregating point clouds, GNSS and IMU data informed the location and orientation of the arcs in each 360 degree scan.
-Three arcs from three channels with sizes determined by approximate vehicle width (~7 ft, Figure \ref{fig:area_example}) were then classified as "asphalt", "gravel", or "unknown", and the averaged coordinates of each arc was aggregated into a single, classified point cloud (Figure \ref{fig:raw_classification_results}).}
-
 			{Detection of unmarked gravel road surfaces using terrain classification requires collecting data, training an RDF, and testing the algorithm for accuracy of unmarked road detection (outline of work in Fig.
-\ref{fig:flowz_7}).
+\ref{fig:flowz_2}).
 Training data was gathered by the Ohio University Autonomous Van (Fig.
@@
 Scanning LiDAR data was consecutively classified and compared to manually defined truth areas, yielding the true-positive accuracy for gravel, asphalt, and non-road surfaces.
-Classified point cloud data was manually examined in order to manually project possible road surfaces using manual and automated methods.
-Projected road surface areas were compared to truth areas to determine the accuracy of asphalt road and intersecting gravel driveway detection.}
+Classified point cloud data was examined without reference to truth areas in order to manually project possible road surfaces.
+Manually projected road surface areas were compared to truth areas to determine the accuracy of asphalt road and intercepting gravel driveway detection.}
 
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/flowz_7}
+				\includegraphics[width=0.9\linewidth]{figures/flowz_2}
 				\caption[Project Flow]{High level view of project work}
-				\label{fig:flowz_7}
+				\label{fig:flowz_2}
 			\end{figure}
@@
 			{Required physical data was gathered by the Ohio University Autonomous Van (Fig.
-\ref{fig:Experimental_Apperatus})  on a sunny, cloudless day with no road obstacles or debris at a speed below 15 miles per hour.
-It is acknowledged that weather or non-ideal road conditions may affect results and is an area of future study.
+\ref{fig:Experimental_Apperatus}).
 LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C.
@@
 GPS data was recorded with a Novatel PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS), and two GNSS-502 antennas by NavtechGPS.
-GNSS was assisted with Real Time Kinematic correction data from the Ohio Continuously Operating Reference Station (CORS) and Ohio Real Time Network (RTN) \cite{noauthor_cors_nodate}, yielding in our tests sub 2cm location accuracy level.
-IMU specifications indicate $0.8 ^{\circ}/hr$ bias instability, $0.06 ^{\circ}/\sqrt{hr}$ angular random walk, and $0.025 m/s/\sqrt{hr}$ velocity random walk, in addition to a Extended Kalman Filter \cite{noauthor_pwrpak7_nodate}.
-Because the IMU was only used to orientate the individual LiDAR scans, IMU drift was not considered a factor that would effect the classification results themselves.
-Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using the Robot Operating System (ROS) to be subsequently examined and categorized later using MATLAB.
+Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using the Robot Operating System (ROS).
 Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot (Fig.
@@
 Scanning LiDAR data of an asphalt surface was gathered from Blackburn Road (Fig.
-\ref{fig:Blackburn_Road_View}) which provides access to the training gravel lot.
+\ref{fig:Blackburn_Road_View}) near the intersecting gravel lot.
 Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot.
 Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection; therefore, grass is labeled as \textit{unknown} in this work.
-While the number of classes is low, this is done to lower computational load.
-It was the expectation for this work that terrain that did not match the asphalt or gravel classes would be classified as unknown, and this would be later confirmed in results.}
+Raw data was gathered using the Robotic Operating System (ROS) to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags.
+Rosbags were unpacked using MATLAB tools to separate data streams.
+Point cloud, GPS, and IMU topics were defined, and point cloud Cartesian coordinates with corresponding intensities, longitude, latitude, altitude, roll, pitch, and yaw messages were read from the appropriate topics and imported.}
 
@@
 				\includegraphics[width=0.75\linewidth]{figures/blackburn_road}
-				\caption[Blackburn Road Camera View]{Camera view of Blackburn Road - an unmarked asphalt road with three intersecting gravel driveways from which asphalt data was gathered.}
+				\caption[Blackburn Road Camera View]{Camera view of Blackburn Road - an unmarked asphalt road with three intercepting gravel driveways from which asphalt data was gathered.}
 				\label{fig:Blackburn_Road_View}
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/gravel_training_lot}
+				\includegraphics[width=0.75\linewidth]{figures/gravel_lot_pic}
 				\caption[Gravel Training Lot]{Camera view of the gravel lot from which gravel training data was gathered.}
@@
 \ref{fig:pc_example}).
-While not as robust as more sophisticated methods of point cloud aggregation due to IMU drift, such as NDT or ICP scan matching, using this method proved to be adequate over shorter distances given the reasons previously mentioned.}
+While not as robust as more sophisticated methods of point cloud aggregation, such as NDT or ICP scan matching, using this method proved to be adequate over shorter distances.}
 
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/combined_pcd_example}
-				\caption[Aggregated Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.
-Intensity informs point color in this example image.}
+				\includegraphics[width=0.75\linewidth]{figures/pc_example}
+				\caption[Aggregated Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.}
 				\label{fig:pc_example}
@@
 
-			{Arcs of 6 degrees from the first three channels of scanning LiDAR data centered directly in front of the vehicle (Fig.
-\ref{fig:nine_arcs_example}) were extracted and exported to a database if coincident with the manually defined areas (Fig.
+			{Manually defined areas were then overlayed onto the point cloud, representing training areas (Fig.
 \ref{fig:test_vs_train_areas}).
-Training data area and number of LiDAR points within the area has no bearing on the amount of training data used.
-Later, nine arcs (three arcs from three channels) will be classified for the prediction of intersecting roads.
-Arc length was chosen based on approximate vehicle width (set to 6 degrees), and the location of each arc was fixed relative to the center line (length-wise) of the vehicle.
-Training data was extracted from a single arc, with approximately 28 LiDAR points per arc, from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area.
-As each arc is fixed to 6 degrees, the number of data points is constant, thus the only factor contributing to point density would be potential elevation changes between the vehicle and arc.
-This study found that the training data sparsity had a standard deviation that was distinct between LiDAR channels and similar between terrain classes in each LiDAR channel, therefore for this work LiDAR sparsity was considered a negligible issue.
+Six degree arcs (Fig.
+\ref{fig:area_example}) from the first three channels of scanning LiDAR data directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.
+Training data was extracted from a single arc from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area.
 Raw spatial and intensity values from the coincident arc were saved to a raw training database.
@@
 \ref{fig:training_data_cluster_2}).
-It was found that each class generally clustered to a certain range with some overlap, which indicates that a RDF may be trained.
-}
+It was found that each class generally clustered to a certain range with some overlap, which indicates that a RDF may be trained.}
 
@@
 				\centering
-				\includegraphics[width=0.25\linewidth]{figures/nine_arcs_example}
-				\caption[Areas to Classify]{Three arcs from three channels were classified per 360 degree scan.
-2C, 3C, and 4C were used exclusively for gathering training data, while the rest were included during testing.
-\textit{Note: Vehicle not to scale.}}
-				\label{fig:nine_arcs_example}
+				\includegraphics[width=0.25\linewidth]{figures/area_example}
+				\caption[Areas to Classify]{Three arcs from three channels were classified per 360 degree scan. }
+				\label{fig:area_example}
 			\end{figure}
@@
 				\includegraphics[width=0.75\linewidth]{figures/test_vs_train_areas_hatch}
-				\caption[Training vs Testing Areas]{Training areas were kept separate from the testing areas, shown is a top-down view of the road of interest along with most of the training and testing area.
-Some of the training area is off-image.
-Training data area and number of LiDAR points within the area has no bearing on the amount of training data used.}
+				\caption[Training vs Testing Areas]{Training areas were kept separate from the testing areas.}
 				\label{fig:test_vs_train_areas}
@@
 
-		\subsection{Surface Classification and Road Detection}
-
-%		It is acknowledged by the authors that the algorithm must classify as one of these three terrain types.
-
-			{Classifying consecutive LiDAR scans was accomplished by examination of nine arcs of interest (three arcs from three separate LiDAR channels) in front and forty-five degree angles left and right from the front of the experimental apparatus (Fig.
-\ref{fig:nine_arcs_example}).
+			{Classifying consecutive LiDAR scans was accomplished by examination of the nine arcs of interest in front and forty-five degree angles left and right from the front of the experimental apparatus (Fig.
+\ref{fig:area_example}).
 Features from each arc were extracted and the arc was classified (Fig.
@@
 Averaged x, y, and z coordinates were taken of each arc, and projected unto a manually defined truth area map (Fig.
-\ref{fig:visual_v_auto}).
+\ref{fig:rm_db_4_toc}).
 Accuracy scores was calculated based on the distribution of classified points with each area, allowing evaluation of unmarked road detection performance.
-Due to the arc locations (Fig.
-\ref{fig:nine_arcs_example}), IMU drift was considered a negligible factor for this work as IMU drift would need to be extreme in order to impact these results, in which case it would be easily found visually.
 Classified point clouds were visually inspected to determine if the manual detection of a road surface may be accomplished.
@@
 Guessed road surface areas were then compared to actual areas to determine the method's ability at unmarked gravel road discovery, indicating the method's efficiency at detecting intersecting gravel roads.
-Automated detection of unmarked road was explored in order to determine if visual detection results may be repeated.
-Class percentages, standard deviation of height, standard deviation of average [x,y] distance to the LiDAR point of origin, and distance trends between classified arcs among other features were exploited to determine if the area best describes a gravel, asphalt, or unknown surface (Fig.
-\ref{fig:visual_v_auto}).
-While a simple model, this will serve as a proof-of-concept for future development.
-Results were compiled into a single classified map and compared to truth areas, then scored based off the number of true/false positive road detection.
 }
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/db1range.png}
+				\includegraphics[width=0.9\linewidth]{figures/range_classified_pc_db_6}
 				\caption[Classified Point Cloud]{Classified Point Cloud.}
@@
 				\centering
-				\includegraphics[width=0.9\linewidth]{figures/visual_v_auto}
-				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed based on trends between channels and compared to actual road-surface areas (left).
-Autonomous projection of road surface areas following developed methods (right).}
-				\label{fig:visual_v_auto}
+				\includegraphics[width=0.9\linewidth]{figures/cross_channel_trend}
+				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed based on trends between channels and compared to actual road-surface areas.}
+				\label{fig:rm_db_4_toc}
 			\end{figure}
-
-%			\begin{figure}[H]
-%				\centering
-%				\includegraphics[width=0.9\linewidth]{figures/auto_guess_v_truth}
-%				\caption[Projected Guess over Time vs Truth]{Aggregated automated road projection based on classification results.}
-%				\label{fig:auto_guess_v_truth}
-%			\end{figure}
 
@@
 
-		{Training and verification LiDAR, GPS, and IMU data collection was completed using the Experimental Apparatus.
-Gravel data was gathered from the Redmen Lodge gravel parking lot, while grass data was gathered beside the lot.
-Asphalt data was gathered Blackburn Road.
-Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training; however, for this study the algorithm that produced the final results used multiple thousands of data sets per class.
+		{Five drive-bys of Blackburn Road, which has three intersecting gravel parking lots and driveways, were completed, and scanning LiDAR, GPS, and IMU data was collected.
 Training data was generated using MATLAB to extract LiDAR data from rosbags, and RDF classification algorithms were generated.
@@
 					Asphalt Road Detection 					& 100.00 	& \% True Pos. Detection Rate \\
-					Gravel Driveway Detection 				& 71.67 	& \% True Pos. Detection Rate \\
-					Misidentified Gravel Road Detection 	& 0 		&  Avg. False Pos. Detection Rate per Run\\
+					Gravel Driveway Detection 				& 93.33 	& \% True Pos. Detection Rate \\
+					Misidentified Gravel Road Detection 	& 0.60 		&  Avg. False Pos. Detection Rate per Run\\
 					\bottomrule
@@
 \ref{fig:dirt_v_gravel}).
-Higher LiDAR point density from a higher resolution LiDAR in conjunction with more descriptive features may yield better results, however for this was outside the scope of this work.
 Exploitation of confidence scores to adjust results by requiring a very high confidence for positive gravel surface classification was explored.
 All gravel surfaces that did not meet a 80\% confidence score was re-labeled as \textit{unknown}, however these adjustment had either a negligible or detrimental impact on the final results.
-Higher confidence score requirements resulted in \textit{unknown} over-classification, while too low resulted in \textit{gravel} over-classification.
-Overall it was found that an average true-positive accuracy of all intersecting gravel surfaces was 74.97\% (Fig.
+Higher confidence score requirements resulted in \textit{unknown} over-classification, while too low resulted in "gravel" over-classification.
+Overall it was found that an average true-positive accuracy of all intercepting gravel surfaces was 74.97\% (Fig.
 \ref{fig:prepostadjust}).
@@
 It is acknowledged by the authors that the classification algorithm must yield a result that most closely matches one of the trained classes, which led to lower accuracy scores in these circumstances.
-Results were compared to algorithms trained with smaller data sets in order to determine if the training data was sufficiently large for this specific scenario.
-Using 25\% and 50\% of total available training data yielded poor results compared to shown, however using 75\% of all available training data yielded results comparable to the shown results, indicating that the training data set is sufficient for this specific scenario.
+It is difficult to determine if these results are comparable to similar research, as found research classifies larger data sets.
+For example, work completed by Laible et.
+al.
+\cite{levi_3d_2012_light} used $\sim1200$ data points, compared to the $\sim28$ data points per arc in this work.
 Overall discovered accuracy levels may be sufficient for the purpose of identifying asphalt and gravel roads by considering trends between classified arcs.}
@@
 			\centering
-			\includegraphics[width=0.75\linewidth]{figures/range_actual_rm_db_4_area_score}
+			\includegraphics[width=0.75\linewidth]{figures/rm_db_1_area_score}
 			\caption[Area Scores]{Per-area scores were calculated, indicating algorithm unmarked road detection accuracy.
-Note that the asphalt area in front of the gravel driveway has significant \textit{gravel} classification due to construction.
+Note that the asphalt area in front of the gravel driveway has significant gravel classification due to construction.
 These results are included with the overall asphalt classification score}
@@
 Two out of three positive gravel identifications must be made on each channel, be evenly spaced, and be relatively co-planar to other road surface guesses in order for a positive guess for an asphalt or gravel surface to be made (Fig.
-\ref{fig:visual_v_auto}).
-Driveways $1$ and $2$ were detected in all five drive-bys, driveway $3$ was detected in four leading to an 91.67\% accuracy in intersecting gravel road detection.
+\ref{fig:rm_db_4_toc}).
+Driveways $1$ and $2$ were detected in all five drive-bys, driveway $3$ was detected in four leading to an 91.67\% accuracy in intercepting gravel road detection.
 It was found that in three drive-bys there was a false positive driveway detection, however the location of the false-positives were not repeated in any of the other drivebys.
 In all drive-bys the asphalt road was accurately detected.
-It is acknowledged that the author's a-priori knowledge of the gravel drive locations may have introduced a bias, however this method proved useful for developing an autonomous method.
-Trends across classified arcs were autonomously exploited to determine the best guess at area terrain type.
-Guessed areas were aggregated into a larger map then compared to truth areas (Fig.
-\ref{fig:guess_grav_intersect}).
-Using autonomous methods to project road surfaces, it was found that a true-positive intersecting gravel road detection rate of 71.66\% with no false positives.
-It was found that the true positive rate may be increased to 100\% at the expense of increasing the number of false positives with a 1.8 average number of false positives per drive-by.}
+}
 
-		\begin{figure}[H]
-			\centering
-			\includegraphics[width=0.95\linewidth]{figures/db1_conf_elim_auto_guess_mca}
-			\caption[Aggregated Automated Road Surface Detection]{Aggregated automated road surface detection was compared to truth areas. }
-			\label{fig:guess_grav_intersect}
-		\end{figure}
-
-		{Classification was completed using Ubuntu 18.04 and an AMD Ryzen 9 5900X 12 core 24 thread CPU.
+		{Classification was completed using Ubuntu 18.04 and an AMD Ryzen 9 5900X 12 core 24 thread.
 Rate of classification for all nine areas of interest in a single scan was found to be an average of 35.19 ms using a single core (Fig.
@@
 		\end{figure}
-
-		{It is difficult to determine if these results are comparable to similar research, as found research classifies using distinctly different sensor configurations,  algorithm input requirements, or have different objectives.
-For example, work completed by Laible et.
-al.
-\cite{levi_3d_2012_light} used a camera as well as LiDAR, compared to the LiDAR-only focus for this work.
-Work completed by Wang et.
-al.
-\cite{wang_road_nodate}, which found a true positive gravel terrain classification rate of 98.5\%; however, the method again used significantly more data from the LiDAR sensor compared to this work, exploiting all LiDAR channels that fall within a 1.3 meter region of interest.
-There is a general lack of open source LiDAR data for the express purpose of comparing road detection and terrain classification methods in an unstructured, rural environment.
-This is in stark contrast to open source data for urban environments.
-Furthermore, to the best of the author's knowledge there is no comparison between road detection methods using third-party LiDAR data of unmarked, unstructured gravel and asphalt roads in found literature.
-To the best of the author's knowledge, this study employed a novel approach for detecting road surface area and type.}
 
@@
 Classified point clouds were examined and gravel areas were projected and compared with actual gravel surface areas.
-Overlapping results between autonomously detected and actual road surface areas resulted in a minimum  71.66\% intersecting gravel road detection accuracy.
+Overlapping results between detected and actual road surface areas resulted in 91.67\% intersecting gravel road detection accuracy.
 In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is promising for future exploration of additional roads and road surfaces.
-Work as described addressed the problem of road surface detection on unmarked gravel and asphalt roads.
-The described work was a method of detecting physical unmarked gravel and asphalt roads by using a terrain classification approach to predicting road surface area.
-The impact of this work is that autonomous vehicles using the developed method may be able to detect unmarked road surfaces and differentiate between gravel and asphalt, allowing autonomous operations on previously undetected rural roads.}
+Work as described addressed the problem of road surface detection on unmarked gravel and chipseal roads.
+The described work was a method of detecting physical unmarked gravel and chipseal roads by using a terrain classification approach to predicting road surface area.
+The impact of this work is that autonomous vehicles using the developed method may be able to detect unmarked gravel and chipseal road surfaces, allowing autonomous operations on 1.5 million miles of previously undetected rural roads.}
 

### 2024-07-12 rebuttal_07092024.txt
added rebuttal wip

REVIEWER 1)

Please find my comments below:
1. The theory section of the paper is explained in depth and the technical contributions are also articulated in depth.
2. The figures however are very inconsistent. For example, figure 7 uses "Train Grav" in the legend whereas figure 8 uses Gravel. In other cases, asphalt is mentioned as "Asph" or "Asphalt". Please choose one terminology and keep it consistent across all figures.

	The authors made the figures more consistent.

3. The font size in the legends are highly inconsistent. Please set the graphic parameters to be the same.

	The authors made the figures more consistent.

4. The terms "grass" and "unknown/unkn" are used interchangeably. As earlier, it is rather confusing when each plot addresses these categories differently.

	The authors made the terminology more consistent.

5. The line color in Figure 9 is almost the same. Please use contrasting colors to distinguish the estimated min classification error and observed min classification error.

	The authors made the figure more legible.

6. Prior to the results section, the "dirt" category was not mentioned in the training phase. But it seems to appear only in the results section. Could the authors clarify the "dirt" category and how it is introduced in this experiment?

	Dirt was not used in the training phase. Rather, data was simply gathered and used to compare against gravel data. This point was clarified in the text.

7. Some citations are not referenced correctly in the introduction section, for example in the line "Furthermore, RDFs are less complex to train, able to be trained with smaller data sets, and are generally
easier to interpret [? ? ? ? ]."

	Citations have been fixed.

8. The references are not consistent. Reference 7 has the title in capital letters whereas the titles are in sentence case in others. Please keep all of the reference consistent in style.

	References have made more consistent.

Thank you for making the revisions and responding. Many of the reviewer’s previous comments were well-addressed. However, some points in the manuscript still require further clarification.

1) Please do provide the area size of each type of the surfaces in square meters. This provides important fact regarding the data used in this work.

	The authors do not believe that this information is pertinent information to add to the text, as it is not only irrelevant to the work, the potential implications and questions would need to be explained in text. The reasons for this were explained in a previous rebuttal. Unless the reviewer provides a reason as to why this information is valuable, the authors will not be adding this information to the work. As stated in the text: "As the training data is independent of the size of each designated training area and the number of LiDAR points in that area, it is not considered in this work".

2) The reviewer did not find the updates in the manuscript related the following comment from the previous round. Please highlight your changes in the manuscript or copy the changes to the response letter. Ideally both.
Figure 16 (“Side-of-road detection accuracy was…”) and the associated description in the Results section. The features the method used do not sufficiently separate the two classes does not mean the task itself is not possible. It could just be the features we engineered do not have sufficient descriptive power. Having this problem fully solved may be beyond the scope of this work. However, the authors could attempt to provide more insights into the nature of the problem. For example, why is this problem not feasible by its nature? Or, if it is possible, how could it be better solved (with a higher resolution lidar or some specific features, etc.)?

	The authors provided a diff file in the same pdf document as the rebuttal - changes in red are deletions, changes in blue are additions. Meanwhile the submitted document is un-annotated. This practice will continue, however any significant changes will be repeated in this rebuttal. As stated in the document, further distinction between the classes may be accomplished but it outside the scope of this present work. A higher resolution LiDAR may provide more detailed or consistent features with a higher point density.

3) Same as above, the reviewer did not find the updates in the manuscript related the following comment from the previous round. Please highlight your changes in the manuscript or copy the changes to the response letter. Ideally both.
In the Results section, “however, for this study, the algorithm that produced the final results used two thousand points of reference per class.” It is still somewhat confusing regarding the definition of “point.” Does a point mean a lidar point or a time instance that corresponds to, say, five lidar points of each lidar? Could you clarify this further in the manuscript?

	The submitted document does not contain the sentence "however, for this study, the algorithm that produced the final results used two thousand points of reference per class". This was revised as requested in the previous revisions, and the changes were highlighted in the annotated document: "Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training, however after thirty percent of the training data was randomly selected and reserved to create a validation data set, approximately two thousand samples of gravel, asphalt, and grass were used to train the algorithm". The authors believe this should alleviate any confusion.

4) Related work: “Lei et al suggest a method using an SVM for the detection of unstructured dirt and gravel roads using a 2D LiDAR and a camera [? ], however the method does not use a 3D scanning LiDAR and cameras may be insufficient for reliable road detection [19]. Fernandez-Delgado et al found that RDFs out-performed other machine learning algorithms, including NNs, when classifying 121 data sets taken from the UCI data base [? ]. Therefore while both Neural Networks and RDFs have been used to classify scanning LiDAR data, a RDF classifier was chosen. Furthermore, RDFs are less complex to train, able to be trained with smaller data sets, and are generally easier to interpret [? ? ? ? ].” Please correct all the [????] in the text. The same problem exists in other sections in the manuscript, e.g., Section Training Data Extraction and Section RDF Training and Testing.

	The authors fixed the citations.

5) Figure 9. Please use different line styles and drastically different colors for the two lines. Also, use a larger marker for the best point. Since the best hyper-parameters are chosen based on the minimum upper confidence interval of the classification error, the confidence interval should be included in the plot as well.

	The authors made the figure more legible.

6) Figure 10. Is the Matlab marker included in the plot by design? If so, please make it more visible. If not, please have it removed.

	Good catch! The authors removed the matlab marker.

7) Table I. Where is this table referenced in the text? Also, add a caption to describe how this table should be read. It is not clear what the three columns of information are currently.

	The table is now referenced in the text. The table was modified to be made more legible.

### 2024-07-12 Huston_Paper.tex

--- 2023-12-18 Huston_Paper.tex
@@
 				\includegraphics[width=0.75\linewidth]{figures/test_vs_train_areas_hatch}
-				\caption[Training vs Testing Areas]{Training areas were kept separate from the testing areas.}
+				\caption[Training vs Testing Areas]{Training areas were kept separate from the testing areas, shown is a top-down view of the road of interest along with most of the training and testing area.
+Some of the training area is off-image.
+Training data area and number of LiDAR points within the area has no bearing on the amount of training data used.}
 				\label{fig:test_vs_train_areas}

### 2024-07-15 LiDAR_Road_Detection_Paper.tex
latest version

--- 2024-07-10 Huston_Paper_07092024.tex
@@
 \usepackage{attachfile}
-%\usepackage{tabularray}
-
+\usepackage{tabularray}
+\hypersetup{
+     colorlinks=true,
+     linkcolor=black,
+     citecolor=black,
+     filecolor=black,
+     urlcolor=black,
+ } 
 % Table Stuff
@@
 Normal Distribution Transform (NDT) LiDAR scan matching may be insufficient for navigating on gravel roads as there may not be enough geometrically distinct features for reliable scan matching.
-Completed work will examine a method of classifying scanning LiDAR spatial and remission data features for explicit detection of unmarked gravel road surfaces.
-Exploration of terrain classification using high resolution scanning LiDAR data of these road surfaces may allow for predicting gravel road boundary locations potentially enabling confident autonomous operations on gravel roads.
+This paper examined a method of classifying scanning LiDAR spatial and remission data features for explicit detection of unmarked gravel road surfaces.
+Exploration of terrain classification using high resolution scanning LiDAR data of specific road surfaces may allow for predicting gravel road boundary locations potentially enabling confident autonomous operations on gravel roads.
 The principal outcome of this work was a method for gravel road terrain detection using LiDAR data for the purpose of predicting potential road boundary locations.
 Random Decision Forests were trained using scanning LiDAR data terrain classification to detect unmarked gravel and asphalt surfaces.
-It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively at an estimated rate of 13.19 ms per 360 degree scan.
-Overlapping results between manually projected and actual road surface areas resulted in 93.33\% intersecting gravel road detection accuracy.
-Automated post-process examination of classification results yielded an  true-positive gravel road detection rate of 71.67\%.
-Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on previously undetected roads.}
+It was found that a true-positive accuracy for gravel and asphalt surfaces was 75\% and 87\% respectively at an estimated rate of 13 ms per 360 degree scan.
+Overlapping results between manually projected and actual road surface areas resulted in 93\% intersecting gravel road detection accuracy.
+Automated post-process examination of classification results yielded an  true-positive gravel road detection rate of 72\%.}
 
@@
 
-	% Current methods of road surface analysis rely on cameras and Light Detection and Ranging (LiDAR) sensors to detect standardized road features, however on gravel roads such attributes are non-existent \cite{skorseth_gravel_nodate}.
+	% Current methods of road surface analysis rely on cameras and Light Detection and Ranging (LiDAR) sensors to detect standardized road features, however on gravel roads such attributes are non-existent \cite{skorseth2000gravel}.
 
-	{Gravel roads comprise 34.8\% \cite{road_stats_2} of all road surfaces in the United States.
-Some nations have predominately unpaved road networks, such as India, where 70.7\% of all roads by mile are categorized as unpaved \cite{malik_lal_2019}.
+	{Gravel roads comprise 34.8\% \cite{mbiyana2023literature} of all road surfaces in the United States.
+Some nations have predominately unpaved road networks, such as India, where 70.7\% of all roads by mile are categorized as unpaved \cite{medhi2015rural}.
 Detection of gravel roads may allow for expansion of current civilian autonomous vehicle capabilities as well as introduce opportunities for industries that operate in rural areas.
-Global Positioning Service (GPS) and high resolution maps should not be relied upon for navigation of rural roads alone as GPS is not accurate enough or may not have a reliable signal for precision driving \cite{noauthor_gpsgov_nodate}, and gravel roads are not marked or mapped.
+Global Positioning Service (GPS) and high resolution maps should not be relied upon for navigation of rural roads alone as GPS is not accurate enough or may not have a reliable signal for precision driving \cite{ort2018autonomous,walters2019rural}, and gravel roads are not marked or mapped.
 Systems that rely on cameras may fail when analyzing gravel roads, as these lack visual cues such as painted lane markings \cite{crisman_scarf_1993} and may in some circumstances closely resemble surrounding terrain.
-LiDAR point cloud processing systems depend upon distinct geometric features, most commonly curbs \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_nodate,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja_road_nodate}, however as curbs are not installed on rural gravel roads \cite{skorseth_gravel_nodate} this method breaks down.
+LiDAR point cloud processing systems depend upon distinct geometric features, most commonly curbs \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_2017,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja2010road}, however as curbs are not installed on rural gravel roads \cite{skorseth2000gravel} this method breaks down.
 Normal Distribution Transform (NDT) scan matching analytically compares two point cloud data sets to track current position, however this model relies on distinct geometric features, and suffers when distinguishing between terrain types making it insufficient for determining road boundaries and type \cite{biber_normal_2003}.
 Using 2D projection of planes unto a 3D point cloud of a road surface is an alternative approach, but likewise requires curbs or painted lines for road boundary definition \cite{fernandes_road_2014, borkar_robust_2009-1, guo_lane_2015}.
-Instead, using LiDAR may overcome difficulties in achieving detection of gravel roads by using terrain classification to predict road surface location and type.}
+Instead, processing LiDAR points that represent x, y, and z coordinates relative to the sensor may overcome difficulties in achieving detection of gravel roads by predicting road surface location and type using terrain classification.}
 
 	{Detection of unmarked road surfaces may utilize roughness properties without distinct markings or topographical features.
-Surface roughness is a measurable property that may be used to characterize a gravel road, as they are typically consistent with gravel type used \cite{skorseth_gravel_nodate} which may be exploited in a search and compare function, and is distinct from chipseal, grass, and other ground types \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain}.
-Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}.
+Surface roughness is a measurable property that may be used to characterize a gravel road, as they are typically consistent with gravel type used \cite{skorseth2000gravel} which may be exploited in a search and compare function, and is distinct from chipseal, grass, and other ground types \cite{wan_road_2007, laible20123d}.
+Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, laible20123d, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}.
 Implementing a Random Decision Forests (RDF) terrain classification model using subsets of a single 360 degree scan of LiDAR data to exploit surface properties was studied to determine the method's feasibility in road surface detection, without using a-priori knowledge of the road, additional sensors, nor memory of previous classifications.
@@
 
-	{Processing rural gravel roads cannot rely on normalized, consistent features found on urban roads such as curbs, as rural roads lack these features \cite{skorseth_gravel_nodate}.
-As such, many of the proposed models are rendered unserviceable, as they rely on curbs or painted lines to offer distinctions in data sets \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_nodate,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja_road_nodate}.}
+	{Processing rural gravel roads cannot rely on normalized, consistent features found on urban roads such as curbs, as rural roads lack these features \cite{skorseth2000gravel}.
+As such, many of the proposed models are rendered unserviceable, as they rely on curbs or painted lines to offer distinctions in data sets \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_2017,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja2010road}.}
 
@@
 	{Classification of the surrounding environment is an integral part to the proposed work, as differentiation of gravel road surfaces and surrounding terrain is necessary to determine road boundaries and type.
-Terrain classification is the analysis of the ground surface in order to specify the ground surface type \cite{laible_3d_2012,laible_terrain_2013,laible_map_building,rasmussen_combining_2002,reymann_improving_2015,walas_terrain_2014,wietrzykowski_boosting_2014,wang_road_nodate}.
-LiDAR and camera based real-time terrain classification was studied by projecting a 2D plane unto the point cloud data using a variant of Random Sample Consensus (RANSAC) called M-estimator Sample Consensus (MSAC) \cite{mijakovska_generating_2014} that is built for a more robust result \cite{laible_3d_2012,laible_map_building,laible_terrain_2013}.
+Terrain classification is the analysis of the ground surface in order to specify the ground surface type \cite{laible_3d_2012,laible_terrain_2013,laible2014building,rasmussen_combining_2002,reymann_improving_2015,walas_terrain_2014,wietrzykowski_boosting_2014,wang_road_2012}.
+LiDAR and camera based real-time terrain classification was studied by projecting a 2D plane unto the point cloud data using a variant of Random Sample Consensus (RANSAC) called M-estimator Sample Consensus (MSAC) \cite{mijakovska_generating_2014} that is built for a more robust result \cite{laible_3d_2012,laible2014building,laible_terrain_2013}.
 After segmenting the plane into cells, roughness and intensity histograms were analyzed on a per-cell basis.
-Markov Random Field (MRF) \cite{chellappa_classification_1985} and Conditional Random Field (CRF) \cite{wallach_conditional_nodate} were compared in their ability to compare individual cells with neighboring cells.
+Markov Random Field (MRF) \cite{chellappa_classification_1985} and Conditional Random Field (CRF) \cite{wallach_conditional_2004} were compared in their ability to compare individual cells with neighboring cells.
 Comparison with neighboring cells increases classification type accuracy, as adjacent cell are likely to have the same terrain type \cite{haselich_terrain_2011,zhao_fusion_2014}.
-It was found that CRF was better suited for classification, however this work did not use the method to study road projection and used cameras which may not be reliable for terrain classification \cite{levi_3d_2012_light}.}
+It was found that CRF was better suited for classification, however this work did not use the method to study road projection and used cameras which may not be reliable for terrain classification \cite{laible20123d}.}
 
@@
 It was found that characterizing the LiDAR intensity as Gaussian distributed noise yielded poor results with only a 49.5\% true positive rate, however a low-resolution LiDAR was used \cite{rauscher_comparison_2016} and this method was not employed in road identification.
-Higher resolution LiDAR that produces a high density point cloud, such as the Velodyne VLP-32 that will be used in the proposed work \cite{vlp_32c}, may allow for better terrain classification.
+Higher resolution LiDAR that produces a high density point cloud, such as the Velodyne VLP-32 that will be used in the proposed work, may allow for better terrain classification.
 Edmond et al.
@@
 Obstacle detection using a single axis LiDAR was studied \cite{manduchi_obstacle_2005}; however, this method could not detect road surfaces, only obstacles in roads.
-Terrain classification using cameras for use in off-road robots was studied \cite{walch_offroad_2022}; however, the method focuses on terrain traversability, and ambient conditions may lower classification accuracy \cite{levi_3d_2012_light}.
+Terrain classification using cameras for use in off-road robots was studied \cite{walch_offroad_2022}; however, the method focuses on terrain traversability, and ambient conditions may lower classification accuracy \cite{laible20123d}.
 }
@@
 	{Neural Networks and similar machine learning tools may be used in terrain and road identification.
-Road classification with LiDAR spacial and remission features was completed with Neural Network classifier, Naive Bayes classifier, and Support Vector Machines (SVM) \cite{wang_road_nodate,wang_two-stage_2018}.
+Road classification with LiDAR spacial and remission features was completed with Neural Network classifier, Naive Bayes classifier, and Support Vector Machines (SVM) \cite{wang_road_2012,wang_two-stage_2018}.
 Gravel surfaces were successfully identified with a 98.5\% true positive identification rate; however, the method was only focused on classifying road surface type directly behind the vehicle, and was not tested for informing vehicle trajectory based on forward-projected road surface areas.
-MATLAB's Neural Network tools were used in autonomous road identification using LiDAR and Camera data \cite{rasmussen_combining_2002}; however, cameras may not be relied upon for ambient conditions may affect classification accuracy \cite{levi_3d_2012_light}.
+MATLAB's Neural Network tools were used in autonomous road identification using LiDAR and Camera data \cite{rasmussen_combining_2002}; however, cameras may not be relied upon for ambient conditions may affect classification accuracy \cite{laible20123d}.
 Terrain classification was completed using Support Vector Machine (SVM), a classification algorithm where a line is drawn between two different categories to differentiate between them \cite{wietrzykowski_boosting_2014}.
-Histograms and averages for image hue, saturation, and color value, along with LiDAR intensity were used as points of classification for different terrain types.
+Histograms and averages for image hue, saturation, and color value, along with LiDAR intensity were used as reference samples for the classification of different terrain types.
 Although a 96\% true positive rate was reported for a rocky surface, the process was not tested for road detection.
-Lei et al suggest a method using an SVM for the detection of unstructured dirt and gravel roads using a 2D LiDAR and a camera \cite{lei_detection_2021}, however the method does not use a 3D scanning LiDAR and cameras may be insufficient for reliable road detection \cite{levi_3d_2012_light}.
+Lei et al suggest a method using an SVM for the detection of unstructured dirt and gravel roads using a 2D LiDAR and a camera \cite{lei_detection_2021}, however the method does not use a 3D scanning LiDAR and cameras may be insufficient for reliable road detection \cite{laible20123d}.
 Fernandez-Delgado et al found that RDFs out-performed other machine learning algorithms, including NNs, when classifying 121 data sets taken from the UCI data base \cite{fernandez2014we}.
@@
 	{RDFs are a supervised machine learning method that operates on a set of decision trees \cite{ho_random_1995}.
-RDFs have been shown to be useful with handling LiDAR data sets \cite{breiman_random_2001}, and have been employed in terrain classification \cite{laible_3d_2012,laible_map_building,laible_terrain_2013,khan_high_2011,reymann_improving_2015,schilling_geometric_2017, wietrzykowski_context-aware_2019}.
+RDFs have been shown to be useful with handling LiDAR data sets \cite{breiman_random_2001}, and have been employed in terrain classification \cite{laible_3d_2012,laible2014building,laible_terrain_2013,khan_high_2011,reymann_improving_2015,schilling_geometric_2017, wietrzykowski_context-aware_2019}.
 Multiple decision trees are created with bootstrapped data sets - randomly selected data sub-sets from a training set.
@@
 While RDFs were used in terrain classification in multiple instances in found literature, to the author's best knowledge they have not been employed in predicting road surface areas and type while also only using small portions of a single LiDAR scan.
-Bayesian Optimization may be used to tune the hyper parameters of a machine algorithm \cite{noauthor_bayesian_nodate, snoek_practical_2012}.
+Bayesian Optimization may be used to tune the hyper parameters of a machine algorithm \cite{snoek2012}.
 }
@@
 
-		{Detection of unmarked gravel road surfaces using terrain classification requires collecting data, training an RDF, and testing the algorithm for accuracy of unmarked road detection (outline of work in Fig.
-\ref{fig:flowz_8}).
-Training data was gathered by the Ohio University Autonomous Van (Fig.
-\ref{fig:Experimental_Apperatus}).
+		{Detection of unmarked gravel road surfaces using terrain classification requires collecting data, training an RDF, and testing the algorithm for accuracy of unmarked road detection (outline of work in Figure \ref{fig:flowz_8}).
+Training data was gathered by the Ohio University Autonomous Van shown in Figure \ref{fig:Experimental_Apperatus}.
 MATLAB was used to extract and parse LiDAR and GPS data.
@@
 
-%	{The experimental apparatus for this work was the Ohio University Autonomous Van van (Fig.
-\ref{fig:Experimental_Apperatus}), which has a Velodyne VLP-32C scanning LiDAR sensor, Novatel PwrPak 7D-E2 GNSS and INS system, and Mako cameras.
+%	{The experimental apparatus for this work was the Ohio University Autonomous Van van (Figure \ref{fig:Experimental_Apperatus}), which has a Velodyne VLP-32C scanning LiDAR sensor, Novatel PwrPak 7D-E2 GNSS and INS system, and Mako cameras.
 Raw data was gathered using the Robotic Operating System (ROS) to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags.
@@
 Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data.
-Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan (Fig.
-\ref{fig:road_areas_annotated_22}).
+Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan (Figure \ref{fig:road_areas_annotated_22}).
 Compiled point cloud maps was examined and compared to camera and satellite data to determine gravel and asphalt surface location.
-Manually defined 2D areas representing gravel and asphalt were projected unto the point cloud (Fig.
-\ref{fig:road_areas_annotated_22}).}
+Manually defined 2D areas representing gravel and asphalt were projected unto the point cloud (Figure \ref{fig:road_areas_annotated_22}).}
 
-			{Required physical data was gathered by the Ohio University Autonomous Van (Fig.
-\ref{fig:Experimental_Apperatus}) on a single sunny, cloudless day with no road obstacles or debris at a speed below 15 miles per hour.
+			{LiDAR, GPS, and IMU data was gathered by the Ohio University Autonomous Van, shown in Figure \ref{fig:Experimental_Apperatus}, on a single sunny, cloudless day with no road obstacles or debris at or below a speed of 15 miles per hour.
 Data collection was at mid-day, which minimized the risk of shadows from the road-side trees interfering with the data collection.
 Vehicle speed was limited to maximize the amount of data collected and to minimize any effect of IMU drift.
-It was assumed that vehicle speed did not affect the gathered data and is considered outside the scope of this work.
-It is acknowledged that weather or non-ideal road conditions may affect results and is an area of future study.
 LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C.
-Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a minimum 0.33$^{\circ}$ vertical angular resolution and 0.2$^{\circ}$ horizontal angular resolution \cite{vlp_32c}.
+The LiDAR sensor utilized 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a minimum 0.33$^{\circ}$ vertical angular resolution and 0.2$^{\circ}$ horizontal angular resolution.
 GPS data was recorded with a Novatel PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS), and two GNSS-502 antennas by NavtechGPS.
-GNSS was assisted with Real Time Kinematic correction data from the Ohio Continuously Operating Reference Station (CORS) and Ohio Real Time Network (RTN) \cite{noauthor_cors_nodate}, yielding in our tests sub 2cm location accuracy level.
-IMU specifications indicate $0.8 ^{\circ}/hr$ bias instability, $0.06 ^{\circ}/\sqrt{hr}$ angular random walk, and $0.025 m/s/\sqrt{hr}$ velocity random walk, in addition to a Extended Kalman Filter \cite{noauthor_pwrpak7_nodate}.
+GNSS was assisted with Real Time Kinematic correction data from the Ohio Continuously Operating Reference Station (CORS) and Ohio Real Time Network (RTN) \cite{grejner2009network,wielgosz2005high}, yielding in our tests sub 2cm location accuracy level.
+IMU specifications indicate $0.8 ^{\circ}/hr$ bias instability, $0.06 ^{\circ}/\sqrt{hr}$ angular random walk, and $0.025 m/s/\sqrt{hr}$ velocity random walk, in addition to an Extended Kalman Filter provided by the Novatel PwrPak 7.
 Because the IMU was only used to orientate the individual LiDAR scans, IMU drift was not considered a factor that would affect the classification results themselves.
-It is assumed that the GNSS was not affected by the noise generated by the other electronic equipment as the antennas are located outside the vehicle, and is considered outside the scope of this work.
 Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using the Robot Operating System (ROS) to be subsequently examined and categorized later using MATLAB.
 Training data collection location was chosen for the clear distinction between gravel and paved surfaces, thus minimizing the risk of class confusion during the training process.
-Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot (Fig.
-\ref{fig:gravel_training_lot}) near Athens Ohio, which was freshly re-surfaced and re-graded.
-While there are a variety of gravel sizes, differentiating between multiple gravel sizes is outside the scope of this work.
-Furthermore, gravel roads are specified to have a consistent type and size \cite{skorseth_gravel_nodate}, therefore it is assumed that a single source for gravel data will be sufficient for demonstrating the ability of this work's method for gravel road detection.
-Two additional driveways intersected the main asphalt road, and while they had the same gravel type they differed from the training lot.
-Condition of one was severely segregated, however the other was in excellent condition.
-Road degradation and debris were not considered and were outside the scope of this work.
-Scanning LiDAR data of an asphalt surface was gathered from Blackburn Road (Fig.
-\ref{fig:Blackburn_Road_View}) which provides access to the training gravel lot.
+Total data collection area encompassed approximately 11,400 square meters.
+Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot (Figure \ref{fig:gravel_training_lot}) near Athens Ohio, which was freshly re-surfaced and re-graded.
+Gravel roads are specified to have a consistent type and size \cite{skorseth2000gravel}, therefore it is assumed that a single source for gravel data will be sufficient for demonstrating the ability of this work's method for gravel road detection.
+Two additional driveways intersected the main asphalt road, and while they had the same gravel type they differed from the training lot in terms of surface degradation.
+Classifying road degradation and debris were not considered for this effort.
+Scanning LiDAR data of an asphalt surface was gathered from Blackburn Road in Athens, Ohio (Figure \ref{fig:Blackburn_Road_View}) which provided access to a training gravel lot.
 Due to the aforementioned re-surfacing and re-grading of the gravel lot, some gravel was in front of the lot entrance on the road.
@@
 Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot.
-Although this work does not evaluate the detection accuracy of grass specifically, differentiation between a road and non-road surface is necessary for evaluating road-surface detection; therefore, grass is labeled as \textit{unknown} in this work.}
+Although this did not evaluate the detection accuracy of grass specifically, differentiation between a road and non-road or unknown surface is necessary for evaluating road-surface detection; therefore, grass is labeled as \textit{unknown} in this work.}
 
@@
 
-			{Scanning LiDAR training data requires extraction from manually defined \textit{gravel}, \textit{asphalt}, and \textit{unknown} terrain areas on aggregated point clouds.
+			{Scanning LiDAR training data required extraction from manually defined \textit{gravel}, \textit{asphalt}, and \textit{unknown} terrain areas on aggregated point clouds.
 Aggregating LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road.
-Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data.
+Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data that never varied more than 10 ms.
 Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan.
@@
 Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU.
-LiDAR origin was found using GPS longitude, latitude, and altitude data.
+The LiDAR origin was found using GPS longitude, latitude, and altitude data.
 Physical distances between the GPS, IMU, and LiDAR sensors were rectified by obtaining reference frames provided by AutonomouStuff.
@@
 GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors.
-Consecutive point cloud scans were then translated and rotated with derived transformation matrices (Fig.
-\ref{fig:pc_example}).
-While not as robust as more sophisticated methods of point cloud aggregation due to IMU drift, such as NDT or ICP scan matching, using this method proved to be adequate over shorter distances given the reasons previously mentioned.}
+Consecutive point cloud scans were then translated and rotated with derived transformation matrices (as shown in Figure \ref{fig:pc_example}) using GPS and IMU data to determine point of origin and orientation.
+While not as robust as more sophisticated methods of point cloud aggregation due to IMU drift, such as NDT or ICP scan matching, this method proved to be successful over shorter distances.}
 
@@
 				\includegraphics[width=0.75\linewidth]{figures/combined_pcd_example}
-				\caption[Aggregated Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.
-See Area 1 in Fig.
-\ref{fig:test_vs_train_areas}.}
+				\caption[Aggregated Point Cloud Data]{Point cloud of scanning LiDAR data from Area 1}
 				\label{fig:pc_example}
@@
 
-			{Discretization of scanning LiDAR data is required for training data extraction and future classification.
+			{Discretization of scanning LiDAR data was required for training data extraction and future classification.
 For this work, arcs of six degrees centered in front of the vehicle were designated, each arc only containing points from one of the thirty two scanning LiDAR channels.
-For this work, only the closest three channels were considered in order to minimize the affect of LiDAR sparsity (Fig \ref{fig:nine_arcs_example}).
-Later, nine arcs (three arcs from the three channels) will be classified for the prediction of intersecting roads (Fig \ref{fig:nine_arcs_example}).
+Only the closest three channels were considered in order to minimize the affect of LiDAR sparsity, see Figure \ref{fig:nine_arcs_example}.
+Later, nine arcs (three arcs from the three channels) were classified for the prediction of intersecting roads (Figure \ref{fig:nine_arcs_example}) due to a need for increased data.
 Arc length was chosen based on approximate vehicle width (set to 6 degrees), and the location of each arc was fixed relative to the center line (length-wise) of the vehicle.
-Training data was extracted from a single arc, with approximately 28 LiDAR points per arc, from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area (Fig.
-\ref{fig:test_vs_train_areas}).
-As the training data is independent of the size of each designated training area and the number of LiDAR points in that area, it is not considered in this work.
-As arc is fixed to 6 degrees, the number of data points is constant, thus the only factor contributing to point density would be potential elevation changes between the vehicle and arc.
-This study found that the training data sparsity had a standard deviation that was distinct between LiDAR channels and similar between terrain classes in each LiDAR channel, therefore for this work LiDAR sparsity was considered a negligible issue.
-Raw spatial and intensity values from the coincident arc were saved to a raw training database.
-Separate training databases were maintained for each individual channel.
-Features that describe the raw spatial and intensity data were extracted.
+Training data was extracted from a single arc, with approximately 28 LiDAR points per arc, from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area shown in Figure \ref{fig:test_vs_train_areas}.
+As the number of \textit{gravel}, \textit{asphalt}, and \textit{unknown} samples used in training the RDF data were independent of the size of each designated training area and the number of LiDAR points in that area, it was not considered in this work.
+As the arc was fixed to 6 degrees, the number of data LiDAR points is constant, thus the only factor contributing to point density would be potential elevation changes between the vehicle and arc.
+This study found that the training data sparsity had a standard deviation that was distinct between LiDAR channels and similar between terrain classes in each LiDAR channel, LiDAR sparsity was considered a negligible issue.
+Raw spatial and intensity values from the coincident arc were saved to a raw training database to separate datasets for each individual channel.
+Features that describe the raw spatial and intensity data were then extracted.
 Thirty percent of the training data base was randomly selected and set aside as a validation data base.
-Training data was visually examined for clustering to verify that a RDF would be able to be trained (Fig.
-\ref{fig:training_data_cluster_2}).
-It was found that each class generally clustered to a certain range with some overlap, which indicates that a RDF may be trained.
-}
+Training data was visually examined for clustering to verify that a RDF would be able to be trained shown in Figure \ref{fig:training_data_cluster_2}.
+It was found that each class generally clustered to a certain range with some overlap, which indicates that a RDF may be trained.}
 
@@
 				\centering
-				\includegraphics[width=0.25\linewidth]{figures/nine_arcs_example}
+				\includegraphics[width=0.35\linewidth]{figures/nine_arcs_example_2.png}
 				\caption[Areas to Classify]{Three arcs from three channels were classified per 360 degree scan.
@@
 				\centering
-				\includegraphics[width=0.95\linewidth]{figures/test_vs_train_areas_hatch_annotated}
-				\caption[Training vs Testing Areas]{Training areas were kept separate from the testing areas.
-While not all are shown, the majority shown here is representative of the bulk of the areas.
-Area 1: Refers to area encompassing plots found in Figs.
-\ref{fig:pc_example}, \ref{fig:raw_classification_results}, \ref{fig:auto_guess_v_truth}, \ref{fig:guess_grav_intersect}.
-Area 2: Refer to area encompassing plots found in Figs.
-\ref{fig:rm_db_6_overlap_2}, \ref{fig:auto_area_guess}, \ref{fig:prepostadjust}
+				\includegraphics[width=0.75\linewidth]{figures/test_vs_train_areas_hatch_annotated_v4}
+				\caption[Training vs Testing Areas]{Training and validation zones, top-down view of the areas.  
 				}
@@
 			\end{figure}
-
+			%Area 3: Refers to location where data was gathered in dirt vs gravel comparison in Fig \ref{fig:dirt_v_gravel}.
 %			\begin{figure}[H]
@@
 			\centering
-			\includegraphics[width=0.75\linewidth]{figures/training_data_cluster_5}
+			\includegraphics[width=0.75\linewidth]{figures/training_data_cluster_6}
 			\caption[Example Clustering]{Example of training data comparison between two features.}
@@
 			{LiDAR returns spatial data represented by Cartesian coordinates with units in meters and remission data represented by a dimensionless ratio of minimum to maximum detected brightness.
-No pre-processing of LiDAR data as extracted from the rosbags was required, however the Cartesian coordinates were converted to Polar by using the Pythagorean Theorem to determine range ($sqrt(x^{2} + y^{2} + z^{2})$) and simple trigonometry to determine the azimuth ($atan2(y,x)$).
-Feature extraction was completed by performing mathematical functions on the gathered spatial training data, yielding the Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial data array.
+No pre-processing of LiDAR data as extracted from rosbags was required, however the Cartesian coordinates were converted to Polar by using the Pythagorean Theorem to determine range ($sqrt(x^{2} + y^{2} + z^{2})$) and simple trigonometry to determine the azimuth ($atan2(y,x)$).
+Feature extraction was completed by performing mathematical functions on the gathered spatial training data, yielding the Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive LiDAR points in the spatial data array.
 Spatial features were non-dimensionalized as necessary by dividing the average range of the arc to LiDAR point of origin to the appropriate power.
@@
 Feature selection closely mimics those suggested in literature \cite{reymann_improving_2015}.
-Future work may be completed to introduce additional features.
+Future work may be completed to introduce additional features, collect data during all hours of the day, at high vehicle speeds, and different weather conditions.
 Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training, however after thirty percent of the training data was randomly selected and reserved to create a validation data set, approximately two thousand samples of \textit{gravel}, \textit{asphalt}, and \textit{unknown} were used to train the algorithm.}
@@
 Listed are the ones available using MATLAB's Classification Learner Application that were able to be tuned using the Bayesian Optimization function.
-Bayesian Optimization automates the manual hyper-parameter tuning process by creating a surrogate model representing the objective function, in this case the upper confidence interval of the classification error objective model (Fig.
-\ref{fig:c2_min_class_error}).
+Bayesian Optimization automates the manual hyper-parameter tuning process by creating a surrogate model representing the objective function, in this case the upper confidence interval of the classification error objective model.
 Initialization of the objective function is completed by training a number of RDFs using randomly selected hyper-parameters.
@@
 Hyperparameter tuning may lead to over-fitting to the training data base.
-Testing for over-fitting was accomplished by comparing the out-of-bag training error to the validation error (Fig.
-\ref{fig:train_vs_valid_overfit_test2}).
-Confusion matrices provide validation database classification accuracy information by classifying each data set in the validation database and comparing actual and predicted classes, resulting in a matrix that visualizes the accuracy for each terrain type as well as which terrain types are more likely to be confused for a different class (Fig.
-\ref{fig:vali_err_conf_mat}).
+Testing for over-fitting was accomplished by comparing the out-of-bag training error to the validation error as shown in Figure \ref{fig:train_vs_valid_overfit_test2}.
+Confusion matrices provide validation database classification accuracy information by classifying each data set in the validation database and comparing actual and predicted classes, resulting in a matrix that visualizes the accuracy for each terrain type as well as which terrain types are more likely to be confused for a different class shown in Figure \ref{fig:vali_err_conf_mat}.
 This process was completed for each of the three considered scanning LiDAR channels used for classification.}
+
+			% \begin{figure}[H]
+			% 	\centering
+			% 	\includegraphics[width=0.75\linewidth]{figures/c2_min_class_error_2}
+			% 	\caption[RDF Training Classification Error]{Classification error during Bayesian Optimization tuning.
+Best hyper-parameters are chosen based on minimum upper confidence interval of the classification error objective model, not minimum classification error.}
+			% 	\label{fig:c2_min_class_error}
+			% \end{figure}
 
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/c2_min_class_error_2}
-				\caption[RDF Training Classification Error]{Classification error during Bayesian Optimization tuning.
-Best hyper-parameters are chosen based on minimum upper confidence interval of the classification error objective model, not minimum classification error.}
-				\label{fig:c2_min_class_error}
-			\end{figure}
-
-			\begin{figure}[H]
-				\centering
-				\includegraphics[width=0.75\linewidth]{figures/train_vs_valid_overfit_test_2}
-				\caption[Training vs Validation Error]{Hyper-parameter tuning may lead to over-fitting.
-Testing out-of-bag versus validation error demonstrates that the model is not over-fitting to the training data base if the validation error does not diverge from the out-of-bag error.}
+				\includegraphics[width=0.75\linewidth]{figures/train_vs_valid_overfit_test_5.png}
+				\caption[Training vs Validation Error]{Training and validation error for increasing number of trees.}
 				\label{fig:train_vs_valid_overfit_test2}
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/chan_2c_conf_VALIDATION_mat222}
+				\includegraphics[width=0.75\linewidth]{figures/true_v_predicted_class_matrix_v2.png}
 				\caption[Validation Error]{Confusion matrix with the validation data set.}
@@
 			\end{figure}
-
+        \newpage
 		\subsection{Surface Classification and Road Detection}
@@
 
-			{Classifying consecutive LiDAR scans was accomplished by examination of nine arcs of interest (three arcs from three separate LiDAR channels) in front and forty-five degree angles left and right from the front of the experimental apparatus (Fig.
-\ref{fig:nine_arcs_example}).
-Features from each arc were extracted and the arc was classified (Fig.
-\ref{fig:raw_classification_results} shows the average spatial coordinates of each classified arc).
-Previously-created manually defined gravel, asphalt, and side-of-road truth areas (denoted as Testing areas in Fig.
-\ref{fig:test_vs_train_areas}) were projected onto the classified point cloud for the determination of classification accuracy.
-Averaged x, y, and z coordinates were taken of each arc, and projected unto a manually defined truth area map (Fig.
-\ref{fig:rm_db_6_overlap_2}).
-Accuracy scores was calculated based on the distribution of classified points with each area, allowing evaluation of unmarked road detection performance.
-Due to the arc locations (Fig.
-\ref{fig:nine_arcs_example}), IMU drift was considered a negligible factor for this work as IMU drift would need to be extreme in order to impact these results, in which case the results would be easily recognized as having experienced significant drift.
-Classified point clouds were visually inspected to determine if the manual detection of a road surface may be accomplished.
-If consecutive points from each channel in any direction had two or more matching gravel or asphalt classification and was evenly spaced, a gravel or road surface area was projected.
+			{Classifying consecutive LiDAR scans was accomplished by examination of nine arcs of interest (three arcs from three separate LiDAR channels) in front and forty-five degree angles left and right from the front of the experimental apparatus as shown in Figure \ref{fig:nine_arcs_example}.
+and the average spatial coordinates of each classified arc shown in Figure \ref{fig:raw_classification_results}.
+Previously-created manually defined gravel, asphalt, and side-of-road truth areas were denoted as Testing areas and identified in Figure \ref{fig:test_vs_train_areas}.
+These areas were projected onto the classified point cloud for the determination of classification accuracy.
+Averaged x, y, and z coordinates were taken of each arc, and projected unto a manually defined truth area map as shown in Figure \ref{fig:rm_db_6_overlap_2}.
+Accuracy scores were calculated based on the distribution of classified points with each area, allowing evaluation of unmarked road detection performance.
+Due to the arc locations defined in Figure \ref{fig:nine_arcs_example}, IMU drift was considered a negligible factor for this work as IMU drift would need to be extreme in order to impact these results, in which case the results would be easily recognized as having experienced significant drift.
+Classified LiDAR point clouds were visually inspected to determine if the manual detection of a road surface may be accomplished.
+If consecutive LiDAR points from each channel in any direction had two or more matching gravel or asphalt classification and was evenly spaced, a gravel or road surface area was projected.
 Guessed road surface areas were then compared to actual areas to determine the method's ability at detecting unmarked asphalt and gravel roads.
 Automated detection of unmarked road was explored in order to determine if visual detection results may be repeated.
-Class percentages, standard deviation of height, standard deviation of average [x,y] distance to the LiDAR point of origin, and distance trends between classified arcs among other features were exploited to determine if the area best describes a gravel, asphalt, or unknown surface (Fig.
-\ref{fig:auto_area_guess}).
+Class percentages, standard deviation of height, standard deviation of average [x,y] distance to the LiDAR point of origin, and distance trends between classified arcs among other features were exploited to determine if the area best describes a gravel, asphalt, or unknown surface with results shown in Figure \ref{fig:auto_area_guess}.
 While a simple model, this will serve as a proof-of-concept for future development.
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/db1range_3.png}
-				\caption[Classified Point Cloud]{Classified Point Cloud. See Area 1 in Fig. \ref{fig:test_vs_train_areas}.}
+				\includegraphics[width=0.75\linewidth]{figures/db1range_4.png}
+				\caption[Classified Point Cloud]{Classified point cloud from area 1.}
 				\label{fig:raw_classification_results}
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/rm_db_6_overlap_2}
-				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed based on trends between channels and compared to actual road-surface areas (left).
-Autonomous projection of road surface areas following developed methods (right).
-See Area 2.
-in Fig.
-\ref{fig:test_vs_train_areas}.}
+				\includegraphics[width=0.75\linewidth]{figures/rm_db_6_overlap_3.png}
+				\caption[Projected Guess vs Truth]{Projected guess vs truth for area 2.}
 				\label{fig:rm_db_6_overlap_2}
@@
 				\centering
-				\includegraphics[width=0.75\linewidth]{figures/auto_area_guess_4}
-				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed based on trends between channels and compared to actual road-surface areas (left).
+				\includegraphics[width=0.75\linewidth]{figures/auto_area_guess_6.png}
+				\caption[Projected Guess vs Truth]{Projected guess vs truth for area 2.}
+    %Gravel and asphalt surfaces were automatically guessed based on trends between channels and compared to actual road-surface areas (left).
 Autonomous projection of road surface areas following developed methods (right).
 See Area 2.
-in Fig.
-\ref{fig:test_vs_train_areas}.}
+in Figure \ref{fig:test_vs_train_areas}.}
 				\label{fig:auto_area_guess}
@@
 
-			\begin{figure}[H]
-				\centering
-				\includegraphics[width=0.75\linewidth]{figures/auto_guess_v_truth_2}
-				\caption[Projected Guess over Time vs Truth]{Aggregated automated road projection based on classification results. See Area 1 in Fig. \ref{fig:test_vs_train_areas}.}
-				\label{fig:auto_guess_v_truth}
-			\end{figure}
+			% \begin{figure}[H]
+			% 	\centering
+			% 	\includegraphics[width=0.75\linewidth]{figures/aggregated_auto_area_guess_v2.png}
+			% 	\caption[Projected Guess over Time vs Truth]{Aggregated automated road projection based on classification results. See Area 1 in Figure \ref{fig:test_vs_train_areas}.}
+			% 	\label{fig:auto_guess_v_truth}
+			% \end{figure}
 
@@
 Asphalt data was gathered Blackburn Road.
-Found research did not make any conclusive arguments for the proper training data set size for Random Decision Forest training; however, for this study the algorithm that produced the final results used two thousand points of reference per class.
 Training data was generated using MATLAB to extract LiDAR data from rosbags, and RDF classification algorithms were generated.
@@
 Translation and rotation was applied to the averaged x, y, and z coordinates of each arc's terrain classification results.
-Consecutive scanning LiDAR data was projected on manually classified truth areas representing road surfaces (Figure \ref{fig:prepostadjust}) and percentages of terrain types per area were calculated.}
-
-		\begin{table}[t]
-
-			\centering{%
-				\begin{tabular}{lrl}
-					\toprule
-					Gravel Areas \% 						& 74.97 	& \% Gravel \\
-					\cline{1-3}
-					Asphalt Area \%							& 87.05   	& \% Asphalt \\
-					\cline{1-3}
-					Side-of-Road \% 						& 77.47 	& \% Unknown\\
-					\specialrule{1pt}{0pt}{0pt}
-					Asphalt Road Detection 					& 100.00 	& \% True Pos. Detection Rate \\
-					\cline{1-3}
-					Gravel Driveway Detection 				& 71.67 	& \% True Pos. Detection Rate \\
-					\cline{1-3}
-					Misidentified Gravel Road Detection 	& 0 		& \% False Pos. Detection Rate\\
-					\bottomrule
-				\end{tabular}
-			}
-			\caption[Table]{Overall results for presented work.
-74.97\% of all samples were classified as \textit{gravel} inside all \textit{gravel} areas, with an overall 71.67\% true positive \textit{gravel} road detection rate and a 0\% false positive \textit{gravel} road detection rate.\label{tab:1}}
-		\end{table}
-
-		{Classification of real-world data outside of the training data set presents difficulties to the generated Random Decision Forest Classifiers generated in this work.
-Due to the side of the road being re-grassed near a portion of the asphalt road, the terrain classification algorithm struggled to differentiate the dirt from gravel, leading to some misclassification of the side-of-road areas.
-Examination of the feature distribution shows a high amount of overlap, Fig.
-\ref{fig:dirt_v_gravel} shows an example of this however all features suffered from class confusion.
-Exploitation of confidence scores to adjust results by requiring a very high confidence for positive gravel surface classification was explored.
-All gravel surfaces that did not meet a 80\% confidence score was re-labeled as \textit{unknown}, however these adjustment had either a negligible or detrimental impact on the final results.
-Higher confidence score requirements resulted in \textit{unknown} over-classification, while too low resulted in \textit{gravel} over-classification (Fig.
-\ref{fig:precision_recall_curve_averaged} shows a typical precision-recall curve.).
-Overall it was found that an average true-positive accuracy of all intersecting gravel surfaces was 74.97\% (Tab.
-\ref{tab:1}, Fig.
-\ref{fig:prepostadjust}).
-If any asphalt miss-classification on the gravel areas is included, the average true-positive road detection accuracy increases to 79.18\%.
-It was found that an average true-positive accuracy of all asphalt surfaces was 87.05\%.
-Due to recent construction, discovered asphalt accuracy was artificially lowered due to the area in front of a gravel driveway being contaminated with gravel.
-It was found that an average of 18.55\% of asphalt areas were classified as gravel because of contamination.}
-
-		{Differentiation between gravel road surfaces and surrounding terrain is necessary to determine boundaries for trajectory tracking and navigation.
-Overall average distribution for \textit{unknown} terrain class within side-of-road areas was found to be 77.94\% (Tab.
-\ref{tab:1}), which is lower than expected due to the side of the road being re-grassed.
-It is acknowledged by the authors that the classification algorithm must yield a result that most closely matches one of the trained classes, which led to lower accuracy scores in these circumstances.
-It is also acknowledged by the authors that using a higher resolution LiDAR scanner in conjunction with a more comprehensive and descriptive feature set may be used to aid in differentiating between dirt and gravel, however this future study is outside the scope of this work.
-Overall discovered accuracy levels may be sufficient for the purpose of identifying asphalt and gravel roads by considering trends between classified arcs.}
+Consecutive scanning LiDAR data was projected on manually classified truth areas representing road surfaces  and is shown in Figure \ref{fig:prepostadjust} for area 2 where percentages of terrain types per area were calculated.
+Note that the asphalt area in front of the gravel driveway had significant \textit{gravel} classification due to construction.
+These results are included with the overall asphalt classification score.}
 
@@
 			\centering
-			\includegraphics[width=0.75\linewidth]{figures/dirt_v_gravel_3}
-			\caption[Re-grassed Dirt vs Gravel]{Side-of-road detection accuracy was diminished due to overlapping features between dirt and gravel. }
-			\label{fig:dirt_v_gravel}
+			\includegraphics[width=0.75\linewidth]{figures/db_4_area_scores_4.png}
+			\caption[Area Scores]{Classified scores from test area}
+            \label{fig:prepostadjust}
 		\end{figure}
+
+		{Classification of real-world data outside of the training data set presents difficulties to the generated Random Decision Forest Classifiers generated in this work.
+Due to the side of the road being re-grassed near a portion of the asphalt road, the terrain classification algorithm struggled to differentiate gravel from unknown, leading to some misclassification of the side-of-road areas.
+Exploitation of confidence scores to adjust results by requiring a very high confidence for positive gravel surface classification was explored.
+All gravel surfaces that did not meet a 80\% confidence score was re-labeled as \textit{unknown}, however these adjustment had either a negligible or detrimental impact on the final results.
+Higher confidence score requirements resulted in \textit{unknown} over-classification, while too low resulted in \textit{gravel} over-classification.
+Figure \ref{fig:precision_recall_curve_averaged} shows a typical precision-recall curve.
+Overall it was found that an average true-positive accuracy of all intersecting gravel surfaces was 757\% (Table \ref{tab:1}, Figure \ref{fig:prepostadjust}).
+If any asphalt miss-classification on the gravel areas is included, the average true-positive road detection accuracy increases to 79\%.
+It was found that an average true-positive accuracy of all asphalt surfaces was 87\%.
+Due to recent construction, discovered asphalt accuracy was artificially lowered due to the area in front of a gravel driveway being contaminated with gravel.
+It was found that an average of 19\% of asphalt areas were classified as gravel most likely due to contamination.}
+
+    % Samples were gathered from Area 3 (Figure \ref{fig:test_vs_train_areas}), and examination of the feature distribution showed a high amount of overlap.
+				\begin{table}[h]
+
+			\centering{
+            %\begin{table}
+                \begin{tabular}{|l|r|l|}
+                \hline
+                Gravel Areas \%                     & 75  & \% Gravel                    \\ \hline
+                Asphalt Area \%                     & 87  & \% Asphalt                   \\ \hline
+                Side-of-Road \%                     & 77  & \% Unknown                   \\ \hline
+                Asphalt Road Detection              & 100.00 & \% True Pos. Detection Rate  \\ \hline
+                Gravel Driveway Detection           & 72  & \% True Pos. Detection Rate  \\ \hline
+                Misidentified Gravel Road Detection & 0      & \% False Pos. Detection Rate \\ \hline
+                \end{tabular}
+            %\end{table}
+
+   %
+				% \begin{tabular}{lrl}
+				% 	\toprule
+				% 	Gravel Areas \% 						& 74.97 	& \% Gravel \\
+				% 	\cline{1-3}
+				% 	Asphalt Area \%							& 87.05   	& \% Asphalt \\
+				% 	\cline{1-3}
+				% 	Side-of-Road \% 						& 77.47 	& \% Unknown\\
+				% 	\specialrule{1pt}{0pt}{0pt}
+				% 	Asphalt Road Detection 					& 100.00 	& \% True Pos. Detection Rate \\
+				% 	\cline{1-3}
+				% 	Gravel Driveway Detection 				& 71.67 	& \% True Pos. Detection Rate \\
+				% 	\cline{1-3}
+				% 	Misidentified Gravel Road Detection 	& 0 		& \% False Pos. Detection Rate\\
+				% 	\bottomrule
+				% \end{tabular}
+			}
+			\caption[Table]{Overall results for presented work.}
+   %74.97\% of all samples were classified as \textit{gravel} inside all \textit{gravel} areas, with an overall 71.67\% true positive \textit{gravel} road detection rate and a 0\% false positive \textit{gravel} road detection rate.
+            \label{tab:1}
+		\end{table}
+		{Differentiation between gravel road surfaces and surrounding terrain is necessary to determine boundaries.
+Overall average distribution for \textit{unknown} terrain class within side-of-road areas was found to be 78\% (Table \ref{tab:1}), which is lower than expected due to the side of the road being re-grassed.
+The classification of type was prioritized over accuracy scores leading to lower numbers but better matching.
+The data collected, sensors utilized, and processes were done to identify qualitatively the road type.
+As such, overall discovered accuracy levels may be sufficient for the purpose of identifying asphalt and gravel roads by considering trends between classified arcs.
+
+  %It is acknowledged by the authors that the classification algorithm must yield a result that most closely matches one of the trained classes, which led to lower accuracy scores in these circumstances.
+
+  %It is also acknowledged by the authors that using a higher resolution LiDAR scanner in conjunction with a more comprehensive and descriptive feature set may be used to aid in differentiating between unknown and gravel, however this future study is outside the scope of this work.
+Overall discovered accuracy levels may be sufficient for the purpose of identifying asphalt and gravel roads by considering trends between classified arcs.}
+
+		% \begin{figure}[H]
+		% 	\centering
+		% 	\includegraphics[width=0.75\linewidth]{figures/dirt_v_gravel_4.png}
+		% 	\caption[Re-grassed Dirt vs Gravel]{Side-of-road detection accuracy was diminished due to overlapping features between dirt and gravel.
+Data was gathered from Area 3 (See Figure \ref{fig:test_vs_train_areas}}
+		% 	\label{fig:dirt_v_gravel}
+		% \end{figure}
 
@@
 			\centering
-			\includegraphics[width=0.75\linewidth]{figures/precision_recall_curve_averaged_marked}
+			\includegraphics[width=0.75\linewidth]{figures/precision_recall_curve_averaged_marked_v2.png}
 			\caption[PR Curve]{Precision-Recall curve of the averaged \textit{gravel} vs \textit{unknown} classification results from all test runs. The red circle indicates the presented method's location.}
@@
 
-		\begin{figure}[H]
-			\centering
-			\includegraphics[width=0.75\linewidth]{figures/db_4_area_scores_2}
-			\caption[Area Scores]{Per-area scores were calculated, indicating algorithm unmarked road detection accuracy.
-Note that the asphalt area in front of the gravel driveway has significant \textit{gravel} classification due to construction.
-These results are included with the overall asphalt classification score.
-See Area 2.
-in Fig.
-\ref{fig:test_vs_train_areas}.}
-			\label{fig:prepostadjust}
-		\end{figure}
-
-		{Classified point clouds were manually examined for 'best guess' locations for gravel and asphalt surfaces.
-Two out of three positive gravel identifications must be made on each channel, be evenly spaced, and be relatively co-planar to other road surface guesses in order for a positive guess for an asphalt or gravel surface to be made (Fig.
-\ref{fig:auto_area_guess}).
+		{Classified LiDAR point clouds were manually examined for 'best guess' locations for gravel and asphalt surfaces.
+Two out of three positive gravel identifications must be made on each channel, be evenly spaced, and be relatively co-planar to other road surface guesses in order for a positive guess for an asphalt or gravel surface to be made.
 Driveways $1$ and $2$ were detected in all five drive-bys, driveway $3$ was detected in four leading to an 91.67\% accuracy in intersecting gravel road detection.
@@
 Trends across classified arcs were autonomously exploited to determine the best guess at area terrain type.
-Guessed areas were aggregated into a larger map then compared to truth areas (Fig.
-\ref{fig:guess_grav_intersect}).
-Using autonomous methods to project road surfaces, it was found that a true-positive intersecting gravel road detection rate of 71.66\% with no false positives (Tab.
-\ref{tab:1}).
+Guessed areas were aggregated into a larger map then compared to truth areas as shown in Figure \ref{fig:guess_grav_intersect}.
+Using autonomous methods to project road surfaces, it was found that a true-positive intersecting gravel road detection rate of 72\% with no false positives (Table \ref{tab:1}).
 It was found that the true positive rate may be increased to 100\% at the expense of increasing the number of false positives with a 1.8 average number of false positives per drive-by.}
@@
 			\centering
-			\includegraphics[width=0.95\linewidth]{figures/db1_conf_elim_auto_guess_mca}
-			\caption[Aggregated Automated Road Surface Detection]{Aggregated automated road surface detection was compared to truth areas. See Area 1 in Fig. \ref{fig:test_vs_train_areas}.}
+			\includegraphics[width=0.75\linewidth]{figures/aggregated_auto_area_guess_v2.png}
+			\caption[Aggregated Automated Road Surface Detection]{Aggregated automated road surface detection was compared to truth areas from area 1.}
 			\label{fig:guess_grav_intersect}
@@
 		{Classification was completed using Ubuntu 18.04 and an AMD Ryzen 9 5900X 12 core 24 thread CPU.
-Rate of classification for all nine areas of interest in a single scan was found to be an average of 35.19 ms using a single core or 316.68 ms per 360 degree scan.
+Rate of classification for all nine areas of interest in a single scan was found to be an average of 35 ms using a single core or 316.68 ms per 360 degree scan.
 When utilizing all 24 cores during classification a rate of 13.19 ms per 360 scan.
-Extrapolating from the distance traveled during a drive-by, a hypothetical 71.39 feet per second or 48.68 miles per hour may be possible.
-Future work may include a path of development for testing real-time performance.
-Future work may include a real-time implementation of the described methods into the vehicle platform, however this is outside the scope of this current work.
+Extrapolating from the distance traveled during a drive-by, a hypothetical 71 feet per second or 49 miles per hour may be possible.
+Future work may include a path of development for real-time implementation of the described methods into the vehicle platform, however this is outside the scope of this current work.
 }
+
+   % Future work may include a path of development for testing real-time performance.
 
@@
 %		\end{figure}
-
-		{It is difficult to determine if these results are comparable to similar research, as found research classifies using distinctly different sensor configurations,  algorithm input requirements, or have different objectives.
+		Comparing our results to others highlights some recent improvements in sensing technologies and extension to use with autonomy for on-road conditions.
+Previous road detection work using LiDAR from \cite{wang_road_2012} showed a 95\% accuracy between gravel, concrete, asphalt, and grass given a sensor looking directly down at the surfaces.
+In addition, \cite{laible20123d} was able to use a camera and LiDAR at close-range (1-2 meters) for classification ~90\% accuracy of similar road types.
+These studies are over ten years old and were using LiDAR only for the classification.
+Our investigation was using a modern, long range (100 m), and high-rate LiDAR sensor that is placed for object detection on an autonomous vehicle.
+The sensor location on the vehicle is vastly different that comparable studies and required the use of a novel data approach to yield similar classification results.
+		%{It is difficult to determine if these results are comparable to similar research, as found research classifies using distinctly different sensor configurations,  algorithm input requirements, or have different objectives.
 For example, work completed by Laible et.
@@
 al.
-\cite{wang_road_nodate}, which found a true positive gravel terrain classification rate of 98.5\%; however, the method again used significantly more data from the LiDAR sensor compared to this work, exploiting all LiDAR channels that fall within a 1.3 meter region of interest leading to a higher point density.
+\cite{wang_road_2012}, which found a true positive gravel terrain classification rate of 98.5\%; however, the method again used significantly more data from the LiDAR sensor compared to this work, exploiting all LiDAR channels that fall within a 1.3 meter region of interest leading to a higher LiDAR point density.
 There is a general lack of open source LiDAR data for the express purpose of comparing road detection and terrain classification methods in an unstructured, rural environment, in stark contrast to open source data for urban environments.
@@
 
-% 		(Fig. \ref{fig:per_scan_classification_rate})		
+% 		(Figure \ref{fig:per_scan_classification_rate})		
 %		\begin{figure}[H]
@@
 Terrain classification of consecutive scans compared to manually detected road surface areas yielded accuracy scores for gauging algorithm capability for gravel surface detection.
-It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively at an estimated rate of 13.19 milliseconds per 360 degree scan.
-Classified point clouds were examined and gravel areas were projected and compared with actual gravel surface areas.
-Overlapping results between autonomously detected and actual road surface areas resulted in a minimum  71.66\% intersecting gravel road detection accuracy.
+It was found that a true-positive accuracy for gravel and asphalt surfaces was 75\% and 87\% respectively at an estimated rate of 13 milliseconds per 360 degree scan.
+Classified LiDAR point clouds were examined and gravel areas were projected and compared with actual gravel surface areas.
+Overlapping results between autonomously detected and actual road surface areas resulted in a minimum  72\% intersecting gravel road detection accuracy.
 In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is promising for future exploration of additional roads and road surfaces.
-Work as described addressed the problem of road surface detection on unmarked gravel and asphalt roads.
-The described work was a method of detecting physical unmarked gravel and asphalt roads by using a terrain classification approach to predicting road surface area.
-The impact of this work is that autonomous vehicles using the developed method may be able to detect unmarked road surfaces and differentiate between gravel and asphalt, allowing autonomous operations on previously undetected rural roads.}
+This work produced and examined a method for detecting unmarked physical gravel and asphalt roads using a terrain classification approach.
+Future autonomous vehicles may be able to use the developed method to detect unmarked road surfaces and differentiate between gravel and asphalt, allowing operations unmapped gravel roads.}
 
@@
 
-	{The authors would like to acknowledge the Ohio Department of Transportation, United States Department of Transportation, and the Drive Ohio Automated Driving Systems Project for making this work possible.}
+	{The authors would like to acknowledge the Ohio Department of Transportation, DriveOhio, and the United States Department of Transportation Federal Highways Administration for making this work possible.}
 
